### 2023-09-13 hrs-selection.qmd
First commit

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
date: "September 2023"
format: 
  pdf:
    geometry:
    - top=2in
    - left=1.5in
    - right=1.5in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

```{r setup}

options(digits = 3)
set.seed(271075)

library(haven)
library(forcats)
# library(Hmisc)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop

shorten <- function (x) {
  x <- str_remove(x, "^(A4|E4)_")
  x_short <- str_remove(x, "_.*")

  duplicates <- table(x_short)
  duplicates <- names(duplicates)[duplicates > 1]
  x_short[x_short %in% duplicates] <- x[x_short %in% duplicates]

  x_short
}

calc_income <- function (x) {
  ages <- x |> select(matches("r\\d+agey_e"))
  income <- x |> select(matches("r\\d+iearn"))
  income <- rowMeans(income, na.rm = TRUE)
  income_r <- resid(lm(income ~ factor(x$rabyear)))
  income_r <- income_r - min(income_r) + 1

  income_r
}

# rand HRS longit data. VERY BIG
# Done just once:
# rand_orig <- haven::read_dta("data/randhrs/randhrs1992_2020v1.dta")
# saveRDS(rand_orig, file = "data/randhrs/rand-orig.rds", compress = FALSE)
rand_orig <- readRDS("data/randhrs/rand-orig.rds")

# ragender: 1 male, 2 female
rand_orig <- filter(rand_orig, 
                    (rabyear <= 1965 & ragender == 1) |
                    (rabyear <= 1970 & ragender == 2))

# african ancestry
pgen4a <- haven::read_dta("data/PGENSCORE4r3/pgenscore4a_r.dta")
# european ancestry
pgen4e <- haven::read_dta("data/PGENSCORE4r3/pgenscore4e_r.dta")

# original HRS tracker file
trk_orig <- haven::read_dta("data/trk2020v3/trk2020tr_r.dta")
trk_weights <- trk_orig |> select(HHID, PN, matches("BIOWGTR"))

eths <- c("white", "black")
pgens <- list(white = pgen4e, black = pgen4a)

rands <- list()

for (eth in eths) {
  pgen <- pgens[[eth]]
  names(pgen) <- shorten(names(pgen))
  pgs <- pgen |>
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -EDU2, -EDU3, -HEIGHT, -BMI,
           -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
           -ADHD_PGC10, MDD
           ) |>
    names()

  pcs <- c(paste0("PC1_5", LETTERS[1:5]), paste0("PC6_10", LETTERS[1:5]))

  rand <- inner_join(rand_orig, pgen,
                     join_by(hhid, pn),
                     unmatched = "drop",
                     relationship = "one-to-one")

  rand <- rand |>
    mutate(
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn,
      .by = rabyear
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")) 
    )

  # did this to check if white larger sample size drove result sizes; 
  # it didn't:
  # rand <- rand |> slice_sample(n = 3100)

  # two of four Alzheimer's variants with awkward names:
  pgs <- str_replace(pgs, "^01", "ALZ_01")
  names(rand) <- str_replace(names(rand), "^01", "ALZ_01")

  # weighting
  rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
  rand$weight <- ifelse(! is.na(rand$MBIOWGTR), rand$MBIOWGTR, rand$NBIOWGTR)

  rands[[eth]] <- rand
}

```

Hugh-Jones and Abdellaoui (2022) explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from Becker (XXX). The theory has two components.

1.  There is a trade-off between time spent working and raising children. This leads people with more human capital and higher expected wages to have fewer children.
2.  The trade-off is sharper for low-income people, people with low human capital, and single parents. As a result, natural selection is stronger among these groups.

The evidence for point 1 is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e. they are being selected against. And vice versa, scores which correlate positively with human capital are being selected for. The evidence for point 2 is that correlations with number of children are larger and more significant among people with lower income, less education and single parents.

Here, I test the same theory in the US population, using the Health and Retirement Survey (HRS; XXX). This gives an independent test of the theory. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. Also, the HRS provides precalculated polygenic scores (XXX Aysu) for both black and (non-hispanic) white participants, so I can check whether the human capital theory of natural selection works in both these ethnicities.

Economic theory treats all humans as the same. Once you have controlled appropriately for people's preferences, resources and beliefs, there should be no difference between ethnicities. So, the economic theory of fertility predicts that patterns of natural selection should be the same in both white and black ethnic groups. Of course, ethnic groups may have different economic characteristics. Also, most of the polygenic scores were created using people of European ancestry. This means they are less predictive for people with African ancestry. If natural selection works via scores' correlation with phenotypes, then we would expect these correlations to be smaller among black people.

# Data

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.

TODO: I think for all these guys, the raevborn question will have been asked by the time they were indeed over 45/40. Double check.

The resulting sample contains `r nrow(rands$white)` genotyped white participants and `r nrow(rands$black)` genotyped black participants. Genotyping took place in 2006, 2008 and subsequent years. I use polygenic scores (PGS) precalculated by XXX. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth).

PGS are rescaled to zero mean and unit variance within each ethnic group, so effect sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, I control for ten principal components of the DNA array data.

I use realised fertility (number of children ever born) as the key dependent variable. I use this rather than relative lifetime reproductive success for ease of interpretation and because I don't have data for dates when children were born. My focus is on examining patterns of natural selection across PGS, rather than estimating their effects on PGS over time; raw fertility is good enough for this.

The HRS contains weights which match survey respondents to the US population. I use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, I weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.

# Results

```{r stat-regressions}

tidy_pgs <- mod_pgs <- list()
tidy_pgs_edyrs <- mod_pgs_edyrs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
cors_fert_edu <- tidy_pgs_joined <- list()

for (eth in eths) {
  rand <- rands[[eth]]

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
  mod_pgs[[eth]] <- map(form_pgs, 
                        \(x) lm(x, data = rand, weights = weight)
                        )

  tidy_pgs[[eth]] <- list_rbind(map(
    mod_pgs[[eth]],
    broom::tidy,
    conf.int = TRUE,
    conf.level = 0.95
  )) |>
    filter(term != "(Intercept)", ! term %in% pcs)

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                              \(x) lm(x, data = rand, weights = weight)
                              )

  tidy_pgs_edyrs[[eth]] <-
    list_rbind(map(mod_pgs_edyrs[[eth]], broom::tidy)) |>
    filter(term != "(Intercept)", ! term %in% pcs)

  tidy_pgs_joined[[eth]] <- inner_join(tidy_pgs[[eth]], tidy_pgs_edyrs[[eth]],
                                by = join_by(term),
                                unmatched = "error",
                                suffix = c(".fert", ".edyrs"),
                                relationship = "one-to-one")
  cors_fert_edu[[eth]] <- cor(tidy_pgs_joined[[eth]]$estimate.edyrs, 
                     tidy_pgs_joined[[eth]]$estimate.fert)

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
        \(x) lm(x, data = rand, weights = weight)
        )

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "edu13-17", pcs)) |>
    mutate(edu = str_remove(term, ".*:"))

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
        \(x) lm(x, data = rand, weights = weight)
        )

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))
}

tidy_pgs_eth <- bind_rows(tidy_pgs,  .id = "ethnicity")

n_tests <- nrow(tidy_pgs_eth)
n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)
if (n_sig_bonf != 1) stop("More/less than 1 score was Bonferroni significant")

cor_fert_edu_phen <- list()
for (eth in eths) {
  cor_fert_edu_phen[[eth]] <- lm(raevbrn ~ raedyrs, data = rands[[eth]],
                                 weights = weight) |> 
    coef() |> 
    pluck("raedyrs")
}

```

@fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group. Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores. (Also, because all scores are differently weighted sums of the same set of alleles, if we can reject the null for any score we should probably reject it for all of them; in other words, effect sizes are more interesting than p values here.)

```{r}
#| label: fig-fertility
#| fig-cap: |
#|   Effects of PGS on realized fertility. Values are effects of a one
#|   standard deviation (within ethnicity) change of the PGS on fertility,
#|   controlling for 10 principal components of genomic array data. Scales
#|   differ between the two charts.
#| fig-height: 6

tidy_pgs_eth |>
  mutate(
    term = fct_reorder(term, estimate, \(x) x[2]),
    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
    # hack to order scales separately from 
    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
    term_mangled = paste(ethnicity, term, sep = "/"),
    term_mangled = fct_reorder(term_mangled, estimate),
  ) |>
  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
    geom_vline(xintercept = 0) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                    position = position_dodge(width = 0.5), 
                    alpha = 0.8, size = 0.3) +
    facet_wrap(vars(ethnicity), scales = "free") +
    theme_light() + 
    scale_y_discrete(labels = \(x) str_remove(x, "^.*/")) +
    scale_color_manual(values = c("orange2", "seagreen")) +
    labs(
      x = "", y = "",
      title = "Effects of polygenic scores on fertility"
    ) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.y = element_text(size = 7),
      strip.text = element_text(colour = "black", size = 10),
      strip.background = element_rect(fill = NA)
    )
```

```{r}
#| label: stat-fertility-edu
#| cache: true

cors_boot <- list()
for (eth in eths) {
  rand_short <- rands[[eth]] |> 
    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight)
  cors_boot[[eth]] <-
    replicate(199, {
      rand <- rand_short |> 
              slice_sample(prop = 1, replace = TRUE)

      coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
        map(\(x) lm(x, data = rand, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

      coef_pgs_edyrs <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
        map(\(x) lm(x, data = rand, weights = weight)) |> 
        map_dbl(\(x) coef(x)[[2]])

      cor(coef_pgs_fert, coef_pgs_edyrs)
    })
}

tails_boot <- map(cors_boot, \(x) quantile(x, c(0.025, 0.975)))

if (! all(tails_boot$white < 0)) {
  stop("Fertility-edu correlation NOT sig at 95% for whites")
}
if (all(tails_boot$black < 0)) {
  stop("Fertility-edu correlation IS sig at 95% for blacks")
}

```

@fig-fertility-edu plots each PGS's correlation with fertility against its correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white`, bootstrapped 95% C.I. `r tails_boot$white[1]` to `r tails_boot$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I. `r tails_boot$black[1]` to `r tails_boot$black[2]`). I use bootstraps to make inferences from the sample of individuals, rather than the "sample" of different PGS.

The relationship is imprecisely estimated in the smaller black sample, but it is also steeper. Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites. One reason for this may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites. In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf. @goldscheider1969minority; @johnson1979minority; @yang2003big]. In turn, the simplest explanation for this fact would be point 2 of the theory. Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them. TODO what does the existing literature say?

TODO if so then presumably controlling for education and income would reduce these differences.

```{r}
#| label: fig-fertility-edu
#| fig-cap: |
#|   Scatterplot of PGS partial correlations with fertility and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")
for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[[eth]],
            x = "Effect on years of education",
            y = "Effect on realized fertility"
          )
  print(ggp)
}

# TODO: 3 scores are outliers among african-americans, with negative effects
# on fertility and education - kidney disease, including the trans-ethnic
# score, and PTSD estimated among european ancestry. Why?
```

# Bibliography

### 2023-09-15 hrs-selection.qmd
Add siblings regressions

@@
 author: "David Hugh-Jones"
+abstract: |
+  I investigate natural selection on polygenic scores
+  in the contemporary US, using the Health and Retirement Study. Results
+  support the economic theory of fertility as an explanation for
+  natural selection: among both white and black respondents,
+  scores which correlate negatively (positively) with education are
+  selected for (against). However, selection effects 
+  are not larger among people with less education. 
+  For whites, selection effects are smaller among married people. 
 date: "September 2023"
@@
 
+# TODO: siblings E5669 "how many have died"
+#   rands$white |> select(matches("r\\d+livsib")) -> tmp
+#   apply(tmp, 1, max)
+
+# TODO: use survey package
+# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd
+
+# TODO: ratio to family poverty threshold, h10inpovr
+
+# TODO: age at first birth. RAND family data provides h*ageokid
+# TODO: RAND family data has r*sbdied for how many siblings have died
+
 options(digits = 3)
@@
 library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
+
+nice_names <- c(
+  SCZ         = "Schizophrenia",
+  WC          = "Waist circ.",
+  WHR         = "Waist-hip ratio",
+  NEUROT      = "Neuroticism",
+  WELLB       = "Well-being",
+  DEPSYMP     = "Depr. symptoms",
+  CD          = "Cor. art. disease",
+  MI          = "Myoc. infarction",
+  CORT        = "Cortisol",
+  T2D         = "Type 2 diabetes",
+  BIP         = "Bipolar",
+  ADHD_PGC17  = "ADHD",
+  XDISORDER   = "Cross-disorder",
+  MENA        = "Age at menarche",
+  MENO        = "Age at menopause",
+  EXTRAVER    = "Extraversion",
+  AUTISM      = "Autism",
+  AB          = "Antisoc. behav.",
+  OCD         = "OCD",
+  AFBC        = "Age 1st birth combn",
+  AFBF        = "Age 1st birth F",
+  AFBM        = "Age 1st birth M",
+  MDD2        = "Maj. depr. disorder",
+  PTSDAA      = "PTSD Afr",
+  PTSDEA      = "PTSD Eur",
+  PTSDC       = "PTSD combined",
+  HDL         = "HDL cholesterol",
+  LDL         = "LDL cholesterol",
+  TC          = "Total cholesterol",
+  ANXFS       = "Anxiety CC", 
+  ANXCC       = "Anxiety factor",
+  BUN         = "Blood urea nitro.",
+  BUNTE       = "Blood urea nitro. TE",
+  CKD         = "Kidney disease",
+  CKDTE       = "Kidney disease TE",
+  DBP         = "Diastolic blood pr.",
+  BMI2        = "BMI",
+  HEIGHT2     = "Height",
+  AI          = "Age smoking init.",
+  CPD_GSCAN19 = "Cigs per day",
+  DPW         = "Drinks per week",
+  SC          = "Smoking cessation",
+  SI          = "Smoking initiation",
+  HTN         = "Hypertension",
+  CANNABIS    = "Cannabis",
+  GWAD2NA     = "Alzheimer's no APOE",
+  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
+  GWAD2WA     = "Alzheimer's",
+  ALZ_01AD2WA = "Alzheimer's p 0.01",
+  ALC         = "Alcohol dependence",
+  PP          = "Pulse pressure",
+  SBP         = "Systolic blood pr.",
+  EGFR        = "Glom. filtr.",
+  EGFRTE      = "Glom. filtr. TE",
+  EA3         = "Educ. attainment",
+  HBA1CAA     = "HbA1c Afr",
+  HBA1CEA     = "HbA1c Eur",
+  GCOG2       = "Gen. cognition"
+)
 
@@
 
+# rand family respondents file
+rand_fam <- haven::read_dta(
+  "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
+# only selected columns to avoid creating suffixes when we join with
+# rand.
+# Age of oldest kid; grandchildren; num siblings who died
+rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)
+
 eths <- c("white", "black")
@@
            -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
-           -ADHD_PGC10, MDD
+           -ADHD_PGC10, -MDD
            ) |>
@@
   rand <- rand |>
-    mutate(
+    mutate(.by = rabyear,
       mean_raevbrn = mean(raevbrn, na.rm = TRUE),
-      rlrs = raevbrn/mean_raevbrn,
-      .by = rabyear
+      rlrs = raevbrn/mean_raevbrn
     ) |>
@@
       raedyrs = as.numeric(raedyrs),
-      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")) 
+      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
+      # wave 10 is 2010
+      # table(rands$white$r10mstat)
+       # 1                 1.married
+       # 2   2.married,spouse absent
+       # 3               3.partnered
+       # 4               4.separated
+       # 5                5.divorced
+       # 6      6.separated/divorced
+       # 7                 7.widowed
+       # 8           8.never married
+      married = factor(r10mstat == 1, 
+                       levels = c(FALSE, TRUE),
+                       labels = c("Other", "Married")),
+    )
+
+  rand <- rand |> left_join(rand_fam, 
+                            by = join_by(hhid, pn),
+                            unmatched = "drop",
+                            relationship = "one-to-one")
+  rand <- rand |> mutate(
+    r10sibs = r10livsib + r10sbdied
   )
@@
 
-Hugh-Jones and Abdellaoui (2022) explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from Becker (XXX). The theory has two components.
+@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.
 
@@
 tidy_pgs <- mod_pgs <- list()
-tidy_pgs_edyrs <- mod_pgs_edyrs <- list()
+tidy_pgs_ed <- mod_pgs_edyrs <- list()
+tidy_pgs_sibs <- mod_pgs_sibs <- list()
 tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
 tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
+tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
 cors_fert_edu <- tidy_pgs_joined <- list()
-
+cors_sibs_edu <- list()
 for (eth in eths) {
@@
 
-  tidy_pgs[[eth]] <- list_rbind(map(
-    mod_pgs[[eth]],
-    broom::tidy,
-    conf.int = TRUE,
-    conf.level = 0.95
-  )) |>
-    filter(term != "(Intercept)", ! term %in% pcs)
+  tidy_pgs[[eth]] <- mod_pgs[[eth]] |> 
+    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
+    list_rbind() |>
+    filter(term %in% pgs)
 
@@
 
-  tidy_pgs_edyrs[[eth]] <-
-    list_rbind(map(mod_pgs_edyrs[[eth]], broom::tidy)) |>
-    filter(term != "(Intercept)", ! term %in% pcs)
+  tidy_pgs_ed[[eth]] <- mod_pgs_edyrs[[eth]] |>  
+    map(broom::tidy) |>
+    list_rbind() |> 
+    filter(term %in% pgs)
 
-  tidy_pgs_joined[[eth]] <- inner_join(tidy_pgs[[eth]], tidy_pgs_edyrs[[eth]],
+  tidy_pgs_joined[[eth]] <- inner_join(tidy_pgs[[eth]], tidy_pgs_ed[[eth]],
                                 by = join_by(term),
@@
 
+  form_pgs_sibs <- map(pgs, 
+                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
+  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
+                             \(x) lm(x, data = rand, 
+                                     weights = weight/(1 + r10sibs)))
+
+  tidy_pgs_sibs[[eth]] <- mod_pgs_sibs[[eth]] |> 
+    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |>
+    list_rbind() |> 
+    filter(term %in% pgs)
+
+  cors_sibs_edu[[eth]] <- cor(tidy_pgs_sibs[[eth]]$estimate, 
+                     tidy_pgs_ed[[eth]]$estimate)
+
   # the colon interaction estimates the effect separately within each group
   form_pgs_x_ed <- paste0(pgs, ":edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
   mod_pgs_x_ed[[eth]]  <-
@@
   form_pgs_x_inc <- paste0(pgs, ":income_med") |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
   mod_pgs_x_inc[[eth]]  <-
@@
     mutate(income_med = str_remove(term, ".*:income_med"))
+
+  form_pgs_x_mar <- paste0(pgs, ":married") |>
+    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
+  mod_pgs_x_mar[[eth]]  <-
+    map(form_pgs_x_mar, 
+        \(x) lm(x, data = rands[[eth]], weights = weight)
+        )
+
+  tidy_pgs_x_mar[[eth]] <-
+    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_mar[[eth]]) <- pgs
+  tidy_pgs_x_mar[[eth]] <-
+    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
+    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
+    mutate(
+      married = str_remove(term, ".*:married")
+    ) 
 }
@@
 #| fig-cap: |
-#|   Effects of PGS on realized fertility. Values are effects of a one
-#|   standard deviation (within ethnicity) change of the PGS on fertility,
-#|   controlling for 10 principal components of genomic array data. Scales
-#|   differ between the two charts.
+#|  Effects of PGS on realized fertility. Values are partial correlations 
+#|  of a one standard deviation (within ethnicity) change of the PGS with
+#|  fertility, controlling for 10 principal components of genomic array 
+#|  data. Scales differ between the two charts.
 #| fig-height: 6
@@
   mutate(
+    term = nice_names[term],
     term = fct_reorder(term, estimate, \(x) x[2]),
@@
     # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
-    term_mangled = paste(ethnicity, term, sep = "/"),
+    term_mangled = paste(ethnicity, term, sep = "//"),
     term_mangled = fct_reorder(term_mangled, estimate),
@@
     theme_light() + 
-    scale_y_discrete(labels = \(x) str_remove(x, "^.*/")) +
+    scale_y_discrete(labels = \(x) str_remove(x, "^.*//")) +
     scale_color_manual(values = c("orange2", "seagreen")) +
@@
 ```{r}
-#| label: stat-fertility-edu
+#| label: stat-bootstraps
 #| cache: true
 
+n_reps <- 599
 cors_boot <- list()
@@
   rand_short <- rands[[eth]] |> 
-    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight)
+    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, live_sibs)
   cors_boot[[eth]] <-
-    replicate(199, {
+    replicate(n_reps, {
       rand <- rand_short |> 
@@
 
+      coef_pgs_sibs <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "live_sibs")) |> 
+        map(\(x) lm(x, data = rand, weights = weight/(live_sibs + 1))) |> 
+        map_dbl(\(x) coef(x)[[2]])
+
       coef_pgs_edyrs <- pgs |> 
@@
 
-      cor(coef_pgs_fert, coef_pgs_edyrs)
+      c(fert_edyrs = cor(coef_pgs_fert, coef_pgs_edyrs),
+        sibs_edyrs = cor(coef_pgs_sibs, coef_pgs_edyrs))
     })
@@
 
-tails_boot <- map(cors_boot, \(x) quantile(x, c(0.025, 0.975)))
+tails_boot_fert <- map(cors_boot, 
+                       \(x) quantile(x["fert_edyrs", ], c(0.025, 0.975)))
 
-if (! all(tails_boot$white < 0)) {
+tails_boot_sibs <- map(cors_boot,
+                       \(x) quantile(x["sibs_edyrs", ], c(0.025, 0.975)))
+
+if (! all(tails_boot_fert$white < 0)) {
   stop("Fertility-edu correlation NOT sig at 95% for whites")
 }
-if (all(tails_boot$black < 0)) {
+if (all(tails_boot_fert$black < 0)) {
   stop("Fertility-edu correlation IS sig at 95% for blacks")
@@
 The relationship is negative among whites (correlation `r cors_fert_edu$white`, bootstrapped 95% C.I.
-`r tails_boot$white[1]` to `r tails_boot$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I.
-`r tails_boot$black[1]` to `r tails_boot$black[2]`).
-I use bootstraps to make inferences from the sample of individuals, rather than the "sample" of different PGS.
+`r tails_boot_fert$white[1]` to `r tails_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I.
+`r tails_boot_fert$black[1]` to `r tails_boot_fert$black[2]`).
+I use bootstraps so as to make inferences from the sample of individuals, rather than the "sample" of different PGS.
 
@@
 Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites.
-One reason for this may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites.
+If this holds in the population, one reason may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites.
 In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf.
@@
 Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them.
-TODO what does the existing literature say?
-
-TODO if so then presumably controlling for education and income would reduce these differences.
 
@@
 #| fig-cap: |
-#|   Scatterplot of PGS partial correlations with fertility and years of
+#|   Scatterplot of PGS effect sizes on fertility and years of
 #|   education. Controls include 10 principal components of genetic array
@@
 
+```{r}
+#| label: stat-siblings
+
+tmp <- 
+  tidy_pgs_sibs$black |> 
+  inner_join(tidy_pgs_ed$black, 
+             by = join_by(term), 
+             suffix = c(".sibs", ".edyrs")) |> 
+  filter(term != "GWAD2NA", term != "GWAD2WA")
+
+cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
+if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
+```
+
+I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010.
+I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
+Parents of no siblings cannot be included, so effect sizes are not comparable across the generations.
+@fig-siblings plots effects on number of siblings versus effects on years of education.
+The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I.
+`r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`).
+It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.
+
+```{r}
+#| label: fig-siblings
+#| fig-cap: |
+#|   Scatterplot of PGS effect sizes on number of live siblings and years of
+#|   education. Controls include 10 principal components of genetic array
+#|   data. Dashed lines show linear regressions.
+#| fig-subcap:
+#|   - ""
+#|   - ""
+#| layout-nrow: 2
+#| column: page
+
+for (eth in eths) {
+  ggp <-  
+    tidy_pgs_sibs[[eth]] |> 
+    inner_join(tidy_pgs_ed[[eth]], 
+               by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
+    ggplot(aes(estimate.edyrs, estimate.sibs)) +
+      geom_hline(yintercept = 0, color = "grey80") +
+      geom_vline(xintercept = 0, color = "grey80") +
+      geom_point() +
+      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
+                  linetype = "dashed", colour = "red", linewidth = 0.5) +
+      coord_cartesian(xlim = c(-.3, .75)) +
+      theme_light() + 
+      theme(panel.grid = element_blank()) +
+      labs(
+          title = eth,
+          x = "Effect on years of education",
+          y = "Effect on number of siblings"
+      )
+  print(ggp)
+}
+
+```
+
+I next test part 2 of the theory more directly by interacting PGS with measures of education, income and marital status.
+Education is years of education, split at 12 years.
+Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split.
+
+```{r}
+#| label: stat-x-edu
+#| cache: true
+
+n_reps <- 199
+
+cross_terms <- function (x) {
+  cx <- coef(x)
+  cx[str_detect(names(cx), ":")]
+}
+
+form_pgs_x_ed <- paste0(pgs, ":edu") |>
+    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
+form_pgs_x_inc <- paste0(pgs, ":income_med")  |>
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
+form_pgs_x_mar <- paste0(pgs, ":married")  |>
+    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
+
+coefs_boot_ed <- coefs_boot_inc <- coefs_boot_mar <- list()
+
+for (eth in eths) {
+  rand <- rands[[eth]]
+  rand_short <- rands[[eth]] |> 
+    select(
+      all_of(pgs), all_of(pcs), 
+      raevbrn, raedyrs, weight, income_med, edu, married
+    )
+
+  coefs_boot_ed[[eth]] <- replicate(n_reps, {
+    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
+    map(form_pgs_x_ed, 
+        \(x) lm(x, data = rand_r, weights = weight)
+        ) |> 
+      map(cross_terms) |> 
+      unlist()
+  })
+
+  coefs_boot_inc[[eth]]  <- replicate(n_reps, {
+    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
+    map(form_pgs_x_inc, 
+          \(x) lm(x, data = rand_r, weights = weight)
+          ) |> 
+      map(cross_terms) |> 
+      unlist()
+  })
+
+  coefs_boot_mar[[eth]]  <- replicate(n_reps, {
+    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
+    map(form_pgs_x_mar, 
+          \(x) lm(x, data = rand_r, weights = weight)
+          ) |> 
+      map(cross_terms) |> 
+      unlist()
+  })
+
+  reorganize <- function (x) {
+   x |> 
+    t() |> 
+    as.data.frame() |> 
+    pivot_longer(everything(), names_to = c("pgs", "group"), 
+                 names_sep = ":") |> 
+    mutate(rep = 1:n_reps, .by = c(pgs, group)) |> 
+    pivot_wider(id_cols = c(pgs, rep), 
+                names_from = group, 
+                values_from = value) 
+  }
+
+  coefs_boot_ed[[eth]] <- reorganize(coefs_boot_ed[[eth]])
+  coefs_boot_inc[[eth]] <- reorganize(coefs_boot_inc[[eth]])
+  coefs_boot_mar[[eth]] <- reorganize(coefs_boot_mar[[eth]])
+}
+
+boot_ci <- function (x) {
+  names(x)[3:4] <- c("lo", "hi")
+  x |> 
+    mutate(
+      abs_diff = abs(hi) - abs(lo)
+    ) |> 
+    summarize(
+      mad = mean(abs_diff),
+      .by = rep
+    ) |> 
+    pull(mad) |> 
+    quantile(c(0.025, 0.975))
+}
+
+ci_inc_w <- boot_ci(coefs_boot_inc$white)
+ci_inc_b <- boot_ci(coefs_boot_inc$black)
+ci_ed_w <- boot_ci(coefs_boot_ed$white)
+ci_ed_b <- boot_ci(coefs_boot_ed$black)
+ci_mar_w <- boot_ci(coefs_boot_mar$white)
+ci_mar_b <- boot_ci(coefs_boot_mar$black)
+
+for (ci in list(ci_inc_w, ci_inc_b, ci_ed_w, ci_ed_b)) {
+  if (prod(sign(ci)) != -1) {
+    stop("effect size IS significantly different across some pair groups")
+  }  
+}
+```
+
+@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education.
+@fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks.
+However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group.
+@fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
+Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I.
+for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`).
+
+```{r}
+#| label: fig-x-edu
+#| fig-cap: |
+#|   Violin plots of polygenic score effects on fertility by education. 
+#|   Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high
+#|   education group. Lines show quartiles.
+
+tidy_pgs_x_ed |> 
+  list_rbind(names_to = "ethnicity") |> 
+  mutate(
+    Ethnicity = unlist(titles[ethnicity]),
+    Education = str_remove(edu, "edu")
+  ) |> 
+  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
+    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
+    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
+    facet_wrap(vars(Ethnicity), scales = "free_y") +
+    labs(
+      x = "Years of education",
+      y = "Absolute effect size"
+    ) +
+    theme_minimal() +
+    theme(
+      legend.position = "none"
+    )
+```
+
+```{r}
+#| label: fig-x-inc
+#| fig-cap: |
+#|   Violin plots of polygenic score effects on fertility by income. 
+#|   Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Lines show quartiles.
+
+tidy_pgs_x_inc |> 
+  list_rbind(names_to = "ethnicity") |> 
+  mutate(
+    Ethnicity = unlist(titles[ethnicity]),
+    Income = fct_relevel(income_med, c("Low", "High"))
+  ) |> 
+  ggplot(aes(Income, abs(estimate), color = Income, fill = Income)) +
+    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
+    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
+    facet_wrap(vars(Ethnicity), scales = "free_y") +
+    labs(
+      x = "Income",
+      y = "Absolute effect size"
+    ) +
+    theme_minimal() +
+    theme(
+      legend.position = "none"
+    )
+```
+
+```{r}
+#| label: fig-x-mar
+#| fig-cap: |
+#|   Violin plots of polygenic score effects on fertility by marital status. 
+#|   Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Lines show quartiles.
+
+tidy_pgs_x_mar |> 
+    list_rbind(names_to = "ethnicity") |> 
+    mutate(
+        Ethnicity = unlist(titles[ethnicity]),
+        married = fct_relevel(married, "Other", "Married")
+    ) |> 
+    ggplot(aes(married, abs(estimate), color = married, fill = married)) +
+      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
+      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
+      facet_wrap(vars(Ethnicity), scales = "free_y") +
+      labs(
+          x = "Married",
+          y = "Absolute effect size"
+      ) +
+      theme_minimal() +
+      theme(
+          legend.position = "none"
+      )
+
+```
+
+# Discussion
+
+The results here provide qualified support for the economic theory of fertility as an explanation for contemporary natural selection in humans.
+PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
+However, there is no evidence that PGS effects on fertility are larger among people with less education, unlike the UK [@hugh2022human].
+And there's not enough evidence to show that effects on fertility are larger for lower-income people.
+Also, the smaller black sample made most tests inconclusive for this population.
+
 # Bibliography

### 2023-09-15 hrs-selection.qmd
Add agefb regs and cohorts bar plot

@@
 
+titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")
+
 shorten <- function (x) {
@@
 
+  rand <- rand |> left_join(rand_fam, 
+                            by = join_by(hhid, pn),
+                            unmatched = "drop",
+                            relationship = "one-to-one")
   rand <- rand |>
@@
                        labels = c("Other", "Married")),
-    )
-
-  rand <- rand |> left_join(rand_fam, 
-                            by = join_by(hhid, pn),
-                            unmatched = "drop",
-                            relationship = "one-to-one")
-  rand <- rand |> mutate(
-    r10sibs = r10livsib + r10sbdied
+    r10sibs = r10livsib + r10sbdied,
+    agefb = r10agey_e - h10ageokid,
+    agefb = ifelse(agefb < 12, NA_real_, agefb),
+    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
   )
@@
 tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
+tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()
+
 cors_fert_edu <- tidy_pgs_joined <- list()
@@
     )
+
+  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
+  mod_pgs_x_agefb[[eth]]  <-
+    map(form_pgs_x_agefb, 
+        \(x) lm(x, data = rands[[eth]], weights = weight)
+        )
+
+  tidy_pgs_x_agefb[[eth]] <-
+    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
+    setNames(pgs) |> 
+    list_rbind(names_to = "pgs") |>
+    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
+    mutate(
+      agefb = str_remove(term, ".*:agefb")
+    )
+
 }
@@
 
-titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")
 for (eth in eths) {
@@
 for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`).
+@fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
+There are no large differences between early and late parents.
+Note that the N is lower here due to missing values.
+TODO test
 
@@
 
+nobs <- map_dbl(eths, 
+                \(x) nobs(mod_pgs_x_agefb[[x]][[1]])) |> 
+  setNames(eths)
+
 tidy_pgs_x_ed |> 
@@
   mutate(
-    Ethnicity = unlist(titles[ethnicity]),
+    Ethnicity = paste0(unlist(titles[ethnicity]), 
+                       " (N = ", nobs[ethnicity], ")"),
     Education = str_remove(edu, "edu")
@@
 
+```{r}
+#| label: fig-x-agefb
+#| fig-cap: |
+#|   Violin plots of polygenic score effects on fertility by age at 
+#|   first birth (median split). Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Lines show quartiles.
+
+tidy_pgs_x_agefb |> 
+    list_rbind(names_to = "ethnicity") |> 
+    mutate(
+        Ethnicity = unlist(titles[ethnicity]),
+        agefb = fct_relevel(agefb, "Low", "High")
+    ) |> 
+    ggplot(aes(agefb, abs(estimate), color = agefb, fill = agefb)) +
+      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
+      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
+      facet_wrap(vars(Ethnicity), scales = "free_y") +
+      labs(
+          x = "Age at first birth",
+          y = "Absolute effect size"
+      ) +
+      theme_minimal() +
+      theme(
+          legend.position = "none"
+      )
+
+```
+
 # Discussion
@@
 
+# Appendix
+
+```{r}
+#| label: fig-cohorts
+#| fig-cap: Distribution of birth years for the sample
+#| fig-subcap:
+#|   - ""
+#|   - ""
+#| layout-nrow: 2
+#| column: page
+
+for (eth in eths) {
+  ggp <- ggplot(rands[[eth]], aes(rabyear)) + 
+    geom_bar() +
+    theme_minimal() +
+    labs(x = "Birth year", y = "Count", title = titles[[eth]])
+  print(ggp)
+}
+```
+
 # Bibliography

### 2023-09-16 hrs-selection.qmd
Move pics to appendix

@@
   pdf:
+    include-in-header:
+      - text: |
+          \usepackage{placeins}
     geometry:
@@
 
-# TODO: siblings E5669 "how many have died"
-#   rands$white |> select(matches("r\\d+livsib")) -> tmp
-#   apply(tmp, 1, max)
-
 # TODO: use survey package
 # here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd
+# TODO (related) move away from lists and towards a single datasource with an
+# "ethnicity" column; results from broom::tidy can still be in lists?
 
@@
 
-# TODO: age at first birth. RAND family data provides h*ageokid
-# TODO: RAND family data has r*sbdied for how many siblings have died
+# TODO: think about your bootstraps, what's the test... given that 
+# the scores are not independent, does it really make sense to aggregate
+# a general size comparison? Or am I secretly just reproducing my own
+# crappy standard errors?
 
@@
 library(forcats)
-# library(Hmisc)
 library(dplyr, warn.conflicts = FALSE)
@@
 
-titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")
+eths <- c("white", "black")
+
+titles <- c(white = "White", black = "Black")
 
@@
 
-eths <- c("white", "black")
 pgens <- list(white = pgen4e, black = pgen4a)
+
+pcs <- c(paste0("PC1_5", LETTERS[1:5]), paste0("PC6_10", LETTERS[1:5]))
 
@@
 
-@fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
+Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
 Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
 But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
-(Also, because all scores are differently weighted sums of the same set of alleles, if we can reject the null for any score we should probably reject it for all of them; in other words, effect sizes are more interesting than p values here.)
-
-```{r}
-#| label: fig-fertility
-#| fig-cap: |
-#|  Effects of PGS on realized fertility. Values are partial correlations 
-#|  of a one standard deviation (within ethnicity) change of the PGS with
-#|  fertility, controlling for 10 principal components of genomic array 
-#|  data. Scales differ between the two charts.
-#| fig-height: 6
-
-tidy_pgs_eth |>
-  mutate(
-    term = nice_names[term],
-    term = fct_reorder(term, estimate, \(x) x[2]),
-    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
-    # hack to order scales separately from 
-    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
-    term_mangled = paste(ethnicity, term, sep = "//"),
-    term_mangled = fct_reorder(term_mangled, estimate),
-  ) |>
-  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
-    geom_vline(xintercept = 0) +
-    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
-                    position = position_dodge(width = 0.5), 
-                    alpha = 0.8, size = 0.3) +
-    facet_wrap(vars(ethnicity), scales = "free") +
-    theme_light() + 
-    scale_y_discrete(labels = \(x) str_remove(x, "^.*//")) +
-    scale_color_manual(values = c("orange2", "seagreen")) +
-    labs(
-      x = "", y = "",
-      title = "Effects of polygenic scores on fertility"
-    ) +
-    theme(
-      panel.grid.major.y = element_blank(),
-      panel.grid.minor.x = element_blank(),
-      legend.position = "none",
-      axis.text.y = element_text(size = 7),
-      strip.text = element_text(colour = "black", size = 10),
-      strip.background = element_rect(fill = NA)
-    )
-```
 
@@
           labs(
-            title = titles[[eth]],
+            title = titles[eth],
             x = "Effect on years of education",
@@
 Parents of no siblings cannot be included, so effect sizes are not comparable across the generations.
-@fig-siblings plots effects on number of siblings versus effects on years of education.
+Appendix @fig-siblings plots effects on number of siblings versus effects on years of education.
 The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I.
@@
 It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.
-
-```{r}
-#| label: fig-siblings
-#| fig-cap: |
-#|   Scatterplot of PGS effect sizes on number of live siblings and years of
-#|   education. Controls include 10 principal components of genetic array
-#|   data. Dashed lines show linear regressions.
-#| fig-subcap:
-#|   - ""
-#|   - ""
-#| layout-nrow: 2
-#| column: page
-
-for (eth in eths) {
-  ggp <-  
-    tidy_pgs_sibs[[eth]] |> 
-    inner_join(tidy_pgs_ed[[eth]], 
-               by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
-    ggplot(aes(estimate.edyrs, estimate.sibs)) +
-      geom_hline(yintercept = 0, color = "grey80") +
-      geom_vline(xintercept = 0, color = "grey80") +
-      geom_point() +
-      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
-                  linetype = "dashed", colour = "red", linewidth = 0.5) +
-      coord_cartesian(xlim = c(-.3, .75)) +
-      theme_light() + 
-      theme(panel.grid = element_blank()) +
-      labs(
-          title = eth,
-          x = "Effect on years of education",
-          y = "Effect on number of siblings"
-      )
-  print(ggp)
-}
-
-```
 
@@
 ```{r}
-#| label: stat-x-edu
+#| label: stat-x-edu-bootstraps
 #| cache: true
@@
 
-nobs <- map_dbl(eths, 
-                \(x) nobs(mod_pgs_x_agefb[[x]][[1]])) |> 
-  setNames(eths)
-
 tidy_pgs_x_ed |> 
@@
   mutate(
-    Ethnicity = paste0(unlist(titles[ethnicity]), 
-                       " (N = ", nobs[ethnicity], ")"),
+    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_ed, ethnicity)),
     Education = str_remove(edu, "edu")
@@
   mutate(
-    Ethnicity = unlist(titles[ethnicity]),
+    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_inc, ethnicity)),
     Income = fct_relevel(income_med, c("Low", "High"))
@@
     mutate(
-        Ethnicity = unlist(titles[ethnicity]),
+      Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_mar, ethnicity)),
       married = fct_relevel(married, "Other", "Married")
@@
     mutate(
-        Ethnicity = unlist(titles[ethnicity]),
+     Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_agefb, ethnicity)),
       agefb = fct_relevel(agefb, "Low", "High")
@@
 
+\FloatBarrier
+
 # Appendix
+
+## Figures
 
@@
 
+```{r}
+#| label: fig-fertility
+#| fig-cap: |
+#|  Effects of PGS on realized fertility. Values are partial correlations 
+#|  of a one standard deviation (within ethnicity) change of the PGS with
+#|  fertility, controlling for 10 principal components of genomic array 
+#|  data. Scales differ between the two charts.
+#| fig-height: 6
+
+tidy_pgs_eth |>
+  mutate(
+    term = nice_names[term],
+    term = fct_reorder(term, estimate, \(x) x[2]),
+    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
+    # hack to order scales separately from 
+    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
+    term_mangled = paste(ethnicity, term, sep = "//"),
+    term_mangled = fct_reorder(term_mangled, estimate),
+  ) |>
+  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
+    geom_vline(xintercept = 0) +
+    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
+                    position = position_dodge(width = 0.5), 
+                    alpha = 0.8, size = 0.3) +
+    facet_wrap(vars(ethnicity), scales = "free") +
+    theme_light() + 
+    scale_y_discrete(labels = \(x) str_remove(x, "^.*//")) +
+    scale_color_manual(values = c("orange2", "seagreen")) +
+    labs(
+      x = "", y = "",
+      title = "Effects of polygenic scores on fertility"
+    ) +
+    theme(
+      panel.grid.major.y = element_blank(),
+      panel.grid.minor.x = element_blank(),
+      legend.position = "none",
+      axis.text.y = element_text(size = 7),
+      strip.text = element_text(colour = "black", size = 10),
+      strip.background = element_rect(fill = NA)
+    )
+```
+
+```{r}
+#| label: fig-siblings
+#| fig-cap: |
+#|   Scatterplot of PGS effect sizes on number of live siblings and years of
+#|   education. Controls include 10 principal components of genetic array
+#|   data. Dashed lines show linear regressions.
+#| fig-subcap:
+#|   - ""
+#|   - ""
+#| layout-nrow: 2
+#| column: page
+
+for (eth in eths) {
+  ggp <-  
+    tidy_pgs_sibs[[eth]] |> 
+    inner_join(tidy_pgs_ed[[eth]], 
+               by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
+    ggplot(aes(estimate.edyrs, estimate.sibs)) +
+      geom_hline(yintercept = 0, color = "grey80") +
+      geom_vline(xintercept = 0, color = "grey80") +
+      geom_point() +
+      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
+                  linetype = "dashed", colour = "red", linewidth = 0.5) +
+      coord_cartesian(xlim = c(-.3, .75)) +
+      theme_light() + 
+      theme(panel.grid = element_blank()) +
+      labs(
+          title = eth,
+          x = "Effect on years of education",
+          y = "Effect on number of siblings"
+      )
+  print(ggp)
+}
+
+```
+
+\FloatBarrier
+
 # Bibliography

### 2023-09-16 hrs-selection.qmd
Create a single dataframe for both ethnicities, split when necessary

@@
 
-```{r setup}
+```{r}
+#| label: setup
 
@@
 
-calc_income <- function (x) {
+calc_income <- function (x, eth) {
+  x <- x |> filter(ethnicity == eth)
   ages <- x |> select(matches("r\\d+agey_e"))
@@
   income_r
+}
+
+calc_n <- function (mods, eths = c("black", "white"), format = "(N = %s)") {
+  nobs <- map_dbl(eths, \(x) nobs(mods[[x]][[1]]) ) 
+  if (! is.null(format)) nobs <- sprintf(format, nobs)
+  names(nobs) <- eths
+  nobs
 }
@@
 
-  pcs <- c(paste0("PC1_5", LETTERS[1:5]), paste0("PC6_10", LETTERS[1:5]))
-
-  rand <- inner_join(rand_orig, pgen,
+  rands[[eth]] <- inner_join(rand_orig, pgen,
                      join_by(hhid, pn),
@@
                      relationship = "one-to-one")
+}
+
+rand <- list_rbind(rands, names_to = "ethnicity")
 
@@
 rand <- rand |>
-    mutate(.by = rabyear,
+  mutate(.by = rabyear, # calculate this pooling ethnicity
     mean_raevbrn = mean(raevbrn, na.rm = TRUE),
@@
   ) |>
-    mutate(
-      income_resid = calc_income(rand),
+  mutate(.by = ethnicity, # calculate these stats within-ethnicity
+    income_resid = calc_income(rand, ethnicity[1]),
     income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
@@
 
-  # did this to check if white larger sample size drove result sizes; 
-  # it didn't:
-  # rand <- rand |> slice_sample(n = 3100)
-
 # two of four Alzheimer's variants with awkward names:
@@
 rand$weight <- ifelse(! is.na(rand$MBIOWGTR), rand$MBIOWGTR, rand$NBIOWGTR)
-
-  rands[[eth]] <- rand
-}
 
@@
 
-```{r stat-regressions}
+```{r}
+#| label: stat-regressions
+#| cache: true
 
@@
 cors_sibs_edu <- list()
+cor_fert_edu_phen <- list()
+
 for (eth in eths) {
-  rand <- rands[[eth]]
+  r_eth <- rand |> filter(ethnicity == eth)
 
@@
   mod_pgs[[eth]] <- map(form_pgs, 
-                        \(x) lm(x, data = rand, weights = weight)
+                        \(x) lm(x, data = r_eth, weights = weight)
                         )
@@
   mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
-                              \(x) lm(x, data = rand, weights = weight)
+                              \(x) lm(x, data = r_eth, weights = weight)
                               )
@@
   mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
-                             \(x) lm(x, data = rand, 
+                             \(x) lm(x, data = r_eth, 
                                      weights = weight/(1 + r10sibs)))
@@
     map(form_pgs_x_ed, 
-        \(x) lm(x, data = rand, weights = weight)
+        \(x) lm(x, data = r_eth, weights = weight)
         )
@@
     map(form_pgs_x_inc, 
-        \(x) lm(x, data = rand, weights = weight)
+        \(x) lm(x, data = r_eth, weights = weight)
         )
@@
     map(form_pgs_x_mar, 
-        \(x) lm(x, data = rands[[eth]], weights = weight)
+        \(x) lm(x, data = r_eth, weights = weight)
         )
@@
     map(form_pgs_x_agefb, 
-        \(x) lm(x, data = rands[[eth]], weights = weight)
+        \(x) lm(x, data = r_eth, weights = weight)
         )
@@
 
+  cor_fert_edu_phen[[eth]] <- 
+    lm(raevbrn ~ raedyrs, data = r_eth, weights = weight) |> 
+    coef() |> 
+    pluck("raedyrs")
 }
@@
 if (n_sig_bonf != 1) stop("More/less than 1 score was Bonferroni significant")
-
-cor_fert_edu_phen <- list()
-for (eth in eths) {
-  cor_fert_edu_phen[[eth]] <- lm(raevbrn ~ raedyrs, data = rands[[eth]],
-                                 weights = weight) |> 
-    coef() |> 
-    pluck("raedyrs")
-}
 
@@
 for (eth in eths) {
-  rand_short <- rands[[eth]] |> 
-    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, live_sibs)
+  rand_short <- rand |> 
+    filter(ethnicity == eth) |> 
+    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, r10sibs)
   cors_boot[[eth]] <-
     replicate(n_reps, {
-      rand <- rand_short |> 
+      r_boot <- rand_short |> 
               slice_sample(prop = 1, replace = TRUE)
@@
         map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
-        map(\(x) lm(x, data = rand, weights = weight)) |> 
+        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
       coef_pgs_sibs <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "live_sibs")) |> 
-        map(\(x) lm(x, data = rand, weights = weight/(live_sibs + 1))) |> 
+        map(\(x) reformulate(c(x, pcs), response = "r10sibs")) |> 
+        map(\(x) lm(x, data = r_boot, weights = weight/(r10sibs + 1))) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
         map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
-        map(\(x) lm(x, data = rand, weights = weight)) |> 
+        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
 for (eth in eths) {
-  rand <- rands[[eth]]
-  rand_short <- rands[[eth]] |> 
+  rand_short <- rand |> 
+    filter(ethnicity == eth) |> 
     select(
@@
 #| fig-cap: Distribution of birth years for the sample
-#| fig-subcap:
-#|   - ""
-#|   - ""
-#| layout-nrow: 2
-#| column: page
 
-for (eth in eths) {
-  ggp <- ggplot(rands[[eth]], aes(rabyear)) + 
-    geom_bar() +
+rand |> 
+    mutate(
+      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
+    ) |> 
+    ggplot(aes(rabyear, fill = Ethnicity)) + 
+      geom_bar(position = "identity") +
+      labs(x = "Birth year", y = "Count") +
+      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
       theme_minimal() +
-    labs(x = "Birth year", y = "Count", title = titles[[eth]])
-  print(ggp)
-}
+      theme(legend.title = element_blank())
+
 ```
@@
 
+for (eth in eths) {}
+
 for (eth in eths) {

### 2023-09-16 hrs-selection.qmd
weight income residualization

@@
   income <- rowMeans(income, na.rm = TRUE)
-  income_r <- resid(lm(income ~ factor(x$rabyear)))
+  income_r <- resid(lm(income ~ factor(x$rabyear), weights = x$weight))
   income_r <- income_r - min(income_r) + 1
@@
   names(pgen) <- shorten(names(pgen))
-  pgs <- pgen |>
+  rands[[eth]] <- inner_join(rand_orig, pgen,
+                     join_by(hhid, pn),
+                     unmatched = "drop",
+                     relationship = "one-to-one")
+}
+
+pgs <- pgen4e |>
+    rename_with(shorten) |> 
+    # weird that I have to operate on whole df to use select semantics
     select(-hhid, -pn, -version,
@@
     names()
-
-  rands[[eth]] <- inner_join(rand_orig, pgen,
-                     join_by(hhid, pn),
-                     unmatched = "drop",
-                     relationship = "one-to-one")
-}
 

### 2023-09-16 hrs-selection.qmd
Use svyglm and create survey design objects

@@
 library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
+library(survey)
 
@@
 rand$weight <- ifelse(! is.na(rand$MBIOWGTR), rand$MBIOWGTR, rand$NBIOWGTR)
+# compensate for fecund parents more likely to have child in survey:
+rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
+
+rand <- rand |> 
+  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, raedyrs,
+         r10sibs, edu, income_med, married, agefb, rabyear,
+         raehsamp, raestrat, parent_weight)
+
+r_parents <- svydesign(
+    id = ~ raehsamp,
+		strata = ~ raestrat,
+		weights = ~ parent_weight, 
+		nest = TRUE,
+		data = rand |> drop_na(parent_weight) 
+		)
+
+rand <- svydesign(
+    id = ~ raehsamp,
+		strata = ~ raestrat,
+		weights = ~ weight, 
+		nest = TRUE,
+		data = rand |> drop_na(weight) 
+		)
 
@@
 for (eth in eths) {
-  r_eth <- rand |> filter(ethnicity == eth)
+  r_eth <- rand |> subset(ethnicity == eth)
+  r_parents_eth <- r_parents |> subset(ethnicity == eth)
 
@@
   mod_pgs[[eth]] <- map(form_pgs, 
-                        \(x) lm(x, data = r_eth, weights = weight)
+                        \(x) svyglm(x, r_eth)
                         )
@@
   mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
-                              \(x) lm(x, data = r_eth, weights = weight)
+                              \(x) svyglm(x, r_eth)
                               )
@@
   mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
-                             \(x) lm(x, data = r_eth, 
-                                     weights = weight/(1 + r10sibs)))
+                             \(x) svyglm(x, r_parents_eth))
 
@@
     map(form_pgs_x_ed, 
-        \(x) lm(x, data = r_eth, weights = weight)
+        \(x) svyglm(x, r_eth)
         )
@@
     map(form_pgs_x_inc, 
-        \(x) lm(x, data = r_eth, weights = weight)
+        \(x) svyglm(x, r_eth)
         )
@@
     map(form_pgs_x_mar, 
-        \(x) lm(x, data = r_eth, weights = weight)
+        \(x) svyglm(x, r_eth)
         )
@@
     map(form_pgs_x_agefb, 
-        \(x) lm(x, data = r_eth, weights = weight)
+        \(x) svyglm(x, r_eth)
         )
@@
   cor_fert_edu_phen[[eth]] <- 
-    lm(raevbrn ~ raedyrs, data = r_eth, weights = weight) |> 
+    svyglm(raevbrn ~ raedyrs, r_eth) |> 
     coef() |> 
@@
 n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)
-if (n_sig_bonf != 1) stop("More/less than 1 score was Bonferroni significant")
+if (n_sig_bonf != 0) stop("More/less than 0 score was Bonferroni significant")
 
@@
 Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
-Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
+Standard errors are large because of the relatively low sample sizes, and no scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
 But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
@@
   rand_short <- rand |> 
-    filter(ethnicity == eth) |> 
-    select(all_of(pgs), all_of(pcs), raevbrn, raedyrs, weight, r10sibs)
+    subset(ethnicity == eth)
   cors_boot[[eth]] <-
     replicate(n_reps, {
-      r_boot <- rand_short |> 
-              slice_sample(prop = 1, replace = TRUE)
+      # survey package automatically does the right thing
+      # possibly I should be using survey::bootweights
+      # or survey::withReplicates and a function
+      r_boot <- rand_short[sample(nrow(rand_short), replace = TRUE), ]
 
@@
         map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
-        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
+        map(\(x) svyglm(x, r_boot)) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
         map(\(x) reformulate(c(x, pcs), response = "r10sibs")) |> 
-        map(\(x) lm(x, data = r_boot, weights = weight/(r10sibs + 1))) |> 
+        # need to add , weights = weight/(r10sibs + 1)
+        map(\(x) svyglm(x, r_boot)) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
         map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
-        map(\(x) lm(x, data = r_boot, weights = weight)) |> 
+        map(\(x) svyglm(x, r_boot)) |> 
         map_dbl(\(x) coef(x)[[2]])
@@
 
-for (eth in eths) {}
-
 for (eth in eths) {

### 2023-09-17 hrs-selection.qmd
Use withReplicates for main bootstraps

@@
 # crappy standard errors?
+# 
+# TODO: is ADHD reverse-coded? What about XDISORDER?
 
@@
 ```
-
 Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
@@
 n_reps <- 599
-cors_boot <- list()
-for (eth in eths) {
-  rand_short <- rand |> 
-    subset(ethnicity == eth)
-  cors_boot[[eth]] <-
-    replicate(n_reps, {
-      # survey package automatically does the right thing
-      # possibly I should be using survey::bootweights
-      # or survey::withReplicates and a function
-      r_boot <- rand_short[sample(nrow(rand_short), replace = TRUE), ]
 
+calc_cors_edu <- function (weights, data, var) {
   coef_pgs_fert <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
-        map(\(x) svyglm(x, r_boot)) |> 
+        map(\(x) reformulate(c(x, pcs), response = var)) |> 
+        map(\(x) lm(x, data, weights = weights)) |> 
         map_dbl(\(x) coef(x)[[2]])
 
-      coef_pgs_sibs <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "r10sibs")) |> 
-        # need to add , weights = weight/(r10sibs + 1)
-        map(\(x) svyglm(x, r_boot)) |> 
+  coef_pgs_edu <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
+        map(\(x) lm(x, data, weights = weights)) |> 
         map_dbl(\(x) coef(x)[[2]])
 
-      coef_pgs_edyrs <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
-        map(\(x) svyglm(x, r_boot)) |> 
-        map_dbl(\(x) coef(x)[[2]])
-
-      c(fert_edyrs = cor(coef_pgs_fert, coef_pgs_edyrs),
-        sibs_edyrs = cor(coef_pgs_sibs, coef_pgs_edyrs))
-    })
+  cor(coef_pgs_edu, coef_pgs_fert)
 }
 
-tails_boot_fert <- map(cors_boot, 
-                       \(x) quantile(x["fert_edyrs", ], c(0.025, 0.975)))
+cors_fert_edu <- cors_sibs_edu <- list()
+for (eth in eths) {
+  rand_boot <- rand |> 
+    subset(ethnicity == eth) |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps)
+  r_parents_boot <- r_parents |> 
+    subset(ethnicity == eth) |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps)
 
-tails_boot_sibs <- map(cors_boot,
-                       \(x) quantile(x["sibs_edyrs", ], c(0.025, 0.975)))
-
-if (! all(tails_boot_fert$white < 0)) {
-  stop("Fertility-edu correlation NOT sig at 95% for whites")
-}
-if (all(tails_boot_fert$black < 0)) {
-  stop("Fertility-edu correlation IS sig at 95% for blacks")
+  cors_fert_edu[[eth]] <- 
+    withReplicates(rand_boot, theta = calc_cors_edu, var = "raevbrn")
+  cors_sibs_edu[[eth]] <- 
+    withReplicates(r_parents_boot, theta = calc_cors_edu,var = "r10sibs")
 }
 
+tails_boot_fert <- map(cors_fert_edu, 
+                       \(x) x + c(-1.96, 1.96) * SE(x))
+
+tails_boot_sibs <- map(cors_sibs_edu, 
+                       \(x) x + c(-1.96, 1.96) * SE(x))
 ```
@@
 
-rand |> 
+rand$variables |> 
     mutate(

### 2023-09-18 hrs-selection.qmd
bugfixes

@@
 
-# TODO: use survey package
-# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd
-# TODO (related) move away from lists and towards a single datasource with an
-# "ethnicity" column; results from broom::tidy can still be in lists?
-
 # TODO: ratio to family poverty threshold, h10inpovr
@@
 library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
-library(survey)
+ # masks dotchart:
+suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))
 
@@
   income <- rowMeans(income, na.rm = TRUE)
-  income_r <- resid(lm(income ~ factor(x$rabyear), weights = x$weight))
-  income_r <- income_r - min(income_r) + 1
+  mod <- lm(income ~ factor(x$rabyear), weights = x$weight, na.action =
+              na.exclude)
+  income_r <- resid(mod)
+  income_r <- income_r - min(income_r, na.rm = TRUE) + 1
 
@@
 }
+
+# avoid warnings about zero weights
+svyglm_quiet <- function (...) suppressWarnings(svyglm(...))
+
+# for a confidence interval
+significant <- function (x) {
+  stopifnot(is.numeric(x), length(x) == 2)
+  x[1] < 0 && x[2] > 0
+}
+
+insignificant <- Negate(significant)
 
@@
 
+rand <- list_rbind(rands, names_to = "ethnicity")
+
 pgs <- pgen4e |>
@@
 
-rand <- list_rbind(rands, names_to = "ethnicity")
-
 rand <- rand |> left_join(rand_fam, 
@@
                           relationship = "one-to-one")
+
+# weighting
+rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
+rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
+# compensate for fecund parents more likely to have child in survey.
+# we use living sibs rather than including dead sibs too here. I think that's
+# the correct approach:
+rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
+
 rand <- rand |>
@@
                      labels = c("Other", "Married")),
-    r10sibs = r10livsib + r10sbdied,
+    r10sibs = r10livsib + r10sbdied + 1, # n. sibs including oneself!
     agefb = r10agey_e - h10ageokid,
@@
 names(rand) <- str_replace(names(rand), "^01", "ALZ_01")
-
-# weighting
-rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
-rand$weight <- ifelse(! is.na(rand$MBIOWGTR), rand$MBIOWGTR, rand$NBIOWGTR)
-# compensate for fecund parents more likely to have child in survey:
-rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
 
@@
 
+n_white <- nrow(rand |> subset(ethnicity == "white"))
+n_black <- nrow(rand |> subset(ethnicity == "black"))
+
+# clean up cache so it doesn't get too big
+rm(rand_orig, trk_orig, rands, rand_fam)
 ```
@@
 
-The resulting sample contains `r nrow(rands$white)` genotyped white participants and `r nrow(rands$black)` genotyped black participants.
+The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.
 Genotyping took place in 2006, 2008 and subsequent years.
@@
 #| cache: true
+#| cache-lazy: false
 
@@
   mod_pgs[[eth]] <- map(form_pgs, 
-                        \(x) svyglm(x, r_eth)
+                        \(x) svyglm_quiet(x, r_eth)
                         )
@@
   mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
-                              \(x) svyglm(x, r_eth)
+                              \(x) svyglm_quiet(x, r_eth)
                               )
@@
   mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
-                             \(x) svyglm(x, r_parents_eth))
+                             \(x) svyglm_quiet(x, r_parents_eth))
 
@@
     map(form_pgs_x_ed, 
-        \(x) svyglm(x, r_eth)
+        \(x) svyglm_quiet(x, r_eth)
         )
@@
     map(form_pgs_x_inc, 
-        \(x) svyglm(x, r_eth)
+        \(x) svyglm_quiet(x, r_eth)
         )
@@
     map(form_pgs_x_mar, 
-        \(x) svyglm(x, r_eth)
+        \(x) svyglm_quiet(x, r_eth)
         )
@@
     map(form_pgs_x_agefb, 
-        \(x) svyglm(x, r_eth)
+        \(x) svyglm_quiet(x, r_eth)
         )
@@
   cor_fert_edu_phen[[eth]] <- 
-    svyglm(raevbrn ~ raedyrs, r_eth) |> 
+    svyglm_quiet(raevbrn ~ raedyrs, r_eth) |> 
     coef() |> 
@@
 ```
+
 Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
@@
                        \(x) x + c(-1.96, 1.96) * SE(x))
+
+stopifnot(significant(tails_boot_fert$white))
+stopifnot(insignificant(tails_boot_fert$black))
+stopifnot(all(map_lgl(tails_boot_sibs, insignificant)))
 ```
@@
 In turn, the simplest explanation for this fact would be point 2 of the theory.
-Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them.
+Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them [cf.
+@maralani2013demography].
 
@@
 Appendix @fig-siblings plots effects on number of siblings versus effects on years of education.
-The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I.
-`r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`).
-It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.
+Correlations are insignificant for both ethnic groups.
 
@@
 ```{r}
-#| label: stat-x-edu-bootstraps
+#| label: stat-x-bootstraps
 #| cache: true
 
-n_reps <- 199
+n_reps_x <- 199
 
-cross_terms <- function (x) {
-  cx <- coef(x)
-  cx[str_detect(names(cx), ":")]
+median_diff_effect <- function(w, data, term) {
+  forms <- paste0(pgs, ":", term) |>
+    map(\(x) reformulate(c(x, term, pcs), response = "raevbrn"))
+
+  mods <- map(forms, \(x) lm(x, data, weights = w))
+  cfs <- map(mods, coef)
+  cross_terms <- map(cfs, \(x) x[str_detect(names(x), ":")])
+  if(any(lengths(cross_terms) != 2)) stop("Extra cross terms")
+  abs_cross_terms <- map(cross_terms, abs)
+  diffs <- map_dbl(abs_cross_terms, diff) 
+  median(diffs)
 }
 
-form_pgs_x_ed <- paste0(pgs, ":edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
-form_pgs_x_inc <- paste0(pgs, ":income_med")  |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
-form_pgs_x_mar <- paste0(pgs, ":married")  |>
-    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
-
-coefs_boot_ed <- coefs_boot_inc <- coefs_boot_mar <- list()
+median_diff_ed <- median_diff_inc <- median_diff_mar <- median_diff_agefb <-
+  list()
 
 for (eth in eths) {
-  rand_short <- rand |> 
-    filter(ethnicity == eth) |> 
-    select(
-      all_of(pgs), all_of(pcs), 
-      raevbrn, raedyrs, weight, income_med, edu, married
-    )
+  rand_boot <- rand |> 
+    subset(ethnicity == eth) |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps_x)
 
-  coefs_boot_ed[[eth]] <- replicate(n_reps, {
-    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
-    map(form_pgs_x_ed, 
-        \(x) lm(x, data = rand_r, weights = weight)
-        ) |> 
-      map(cross_terms) |> 
-      unlist()
-  })
-
-  coefs_boot_inc[[eth]]  <- replicate(n_reps, {
-    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
-    map(form_pgs_x_inc, 
-          \(x) lm(x, data = rand_r, weights = weight)
-          ) |> 
-      map(cross_terms) |> 
-      unlist()
-  })
-
-  coefs_boot_mar[[eth]]  <- replicate(n_reps, {
-    rand_r <- slice_sample(rand_short, prop = 1, replace = TRUE)
-    map(form_pgs_x_mar, 
-          \(x) lm(x, data = rand_r, weights = weight)
-          ) |> 
-      map(cross_terms) |> 
-      unlist()
-  })
-
-  reorganize <- function (x) {
-   x |> 
-    t() |> 
-    as.data.frame() |> 
-    pivot_longer(everything(), names_to = c("pgs", "group"), 
-                 names_sep = ":") |> 
-    mutate(rep = 1:n_reps, .by = c(pgs, group)) |> 
-    pivot_wider(id_cols = c(pgs, rep), 
-                names_from = group, 
-                values_from = value) 
+  median_diff_ed[[eth]] <- 
+    withReplicates(rand_boot, theta = median_diff_effect, term = "edu")
+  median_diff_inc[[eth]] <- 
+    withReplicates(rand_boot, theta = median_diff_effect, term = "income_med")
+  median_diff_mar[[eth]] <- 
+    withReplicates(rand_boot, theta = median_diff_effect, term = "married")
+  median_diff_agefb[[eth]] <-
+    withReplicates(rand_boot, theta = median_diff_effect, term = "agefb")
 }
 
-  coefs_boot_ed[[eth]] <- reorganize(coefs_boot_ed[[eth]])
-  coefs_boot_inc[[eth]] <- reorganize(coefs_boot_inc[[eth]])
-  coefs_boot_mar[[eth]] <- reorganize(coefs_boot_mar[[eth]])
+confint_workaround <- function (x) {
+  as.numeric(x) + c(-1.96, 1.96) * SE(x)
 }
 
-boot_ci <- function (x) {
-  names(x)[3:4] <- c("lo", "hi")
-  x |> 
-    mutate(
-      abs_diff = abs(hi) - abs(lo)
-    ) |> 
-    summarize(
-      mad = mean(abs_diff),
-      .by = rep
-    ) |> 
-    pull(mad) |> 
-    quantile(c(0.025, 0.975))
-}
+ci_ed <- map(median_diff_ed, confint_workaround)
+ci_inc <- map(median_diff_inc, confint_workaround)
+ci_mar <- map(median_diff_mar, confint_workaround)
+ci_agefb <- map(median_diff_agefb, confint_workaround)
 
-ci_inc_w <- boot_ci(coefs_boot_inc$white)
-ci_inc_b <- boot_ci(coefs_boot_inc$black)
-ci_ed_w <- boot_ci(coefs_boot_ed$white)
-ci_ed_b <- boot_ci(coefs_boot_ed$black)
-ci_mar_w <- boot_ci(coefs_boot_mar$white)
-ci_mar_b <- boot_ci(coefs_boot_mar$black)
+# TODO: is median_diff really a sensible test to run?
 
-for (ci in list(ci_inc_w, ci_inc_b, ci_ed_w, ci_ed_b)) {
-  if (prod(sign(ci)) != -1) {
-    stop("effect size IS significantly different across some pair groups")
-  }  
-}
 ```
@@
 @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
-Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I.
-for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`).
+Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents.
 @fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
@@
 Note that the N is lower here due to missing values.
-TODO test
 

### 2023-09-18 hrs-selection.qmd

@@
 # TODO: ratio to family poverty threshold, h10inpovr
-
-# TODO: think about your bootstraps, what's the test... given that 
-# the scores are not independent, does it really make sense to aggregate
-# a general size comparison? Or am I secretly just reproducing my own
-# crappy standard errors?
 # 
@@
 
-Here, I test the same theory in the US population, using the Health and Retirement Survey (HRS; XXX).
+Here, I test the same theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
 This gives an independent test of the theory.
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-Also, the HRS provides precalculated polygenic scores (XXX Aysu) for both black and (non-hispanic) white participants, so I can check whether the human capital theory of natural selection works in both these ethnicities.
+Also, the HRS provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the human capital theory of natural selection works in both these ethnicities.
 
@@
 Genotyping took place in 2006, 2008 and subsequent years.
-I use polygenic scores (PGS) precalculated by XXX.
 I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
@@
 
-median_diff_effect <- function(w, data, term) {
+calc_diff_cor_ed <- function (w, data, term) {
+  groups <- switch(term,
+                edu = c("0-12", "13-17"),
+                income_med = c("Low", "High"),
+                married = c("Other", "Married"),
+                agefb = c("Low", "High"))
+  groups <- paste0(":", term, groups) # colon for the cross term
+  low_group <- groups[1]
+  high_group <- groups[2]
+
   forms <- paste0(pgs, ":", term) |>
     map(\(x) reformulate(c(x, term, pcs), response = "raevbrn"))
-
   mods <- map(forms, \(x) lm(x, data, weights = w))
   cfs <- map(mods, coef)
-  cross_terms <- map(cfs, \(x) x[str_detect(names(x), ":")])
-  if(any(lengths(cross_terms) != 2)) stop("Extra cross terms")
-  abs_cross_terms <- map(cross_terms, abs)
-  diffs <- map_dbl(abs_cross_terms, diff) 
-  median(diffs)
+  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
+  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])
+
+  forms_ed <- paste0(pgs, ":", term)  |> 
+    map(\(x) reformulate(c(x, term, pcs), response = "raedyrs"))
+  mods_ed <- forms_ed |> map(\(x) lm(x, data, weights = w))
+  cfs_ed <- map(mods_ed, coef)
+  cfs_ed_low <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), low_group)])
+  cfs_ed_high <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), high_group)])
+
+  cor_ed_low <- cor(cfs_low, cfs_ed_low)
+  cor_ed_high <- cor(cfs_high, cfs_ed_high)
+
+  cor_ed_low - cor_ed_high
 }
 
-median_diff_ed <- median_diff_inc <- median_diff_mar <- median_diff_agefb <-
+diff_cor_ed_ed <- diff_cor_ed_inc <- diff_cor_ed_mar <- diff_cor_ed_agefb <-
   list()
@@
 
-  median_diff_ed[[eth]] <- 
-    withReplicates(rand_boot, theta = median_diff_effect, term = "edu")
-  median_diff_inc[[eth]] <- 
-    withReplicates(rand_boot, theta = median_diff_effect, term = "income_med")
-  median_diff_mar[[eth]] <- 
-    withReplicates(rand_boot, theta = median_diff_effect, term = "married")
-  median_diff_agefb[[eth]] <-
-    withReplicates(rand_boot, theta = median_diff_effect, term = "agefb")
+  diff_cor_ed_ed[[eth]] <- 
+    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "edu")
+  diff_cor_ed_inc[[eth]] <- 
+    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "income_med")
+  diff_cor_ed_mar[[eth]] <- 
+    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "married")
+  diff_cor_ed_agefb[[eth]] <-
+    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "agefb")
 }
@@
 
-ci_ed <- map(median_diff_ed, confint_workaround)
-ci_inc <- map(median_diff_inc, confint_workaround)
-ci_mar <- map(median_diff_mar, confint_workaround)
-ci_agefb <- map(median_diff_agefb, confint_workaround)
+ci_ed <- map(diff_cor_ed_ed, confint_workaround)
+ci_inc <- map(diff_cor_ed_inc, confint_workaround)
+ci_mar <- map(diff_cor_ed_mar, confint_workaround)
+ci_agefb <- map(diff_cor_ed_agefb, confint_workaround)
 
-# TODO: is median_diff really a sensible test to run?
+# TODO: does splitting by education to 12 and education 13-16 mechanically
+#   cause a lower correlation with edyrs in the 13-16 group?
+stopifnot(significant(ci_ed$white))
+stopifnot(insignificant(ci_ed$black))
+stopifnot(all(map_lgl(ci_inc, insignificant)))
+stopifnot(all(map_lgl(ci_mar, insignificant)))
+stopifnot(all(map_lgl(ci_agefb, insignificant)))
 
@@
 @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks.
-However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group.
 @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
@@
 Note that the N is lower here due to missing values.
+
+The economic theory implies that correlations between education and fertility should be higher for lower-income/education people, single parents and people who have children earlier.
+This in turn implies that correlations between PGS effects on education and on fertility should be higher among these groups.
+I test this with bootstraps, calculating effect sizes and correlations within each pair of groups, and taking the difference between them.
+Correlations are indeed higher among low- than high-education whites (bootstrapped C.I.
+for difference in correlations `r ci_ed$white[1]` to `r ci_ed$white[2]`) but for all other pairs of groups the difference is not significantly different from zero.
+In fact, the width of confidence intervals shows that we lack power to detect all but large differences.
 
@@
 PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
-However, there is no evidence that PGS effects on fertility are larger among people with less education, unlike the UK [@hugh2022human].
-And there's not enough evidence to show that effects on fertility are larger for lower-income people.
-Also, the smaller black sample made most tests inconclusive for this population.
+However, there is little evidence that PGS effects on fertility are larger, or more correlated with effects on education, among low-education or low-income individuals, unmarried parents, or younger parents.
+This is partly due to the low sample size.
+But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible in this data.
+Lastly, the smaller black sample makes most tests inconclusive for this population.
 
@@
 # Appendix
+
+## Acknowledgements
+
+The HRS (Health and Retirement Study) is sponsored by the National Institute on Aging (grant number NIA U01AG009740) and is conducted by the University of Michigan.
 

### 2023-09-18 hrs-selection.qmd
Bugfixes

@@
 # 
-# TODO: is ADHD reverse-coded? What about XDISORDER?
+# Is ADHD reverse-coded? - Doesn't seem to be, correlates negatively
+#       with income
+# What about XDISORDER?
+# ADHD doesn't seem to be. Nor does XDISORDER (correlates with e.g. r*depres)
 
@@
   stopifnot(is.numeric(x), length(x) == 2)
-  x[1] < 0 && x[2] > 0
+  all(x < 0) || all(x > 0)
 }
@@
 insignificant <- Negate(significant)
+
+confint_workaround <- function (x) {
+  as.numeric(x) + c(-1.96, 1.96) * SE(x)
+}
 
@@
 This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.
+Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].
 
@@
 
-tails_boot_fert <- map(cors_fert_edu, 
-                       \(x) x + c(-1.96, 1.96) * SE(x))
+tails_boot_fert <- map(cors_fert_edu, confint_workaround)
 
-tails_boot_sibs <- map(cors_sibs_edu, 
-                       \(x) x + c(-1.96, 1.96) * SE(x))
+tails_boot_sibs <- map(cors_sibs_edu, confint_workaround)
 
@@
     withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "agefb")
-}
-
-confint_workaround <- function (x) {
-  as.numeric(x) + c(-1.96, 1.96) * SE(x)
 }

### 2023-09-18 hrs-selection.qmd
Tweaks; use fewer bootstraps when results are very obvious

@@
 #| label: stat-regressions
-#| cache: true
-#| cache-lazy: false
+
+# cache seems to cause trouble even with cache-lazy: FALSE
 
@@
 
-n_reps_x <- 199
+n_reps_x <- 99
 
@@
 
-The economic theory implies that correlations between education and fertility should be higher for lower-income/education people, single parents and people who have children earlier.
-This in turn implies that correlations between PGS effects on education and on fertility should be higher among these groups.
+The economic theory implies that correlations between education and fertility should be higher for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS effects on education and on fertility should be higher among these groups.
 I test this with bootstraps, calculating effect sizes and correlations within each pair of groups, and taking the difference between them.
@@
 for difference in correlations `r ci_ed$white[1]` to `r ci_ed$white[2]`) but for all other pairs of groups the difference is not significantly different from zero.
-In fact, the width of confidence intervals shows that we lack power to detect all but large differences.
+In fact, the confidence intervals are so wide that we lack power to detect all but large differences.
+
+[^1]: See @hugh2022human equation (6) and following.
 
@@
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible in this data.
+
 Lastly, the smaller black sample makes most tests inconclusive for this population.
+Most effect sizes, measured in within-ethnicity standard deviations, are larger among black people, which implies that the *proportional* effect of natural selection on PGS scores is larger in this subgroup.
+But since most PGS are also less predictive of their target phenotypes among black people, this fact is hard to interpret.
 

### 2023-09-19 hrs-selection.qmd
Remove most tests and plots on black subsample. Use boxplots.

--- 2023-09-18 hrs-selection.qmd
@@
   in the contemporary US, using the Health and Retirement Study. Results
-  support the economic theory of fertility as an explanation for
+  partially support the economic theory of fertility as an explanation for
   natural selection: among both white and black respondents,
   scores which correlate negatively (positively) with education are
-  selected for (against). However, selection effects 
-  are not larger among people with less education. 
-  For whites, selection effects are smaller among married people. 
+  selected for (against). However, unlike in the UK, there is little evidence
+  that selection effects are larger among low-income or less educated people.
 date: "September 2023"
@@
   pdf:
+    keep-tex: true
     include-in-header:
@@
     - top=2in
-    - left=1.5in
-    - right=1.5in
+    - left=1.25in
+    - right=1.25in
     - heightrounded
@@
 # TODO: ratio to family poverty threshold, h10inpovr
-# 
+# TODO: I think for all these guys, the raevborn question will have been 
+#.   asked by the time they were indeed over 45/40. Double check.
+
 # Is ADHD reverse-coded? - Doesn't seem to be, correlates negatively
@@
 set.seed(271075)
+knitr::knit_hooks$set("inline", 
+  \(x) {
+    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
+      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
+    }
+    x
+  }
+)
 
@@
 
-calc_n <- function (mods, eths = c("black", "white"), format = "(N = %s)") {
+calc_n <- function (mods, eths = "white", format = "(N = %s)") {
   nobs <- map_dbl(eths, \(x) nobs(mods[[x]][[1]]) ) 
@@
 confint_workaround <- function (x) {
-  as.numeric(x) + c(-1.96, 1.96) * SE(x)
+  cbind(
+    as.numeric(x) - 1.96 * SE(x), 
+    as.numeric(x) + 1.96 * SE(x))
 }
@@
 
-TODO: I think for all these guys, the raevborn question will have been asked by the time they were indeed over 45/40. Double check.
-
 The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.
@@
 I also discard PGS for number of children ever born (but keep scores for age at first birth).
+This leaves a total of `r length(pgs)` scores.
 
@@
 
-cors_fert_edu <- tidy_pgs_joined <- list()
+coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
 cors_sibs_edu <- list()
@@
                      tidy_pgs_joined[[eth]]$estimate.fert)
+  coefs_fert_edu[[eth]] <- lm(estimate.fert ~ estimate.edyrs,
+                              data = tidy_pgs_joined[[eth]]) |> 
+                           coef() |> 
+                           pluck("estimate.edyrs")
 
@@
 
-Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group.
+```{r}
+#| label: stat-bootstrap-eth-cor
+
+n_reps_eth_diff <- 99
+rand_boot <- rand |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps_eth_diff)
+
+calc_eth_cor <- function (weights, data) {
+  coef_pgs_white <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) lm(x, data, weights = weights, 
+                    subset = ethnicity == "white")) |> 
+        map2_dbl(pgs, \(x, y) coef(x)[[y]])
+  coef_pgs_black <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) lm(x, data, weights = weights, 
+                    subset = ethnicity == "black")) |> 
+        map2_dbl(pgs, \(x, y) coef(x)[[y]])
+
+  cor(coef_pgs_black, coef_pgs_white)
+}
+
+cor_eths <- cor(tidy_pgs$white$estimate, tidy_pgs$black$estimate)
+cor_eths_boot <- withReplicates(rand_boot, theta = calc_eth_cor)
+ci_eths <- confint_workaround(cor_eths_boot)[1,]
+```
+
+I estimated effect sizes of PGS correlations with realised fertility among black and white respondents separately.[^1] Appendix @fig-fertility shows effect sizes for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
 Standard errors are large because of the relatively low sample sizes, and no scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
 But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
+
+[^1]: Throughout, I use "effect" as shorthand for "regression coefficient".
+No claim about causality is implied.
+Recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
 
@@
 
-  cor(coef_pgs_edu, coef_pgs_fert)
+  c(
+    cor(coef_pgs_edu, coef_pgs_fert),
+    coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
+  )
 }
@@
   cors_sibs_edu[[eth]] <- 
-    withReplicates(r_parents_boot, theta = calc_cors_edu,var = "r10sibs")
+    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
 }
 
-tails_boot_fert <- map(cors_fert_edu, confint_workaround)
-
-tails_boot_sibs <- map(cors_sibs_edu, confint_workaround)
-
-stopifnot(significant(tails_boot_fert$white))
-stopifnot(insignificant(tails_boot_fert$black))
-stopifnot(all(map_lgl(tails_boot_sibs, insignificant)))
 ```
 
-@fig-fertility-edu plots each PGS's correlation with fertility against its correlation with educational attainment.
-The relationship is negative among whites (correlation `r cors_fert_edu$white`, bootstrapped 95% C.I.
-`r tails_boot_fert$white[1]` to `r tails_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black`, bootstrapped 95% C.I.
-`r tails_boot_fert$black[1]` to `r tails_boot_fert$black[2]`).
-I use bootstraps so as to make inferences from the sample of individuals, rather than the "sample" of different PGS.
+```{r}
+#| label: stat-bootstrap-eth-diff
+#| eval: false
 
-The relationship is imprecisely estimated in the smaller black sample, but it is also steeper.
-Put another way, if we measured PGS in units of "years of associated education", estimated selection effects would be larger among blacks than whites.
-If this holds in the population, one reason may be that the correlation between phenotypic years of education and fertility is larger for blacks than for whites.
-In bivariate regressions, an extra year of education reduces fertility by `r cor_fert_edu_phen$white` for whites and `r cor_fert_edu_phen$black` for blacks [cf.
-@goldscheider1969minority; @johnson1979minority; @yang2003big].
-In turn, the simplest explanation for this fact would be point 2 of the theory.
-Because black people on average have lower incomes and fewer years of education, and are more likely to be single parents, the fertility-wage tradeoff is especially acute for them [cf.
-@maralani2013demography].
+# this was an attempt to see if black coefs were larger in abs size than
+# white coefs. Results were always absolutely insignificant (and conf
+# intervals too wide to be informative).
+
+n_reps_eth_diff <- 99
+rand_boot <- rand |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps_eth_diff)
+
+calc_eth_diff <- function (weights, data) {
+  coef_pgs_white <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) lm(x, data, weights = weights, 
+                    subset = ethnicity == "white")) |> 
+        map_dbl(\(x) coef(x)[[2]])
+  coef_pgs_black <- pgs |> 
+        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) lm(x, data, weights = weights, 
+                    subset = ethnicity == "black")) |> 
+        map_dbl(\(x) coef(x)[[2]])
+
+  # this is positive if white was "bigger" than black taking the white sign 
+  # as the correct sign. Negative if b was "bigger"  than w.
+  diff_coefs <- (coef_pgs_white - coef_pgs_black)*sign(coef_pgs_white)
+  names(diff_coefs) <- pgs
+  mean(diff_coefs)
+}
+
+diff_eths <- withReplicates(rand_boot, theta = calc_eth_diff)
+```
+
+```{r}
+#| label: stat-post-bootstraps
+
+# separate so we don't interfere with the bootstrap cache
+
+ci_boot_fert <- map(cors_fert_edu, \(x) confint_workaround(x)[1,])
+ci_boot_sibs <- map(cors_sibs_edu, \(x) confint_workaround(x)[1,])
+
+ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint_workaround(x)[2,])
+ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint_workaround(x)[2,])
+
+stopifnot(significant(ci_boot_fert$white))
+stopifnot(insignificant(ci_boot_fert$black))
+stopifnot(insignificant(ci_boot_sibs$white))
+stopifnot(insignificant(ci_boot_sibs$black))
+```
+
+@fig-fertility-edu plots each PGS's partial correlation with fertility against its partial correlation with educational attainment.
+The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrapped 95% C.I.
+`r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
+Bootstraps are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).
 
@@
 Appendix @fig-siblings plots effects on number of siblings versus effects on years of education.
-Correlations are insignificant for both ethnic groups.
+Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).
 
-I next test part 2 of the theory more directly by interacting PGS with measures of education, income and marital status.
+I next test part 2 of the theory more directly by interacting PGS with measures of education, income, marital status, and age at first birth.
 Education is years of education, split at 12 years.
 Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split.
+From here on I only use the white subsample: there are too few black respondents to be informative.
 
@@
 #| cache-lazy: false
+#| eval: false
 
-n_reps_x <- 199
+# all confindence intervals were too wide to be informative
+n_reps_x <- 99
 
@@
 
-  cor_ed_low - cor_ed_high
+  c(low = cor_ed_low, high = cor_ed_high, diff = cor_ed_low - cor_ed_high)
 }
 
-diff_cor_ed_ed <- diff_cor_ed_inc <- diff_cor_ed_mar <- diff_cor_ed_agefb <-
-  list()
+diff_cor_ed <- ci <- est <- tbl <- list()
 
+terms <- c("edu", "income_med", "married", "agefb")
 for (eth in eths) {
@@
 
-  diff_cor_ed_ed[[eth]] <- 
-    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "edu")
-  diff_cor_ed_inc[[eth]] <- 
-    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "income_med")
-  diff_cor_ed_mar[[eth]] <- 
-    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "married")
-  diff_cor_ed_agefb[[eth]] <-
-    withReplicates(rand_boot, theta = calc_diff_cor_ed, term = "agefb")
+  for (term in terms) {
+    diff_cor_ed[[eth]][[term]] <- 
+      withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
+    ci[[eth]][[term]] <- confint_workaround(diff_cor_ed[[eth]][[term]])
+    est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
   }
-
-ci_ed <- map(diff_cor_ed_ed, confint_workaround)
-ci_inc <- map(diff_cor_ed_inc, confint_workaround)
-ci_mar <- map(diff_cor_ed_mar, confint_workaround)
-ci_agefb <- map(diff_cor_ed_agefb, confint_workaround)
+}
 
@@
 #   cause a lower correlation with edyrs in the 13-16 group?
-stopifnot(significant(ci_ed$white))
-stopifnot(insignificant(ci_ed$black))
-stopifnot(all(map_lgl(ci_inc, insignificant)))
-stopifnot(all(map_lgl(ci_mar, insignificant)))
-stopifnot(all(map_lgl(ci_agefb, insignificant)))
 
@@
 
-@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education.
-@fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks.
-@fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
-Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents.
-@fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
+Appendix @fig-x-edu shows essentially no difference in the distribution of absolute effect sizes between respondents with more and less education.
+Appendix @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents.
+Appendix @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
+Effect sizes appear larger for unmarried respondents.
+Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
 There are no large differences between early and late parents.
@@
 
-The economic theory implies that correlations between education and fertility should be higher for lower-income/education people, single parents and people who have children earlier.
-This in turn implies that correlations between PGS effects on education and on fertility should be higher among these groups.
-I test this with bootstraps, calculating effect sizes and correlations within each pair of groups, and taking the difference between them.
-Correlations are indeed higher among low- than high-education whites (bootstrapped C.I.
-for difference in correlations `r ci_ed$white[1]` to `r ci_ed$white[2]`) but for all other pairs of groups the difference is not significantly different from zero.
-In fact, the width of confidence intervals shows that we lack power to detect all but large differences.
+The economic theory of fertility in @hugh2022human implies that correlations between education and fertility should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS effects on education and on fertility should be more negative among these groups.
+I tested this with bootstraps, but confidence intervals were always too wide to be informative.
 
-```{r}
-#| label: fig-x-edu
-#| fig-cap: |
-#|   Violin plots of polygenic score effects on fertility by education. 
-#|   Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high
-#|   education group. Lines show quartiles.
-
-tidy_pgs_x_ed |> 
-  list_rbind(names_to = "ethnicity") |> 
-  mutate(
-    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_ed, ethnicity)),
-    Education = str_remove(edu, "edu")
-  ) |> 
-  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
-    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
-    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
-    facet_wrap(vars(Ethnicity), scales = "free_y") +
-    labs(
-      x = "Years of education",
-      y = "Absolute effect size"
-    ) +
-    theme_minimal() +
-    theme(
-      legend.position = "none"
-    )
-```
-
-```{r}
-#| label: fig-x-inc
-#| fig-cap: |
-#|   Violin plots of polygenic score effects on fertility by income. 
-#|   Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Lines show quartiles.
-
-tidy_pgs_x_inc |> 
-  list_rbind(names_to = "ethnicity") |> 
-  mutate(
-    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_inc, ethnicity)),
-    Income = fct_relevel(income_med, c("Low", "High"))
-  ) |> 
-  ggplot(aes(Income, abs(estimate), color = Income, fill = Income)) +
-    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
-    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
-    facet_wrap(vars(Ethnicity), scales = "free_y") +
-    labs(
-      x = "Income",
-      y = "Absolute effect size"
-    ) +
-    theme_minimal() +
-    theme(
-      legend.position = "none"
-    )
-```
-
-```{r}
-#| label: fig-x-mar
-#| fig-cap: |
-#|   Violin plots of polygenic score effects on fertility by marital status. 
-#|   Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Lines show quartiles.
-
-tidy_pgs_x_mar |> 
-    list_rbind(names_to = "ethnicity") |> 
-    mutate(
-      Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_mar, ethnicity)),
-      married = fct_relevel(married, "Other", "Married")
-    ) |> 
-    ggplot(aes(married, abs(estimate), color = married, fill = married)) +
-      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
-      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
-      facet_wrap(vars(Ethnicity), scales = "free_y") +
-      labs(
-          x = "Married",
-          y = "Absolute effect size"
-      ) +
-      theme_minimal() +
-      theme(
-          legend.position = "none"
-      )
-
-```
-
-```{r}
-#| label: fig-x-agefb
-#| fig-cap: |
-#|   Violin plots of polygenic score effects on fertility by age at 
-#|   first birth (median split). Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Lines show quartiles.
-
-tidy_pgs_x_agefb |> 
-    list_rbind(names_to = "ethnicity") |> 
-    mutate(
-     Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_agefb, ethnicity)),
-      agefb = fct_relevel(agefb, "Low", "High")
-    ) |> 
-    ggplot(aes(agefb, abs(estimate), color = agefb, fill = agefb)) +
-      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
-      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
-      facet_wrap(vars(Ethnicity), scales = "free_y") +
-      labs(
-          x = "Age at first birth",
-          y = "Absolute effect size"
-      ) +
-      theme_minimal() +
-      theme(
-          legend.position = "none"
-      )
-
-```
+[^2]: See @hugh2022human equation (6) and following.
 
@@
 This is partly due to the low sample size.
-But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible in this data.
-Lastly, the smaller black sample makes most tests inconclusive for this population.
-
-\FloatBarrier
+But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
+Lastly, the smaller black sample makes most tests inconclusive for this population; I can only say that the data do not reject a negative association between PGS correlations with fertility and on education.
 
@@
 #| fig-cap: |
-#|  Effects of PGS on realized fertility. Values are partial correlations 
-#|  of a one standard deviation (within ethnicity) change of the PGS with
-#|  fertility, controlling for 10 principal components of genomic array 
-#|  data. Scales differ between the two charts.
+#|  Effects of PGS on realized fertility among white respondents. 
+#|  Values are partial correlations of a one standard deviation 
+#|  (within ethnicity) change of the PGS with fertility, controlling 
+#|  for 10 principal components of genomic array data. 
 #| fig-height: 6
 
-tidy_pgs_eth |>
+tidy_pgs$white |>
   mutate(
     term = nice_names[term],
-    term = fct_reorder(term, estimate, \(x) x[2]),
-    ethnicity = ifelse(ethnicity == "white", "Whites (non-Hispanic)", "Blacks"),
-    # hack to order scales separately from 
-    # https://stackoverflow.com/questions/26238687/r-reorder-facet-wrapped-x-axis-with-free-x-in-ggplot2
-    term_mangled = paste(ethnicity, term, sep = "//"),
-    term_mangled = fct_reorder(term_mangled, estimate),
+    term = fct_reorder(term, estimate)
   ) |>
-  ggplot(aes(estimate, term_mangled, color = ethnicity)) +
+  ggplot(aes(estimate, term)) +
     geom_vline(xintercept = 0) +
-    geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
-                    position = position_dodge(width = 0.5), 
-                    alpha = 0.8, size = 0.3) +
-    facet_wrap(vars(ethnicity), scales = "free") +
+    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.3) +
     theme_light() + 
-    scale_y_discrete(labels = \(x) str_remove(x, "^.*//")) +
-    scale_color_manual(values = c("orange2", "seagreen")) +
     labs(
@@
     theme(
+      axis.text.y = element_text(size = 7),
       panel.grid.major.y = element_blank(),
       panel.grid.minor.x = element_blank(),
-      legend.position = "none",
-      axis.text.y = element_text(size = 7),
-      strip.text = element_text(colour = "black", size = 10),
-      strip.background = element_rect(fill = NA)
+      legend.position = "none"
     )
@@
 
+```{r}
+#| label: prog-x-plots
+
+geom_x_boxplot <- geom_boxplot(color = "grey50", outlier.shape = NA)
+geom_x_point <- geom_point(aes(shape = pgs=="EA3"), 
+                           position = position_jitter(width = 0.03), 
+                           alpha = 0.5)
+
+themes_x <- theme_minimal() +
+      theme(
+          legend.position = "none",
+          panel.grid.major.x = element_blank()
+      ) 
+```
+
+```{r}
+#| label: fig-x-edu
+#| fig-cap: |
+#|   Boxplots of polygenic score effects on fertility by education
+#|   among white respondents. Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high
+#|   education group. Boxes show quartiles.
+
+tidy_pgs_x_ed$white |> 
+  mutate(
+    Education = str_remove(edu, "edu")
+  ) |> 
+  ggplot(aes(Education, abs(estimate), color = Education)) +
+    geom_x_boxplot + geom_x_point +
+    labs(
+      x = "Years of education",
+      y = "Absolute effect size",
+      title = paste("PGS effects on fertility by education", 
+                    calc_n(mod_pgs_x_ed))
+    ) +
+    themes_x
+```
+
+```{r}
+#| label: fig-x-inc
+#| fig-cap: |
+#|   Boxplots of polygenic score effects on fertility by income
+#|   among white respondents. Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Boxes show quartiles.
+
+tidy_pgs_x_inc$white |> 
+  mutate(
+    Income = fct_relevel(income_med, c("Low", "High"))
+  ) |> 
+  ggplot(aes(Income, abs(estimate), color = Income)) +
+    geom_x_boxplot + geom_x_point +
+    labs(
+      x = "Income",
+      y = "Absolute effect size",
+      title = paste("PGS effects on fertility by income", 
+                    calc_n(mod_pgs_x_inc))
+    ) +
+    themes_x
+```
+
+```{r}
+#| label: fig-x-mar
+#| fig-cap: |
+#|   Boxplots of polygenic score effects on fertility by marital status
+#|   among white respondents. Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Boxes show quartiles.
+
+tidy_pgs_x_mar$white |> 
+    mutate(
+      married = fct_relevel(married, "Other", "Married")
+    ) |> 
+    ggplot(aes(married, abs(estimate), color = married)) +
+      geom_x_boxplot + geom_x_point +
+      labs(
+          x = "Married",
+          y = "Absolute effect size",
+          title = paste("PGS effects on fertility by marital status", 
+                        calc_n(mod_pgs_x_mar))
+      ) +
+      themes_x
+
+```
+
+```{r}
+#| label: fig-x-agefb
+#| fig-cap: |
+#|   Boxplots of polygenic score effects on fertility by age at 
+#|   first birth (median split) among white respondents. Effect sizes are
+#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
+#|   change of the PGS on fertility, controlling for 10 principal 
+#|   components of genetic array data, estimated within the low/high income
+#|   group. Boxes show quartiles.
+
+tidy_pgs_x_agefb$white |> 
+    mutate(
+      agefb = fct_relevel(agefb, "Low", "High")
+    ) |> 
+    ggplot(aes(agefb, abs(estimate), color = agefb)) +
+      geom_x_boxplot + geom_x_point +
+      labs(
+          x = "Age at first birth",
+          y = "Absolute effect size",
+          title = paste("PGS effects on fertility by age at first birth", 
+                        calc_n(mod_pgs_x_agefb))
+      ) +
+      themes_x
+
+```
+
 \FloatBarrier

### 2023-09-19 hrs-selection.qmd
Add bootstraps and table of mean absolute effect size differences

@@
 #| label: stat-bootstrap-eth-cor
+#| cache: true
+#| cache-lazy: false
 
@@
 ```{r}
-#| label: stat-x-bootstraps
+#| label: stat-x-bootstrap-cors
 #| cache: true
@@
 
+```{r}
+#| label: stat-x-bootstrap-effect-size
+#| cache: true
+#| cache-lazy: false
+
+n_reps_x <- 99
+
+calc_diff_effect_size <- function (w, data, term) {
+  groups <- switch(term,
+                edu = c("0-12", "13-17"),
+                income_med = c("Low", "High"),
+                married = c("Other", "Married"),
+                agefb = c("Low", "High"))
+  groups <- paste0(":", term, groups) # colon for the cross term
+  low_group <- groups[1]
+  high_group <- groups[2]
+
+  forms <- paste0(pgs, ":", term) |>
+    map(\(x) reformulate(c(x, term, pcs), response = "raevbrn"))
+  mods <- map(forms, \(x) lm(x, data, weights = w))
+  cfs <- map(mods, coef)
+  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
+  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])
+
+  abs_eff_low <- mean(abs(cfs_low))
+  abs_eff_high <- mean(abs(cfs_high))
+  diff_abs_eff <- abs_eff_low - abs_eff_high
+  c(low = abs_eff_low, high = abs_eff_high, diff = diff_abs_eff)
+}
+
+diff_effect_size <- ci <- est <- tbl <- list()
+
+terms <- c("edu", "income_med", "married", "agefb")
+
+rand_boot <- rand |> 
+  subset(ethnicity == "white") |> 
+  as.svrepdesign(type = "bootstrap", replicates = n_reps_x)
+
+for (term in terms) {
+  diff_effect_size[[term]] <- 
+    withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
+  ci[[term]] <- confint_workaround(diff_effect_size[[term]])
+  est[[term]] <- as.numeric(diff_effect_size[[term]])
+  tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
+                         ci[[term]][, 1], ci[[term]][, 2])
+}
+
+```
+
+```{r}
+#| label: stat-post-x-bootstrap
+
+stopifnot(insignificant(ci$edu[3,]))
+stopifnot(significant(ci$income_med[3,]))
+stopifnot(significant(ci$married[3,]))
+stopifnot(insignificant(ci$agefb[3,]))
+```
+
 Appendix @fig-x-edu shows essentially no difference in the distribution of absolute effect sizes between respondents with more and less education.
@@
 Note that the N is lower here due to missing values.
+@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute effect sizes in the two groups, and the difference between them, using bootstraps.
+Differences for income and marriage are significant.
+
+|               | Low group             | High group            | Difference            |
+|------------------|------------------|------------------|------------------|
+| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
+| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
+| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
+| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
+
+: Bootstrap estimates and 95% confidence intervals for mean absolute effect sizes of PGS on fertility in "low" and "high" groups among white respondents.
+Groups are: 0-12 years education vs.
+13-17 years; below vs.
+above median income; all others vs.
+married; below vs.
+above median age at first birth.
+`r n_reps_x` bootstraps.
+{#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}
 
@@
 PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
-However, there is little evidence that PGS effects on fertility are larger, or more correlated with effects on education, among low-education or low-income individuals, unmarried parents, or younger parents.
-This is partly due to the low sample size.
+PGS effects on fertility also appear larger for low-income groups and unmarried respondents.
+However, there is little evidence for larger effects among people with lower education or younger parents.
+This may be due to the low sample size.
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
-Lastly, the smaller black sample makes most tests inconclusive for this population; I can only say that the data do not reject a negative association between PGS correlations with fertility and on education.
+Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with fertility and on education.
 

### 2023-09-19 hrs-selection.qmd
Return to RLRS as a dep var

@@
   scores which correlate negatively (positively) with education are
-  selected for (against). However, unlike in the UK, there is little evidence
-  that selection effects are larger among low-income or less educated people.
+  selected for (against). Selection effects are larger among low-income
+  and unmarried parents, but not among younger parents or those with less 
+  education.
 date: "September 2023"
@@
 rand <- rand |> 
-  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, raedyrs,
-         r10sibs, edu, income_med, married, agefb, rabyear,
+  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
+         raedyrs, r10sibs, edu, income_med, married, agefb, rabyear,
          raehsamp, raestrat, parent_weight)
@@
 
-Here, I test the same theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
-This gives an independent test of the theory.
+Here, I make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-Also, the HRS provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the human capital theory of natural selection works in both these ethnicities.
-
-Economic theory treats all humans as the same.
-Once you have controlled appropriately for people's preferences, resources and beliefs, there should be no difference between ethnicities.
-So, the economic theory of fertility predicts that patterns of natural selection should be the same in both white and black ethnic groups.
-Of course, ethnic groups may have different economic characteristics.
-Also, most of the polygenic scores were created using people of European ancestry.
-This means they are less predictive for people with African ancestry.
-If natural selection works via scores' correlation with phenotypes, then we would expect these correlations to be smaller among black people.
+The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the human capital theory predicts patterns of natural selection in both these ethnicities.
 
@@
 PGS are rescaled to zero mean and unit variance within each ethnic group, so effect sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
-In all regressions using PGS, I control for ten principal components of the DNA array data.
-
-I use realised fertility (number of children ever born) as the key dependent variable.
-I use this rather than relative lifetime reproductive success for ease of interpretation and because I don't have data for dates when children were born.
-My focus is on examining patterns of natural selection across PGS, rather than estimating their effects on PGS over time; raw fertility is good enough for this.
+In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.
+
+The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by mean number of children of people born in the same year.
+RLRS is calculated pooling ethnicities, i.e.
+treating them as members of the same biological population.
 
@@
 
-  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
+  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
   mod_pgs[[eth]] <- map(form_pgs, 
@@
   form_pgs_x_ed <- paste0(pgs, ":edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "raevbrn"))
+    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
   mod_pgs_x_ed[[eth]]  <-
@@
   form_pgs_x_inc <- paste0(pgs, ":income_med") |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "raevbrn"))
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
   mod_pgs_x_inc[[eth]]  <-
@@
   form_pgs_x_mar <- paste0(pgs, ":married") |>
-    map(\(x) reformulate(c(x, "married", pcs), response = "raevbrn"))
+    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
   mod_pgs_x_mar[[eth]]  <-
@@
   form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
-    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
   mod_pgs_x_agefb[[eth]]  <-
@@
   cor_fert_edu_phen[[eth]] <- 
-    svyglm_quiet(raevbrn ~ raedyrs, r_eth) |> 
+    svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
     coef() |> 
@@
 n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)
-if (n_sig_bonf != 0) stop("More/less than 0 score was Bonferroni significant")
 
@@
   coef_pgs_white <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
   coef_pgs_black <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "raevbrn")) |> 
+        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
 
-I estimated effect sizes of PGS correlations with realised fertility among black and white respondents separately.[^1] Appendix @fig-fertility shows effect sizes for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
-Standard errors are large because of the relatively low sample sizes, and no scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
+I estimated effect sizes of PGS correlations with RLRS among black and white respondents separately.[^1] Appendix @fig-rlrs shows effect sizes for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
+Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
 But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
 
-[^1]: Throughout, I use "effect" as shorthand for "regression coefficient".
+[^1]: Throughout, "effect" is shorthand for "regression coefficient".
 No claim about causality is implied.
@@
   cors_fert_edu[[eth]] <- 
-    withReplicates(rand_boot, theta = calc_cors_edu, var = "raevbrn")
+    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
   cors_sibs_edu[[eth]] <- 
@@
 
-@fig-fertility-edu plots each PGS's partial correlation with fertility against its partial correlation with educational attainment.
+@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment.
 The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I.
@@
 ```{r}
-#| label: fig-fertility-edu
+#| label: fig-rlrs-edu
 #| fig-cap: |
-#|   Scatterplot of PGS effect sizes on fertility and years of
+#|   Scatterplot of PGS effect sizes on RLRS and years of
 #|   education. Controls include 10 principal components of genetic array
@@
             x = "Effect on years of education",
-            y = "Effect on realized fertility"
+            y = "Effect on RLRS"
           )
@@
   forms <- paste0(pgs, ":", term) |>
-    map(\(x) reformulate(c(x, term, pcs), response = "raevbrn"))
+    map(\(x) reformulate(c(x, term, pcs), response = "rlrs"))
   mods <- map(forms, \(x) lm(x, data, weights = w))
@@
 
-: Bootstrap estimates and 95% confidence intervals for mean absolute effect sizes of PGS on fertility in "low" and "high" groups among white respondents.
+: Bootstrap estimates and 95% confidence intervals for mean absolute effect sizes of PGS on RLRS in "low" and "high" groups among white respondents.
 Groups are: 0-12 years education vs.
@@
 
-The economic theory of fertility in @hugh2022human implies that correlations between education and fertility should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS effects on education and on fertility should be more negative among these groups.
+The economic theory of fertility in @hugh2022human implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS effects on education and on RLRS should be more negative among these groups.
 I tested this with bootstraps, but confidence intervals were always too wide to be informative.
@@
 PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
-PGS effects on fertility also appear larger for low-income groups and unmarried respondents.
+PGS effects on RLRS also appear larger for low-income groups and unmarried respondents.
 However, there is little evidence for larger effects among people with lower education or younger parents.
@@
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
-Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with fertility and on education.
+Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
 
@@
 ```{r}
-#| label: fig-fertility
+#| label: fig-rlrs
 #| fig-cap: |
-#|  Effects of PGS on realized fertility among white respondents. 
+#|  Effects of PGS on RLRS among white respondents. 
 #|  Values are partial correlations of a one standard deviation 
-#|  (within ethnicity) change of the PGS with fertility, controlling 
+#|  (within ethnicity) change of the PGS with RLRS, controlling 
 #|  for 10 principal components of genomic array data. 
@@
       x = "", y = "",
-      title = "Effects of polygenic scores on fertility"
+      title = "Effects of polygenic scores on RLRS"
     ) +
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on fertility by education
+#|   Boxplots of polygenic score effects on RLRS by education
 #|   among white respondents. Effect sizes are
 #|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
+#|   change of the PGS on RLRS, controlling for 10 principal 
 #|   components of genetic array data, estimated within the low/high
@@
       y = "Absolute effect size",
-      title = paste("PGS effects on fertility by education", 
+      title = paste("PGS effects on RLRS by education", 
                     calc_n(mod_pgs_x_ed))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on fertility by income
+#|   Boxplots of polygenic score effects on RLRS by income
 #|   among white respondents. Effect sizes are
 #|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
+#|   change of the PGS on RLRS, controlling for 10 principal 
 #|   components of genetic array data, estimated within the low/high income
@@
       y = "Absolute effect size",
-      title = paste("PGS effects on fertility by income", 
+      title = paste("PGS effects on RLRS by income", 
                     calc_n(mod_pgs_x_inc))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on fertility by marital status
+#|   Boxplots of polygenic score effects on RLRS by marital status
 #|   among white respondents. Effect sizes are
 #|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
+#|   change of the PGS on RLRS, controlling for 10 principal 
 #|   components of genetic array data, estimated within the low/high income
@@
           y = "Absolute effect size",
-          title = paste("PGS effects on fertility by marital status", 
+          title = paste("PGS effects on RLRS by marital status", 
                         calc_n(mod_pgs_x_mar))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on fertility by age at 
+#|   Boxplots of polygenic score effects on RLRS by age at 
 #|   first birth (median split) among white respondents. Effect sizes are
 #|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on fertility, controlling for 10 principal 
+#|   change of the PGS on RLRS, controlling for 10 principal 
 #|   components of genetic array data, estimated within the low/high income
@@
           y = "Absolute effect size",
-          title = paste("PGS effects on fertility by age at first birth", 
+          title = paste("PGS effects on RLRS by age at first birth", 
                         calc_n(mod_pgs_x_agefb))

### 2023-09-19 hrs-selection.qmd
Use "coefficient" not "effect"

@@
   scores which correlate negatively (positively) with education are
-  selected for (against). Selection effects are larger among low-income
+  selected for (against). Selection coefficients are larger among low-income
   and unmarried parents, but not among younger parents or those with less 
@@
 
-PGS are rescaled to zero mean and unit variance within each ethnic group, so effect sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
+PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
 In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.
@@
 
-I estimated effect sizes of PGS correlations with RLRS among black and white respondents separately.[^1] Appendix @fig-rlrs shows effect sizes for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
+I estimated coefficients of PGS on RLRS among black and white respondents separately.
+These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
+Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
 Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
 But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
-
-[^1]: Throughout, "effect" is shorthand for "regression coefficient".
-No claim about causality is implied.
-Recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
 
@@
 #| fig-cap: |
-#|   Scatterplot of PGS effect sizes on RLRS and years of
+#|   Scatterplot of PGS regression coefficients on RLRS and years of
 #|   education. Controls include 10 principal components of genetic array
@@
             title = titles[eth],
-            x = "Effect on years of education",
-            y = "Effect on RLRS"
+            x = "Coefficient on years of education",
+            y = "Coefficient on RLRS"
           )
@@
 I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
-Parents of no siblings cannot be included, so effect sizes are not comparable across the generations.
-Appendix @fig-siblings plots effects on number of siblings versus effects on years of education.
+Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations.
+Appendix @fig-siblings plots coefficents on number of siblings versus coefficients on years of education.
 Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
@@
 
-Appendix @fig-x-edu shows essentially no difference in the distribution of absolute effect sizes between respondents with more and less education.
-Appendix @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents.
-Appendix @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses.
-Effect sizes appear larger for unmarried respondents.
+Appendix @fig-x-edu shows essentially no difference in the distribution of absolute coefficient sizes between respondents with more and less education.
+Appendix @fig-x-inc, on the other hand, shows that coefficients appear larger among lower-income respondents.
+Appendix @fig-x-mar shows absolute coefficients for respondents who were married in 2010 against all other statuses.
+Coefficients appear larger for unmarried respondents.
 Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
@@
 Note that the N is lower here due to missing values.
-@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute effect sizes in the two groups, and the difference between them, using bootstraps.
+@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in the two groups, and the difference between them, using bootstraps.
 Differences for income and marriage are significant.
@@
 
-: Bootstrap estimates and 95% confidence intervals for mean absolute effect sizes of PGS on RLRS in "low" and "high" groups among white respondents.
+: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents.
 Groups are: 0-12 years education vs.
@@
 
-The economic theory of fertility in @hugh2022human implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS effects on education and on RLRS should be more negative among these groups.
+The economic theory of fertility in @hugh2022human implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups.
 I tested this with bootstraps, but confidence intervals were always too wide to be informative.
@@
 PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
-PGS effects on RLRS also appear larger for low-income groups and unmarried respondents.
-However, there is little evidence for larger effects among people with lower education or younger parents.
+PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents.
+But there is little evidence for larger coefficients among people with lower education, or younger parents.
 This may be due to the low sample size.
@@
 #| fig-cap: |
-#|  Effects of PGS on RLRS among white respondents. 
-#|  Values are partial correlations of a one standard deviation 
-#|  (within ethnicity) change of the PGS with RLRS, controlling 
-#|  for 10 principal components of genomic array data. 
+#|  Coefficients of PGS on RLRS among white respondents, controlling for 10
+#|  principal components of
+#|  genomic array data. Lines are 95% confidence intervals.
 #| fig-height: 6
@@
       x = "", y = "",
-      title = "Effects of polygenic scores on RLRS"
+      title = "Coefficients of polygenic scores on RLRS"
     ) +
@@
 #| fig-cap: |
-#|   Scatterplot of PGS effect sizes on number of live siblings and years of
+#|   Scatterplot of PGS coefficients on number of live siblings and years of
 #|   education. Controls include 10 principal components of genetic array
@@
           title = eth,
-          x = "Effect on years of education",
-          y = "Effect on number of siblings"
+          x = "Coefficient on years of education",
+          y = "Coefficient on number of siblings"
       )
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on RLRS by education
-#|   among white respondents. Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on RLRS, controlling for 10 principal 
+#|   Boxplots of polygenic score coefficients on RLRS 
+#|   among white respondents, controlling for 10 principal 
 #|   components of genetic array data, estimated within the low/high
@@
       x = "Years of education",
-      y = "Absolute effect size",
-      title = paste("PGS effects on RLRS by education", 
+      y = "Absolute coefficient",
+      title = paste("PGS coefficients on RLRS by education", 
                     calc_n(mod_pgs_x_ed))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on RLRS by income
-#|   among white respondents. Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on RLRS, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Boxes show quartiles.
+#|   Boxplots of polygenic score coefficients on RLRS 
+#|   among white respondents, controlling for 10 principal 
+#|   components of genetic array data, estimated within the below/above
+#|   median income group. Boxes show quartiles.
 
@@
       x = "Income",
-      y = "Absolute effect size",
-      title = paste("PGS effects on RLRS by income", 
+      y = "Absolute coefficient",
+      title = paste("PGS coefficients on RLRS by income", 
                     calc_n(mod_pgs_x_inc))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on RLRS by marital status
-#|   among white respondents. Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on RLRS, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Boxes show quartiles.
+#|   Boxplots of polygenic score coefficients on RLRS 
+#|   among white respondents, controlling for 10 principal 
+#|   components of genetic array data, estimated within married 
+#|   and other respondents. Boxes show quartiles.
 
@@
           x = "Married",
-          y = "Absolute effect size",
-          title = paste("PGS effects on RLRS by marital status", 
+          y = "Absolute coefficient",
+          title = paste("PGS coefficients on RLRS by marital status", 
                         calc_n(mod_pgs_x_mar))
@@
 #| fig-cap: |
-#|   Boxplots of polygenic score effects on RLRS by age at 
-#|   first birth (median split) among white respondents. Effect sizes are
-#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
-#|   change of the PGS on RLRS, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high income
-#|   group. Boxes show quartiles.
+#|   Boxplots of polygenic score coefficients on RLRS 
+#|   among white respondents, controlling for 10 principal 
+#|   components of genetic array data, estimated within below/above
+#|   median age at first birth respondents. Boxes show quartiles.
 
@@
           x = "Age at first birth",
-          y = "Absolute effect size",
-          title = paste("PGS effects on RLRS by age at first birth", 
+          y = "Absolute coefficient",
+          title = paste("PGS coefficients on RLRS by age at first birth", 
                         calc_n(mod_pgs_x_agefb))

### 2023-09-19 hrs-selection.qmd
Try out the futureverse...

@@
 library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
+library(future)
+library(furrr)
  # masks dotchart:
 suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))
+
+plan(multisession, workers = 6)
 
@@
 
+# workaround failure in confint.svrepdesign
 confint_workaround <- function (x) {
@@
   form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
-  mod_pgs[[eth]] <- map(form_pgs, 
-                        \(x) svyglm_quiet(x, r_eth)
-                        )
+  mod_pgs[[eth]] <- future_map(form_pgs, 
+                               \(x) svyglm_quiet(x, r_eth))
 
@@
                                               response = "raedyrs"))
-  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
-                              \(x) svyglm_quiet(x, r_eth)
-                              )
+  mod_pgs_edyrs[[eth]] <- future_map(form_pgs_edyrs, 
+                                     \(x) svyglm_quiet(x, r_eth))
 
@@
                        \(x) reformulate(c(x, pcs), response = "r10sibs"))
-  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
+  mod_pgs_sibs[[eth]] <- future_map(form_pgs_sibs,
                                     \(x) svyglm_quiet(x, r_parents_eth))
@@
   mod_pgs_x_ed[[eth]]  <-
-    map(form_pgs_x_ed, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+    future_map(form_pgs_x_ed, 
+               \(x) svyglm_quiet(x, r_eth))
 
@@
   mod_pgs_x_inc[[eth]]  <-
-    map(form_pgs_x_inc, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+    future_map(form_pgs_x_inc, 
+               \(x) svyglm_quiet(x, r_eth))
 
@@
   mod_pgs_x_mar[[eth]]  <-
-    map(form_pgs_x_mar, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+    future_map(form_pgs_x_mar, 
+               \(x) svyglm_quiet(x, r_eth))
 
@@
   mod_pgs_x_agefb[[eth]]  <-
-    map(form_pgs_x_agefb, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+    future_map(form_pgs_x_agefb, 
+               \(x) svyglm_quiet(x, r_eth))
 
@@
 #| cache-lazy: false
+#| eval: false
+
+# I don't mention this below so turning it off for now
+# TODO: should I mention it? It's a low correlation.
 
@@
 #| cache-lazy: false
+
+# use this rather than future_map, to avoid setup costs.
+# Not sure it actually does anything :-)
+options(survey.multicore = TRUE)
 

### 2023-09-20 hrs-selection.qmd
Don't use dead siblings (N reduced too much). Report agefb negative correlation in scores.
Give up on furrr. Figure out bug in confint.svrepdesign. Save time reading rand_fam. Skip many regressions in black sample.

--- 2023-09-19 hrs-selection.qmd
@@
 
-confint_workaround <- function (x) {
-  cbind(
-    as.numeric(x) - 1.96 * SE(x), 
-    as.numeric(x) + 1.96 * SE(x))
-}
-
 # rand HRS longit data. VERY BIG
@@
 # rand family respondents file
-rand_fam <- haven::read_dta(
-  "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
+# Done just once:
+# rand_fam <- haven::read_dta(
+#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
+# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
+rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
+
 # only selected columns to avoid creating suffixes when we join with
@@
                      labels = c("Other", "Married")),
-    r10sibs = r10livsib + r10sbdied + 1, # n. sibs including oneself!
+    r10sibs = r10livsib + 1, # n. sibs including oneself!
     agefb = r10agey_e - h10ageokid,
@@
   mod_pgs[[eth]] <- map(form_pgs, 
-                        \(x) svyglm_quiet(x, r_eth)
-                        )
+                               \(x) svyglm_quiet(x, r_eth))
 
@@
   mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
-                              \(x) svyglm_quiet(x, r_eth)
-                              )
+                                     \(x) svyglm_quiet(x, r_eth))
 
@@
 
+  cor_fert_edu_phen[[eth]] <- 
+    svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
+    coef() |> 
+    pluck("raedyrs")
+}
+
+# we only look at interactions among whites, the rest is pointless
+for (eth in "white") {
+  r_eth <- rand |> subset(ethnicity == eth)
+
   # the colon interaction estimates the effect separately within each group
@@
     map(form_pgs_x_ed, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+               \(x) svyglm_quiet(x, r_eth))
 
@@
     map(form_pgs_x_inc, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+               \(x) svyglm_quiet(x, r_eth))
 
@@
     map(form_pgs_x_mar, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+               \(x) svyglm_quiet(x, r_eth))
 
@@
     map(form_pgs_x_agefb, 
-        \(x) svyglm_quiet(x, r_eth)
-        )
+               \(x) svyglm_quiet(x, r_eth))
 
@@
     ) 
-
-  cor_fert_edu_phen[[eth]] <- 
-    svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
-    coef() |> 
-    pluck("raedyrs")
-}
-
-tidy_pgs_eth <- bind_rows(tidy_pgs,  .id = "ethnicity")
-
+}
+
+tidy_pgs_eth <- list_rbind(tidy_pgs)
 n_tests <- nrow(tidy_pgs_eth)
@@
 
-  cor(coef_pgs_black, coef_pgs_white)
+  c(cor = cor(coef_pgs_black, coef_pgs_white))
 }
@@
 cor_eths_boot <- withReplicates(rand_boot, theta = calc_eth_cor)
-ci_eths <- confint_workaround(cor_eths_boot)[1,]
+ci_eths <- confint(cor_eths_boot)[1,]
 ```
@@
 #| cache-lazy: false
-
+#| cache-comments: false
+
+# Not sure this actually does anything :-)
+old_opts <- options(survey.multicore = TRUE)
+
+# The large n_reps is to get an accurate upper tail on the black sample, 
+# since it is close to 0:
 n_reps <- 599
@@
   c(
-    cor(coef_pgs_edu, coef_pgs_fert),
-    coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
+    cor = cor(coef_pgs_edu, coef_pgs_fert),
+    coef = coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
   )
@@
 
+options(old_opts)
 ```
@@
   names(diff_coefs) <- pgs
-  mean(diff_coefs)
+  c(diff = mean(diff_coefs))
 }
@@
 
-ci_boot_fert <- map(cors_fert_edu, \(x) confint_workaround(x)[1,])
-ci_boot_sibs <- map(cors_sibs_edu, \(x) confint_workaround(x)[1,])
-
-ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint_workaround(x)[2,])
-ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint_workaround(x)[2,])
-
-stopifnot(significant(ci_boot_fert$white))
-stopifnot(insignificant(ci_boot_fert$black))
-stopifnot(insignificant(ci_boot_sibs$white))
-stopifnot(insignificant(ci_boot_sibs$black))
+ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
+ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
+
+ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
+ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
+
+if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
+if (! insignificant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
+if (significant(ci_boot_sibs$white)) warning("ci_boot_sibs$white significant")
+if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
 ```
@@
 
-I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010.
-I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
+I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010.
+(Data for dead siblings has too many missing values to use.) I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
 Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations.
-Appendix @fig-siblings plots coefficents on number of siblings versus coefficients on years of education.
+Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
 Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
@@
 #| eval: false
+
+old_opts <- options(survey.multicore = TRUE)
 
@@
       withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
-    ci[[eth]][[term]] <- confint_workaround(diff_cor_ed[[eth]][[term]])
+    ci[[eth]][[term]] <- confint(diff_cor_ed[[eth]][[term]])
     est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
@@
 
-# TODO: does splitting by education to 12 and education 13-16 mechanically
-#   cause a lower correlation with edyrs in the 13-16 group?
+options(old_opts)
 
@@
 #| cache-lazy: false
-
-n_reps_x <- 99
+#| cache-comments: false
+
+old_opts <- options(survey.multicore = TRUE)
+
+n_reps_x_es <- 99
 
@@
   subset(ethnicity == "white") |> 
-  as.svrepdesign(type = "bootstrap", replicates = n_reps_x)
+  as.svrepdesign(type = "bootstrap", replicates = n_reps_x_es)
 
@@
     withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
-  ci[[term]] <- confint_workaround(diff_effect_size[[term]])
+  ci[[term]] <- confint(diff_effect_size[[term]])
   est[[term]] <- as.numeric(diff_effect_size[[term]])
@@
 }
+
+options(old_opts)
+
+```
+
+```{r}
+#| label: stat-agefb-bootstrap
+
+n_reps_agefb <- 9
+calc_cor_agefb <- function (w, data) {
+  forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
+  mods_low <- map(forms, \(x) lm(x, data, subset = agefb == "Low",
+                                 weights = w))
+  mods_high <- map(forms, \(x) lm(x, data, subset = agefb == "High",
+                                  weights = w))
+  coefs_low <- map2_dbl(mods_low, pgs, \(x, y) coef(x)[[y]])
+  coefs_high <- map2_dbl(mods_high, pgs, \(x, y) coef(x)[[y]])
+
+  c(
+    cor = cor(coefs_low, coefs_high), 
+    discordant = sum(sign(coefs_low) != sign(coefs_high)) # discordant pairs
+  )
+}
+
+rand_boot <- rand |> 
+  subset(ethnicity == "white") |> 
+  as.svrepdesign(type = "bootstrap", replicates = n_reps_agefb)
+
+cor_agefb <- withReplicates(rand_boot, theta = calc_cor_agefb)
+
+stopifnot(insignificant(confint(cor_agefb)[1,]))
+ci_discordant <- confint(cor_agefb)[2,]
+# significantly different from 1/2?
+stopifnot(ci_discordant[1] <= length(pgs)/2 && 
+            ci_discordant[2] >= length(pgs)/2)
 
@@
 stopifnot(insignificant(ci$agefb[3,]))
+
+tidy_agefb_wide <- tidy_pgs_x_agefb$white |> 
+  pivot_wider(id_cols = pgs, names_from = agefb, values_from = estimate)
+
+cor_agefb <- cor(tidy_agefb_wide$Low, tidy_agefb_wide$High)
+n_discordant <- sum(sign(tidy_agefb_wide$Low) != sign(tidy_agefb_wide$High))
 ```
@@
 Coefficients appear larger for unmarried respondents.
+
 Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
-There are no large differences between early and late parents.
+There are no large differences between younger and older parents.
+As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs.
+However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively.
 Note that the N is lower here due to missing values.
-@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in the two groups, and the difference between them, using bootstraps.
+
+@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in each pair of groups, and the difference between them, using bootstraps.
 Differences for income and marriage are significant.
@@
 above median age at first birth.
-`r n_reps_x` bootstraps.
+`r n_reps_x_es` bootstraps.
 {#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}

### 2023-09-21 hrs-selection.qmd
Add conclusion para; mention regs controlling for age at first birth

@@
     r10sibs = r10livsib + 1, # n. sibs including oneself!
-    agefb = r10agey_e - h10ageokid,
-    agefb = ifelse(agefb < 12, NA_real_, agefb),
-    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
+    agefbn = r10agey_e - h10ageokid,
+    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
+  ) |> 
+  mutate(.by = c(ethnicity, ragender),
+    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
   )
@@
   select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
-         raedyrs, r10sibs, edu, income_med, married, agefb, rabyear,
-         raehsamp, raestrat, parent_weight)
+         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
+         rabyear, raehsamp, raestrat, parent_weight, ragender)
 
@@
 
-1.  There is a trade-off between time spent working and raising children. This leads people with more human capital and higher expected wages to have fewer children.
-2.  The trade-off is sharper for low-income people, people with low human capital, and single parents. As a result, natural selection is stronger among these groups.
-
-The evidence for point 1 is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e.
-they are being selected against.
-And vice versa, scores which correlate positively with human capital are being selected for.
-The evidence for point 2 is that correlations with number of children are larger and more significant among people with lower income, less education and single parents.
+1.
+There is a trade-off between time spent working and raising children.
+This leads people with more human capital and higher expected wages to have fewer children.
+Evidence for this is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e.
+they are being selected against; conversely, scores which correlate positively with human capital are being selected for.
+2.
+The trade-off is sharper for low-income people, people with low human capital, and single parents.
+As a result, natural selection is stronger among these groups.
+Evidence for this is that scores' regression coefficients on number of children are larger among people with lower income or less education, and single parents.
 
@@
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the human capital theory predicts patterns of natural selection in both these ethnicities.
+The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the theory predicts patterns of natural selection in both these ethnicities.
 
@@
 cors_sibs_edu <- list()
-cor_fert_edu_phen <- list()
+tidy_fert_edu_phen <- list()
 
@@
 
-  cor_fert_edu_phen[[eth]] <- 
+  tidy_fert_edu_phen[[eth]] <- 
     svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
-    coef() |> 
-    pluck("raedyrs")
+    broom::tidy(conf.int = TRUE) |> 
+    filter(term == "raedyrs")
 }
@@
 
-Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity).
+Appendix @fig-x-agefb splits respondents by age at first birth (median within each gender).
+Note that the N is lower here due to missing values.
 There are no large differences between younger and older parents.
@@
 However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively.
-Note that the N is lower here due to missing values.
+Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with uncontrolled coefficient; here, that isn't true, with coefficients remaining positively correlated.
+
+```{r}
+#| label: stat-control-agefb
+#| eval: false
+
+# cselection effects ontrolling for age at first birth
+# 
+form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
+                                                    response = "rlrs"))
+mod_pgs_control_agefb <- map(form_pgs_control_agefb,
+                             \(x) svyglm_quiet(x, rand, subset = 
+                                                 ethnicity == "white"))
+tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
+                          list_rbind() |> 
+                          filter(term %in% pgs)
+
+cor(tidy_pgs$white$estimate, tidy_pgs_control_agefb$estimate)
+
+```
 
@@
 Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
+
+Why does the US data show fewer differences by socio-economic status (SES) than the UK?
+One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
+The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
+@goldscheider1969minority, @johnson1979minority, @yang2003big).
+And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu).
+But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and by their different predictive power, so this hypothesis can only be speculative.
 

### 2023-09-22 hrs-selection.qmd
Rearrange a para. Add R file extrapolating to effect sizes of true pgs

@@
                                      \(x) svyglm_quiet(x, r_eth))
+  names(mod_pgs_edyrs[[eth]]) <- pgs
 
@@
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
-Bootstraps are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).
+Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).
 
@@
 
-I next test part 2 of the theory more directly by interacting PGS with measures of education, income, marital status, and age at first birth.
+I next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth.
 Education is years of education, split at 12 years.
@@
 
+```{r}
+#| label: stat-pgs-over-time
+#| eval: false
+
+# Plot changes in a couple of variables by decade
+# Below: "Changes in PGS over time are small for both groups..."
+
+rand$variables$birth_decade <- chop_width(rand$variables$rabyear, 10, 1920)
+decades_adhd <- svyby(~ADHD_PGC17, ~ birth_decade + ethnicity, rand, 
+                      FUN = svymean) 
+decades_ea3 <- svyby(~EA3, ~ birth_decade + ethnicity, rand, 
+                     FUN = svymean)
+
+# example plot
+ggplot(decades_ea3, aes(birth_decade, EA3, color = ethnicity, 
+                        group = ethnicity)) + 
+  geom_line() +
+  geom_pointrange(aes(ymax = EA3 + 1.96 * se, ymin = EA3 - 1.96 * se))
+
+```
+
+Why does the US data show fewer differences by socio-economic status (SES) than the UK?
+One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
+The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
+@goldscheider1969minority, @johnson1979minority, @yang2003big).
+And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu).
+But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
+Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality.
+
 # Discussion
@@
 Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
-
-Why does the US data show fewer differences by socio-economic status (SES) than the UK?
-One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority, @johnson1979minority, @yang2003big).
-And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu).
-But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and by their different predictive power, so this hypothesis can only be speculative.
 
@@
     geom_vline(xintercept = 0) +
-    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.3) +
+    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.3,
+                    color = "steelblue3") +
     theme_light() + 

### 2023-09-22 explore-extrapolate-effect-sizes.R

@@ -1,0 +1,323 @@
+
+# == Setup ====
+
+library(ggplot2)
+library(purrr)
+library(dplyr)
+library(stringr)
+library(survey)
+library(tidyr)
+
+rand_orig <- readRDS("data/randhrs/rand-orig.rds")
+
+# ragender: 1 male, 2 female
+rand_orig <- rand_orig |>
+  filter(
+    (rabyear <= 1965 & as.numeric(ragender) == 1) |
+    (rabyear <= 1970 & as.numeric(ragender) == 2)
+  )
+
+# african ancestry
+pgen4a <- haven::read_dta("data/PGENSCORE4r3/pgenscore4a_r.dta")
+# european ancestry
+pgen4e <- haven::read_dta("data/PGENSCORE4r3/pgenscore4e_r.dta")
+
+pgens <- list(white = pgen4e, black = pgen4a)
+
+pcs <- c(paste0("PC1_5", LETTERS[1:5]), paste0("PC6_10", LETTERS[1:5]))
+
+rands <- list()
+
+eths <- c("black", "white")
+for (eth in eths) {
+  pgen <- pgens[[eth]]
+  names(pgen) <- str_replace(names(pgen), "[EA]4_", "")
+  names(pgen) <- str_replace(names(pgen), "^01", "GW01") # invalid names
+  rands[[eth]] <- inner_join(rand_orig, pgen,
+                     join_by(hhid, pn),
+                     unmatched = "drop",
+                     relationship = "one-to-one")
+}
+
+rand <- list_rbind(rands, names_to = "ethnicity")
+
+trk_orig <- haven::read_dta("data/trk2020v3/trk2020tr_r.dta")
+trk_weights <- trk_orig |> select(HHID, PN, matches("BIOWGTR"))
+
+rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
+rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
+
+all_pgs <- pgen4e |>
+    rename_with(\(x) str_replace(x, "[EA]4_", "")) |>
+    rename_with(\(x) str_replace(x, "^01", "GW01")) |>
+    # weird that I have to operate on whole df to use select semantics
+    select(-hhid, -pn, -version,
+           -matches("^PC\\d"),
+           ) |>
+    names()
+
+# scores and phenotypes:
+# EDU2, EDU3, EA3: raedyrs -- educational attainment
[…]

### 2023-09-26 hrs-selection.qmd
Add discussion of true effect sizes and calculations

@@
 set.seed(271075)
-knitr::knit_hooks$set("inline", 
-  \(x) {
-    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
-      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
-    }
-    x
-  }
-)
-
+
+library(tibble)
 library(haven)
@@
 library(purrr)
+library(carData)
+library(car, warn.conflicts = FALSE) # masks some
 library(ggplot2)
@@
 suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))
+
+knitr::knit_hooks$set("inline", 
+  \(x) {
+    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
+      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
+    }
+    x
+  }
+)
 
@@
   TC          = "Total cholesterol",
-  ANXFS       = "Anxiety CC", 
-  ANXCC       = "Anxiety factor",
+  ANXFS       = "Anxiety factor", 
+  ANXCC       = "Anxiety case-control",
   BUN         = "Blood urea nitro.",
@@
 
+  x_short <- str_replace(x_short, "^01", "ALZ_01") # two alzheimer's variants
   x_short
@@
 rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
-
 # only selected columns to avoid creating suffixes when we join with
@@
 rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)
+
+# 2010 (and others) record "age started smoking"
+# rand_2010 <- haven::read_dta("data/hd10f6a_STATA/hd10f6a.dta")
+# saveRDS(rand_2010, "data/hd10f6a_STATA/rand_2010.rds")
+rand_2010 <- readRDS("data/hd10f6a_STATA/rand_2010.rds")
+rand_2010 <- rand_2010 |> 
+  select(hhid, pn, mc120, matches("mlb041")) |> 
+  mutate(
+    age_smoked = ifelse(mc120 %in% c(95, 98, 99), NA_real_, mc120)
+  )
+rand_2010$anxiety <- rand_2010 |> 
+  select(matches("mlb041")) |> 
+  rowMeans(na.rm = TRUE)
+rand_2010$anx_na <- rand_2010 |> 
+  select(matches("mlb041")) |> 
+  is.na() |> 
+  rowSums()
+# matches the documentation at 
+# https://hrs.isr.umich.edu/sites/default/files/biblio/HRS2006LBQscale.pdf :
+rand_2010$anxiety[rand_2010$anx_na > 2] <- NA_real_
+
+# 2016 survey had a module with 18 questions diagnosing adult ADHD
+# rand_2016 <- haven::read_dta("data/h16f2c_STATA/h16f2c.dta")
+# saveRDS(rand_2016, "data/h16f2c_STATA/rand_2016.rds")
+rand_2016 <- readRDS("data/h16f2c_STATA/rand_2016.rds")
+rand_2016 <- rand_2016 |> 
+  select(hhid, pn, pv001:pv018) |> 
+  mutate(
+    across(matches("pv\\d\\d\\d"), \(x) ifelse(x %in% 8:9, NA, x)),
+    across(matches("pv\\d\\d\\d"), \(x) x == 1) # 1 is "yes', 5 is "no"
+  )
+rand_2016$adhd_score <- rand_2016 |> 
+  select(pv001:pv018) |> 
+  rowSums(na.rm = TRUE)
+rand_2016$adhd_nas <- rand_2016 |> 
+  select(pv001:pv018) |> 
+  is.na() |> 
+  rowSums(na.rm = TRUE)
+rand_2016$adhd_score <- rand_2016$adhd_score/(18 - rand_2016$adhd_nas)
 
@@
 
+old_pgs <- c("EDU2", "EDU3", "HEIGHT", "BMI",
+           "AD", "AD2", "GENCOG", "EVRSMK", "LONG", "CPD_TAG10",
+           "ADHD_PGC10", "MDD")
 pgs <- pgen4e |>
@@
            # outdated, or EDU3 without 23andMe:
-           -EDU2, -EDU3, -HEIGHT, -BMI,
-           -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
-           -ADHD_PGC10, -MDD
+           -any_of(old_pgs)
            ) |>
     names()
-
-rand <- rand |> left_join(rand_fam, 
-                          by = join_by(hhid, pn),
-                          unmatched = "drop",
-                          relationship = "one-to-one")
 
@@
 rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
+
+rand <- rand |> left_join(rand_fam, 
+                          by = join_by(hhid, pn),
+                          unmatched = "drop",
+                          relationship = "one-to-one")
+
+rand <- rand |> left_join(rand_2016, 
+                          by = join_by(hhid, pn),
+                          unmatched = "drop",
+                          relationship = "one-to-one")
+
+rand <- rand |> left_join(rand_2010, 
+                          by = join_by(hhid, pn),
+                          unmatched = "drop",
+                          relationship = "one-to-one")
 
@@
     raedyrs = as.numeric(raedyrs),
+    r10hearte = as.numeric(r10hearte),
+    r10diabe = as.numeric(r10diabe),
     edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
@@
 
+rand$neurot <- rand |> 
+  select(matches("r\\d+lbneur")) |> 
+  rowMeans(na.rm = TRUE)
+
 # two of four Alzheimer's variants with awkward names:
@@
 rand <- rand |> 
-  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
+  select(ethnicity, all_of(pgs), all_of(old_pgs), 
+         all_of(pcs), weight, raevbrn, rlrs,
          raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
-         rabyear, raehsamp, raestrat, parent_weight, ragender)
+         rabyear, raehsamp, raestrat, parent_weight, ragender,
+         adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
+         anxiety, r10cogtot)
 
@@
 # clean up cache so it doesn't get too big
-rm(rand_orig, trk_orig, rands, rand_fam)
+rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
 ```
@@
 @fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment.
-The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I.
-`r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrapped 95% C.I.
+The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I.
+`r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I.
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
-Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).
+Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.
 
@@
 `r n_reps_x_es` bootstraps.
-{#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}
-
-The economic theory of fertility in @hugh2022human implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups.
+{#tbl-x-bootstraps tbl-colwidths="\[16,28,28,28\]"}
+
+The economic theory of fertility also implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups.
 I tested this with bootstraps, but confidence intervals were always too wide to be informative.
 
-[^2]: See @hugh2022human equation (6) and following.
+[^1]: See @hugh2022human equation (6) and following.
 
@@
 Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality.
+
+Lastly, we would like to know natural selection's effect sizes.
+The bivariate correlations of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
+Polygenic scores contain error, so estimated correlations are biased towards zero compared to the correlation of the true PGS.
+They can be scaled up by
+
+$$
+\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}
+}$$
+
+where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the $R^2$ of the measured PGS on the target phenotype [@becker2021resource].
+I make this calculation using heritabilities from twin studies, and SNP- or chip-heritabilities, reported in @abdellaoui2021dissecting (chip-heritability is the total variance explained by common SNPs recorded by array chips).
+
+@fig-true-effects reports estimates for the true PGS as well as the "true chip-PGS" - the true best predictor of the phenotype using only variants recorded on the array chip, for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS.
+It's important to understand that estimates for the true PGS are more uncertain than for the true chip-PGS.
+Ideally, the true chip-PGS differs only from the recorded PGS because of sampling error in calculating the underlying summary statistics.[^2] The true PGS would use many more genetic variants, perhaps including e.g.
+*de novo* mutations, and the relationship between these variants' effect on fertility and on the target phenotype may not be the same as for genetic variants captured on array chips.
+
+[^2]: But even the chip-PGS might contain non-random noise from correlated environments.
+
+The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time.
+Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
+Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval.
+One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.
+
+As a rough rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
+Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
+Given the uncertainties about the exact nature of "noise" in existing PGS, the estimates -- even for chip heritability -- mostly show the limits of our knowledge, and should be treated as best guesses only.
+
+```{r}
+#| label: stat-boot-true-effects
+
+#' Calculate corrected PGS
+#'
+#' @param w Weights
+#' @param data Data frame
+#' @param pgs Name(s) of polygenic score(s) columns in  `data`.
+#' @param phenotype Name of target phenotype in `data`
+#' @param chip_h2 Chip-heritability
+#' @param twin_h2 Twin-study heritability
+#'
+#' @return A vector of corrected estimates of effects on fertility, 
+#'   named like `source.to.target` 
+#'   where `source` is one of `pgs` and target is either a `pgs` or the
+#'   chip-PGS or true PGS. If `pgs` has more than one element then
+#'   the function will correct pgs using each other as a sanity check. If
+#'   `source == target` then that is the original uncorrected score.
+#' @noRd
+#'
+#' @examples
+boot_correct <- function (w, data, pgs, phenotype, chip_h2, twin_h2) {
+
+  fs_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
+  fs_phen <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))
+
+  ms_fert <- map(fs_fert, \(x) lm(x, data = data, weights = w))
+  ms_phen <- map(fs_phen, \(x) lm(x, data = data, weights = w))
+  bhat_fert <- map2_dbl(ms_fert, pgs, \(x, y) coef(x)[[y]])
+  bhat_phen <- map2_dbl(ms_phen, pgs, \(x, y) coef(x)[[y]])
+
+  r2_phen <- map_dbl(ms_phen, \(x) {
+    an <- car::Anova(x)
+    an$`Sum Sq`[1]/sum(an$`Sum Sq`)  # incremental R2 of first term
+  })
+
+  r2_phen <- c(r2_phen, chip_h2, twin_h2)
+  # the correction factor
+  # lower triangular part of this shows how much you multiply to get from
+  # COLNAME to ROWNAME
+  rhos <- outer(r2_phen, r2_phen, \(x, y) sqrt(x/y)) 
+  rownames(rhos) <- colnames(rhos) <- c(pgs, "Chip", "Twin")
+
+  # multiply first column by bhat_edu[1] to give predictions from the
+  # first PGS to the next ones; etc.
+  # The diagonals of b_edu are exactly equal to bhat_edu
+  # arow,column gives the predicted value of b by ROWNAME for COLNAME
+  b_phen <- apply(rhos, 1, \(x) x * c(bhat_phen, NA, NA))
+  b_fert <- apply(rhos, 1, \(x) x * c(bhat_fert, NA, NA))
+
+  # don't include the last two rows of NA predictions "from chip/twin h2"
+  b_fert <- b_fert[seq_along(pgs), , drop = FALSE]
+
+  res <- c(b_fert)
+  # TODO: are the "backward estimates" wrong?
+  # TODO: check against the python tool from the PGI repository paper
+  res_names <- outer(rownames(b_fert), colnames(b_fert),
+                      \(x, y) paste0(x, ".to.", y))
+  names(res) <- c(res_names)
+
+  res
+}
+
+n_reps_true <- 99
+rand_boot <- rand |> as.svrepdesign(type = "bootstrap", 
+                                    replicates = n_reps_true)
+
+correct_scores <- tibble::tribble(
+  ~pgs                         , ~phenotype  , ~chip_h2, ~twin_h2,
+  c("ADHD_PGC10", "ADHD_PGC17"), "adhd_score", 0.21    , 0.75    ,
+  c("EDU2", "EDU3", "EA3")     , "raedyrs"   , 0.10    , 0.43    ,  
+  "AFBC"                       , "agefbn"    , 0.05    , 0.26    ,
+  "NEUROT"                     , "neurot"    , 0.10    , 0.37    ,
+  "CD"                         , "r10hearte" , 0.22    , 0.4     ,
+  "T2D"                        , "r10diabe"  , 0.18    , 0.22    , 
+  c("MDD", "MDD2")             , "r10cesd"   , 0.08    , 0.37    ,
+#  "ANXCC"                      , "anxiety"   , 0.31    , 0.45    ,
+  "AI"                         , "age_smoked", 0.05    , 0.50    ,
+  c("GENCOG", "GCOG2")         , "r10cogtot" , 0.19    , 0.86
+)
+
+correct_scores <- drop_na(correct_scores)
+# Sources for chip/twin h2; Abdel.
+# general cognition used h2 for "adult IQ"
+# Except:
+# heart disease 0.4 to 0.6 from https://www.nejm.org/doi/full/10.1056/nejm199404143301503
+# via https://www.jacc.org/doi/full/10.1016/j.jacc.2014.12.033
+# T2D  0.22 to 0.73...
+# https://www.cell.com/ajhg/pdf/S0002-9297(12)00031-6.pdf
+# https://www.cambridge.org/core/services/aop-cambridge-core/content/view/94E0005D626FB0399AA8FDDAC82F970E/S1832427415000833a.pdf/div-class-title-the-concordance-and-heritability-of-type-2-diabetes-in-34-166-twin-pairs-from-international-twin-registers-the-discordant-twin-discotwin-consortium-div.pdf
+# anxiety (nb not social anxiety!)
+# https://ajp.psychiatryonline.org/doi/full/10.1176/ajp.156.2.246
+# generalized anxiet disorder 42%:
+# https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0134865
+# age of smoking initiation, including chip heritability
+# http://www.matthewckeller.com/publications/Evans_Addiction_2021.pdf
+# CAD chip heritability
+# https://academic.oup.com/cardiovascres/article/113/8/973/2965332?login=false
+# type ii diabetes chip heritability
+# https://www.nature.com/articles/s41588-018-0241-6
+# 'current anxiety symptoms' chip heritability
+# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7237282/
+
+results_df <- list()
+for (i in seq(nrow(correct_scores))) {
+  source_pgs <- correct_scores$pgs[[i]]
+  phenotype <- correct_scores$phenotype[[i]]
+  chip_h2 <- correct_scores$chip_h2[[i]]
+  twin_h2 <- correct_scores$twin_h2[[i]]
+
+  results <- withReplicates(rand_boot, theta = boot_correct, 
+                            pgs = source_pgs, phenotype = phenotype, 
+                            chip_h2 = chip_h2, twin_h2 = twin_h2)
+  results_df[[phenotype]] <- results |>
+    as.data.frame() |>
+    tibble::rownames_to_column(var = "rn") |>
+    cbind(confint(results)) |>
+    tibble::remove_rownames() |>
+    setNames(c("rn", "mean", "SE", "conf.low", "conf.high")) |> 
+    mutate(
+      phenotype = phenotype,
+      source.pgs = str_remove(rn, ".to.*"),
+      target.pgs = str_remove(rn, ".*.to."),
+      rn = NULL,
+      .before = 1
+    )
+}
+
+results_df <- list_rbind(results_df)
+```
+
+```{r}
+#| label: fig-true-effects
+#| fig-cap: |
+#|   Estimated effects on RLRS of true polygenic scores calculated
+#|   using estimates of chip- and twin-heritability and the most recent
+#|   polygenic score.
+
+# TODO: maybe just use the polygenic score name not the phenotype?
+# TODO: why are some result bounds negative? e.g. ADHD? Presumably
+#. because can have negative effect on RLRS?
+# TODO: check the original phenotypes from which h2 ests were derived...
+#. - is CAD the same as coronary heart disease (r10hearte)?
+# TODO: why is ANXCC not predicting anxiety?
+results_df |> 
+  mutate(
+    Phenotype = nice_names[source.pgs],
+    Heritability = target.pgs
+  ) |> 
+  filter(Heritability %in% c("Chip", "Twin"), ! source.pgs %in% old_pgs) |> 
+  ggplot(aes(y = Phenotype, x = mean, xmin = conf.low, xmax = conf.high,
+             color = Heritability)) +
+    geom_vline(xintercept = 0, color = "grey60") +
+    geom_pointrange(aes(linetype = Heritability == "Twin"),
+                    position = position_dodge(0.2)) +
+    scale_linetype(guide = "none") +
+    coord_cartesian(xlim = c(- 0.3, 0.3)) +
+    labs(x = "Corrected effect on RLRS", y = "") +
+    theme_minimal()
+
+```
 
@@
 
+```{r}
+#| label: fig-check-correction
+#| fig-cap: |
+#|   Estimated effects of series of PGS using errors-in-variables 
+#|   correction. "Own" is the uncorrected estimate using the PGS
+#|   itself. Other estimates are from earlier or later PGS targeting
+#|   the same phenotype and corrected via the method in @becker2021resource.
+#| fig-height: 6
+
+results_df |> 
+  filter(length(unique(source.pgs)) > 1, .by = phenotype) |> 
+  filter(! target.pgs %in% c("Chip", "Twin")) |> 
+  mutate(
+    Phenotype = case_match(phenotype,
+                           "adhd_score" ~ "ADHD",
+                           "r10cesd" ~ "Depression",
+                           "r10cogtot" ~ "Cognition",
+                           "raedyrs" ~ "Educ. attainment"
+                           ),
+    Target = fct_relevel(target.pgs,
+                         "ADHD_PGC17", "ADHD_PGC10", "GCOG2", "GENCOG",
+                         "MDD2", "MDD", "EA3", "EDU3", "EDU2"),
+    Source = ifelse(source.pgs == target.pgs, "Own", source.pgs),
+  ) |> 
+  ggplot(aes(y = Target, x = mean, xmin = conf.low, xmax = conf.high,
+             color = Source)) +
+    geom_vline(xintercept = 0, color = "grey60") +
+    geom_pointrange(position = position_dodge(0.7)) +
+    geom_text(aes(x = mean, label = Source), hjust = "center", 
+              vjust = 1.5,
+              position = position_dodge(0.7), size = 3) + 
+    scale_color_manual(values = c("Own" = "red"), guide = "none") +
+    facet_grid(rows = vars(Phenotype), scales = "free", space = "free") +
+    labs(x = "Estimated effect on RLRS with correction", y = "Target PGS") +
+    coord_cartesian(xlim = c(-0.05, 0.05)) + 
+    theme_minimal()
+
+```
+
 \FloatBarrier

### 2023-09-26 explore-extrapolate-effect-sizes.R

@@ -70,7 +70,8 @@ all_pgs <- pgen4e |>
 # CPD, CPD_GSCAN19, available for now and also (some years) "when you were
 # smoking the most". But only in original file, not in RAND.
 # See question concordance. -- cigarettes per day (average or max)
-# ADHD_PGC10, ADHD_PGC17, ADHD
+# ADHD_PGC10, ADHD_PGC17: PV001 to PV016 in 2016 core was a module on "difficult
+#   to sustain attention"! Not in tracker ADHD
 # MDD_PGC13, MDD2, r*cesd -- major depressive disorder
 pgs_list <- list(
   # target phenotype = list of successive pgs names
@@ -128,56 +129,6 @@ randsv <- svydesign(
 		data = rand |> drop_na(weight)
 		)
 randsv <- randsv |> subset(ethnicity == "white")
-
-
-# == regressions ====
-
-
-# from the man himself
-# https://stackoverflow.com/a/73956738/10522567
-rsquared <- function (dep_var, design, mod) {
-  dep_var <- reformulate(dep_var)
-  total_var <-svyvar(dep_var, design, na.rm = TRUE)
-  resids <-
-  resid_var <- summary(mod)$dispersion
-  1 - resid_var/total_var
-}
-
-
-phenotypes <- intersect(names(pgs_list), names(randsv$variables))
-rsquares_phen <- list()
-rsquares_fert <- list()
-for (phenotype in phenotypes) {
-  pgs <- pgs_list[[phenotype]]
-
-  fmls <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))
-  mods <- map(fmls, \(x) svyglm(x, design = randsv))
-  rsqs <- map_dbl(mods, \(m) rsquared(phenotype, design = randsv, mod = m))
-
-  fmls_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
-  mods_fert <- map(fmls_fert, \(x) svyglm(x, design = randsv))
-  rsqs_fert <- map_dbl(mods_fert, \(m) rsquared("raevbrn", design = randsv, mod = m))
-
-  names(rsqs) <- names(rsqs_fert) <- pgs
-  rsquares_phen[[phenotype]] <- rsqs
-  rsquares_fert[[phenotype]] <- rsqs_fert
-}
-
-for (phenotype in phenotypes) {
[…]

### 2023-09-27 hrs-selection.qmd
Tweak discussion

@@
 
-@fig-true-effects reports estimates for the true PGS as well as the "true chip-PGS" - the true best predictor of the phenotype using only variants recorded on the array chip, for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS.
-It's important to understand that estimates for the true PGS are more uncertain than for the true chip-PGS.
-Ideally, the true chip-PGS differs only from the recorded PGS because of sampling error in calculating the underlying summary statistics.[^2] The true PGS would use many more genetic variants, perhaps including e.g.
-*de novo* mutations, and the relationship between these variants' effect on fertility and on the target phenotype may not be the same as for genetic variants captured on array chips.
-
-[^2]: But even the chip-PGS might contain non-random noise from correlated environments.
+@fig-true-effects reports estimates for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS.
+As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
+Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
+The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$ , limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip.
+Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
 
 The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time.
-Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
+Appendix @tbl-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
 Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval.
 One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.
-
-As a rough rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
-Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
-Given the uncertainties about the exact nature of "noise" in existing PGS, the estimates -- even for chip heritability -- mostly show the limits of our knowledge, and should be treated as best guesses only.
 

### 2023-09-27 hrs-selection.qmd
More tweaks. First draft.

@@
 The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time.
-Appendix @tbl-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
+Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
 Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval.
-One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.
 
@@
 #| label: stat-boot-true-effects
+#| cache: true
 
@@
 
-n_reps_true <- 99
+n_reps_true <- 199
 rand_boot <- rand |> as.svrepdesign(type = "bootstrap", 
@@
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
-Lastly, the smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
+The smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
+Lastly, true effect sizes are estimated and are typically between 0-0.2 standard deviations.
 
@@
 #| fig-cap: |
-#|   Estimated effects of series of PGS using errors-in-variables 
-#|   correction. "Own" is the uncorrected estimate using the PGS
-#|   itself. Other estimates are from earlier or later PGS targeting
-#|   the same phenotype and corrected via the method in @becker2021resource.
+#|   Estimated coefficients of series of PGS on fertility 
+#|   using errors-in-variables correction. "Own" is the uncorrected estimate
+#|   using the PGS itself. Other estimates are from earlier PGS targeting
+#|   the same phenotype, corrected via the method in @becker2021resource.
 #| fig-height: 6
@@
   filter(! target.pgs %in% c("Chip", "Twin")) |> 
+  filter(
+    ! target.pgs %in% old_pgs
+  ) |> 
   mutate(
@@
                            ),
-    Target = fct_relevel(target.pgs,
-                         "ADHD_PGC17", "ADHD_PGC10", "GCOG2", "GENCOG",
-                         "MDD2", "MDD", "EA3", "EDU3", "EDU2"),
+    Target = target.pgs,
     Source = ifelse(source.pgs == target.pgs, "Own", source.pgs),
@@
               vjust = 1.5,
-              position = position_dodge(0.7), size = 3) + 
+              position = position_dodge(0.7), size = 2.5) + 
     scale_color_manual(values = c("Own" = "red"), guide = "none") +
     facet_grid(rows = vars(Phenotype), scales = "free", space = "free") +
-    labs(x = "Estimated effect on RLRS with correction", y = "Target PGS") +
+    labs(x = "Corrected coefficient on RLRS", y = "Target PGS") +
     coord_cartesian(xlim = c(-0.05, 0.05)) + 

### 2023-09-27 hrs-selection.qmd
Trivia

@@
   and unmarried parents, but not among younger parents or those with less 
-  education.
+  education. I also estimate effect sizes corrected for noise in the 
+  polygenic scores. 
 date: "September 2023"
@@
 where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the $R^2$ of the measured PGS on the target phenotype [@becker2021resource].
-I make this calculation using heritabilities from twin studies, and SNP- or chip-heritabilities, reported in @abdellaoui2021dissecting (chip-heritability is the total variance explained by common SNPs recorded by array chips).
+I make this calculation using heritabilities from twin studies, and SNP- or chip-heritabilities (the total variance explained by SNPs recorded on array chips).[^2]
+
+[^2]: Heritability estimates are from @marenberg1994genetic and @nikpay2017partitioning for coronary heart disease); @willemsen2015concordance and @mahajan2018fine for Type II diabetes; @evans2021genetic for age of smoking initiation; and @abdellaoui2021dissecting for everything else.
+I used low-end heritability estimates for Type II diabetes (22%) and heart disease (40%).
 
@@
 
-results_df <- list()
+corrected_effects <- list()
 for (i in seq(nrow(correct_scores))) {
@@
                             chip_h2 = chip_h2, twin_h2 = twin_h2)
-  results_df[[phenotype]] <- results |>
+  corrected_effects[[phenotype]] <- results |>
     as.data.frame() |>
@@
 
-results_df <- list_rbind(results_df)
+corrected_effects <- list_rbind(corrected_effects)
 ```
@@
 #| fig-cap: |
-#|   Estimated effects on RLRS of true polygenic scores calculated
+#|   Estimated coefficients of true polygenic scores on RLRS calculated
 #|   using estimates of chip- and twin-heritability and the most recent
@@
 # TODO: why is ANXCC not predicting anxiety?
-results_df |> 
+corrected_effects |> 
   mutate(
@@
     coord_cartesian(xlim = c(- 0.3, 0.3)) +
-    labs(x = "Corrected effect on RLRS", y = "") +
+    labs(x = "Corrected coefficient on RLRS", y = "") +
     theme_minimal()
@@
 
-results_df |> 
+corrected_effects |> 
   filter(length(unique(source.pgs)) > 1, .by = phenotype) |> 

### 2023-09-27 explore-extrapolate-effect-sizes.R
add renv

@@ -1,326 +1,0 @@
-
-# == Setup ====
-
-library(ggplot2)
-library(purrr)
-library(dplyr)
-library(stringr)
-library(survey)
-library(tidyr)
-
-rand_orig <- readRDS("data/randhrs/rand-orig.rds")
-
-# ragender: 1 male, 2 female
-rand_orig <- rand_orig |>
-  filter(
-    (rabyear <= 1965 & as.numeric(ragender) == 1) |
-    (rabyear <= 1970 & as.numeric(ragender) == 2)
-  )
-
-# african ancestry
-pgen4a <- haven::read_dta("data/PGENSCORE4r3/pgenscore4a_r.dta")
-# european ancestry
-pgen4e <- haven::read_dta("data/PGENSCORE4r3/pgenscore4e_r.dta")
-
-pgens <- list(white = pgen4e, black = pgen4a)
-
-pcs <- c(paste0("PC1_5", LETTERS[1:5]), paste0("PC6_10", LETTERS[1:5]))
-
-rands <- list()
-
-eths <- c("black", "white")
-for (eth in eths) {
-  pgen <- pgens[[eth]]
-  names(pgen) <- str_replace(names(pgen), "[EA]4_", "")
-  names(pgen) <- str_replace(names(pgen), "^01", "GW01") # invalid names
-  rands[[eth]] <- inner_join(rand_orig, pgen,
-                     join_by(hhid, pn),
-                     unmatched = "drop",
-                     relationship = "one-to-one")
-}
-
-rand <- list_rbind(rands, names_to = "ethnicity")
-
-trk_orig <- haven::read_dta("data/trk2020v3/trk2020tr_r.dta")
-trk_weights <- trk_orig |> select(HHID, PN, matches("BIOWGTR"))
-
-rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
-rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
-
-all_pgs <- pgen4e |>
-    rename_with(\(x) str_replace(x, "[EA]4_", "")) |>
-    rename_with(\(x) str_replace(x, "^01", "GW01")) |>
-    # weird that I have to operate on whole df to use select semantics
-    select(-hhid, -pn, -version,
-           -matches("^PC\\d"),
-           ) |>
-    names()
-
-# scores and phenotypes:
-# EDU2, EDU3, EA3: raedyrs -- educational attainment
[…]

### 2023-09-29 hrs-selection.qmd
Add intergenerational correlation & bootstraps

@@
 I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
-
 The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.
+
 Genotyping took place in 2006, 2008 and subsequent years.
@@
 This leaves a total of `r length(pgs)` scores.
-
 PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
@@
 #| fig-cap: |
-#|   Scatterplot of PGS regression coefficients on RLRS and years of
-#|   education. Controls include 10 principal components of genetic array
-#|   data. Dashed lines show linear regressions.
+#|   Scatterplot of PGS regression coefficients on RLRS against 
+#|   coefficients on years of education. Each dot is one polygenic score.
+#|   Controls include 10 principal components of genetic array
+#|   data. Dashed lines are fitted from linear regressions.
 #| fig-subcap:
@@
 
+```{r}
+#| label: stat-bootstrap-gen-cor
+#| cache: true
+
+n_reps_gen_cor <- 199
+
+f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
+f_sibs <- map(pgs, \(x) reformulate(c(x, pcs), response = "r10sibs"))
+
+estimate_gen_cor <- function (w, data) {
+  data$w <- w # avoid w getting lost by map and lm
+  m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
+  m_sibs <- map(f_sibs, \(x) lm(x, data, weights = w/r10sibs))
+
+  coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
+  coef_sibs <- map2_dbl(m_sibs, pgs, \(x, y) coef(x)[[y]])
+  c(gen_cor = cor(coef_resp, coef_sibs))
+}
+
+rand_boot <- rand |> 
+    subset(ethnicity == "white") |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps_gen_cor)
+
+gen_cor <- withReplicates(rand_boot, theta = estimate_gen_cor)
+ci_gen_cor <- confint(gen_cor)
+```
+
 I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010.
@@
 Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
-Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
+Correlations are insignificant for both ethnic groups, with large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).
+However, there is a positive and significant correlation across generations, i.e.
+between PGS coefficients on number of siblings and number of children, of `r gen_cor`, bootstrapped 95% C.I.
+`r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`.
 

### 2023-09-29 hrs-selection.qmd
Fixes

@@
                       linetype = "dashed", colour = "red", linewidth = 0.5) +
-          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
+          scale_y_continuous(breaks = seq(-0.2, 0.2, 0.02),
                              labels = scales::label_number(0.01)) +
@@
 However, there is a positive and significant correlation across generations, i.e.
-between PGS coefficients on number of siblings and number of children, of `r gen_cor`, bootstrapped 95% C.I.
+between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I.
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`.
@@
 The HRS (Health and Retirement Study) is sponsored by the National Institute on Aging (grant number NIA U01AG009740) and is conducted by the University of Michigan.
+
+Code to reproduce this paper is available at <https://github.com/hughjonesd/hrs-selection>.
 

### 2023-09-29 hrs-selection.qmd
Add split by birthyear

@@
     edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
+    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
     # wave 10 is 2010
@@
          raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
-         rabyear, raehsamp, raestrat, parent_weight, ragender,
+         rabyear, born, raehsamp, raestrat, parent_weight, ragender,
          adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
@@
 
-n_reps_eth_diff <- 99
+n_reps_eth_diff <- 199
 rand_boot <- rand |> 
@@
 # all confindence intervals were too wide to be informative
-n_reps_x <- 99
+n_reps_x <- 199
 
@@
 
-old_opts <- options(survey.multicore = TRUE)
-
-n_reps_x_es <- 99
+n_reps_x_es <- 199
 
@@
                 married = c("Other", "Married"),
-                agefb = c("Low", "High"))
+                agefb = c("Low", "High"),
+                born = c("pre-1942", "post-1942"))
   groups <- paste0(":", term, groups) # colon for the cross term
@@
 
-diff_effect_size <- ci <- est <- tbl <- list()
-
-terms <- c("edu", "income_med", "married", "agefb")
+effect_sizes <- ci <- est <- tbl <- list()
+
+terms <- c("edu", "income_med", "married", "agefb", "born")
 
@@
 for (term in terms) {
-  diff_effect_size[[term]] <- 
+  effect_sizes[[term]] <- 
     withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
-  ci[[term]] <- confint(diff_effect_size[[term]])
-  est[[term]] <- as.numeric(diff_effect_size[[term]])
+  ci[[term]] <- confint(effect_sizes[[term]])
+  est[[term]] <- as.numeric(effect_sizes[[term]])
   tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
@@
 
-options(old_opts)
-
-```
-
-```{r}
-#| label: stat-agefb-bootstrap
-
-n_reps_agefb <- 9
+```
+
+```{r}
+#| label: stat-bootstrap-agefb
+#| cache: true
+
+n_reps_agefb <- 199
+
 calc_cor_agefb <- function (w, data) {
@@
 Note that the N is lower here due to missing values.
-There are no large differences between younger and older parents.
+There are no large differences in absolute effect size between younger and older parents.
 As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs.
 However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively.
-Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with uncontrolled coefficient; here, that isn't true, with coefficients remaining positively correlated.
+Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with the uncontrolled coefficients; here, that isn't true, with coefficients remaining positively correlated.
 
@@
 #| label: stat-control-agefb
-#| eval: false
-
-# cselection effects ontrolling for age at first birth
+
+# selection effects controlling for age at first birth
 # 
@@
 
-cor(tidy_pgs$white$estimate, tidy_pgs_control_agefb$estimate)
+cor_raw_controlled <- cor(tidy_pgs$white$estimate,
+                          tidy_pgs_control_agefb$estimate)
+
+stopifnot(cor_raw_controlled >= 0)
 
@@
 | Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
+| Birth year | `r tbl$born[1]`      | `r tbl$born[2]`      | `r tbl$born[3]`      |
 
@@
 married; below vs.
-above median age at first birth.
+above median age at first birth; born before vs.
+after 1942.
 `r n_reps_x_es` bootstraps.
-{#tbl-x-bootstraps tbl-colwidths="\[16,28,28,28\]"}
+{#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}
 
@@
 
+```{r}
+#| label: stat-x-bootstrap-birthyear
+#| cache: true
+
+n_reps_birthyear <- 199
+
+forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
+
+calc_birthyear <- function (w, data) {
+  data$w <- w
+  mods_young <- map(forms, \(x) lm(x, data, subset = rabyear > 1945,
+                                 weights = w))
+  mods_old <- map(forms, \(x) lm(x, data, subset = rabyear <= 1945,
+                                  weights = w))
+  coefs_young <- map2_dbl(mods_young, pgs, \(x, y) coef(x)[[y]])
+  coefs_old <- map2_dbl(mods_old, pgs, \(x, y) coef(x)[[y]])
+
+  c(cor = cor(coefs_young, coefs_old))
+}
+
+rand_boot <- rand |> 
+  subset(ethnicity == "white") |> 
+  as.svrepdesign(type = "bootstrap", replicates = n_reps_birthyear)
+
+cor_agefb <- withReplicates(rand_boot, theta = calc_birthyear)
+
+```
+
 Why does the US data show fewer differences by socio-economic status (SES) than the UK?
 One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
 `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
@@
 Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality.
+Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
+The last line of @tbl-x-bootstraps shows that effect sizes are larger for respondents born after 1942, but the difference is imprecisely estimated and not significant.
 
@@
 ```{r}
-#| label: stat-boot-true-effects
+#| label: stat-bootstrap-true-effects
 #| cache: true

### 2023-10-02 hrs-selection.qmd
Rewrite conclusion

@@
 
+```
+
+```{r}
+#| label: prog-cleanup
+
+# clean up cache so it doesn't get too big
+rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
+
 rand <- rand |> 
@@
 
-# clean up cache so it doesn't get too big
-rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
 ```
@@
 between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I.
-`r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`.
+`r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
 
@@
 
-The results here provide qualified support for the economic theory of fertility as an explanation for contemporary natural selection in humans.
-PGS which predict less education are being selected for, and PGS which predict more education are being selected against.
+The economic theory of fertility is driven by the trade-off between children and income.
+On the one hand, a higher hourly wage makes time raising children more costly (a "substitution effect").
+On the other hand, higher expected income makes children more affordable (an "income effect").
+Prediction 1 of the theory holds when substitution effects dominate income effects.
+Prediction 2 is driven by the specific form of individuals' preferences for income: when utility for income is sharply curved, the substitution effect is stronger for those who expect to earn less.
+
+The results here support prediction 1 but are more ambiguous for prediction 2.
 PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents.
@@
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
-The smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and with education.
-Lastly, true effect sizes are estimated and are typically between 0-0.2 standard deviations.
+The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason.
+But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.
+
+[^3]: See @balbo2013fertility for a broad review of fertility theories.
+
+The smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.
+Lastly, estimated effect sizes of "true" polygenic scores are typically between 0-0.2 standard deviations.
+To know more, we must await more accurate scores.
 

### 2023-11-22 hrs-selection.qmd
TE changes adding PGI repository scores and grandkid regression

@@
 title: "Natural selection in the Health and Retirement Study"
-author: "David Hugh-Jones"
+author: 
+  - David Hugh-Jones
+  - Tobias Edwards
 abstract: |
-  I investigate natural selection on polygenic scores
+  We investigate natural selection on polygenic scores
   in the contemporary US, using the Health and Retirement Study. Results
   partially support the economic theory of fertility as an explanation for
-  natural selection: among both white and black respondents,
   scores which correlate negatively (positively) with education are
@@
   and unmarried parents, but not among younger parents or those with less 
-  education. I also estimate effect sizes corrected for noise in the 
+  education. We also estimate effect sizes corrected for noise in the 
   polygenic scores. 
-date: "September 2023"
+  natural selection: among both white and black respondents,
+date: "November 2023"
 format: 
@@
 
-# TODO: ratio to family poverty threshold, h10inpovr
-# TODO: I think for all these guys, the raevborn question will have been 
-#.   asked by the time they were indeed over 45/40. Double check.
-
-# Is ADHD reverse-coded? - Doesn't seem to be, correlates negatively
-#       with income
-# What about XDISORDER?
-# ADHD doesn't seem to be. Nor does XDISORDER (correlates with e.g. r*depres)
-
+# TE quick TODOs change reference to time being insignificant. Fix the footnote citations
 options(digits = 3)
@@
   HBA1CEA     = "HbA1c Eur",
-  GCOG2       = "Gen. cognition"
+  GCOG2       = "Gen. cognition",
+  ACTIVITY="Physical Activity", #pgirep here
+  ADHD="ADHD",
+  ADVENTURE="Adventurous",
+  AFB="Age First Birth",
+  ALLERGYCAT="Cat Allergy",
+  ALLERGYDUST="Dust Allergy",
+  ALLERGYPOLLEN="Pollen Allergy",
+  ASTECZRHI="Asthma/Eczema/Rhinitis",
+  ASTHMA="Asthma",
+  AUDIT="Alcohol Misuse",
+  BMI="Body Mass Index",
+  CANNABISpgirep="Cannabis Use",
+  COGEMP="Cognitive Empathy",
+  COPD="Chronic Obstructive Pulmonary Disease",
+  CPD="Cigarettes per Day",
+  CP="Cognitive Performance",
+  DELAYDISC="Delay Discounting",
+  DEP="Depressive Symptoms",
+  DPWpgirep="Drinks per Week",
+  EA="Educational Attainment",
+  EVERSMOKE="Ever Smoker",
+  EXTRA="Exraversion",
+  FAMSAT="Satisfaction with Family",
+  FINSAT="Satisfaction with Finances",
+  FRIENDSAT="Satisfaction with Friends",
+  HAYFEVER="Hayfever",
+  HEIGHTpgirep="Height",
+  HIGHMATH="Highest Math",
+  LEFTOUT="Left Out of Social Activity",
+  LONELY="Loneliness",
+  MENARCHE="Age First Menses",
+  MIGRAINE="Migraine",
+  MORNING="Morning Person",
+  NARCIS="Narcissism",
+  NEARSIGHTED="Nearsightedness",
+  NEBmen="Number of Children (men)",
+  NEBwomen="Number of Children (women)",
+  NEURO="Neuroticism",
+  OPEN="Openness",
+  READING="Age Started Reading",
+  RELIGATT="Religious Attendance",
+  RISK="Risk Tolerance",
+  SELFHEALTH="Self-Rated Health",
+  SELFMATH="Self-Rated Math Ability",
+  SWB="Subjective Well-Being",
+  VOICEDEEP="Age Voice Deepened",
+  WORKSAT="Satisfaction with Work"
 )
@@
 pgen4e <- haven::read_dta("data/PGENSCORE4r3/pgenscore4e_r.dta")
+
+# TE edit PGI repository EUR only
+pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt", show_col_types = FALSE)
+names(pgirep)[1:2] <- c("hhid","pn")
+
+# TE Maybe not the most efficient method, but this code removes MTAG PGS
+pgirep <- select(pgirep,"hhid","pn",matches("PC"),matches("single"))
+
+# TE Going to remove the NEB PGS here
+pgirep <- select(pgirep, -c(PGI_NEBwomen_single) )
+
+# TE this creates simple names for PGS in PGI Rep
+pgi_rep_names <- str_split_fixed(names(pgirep)[23:57],"_",n=3)[,2]
+names(pgirep)[23:57] <- pgi_rep_names
+
+# TE PGI rep does not come pre-standardised. Going to standardize over whole sample to match
+# the approach taken with PGENscores
+pgirep[,3:ncol(pgirep)] <- scale(pgirep[,3:ncol(pgirep)])
 
@@
            "ADHD_PGC10", "MDD")
-pgs <- pgen4e |>
+
+# TE going to simplify names of pgen4e. Then merge, then create pgs.
+pgen4e_new_names <- pgen4e |>
     rename_with(shorten) |> 
-    # weird that I have to operate on whole df to use select semantics
+    # weird that we have to operate on whole df to use select semantics
+    select(-version,
+           -matches("^PC\\d"),
+           -matches("^NEB"), # number ever born
+           # outdated, or EDU3 without 23andMe:
+           -any_of(old_pgs)
+           )
+
+pgirep <- select(pgirep, -matches("^PC\\d"))
+#|>
+#|
+pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, by = join_by(hhid == hhid, pn == pn),suffix=c("","pgirep"))
+
+# TE Height is a duplicated name, but it's with the old PGS, so fixing this manually
+colnames(pgen4e_new_names)[which(names(pgen4e_new_names) == "HEIGHT")] <- "HEIGHTpgirep"
+
+# TE here we removed pgenscores that are duplicated in the pgi repository
+# we also remove trans ethnic and African ancestry PGS
+dupe_PGS <- c("EA3", # TE when I've tested it the pgirep one is slightly better...
+              "GCOG2", #CP
+              "BMI2",
+              "NEUROT",
+              "WELLB",
+              "DEPSYMP",
+              "ADHD_PGC17",
+              "EXTRAVER",
+              "AFBC", 
+              "PTSDAA", # African    
+              "PTSDC", # Combined Euro + African.
+              #"ANXFS"  ,"ANXCC", continuous vs case-control anxiety 
+              # r = 0.6 let's keep them
+              # we suppose it's equivalent to having CP and EA
+              "BUNTE", #trans ethnic BUNTE is kept 
+              "CKDTE", 
+              "BMI2",
+              "HEIGHT2",
+              "CPD_GSCAN19",
+              "DPW",
+              "SI",
+              "CANNABIS", 
+              "GWAD2NA", 
+              "ALZ_01AD2NA",
+              "GWAD2WA", # Three alzheimers deleted. Ones using p value 1 threshold
+              # and alzheimers that remove APOE variant region
+              "ALC", #AUDIT is alcohol misuse in pgirep
+ 	           "EGFR",
+	          "HBA1CEA" #African ancestry vs European ancestry. No TE
+              )
+
+dupe_PGS_black <- c(
+              "PTSDAA", # African    
+              "PTSDC", # Combined Euro + African.
+              #"ANXFS"  ,"ANXCC", continuous vs case-control anxiety 
+              # r = 0.6 let's keep them
+              # we suppose it's equivalent to having CP and EA
+              "BUNTE", #trans ethnic BUNTE is kept 
+              "CKDTE", 
+              "BMI2",
+              "CANNABIS", 
+              "GWAD2NA", 
+              "ALZ_01AD2NA",
+              "GWAD2WA", # Three alzheimers deleted. Ones using p value 1 threshold
+              # and alzheimers that remove APOE variant region
+ 	           "EGFR",
+	          "HBA1CEA" #African ancestry vs European ancestry. No TE
+              )
+
+pgen4e_new_names <- select(pgen4e_new_names,-hhid,-pn,-any_of(dupe_PGS),-any_of(old_pgs))
+pgs_white <- names(pgen4e_new_names)
+pgs_white[which(pgs_white=="HEIGHT")] <- "HEIGHTpgirep"
+
+pgs_black <- pgen4a |>
+    rename_with(shorten) |> 
     select(-hhid, -pn, -version,
@@
            # outdated, or EDU3 without 23andMe:
-           -any_of(old_pgs)
+           -any_of(old_pgs),
+           -any_of(dupe_PGS_black)
            ) |>
     names()
+
+rand <- left_join(rand, pgirep, by = join_by(hhid, pn),suffix=c("","pgirep"))
 
@@
 rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
+
+# TE I'm replacing 0 weights with NA because it can create odd problems
+rand$weight <- na_if(rand$weight, 0)
+
 # compensate for fecund parents more likely to have child in survey.
-# we use living sibs rather than including dead sibs too here. I think that's
+# we use living sibs rather than including dead sibs too here. we think that's
 # the correct approach:
 rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
+rand$child_weight <- rand$weight*rand$raevbrn
 
@@
 
+# TE this is average number of kids that your kids have
+rand$div_h10gkid <- rand$h10gkid/rand$raevbrn
+
+# TE 283 people report having grandkids, but having had no children. Weird!
+gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn==0,1,0)))
+
+gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn==0,1,0)))
+
+# If grandkids but no kids div_gkid is infinite. Replacing with NA
+rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA
+
+# TE added some variables for grandkid analysis
+# TE I've also added sampling weights in calculation of mean values
+# note the child weight is just weight* number of kids. so mean_h10gkid 
+# and mean_div_h10gkid are actually the same
 rand <- rand |>
   mutate(.by = rabyear, # calculate this pooling ethnicity
-    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
-    rlrs = raevbrn/mean_raevbrn
+    mean_weight = mean(weight, na.rm = TRUE),
+    mean_child_weight = mean(child_weight, na.rm = TRUE),
+    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
+    rlrs = raevbrn/mean_raevbrn,
+    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
+    rlrs_h10gkid = h10gkid/mean_h10gkid,
+    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
+    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
   ) |>
@@
   )
+# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. we don't think this is necessary
 
@@
   rowMeans(na.rm = TRUE)
-
 # two of four Alzheimer's variants with awkward names:
-pgs <- str_replace(pgs, "^01", "ALZ_01")
+#pgs <- str_replace(pgs, "^01", "ALZ_01")
+pgs_white <- str_replace(pgs_white, "^01", "ALZ_01")
+pgs_black <- str_replace(pgs_black, "^01", "ALZ_01")
 names(rand) <- str_replace(names(rand), "^01", "ALZ_01")
+pgs <- unique(c(pgs_white,pgs_black))
+
+# Child dataset testing
+# TE sample size not too small... But mum and dad often report different values of grandkids
+# and big questions of sampling bias. Given the weakness of the data I don't think the approach can be justified
+
+#rand_male <- subset(rand,ragender == 1)
+#rand_female <- subset(rand,ragender == 2)
+
+#rand_mid <- merge(rand_male,rand_female, by = "hhid",suffixes=c(".m",".f"),all.x=FALSE,all.y=FALSE)
+#rand_mid <- subset(rand_mid,ethnicity.m=="white"))
+# 3588
+
+#rand_mid$EA <- rand_mid$EA.m + rand_mid$EA.f
+#install.packages("fixest")
+#library(fixest)
+
+#cor(rand_mid$EA3.f,rand_mid$EA3.m)
+#reg <- lm(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
+#reg <- feols(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
+#reg <- feols(rlrs.m ~ EA3.m, data=rand_mid)
+#rand_mid$gkid <- rand_mid$h10gkid.m/rand_mid$raevbrn.m
+#rand_mid$gkidf <- rand_mid$h10gkid.f/rand_mid$raevbrn.f
+
+#reg <- feols(gkid ~ EA, data=rand_mid, ~ child_weight.m)
+
+#reg <- feols(gkidf ~ EA, data=rand_mid, ~ child_weight.f)
+
+#reg <- feols(h10gkid.m ~ EA, data=rand_mid, ~ child_weight.m)
+#reg <- feols(h10gkid.f ~ EA, data=rand_mid, ~ child_weight.f)
 
@@
 rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
+# TE we can delete more
+rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)
 
@@
          adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
-         anxiety, r10cogtot)
+         anxiety, r10cogtot, rlrs_div_h10gkid, child_weight)
+# added my grandkid rlrs
 
@@
 
+r_child <- svydesign(
+    id = ~ raehsamp,
+		strata = ~ raestrat,
+		weights = ~ child_weight, 
+		nest = TRUE,
+		data = rand |> drop_na(child_weight) 
+		)
+
 rand <- svydesign(
@@
 
-@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.
+@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from \@becker1976child. The theory has two components.
 
@@
 
-Here, I make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
+Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey \[\@hrsrandfam2023; \@hrsrand2023\].
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the theory predicts patterns of natural selection in both these ethnicities.
+The HRS also provides precalculated polygenic scores \[\@hrspgs2020\] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.
 
@@
 The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants.
-I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
+we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
 The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.
 
+<!--#  TE sources of pgs are not mentioned in the data section. I've added that since we are now using two sources -->
+
 Genotyping took place in 2006, 2008 and subsequent years.
-I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
-I also discard PGS for number of children ever born (but keep scores for age at first birth).
-This leaves a total of `r length(pgs)` scores.
+PGS were taken from those pre-calculated by the HRS [@hrspgs2020] and those produced by the Social Science Genetic Association Consortium, as part of their Polygenic Index Repository [@becker2021resource].
+Scores created by the HRS were provided for black and white participants, but Polygenic Index Repository scores were only created for white participants.
+
+Multiple steps were taken to prune the polygenic scores.
+For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository.
+For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry.
+We choose to use polygenic scores trained only on individuals of European ancestry.
+we discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
+we also discard PGS for number of children ever born (but keep scores for age at first birth).
+This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants.
 PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
-In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.
+In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.
 
@@
 treating them as members of the same biological population.
+The mean number of children of people born in the same year was calculated using sampling weights.
 
 The HRS contains weights which match survey respondents to the US population.
-I use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file).
-Since half the sample enters the extended interview including biomarker data in each biannual survey, I weight individuals by either their 2010 weight or their 2012 weight.
+We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file).
+Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight.
 This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.
-Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].
+Statistical tests are adjusted for clustering and stratification using the R "survey" package \[\@lumleysurvey2023\].
 
@@
 tidy_pgs_sibs <- mod_pgs_sibs <- list()
+tidy_pgs_gkids <- mod_pgs_gkids <- list()
 tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
@@
 cors_sibs_edu <- list()
+cors_gkids_edu <- list()
 tidy_fert_edu_phen <- list()
@@
 for (eth in eths) {
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
   r_eth <- rand |> subset(ethnicity == eth)
   r_parents_eth <- r_parents |> subset(ethnicity == eth)
+  r_child_eth <- r_child |> subset(ethnicity == eth)
 
@@
 
+    form_pgs_gkids <- map(pgs, 
+                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
+  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
+                                    \(x) svyglm_quiet(x, r_child_eth))
+
+  tidy_pgs_gkids[[eth]] <- mod_pgs_gkids[[eth]] |> 
+    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |>
+    list_rbind() |> 
+    filter(term %in% pgs)
+
+  cors_gkids_edu[[eth]] <- cor(tidy_pgs_gkids[[eth]]$estimate, 
+                     tidy_pgs_ed[[eth]]$estimate)
+
   tidy_fert_edu_phen[[eth]] <- 
@@
   r_eth <- rand |> subset(ethnicity == eth)
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
 
@@
 n_sig_bonf <- sum(tidy_pgs_eth$p.value < 0.05/n_tests)
+
+# TE we don't think it makes sense to include black estimates in the calculation. Since we know PGS in blacks are extremely weak
+# and the tests overlap near 1:1 with the polygenic scores in whites
+# Could just report the results for whites? 
+n_tests_white <- nrow(tidy_pgs$white)
+n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests)
+
+n_tests_black <- nrow(tidy_pgs$black)
+n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests)
 
@@
 
-I estimated coefficients of PGS on RLRS among black and white respondents separately.
+<!--# TE edited because 3 scores are bonferroni significant in whites.
+As of writing the same same three scores were signficiant regardlesss of whether we pool the blacks and whites in the bonferroni calculation.-->
+
+We estimated coefficients of PGS on RLRS among black and white respondents separately.
 These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
-Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
-Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
-But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.
+Appendix \@fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
+Standard errors are large because of the relatively low sample sizes.
+`r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
+The scores are age at first birth, educational attainment and attention deficit disorder.
+But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.
 
@@
 for (eth in eths) {
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
   rand_boot <- rand |> 
@@
     as.svrepdesign(type = "bootstrap", replicates = n_reps)
+  r_child_boot <- r_child |> 
+    subset(ethnicity == eth) |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps)
 
@@
     withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
+  cors_gkids_edu[[eth]] <- 
+    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h10gkid")
 }
@@
 ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
+ci_boot_gkids <- map(cors_gkids_edu, \(x) confint(x)[1,])
 
@@
 ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
+ci_boot_gkids_lm <- map(cors_gkids_edu, \(x) confint(x)[2,])
 
 if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
-if (! insignificant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
-if (significant(ci_boot_sibs$white)) warning("ci_boot_sibs$white significant")
+if (significant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
+if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
 if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
+if (insignificant(ci_boot_gkids$white)) warning("ci_boot_gkids$white insignificant")
+if (significant(ci_boot_gkids$black)) warning("ci_boot_gkids$black significant")
 ```
@@
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
-Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.
+Survey bootstraps [\@canty1999resampling] are used so as to make inferences from the sample of respondents.
 
@@
 
-# TODO: 3 scores are outliers among african-americans, with negative effects
-# on fertility and education - kidney disease, including the trans-ethnic
-# score, and PTSD estimated among european ancestry. Why?
 ```
@@
 cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
-if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
+#if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
+# TE correlation without outliers is not NEG
 ```
@@
 n_reps_gen_cor <- 199
+pgs <- pgs_white
 
@@
 
-I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010.
-(Data for dead siblings has too many missing values to use.) I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
+```{r}
+#| label: stat-bootstrap-gen-cor-gkids
+#| cache: true
+
+n_reps_gen_cor <-  199
+pgs <- pgs_white
+
+f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
+f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
+
+estimate_gen_cor <- function (w, data) {
+  data$w <- w # avoid w getting lost by map and lm
+  m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
+  m_gkids <- map(f_gkids, \(x) lm(x, data, weights = w*raevbrn))
+
+  coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
+  coef_gkids <- map2_dbl(m_gkids, pgs, \(x, y) coef(x)[[y]])
+  c(gen_cor = cor(coef_resp, coef_gkids))
+}
+
+rand_boot <- rand |> 
+    subset(ethnicity == "white") |> 
+    as.svrepdesign(type = "bootstrap", replicates = n_reps_gen_cor)
+
+gen_cor_gkids <- withReplicates(rand_boot, theta = estimate_gen_cor)
+ci_gen_cor_gkids <- confint(gen_cor_gkids)
+
+```
+
+<!--# TE Added grandkids and fixed minor typos -->
+
+We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively.
+We regress PGS on respondents' number of living siblings in 2010.
+Data for dead siblings has too many missing values to use.
+We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
 Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations.
 Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
-Correlations are insignificant for both ethnic groups, with large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
+Correlations are only significant and negative in the white respondents, although ethnicities have large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).
-However, there is a positive and significant correlation across generations, i.e.
-between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I.
+There is a positive and significant correlation across generations, i.e.
+between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`,cbootstrapped 95% C.I.
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
 
-I next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth.
+To examine selection in the succeeding generation, we divide the respondents' number of grandchildren by their number of children.
+In other words, we calculate the average number of children the respondent's children have.
+This measures the reproductive success in the second generation which we can call RS2.
+`r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having had no children.
+These individuals were removed for this test.
+In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.
+Older grandparents will have more time for their number of grandkids to accumulate.
+To deal with this time trend, we divide RS2 by the mean RS2 of respondents born in the same year, creating relative RS2 (RRS2).
+
+We regress the respondent's (the grandparent's) PGS on RRS2.
+The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure.
+The grandparent's PGS indicates, but does not determine the PGS of the parent.
+To know the expected parent's PGS, we would need both grandparent's PGS.
+Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, such an approach is untenable in our sample.
+We expect the effect of the grandparent's PGS to also be dependent upon the level of assortative mating for the trait.
+For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.
+
+Appendix @fig-gkids plots coefficients on RRS2 versus coefficients on years of education.
+Correlations are significant for whites but not blacks.
+Standard errors were moderate in whites, but large in blacks (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I.
+`r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`).
+There is a strong positive correlation between the parent and grandchild regressions, i.e.
+between PGS coefficients on RRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I.
+`r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`).
+
+We next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth.
 Education is years of education, split at 12 years.
 Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split.
-From here on I only use the white subsample: there are too few black respondents to be informative.
+From here on we only use the white subsample: there are too few black respondents to be informative.
 
@@
 `r n_reps_x_es` bootstraps.
-{#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}
+{#tbl-x-bootstraps tbl-colwidths="\[19,28,28,33\]"}
 
 The economic theory of fertility also implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups.
-I tested this with bootstraps, but confidence intervals were always too wide to be informative.
+we tested this with bootstraps, but confidence intervals were always too wide to be informative.
 
@@
 
-where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the $R^2$ of the measured PGS on the target phenotype [@becker2021resource].
-I make this calculation using heritabilities from twin studies, and SNP- or chip-heritabilities (the total variance explained by SNPs recorded on array chips).[^2]
-
-[^2]: Heritability estimates are from @marenberg1994genetic and @nikpay2017partitioning for coronary heart disease); @willemsen2015concordance and @mahajan2018fine for Type II diabetes; @evans2021genetic for age of smoking initiation; and @abdellaoui2021dissecting for everything else.
-I used low-end heritability estimates for Type II diabetes (22%) and heart disease (40%).
-
-@fig-true-effects reports estimates for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS.
+where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype \[\@becker2021resource\].
+Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters.
+To do this we use the error-corrected estimator developed by \@becker2021resource.
+
+SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository.
+The authors use GCTA to estimate heritability.
+When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead.
+We also perform error correction with twin heritabilites.
+To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2].
+We remove polygenic scores with $R^2_{PGS} > 0.005$ to focus on PGS with adequate power.
+This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction.
+
+[^2]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin.
+For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].
+
+@fig-true-effects reports the error corrected estimates of selection.
 As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
 Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
-The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$ , limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip.
+Age of First Birth has a twin corrected heritability of 0.15, implying that it maybe under fast selection.
+The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true \$h\^2\$ , the true \$R\^2\_{PGS}\$, limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip.
 Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
 
-The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time.
-Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula.
-Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval.
+Results from a few traits show substantial uncertainty regarding the true effect size.
+For asthma, when chip heritability is used, the effect is positive and significant, but negative and significant when twin heritability is used!
+Such a large disparity is because @becker2021resource estimate the chip heritability at 0.015 and we use a twin heritability 0f 0.55.
+For personality traits, chip heritability is often low (e.g.
+$<0.05$) but twin heritabilities are often much higher ($≈0.4$) implying substantial missing heritability.
+For these traits, the confidence intervals on twin heritabilities are enormous.
+Error-corrected effect sizes and the $R^2$ and $h^2$ parameters used can be found on the Github page for this paper.
+
+<!--# TE The lay public want to know how much IQ is changing. So I've added in a calculation of this. -->
+
+To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability.
+For cognitive performance we have estimated a genetic change of $-0.066$ standard deviations per generation, assuming a twin heritability of $0.51$.
+This implies a phenotypic change of $-0.066\times0.51^\frac{1}{2}=-0.047$ standard deviations per generation, equivalent to $-0.71$ points in the units of IQ, where a standard deviation is equal to $15$ points.
+This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype.
+Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.
+
+<!--# TE It's hard to keep in the sanity check since we are jumping from LDSC back to GCTA to calcualte snp heritability with pgenscores.
+The above also uses PGI repository scores only.
+I can add the sanity check back in.
+To be fair, I'm only aware of the method being tested with different discovery sample sizes using simulated data (link below) the sanity check is useul.
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10368647/-->
+
+```{r}
+#| label: TE-pgicorrect
+#| cache: true
+
+# Writing data for pgi_correct to use
+write.table(subset(rand$variables, ethnicity == "white"),"pgi_correct_data.txt",
+            quote=FALSE, sep="\t")
+
+correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx",na = "NA", range="B1:U53")
+
+# Where h2 and R2 are missing, we take values from a similar sample - the WLS
+correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2) , correct_scores$WLS_h2,
+                                 correct_scores$HRS_h2)
+
+correct_scores$R2 <- ifelse(is.na(correct_scores$HRS_R2) , correct_scores$WLS_R2,
+                                 correct_scores$HRS_R2)
+
+correct_scores$twin_h2 <- correct_scores$TwinH2
+correct_scores$pgs <- correct_scores$Abr
+
+correct_scores[which(correct_scores[,"pgs"]=="CANNABIS"),"pgs"] <- "CANNABISpgirep"
+correct_scores[which(correct_scores[,"pgs"]=="DPW"),"pgs"] <- "DPWpgirep"
+correct_scores[which(correct_scores[,"pgs"]=="HEIGHT"),"pgs"] <- "HEIGHTpgirep"
+
+# some of the PGS have estimates rounded to zero and are thus useless
+correct_scores <- subset(correct_scores, chip_h2 > 0)
+correct_scores <- subset(correct_scores, R2 > 0)
+
+# TE I originally used 0.28 from the TED sample. But that leads to especially absurd estimates
+# 0.15 was estiamted in the American MIDUS sample from the same paper.
+Leading to slightly less absurd estimates.
+Perhaps more appropriate in the American context.
+Uncertainty regarding parameters means results should be interepreted with CAUTION!!!
+correct_scores[which(correct_scores$pgs=="AFB"),"twin_h2"] <- 0.15
+#adjustment to use Midus values for NEB is also used
+correct_scores[which(correct_scores$pgs=="NEBwomen"),"twin_h2"] <- 0.28
+
+# PGI correct is quite fast, but can be made faster if the dataset is made smaller
+#loop for chip results
+map_df(1:nrow(correct_scores), function(i) {
+
+system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var',correct_scores[i,"pgs"],' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2',correct_scores[i,"R2"],' --h2 ',correct_scores[i,"chip_h2"],' --weights "weight"',sep=" "))
+
+system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
+reg_snp <- read.table("reg_results",header=TRUE)
+
+tibble(
+  trait = as.character(correct_scores[i,"pgs"]), #Ugly way of removing the column name...
+  beta = as.numeric(reg_snp[1,4]),
+  SE = as.numeric(reg_snp[1,5]),
+  Heritability = "Chip"
+)
+}
+) ->
+  corrected_effects_snp
+
+# Loop for twin results
+map_df(1:nrow(correct_scores), function(i) {
+
+system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var',correct_scores[i,"pgs"],' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2',correct_scores[i,"R2"],' --h2 ',correct_scores[i,"twin_h2"],' --weights "weight"',sep=" "))
+
+system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
+reg_twin <- read.table("reg_results",header=TRUE)
+
+tibble(
+  trait = as.character(correct_scores[i,"pgs"]), 
+  beta = as.numeric(reg_twin[1,4]),
+  SE = as.numeric(reg_twin[1,5]),
+  Heritability = "Twin"
+)
+}
+) ->
+  corrected_effects_twin
+
+corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
+corrected_effects$conf.low <- corrected_effects$beta - 1.96*corrected_effects$SE
+corrected_effects$conf.high <- corrected_effects$beta + 1.96*corrected_effects$SE
+
+# Deleting data created for PGI correct
+system("rm pgi_correct_data.txt")
+
+corrected_effects <- merge(corrected_effects, correct_scores, by.x="trait", by.y="Abr",prefix=c("",".y"))
+
+# The odd looking results are for traits with large missing heritability e.g.
+Asthma.
+My guess is that the rare variants act somewhat differently as DHJ as argued, meaning traits with large missing heritability show nonsense twin heritability estimates.
+
+corrected_effects$Category <- case_match(corrected_effects$Category, 
+         "Health and health behaviours" ~ "Health",
+         "Personality and well-being" ~ "Personality",
+         "Fertility and sexual development" ~ "Health",
+         "Anthropometric" ~ "Health",
+         "Cognition and education" ~ "Cognition")
+
+```
+
+```{r}
+#| label: fig-true-effects
+#| warning: false
+#| fig-cap: |
+#|   Estimated coefficients of true polygenic scores on RLRS calculated
+#|   using estimates of chip- and twin-heritability and the most recent
+#|   polygenic score.
+#| fig-height: 6.79
+#| fig-width:  5.98
+
+# TE TODO
+#opts_chunk$set(fig.width = 12, fig.height = 8)
+#570, 600
+# or 5.98 and 6.79 inches
+
+# This removes ADHD due to the low power of the score. Note ADHD predict RLRS after bonferroni! 
+# It also removes NEB
+corrected_effects <- subset(corrected_effects, R2 > 0.005)
+
+# TE Note that under Fisher's fundamental theorem of natural selection, the selection
+# for fitness should be equal to it's heritability... But in HRS the
+# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
+# is much lower than what the heritability of NEB should be.
+# The low R2 of NEB causes it to be removed, but the genetics for NEB could just be a little different
+# in HRS versus other samples!
+#NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]
+
+  my_labeller <- function(variable, value) {
+  return(value)
+  }
+
+  corrected_effects |>
+  mutate(
+    trait = nice_names[trait],
+    trait = fct_reorder(trait, beta)
+  )|>
+  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
+           color = Heritability)) +
+  geom_vline(xintercept = 0, color = "grey60") +
+  geom_pointrange(aes(linetype = Heritability == "Twin"),
+                  position = position_dodge(0.2)) +
+  scale_linetype(guide = "none") +
+  coord_cartesian(xlim = c(-0.3, 0.3)) +
+  labs(x = "Corrected coefficient on RLRS", y = "") +
+  theme_minimal() +
+  facet_grid(Category ~ ., scales = "free_y", space = "free_y", labeller = my_labeller)+
+  theme(strip.text.y = element_text(angle = -90,face = "bold"))
+
+  corrected_effects$TwinH2_Source <- corrected_effects$Source
+  corrected_effects <- corrected_effects[,c("trait","beta","SE","Heritability","chip_h2",
+                                            "twin_h2","R2","TwinH2_Source","PoldermanICF_ICD10_subchapter","Category")]
+
+ write.csv(corrected_effects,"Corrected_effects.csv")
+
+```
 
@@
 #| cache: true
+#| eval: false
 
@@
 ```{r}
-#| label: fig-true-effects
+#| label: fig-true-effects-old
+#| eval: false
 #| fig-cap: |
@@
 
-The smaller black sample makes most tests inconclusive for this population: I can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.
+The smaller black sample makes most tests inconclusive for this population: we can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.
 Lastly, estimated effect sizes of "true" polygenic scores are typically between 0-0.2 standard deviations.
@@
     )
+
 ```
@@
 }
+```
+
+```{r}
+#| label: fig-gkids
+#| fig-cap: |
+#|   Scatterplot of PGS coefficients on RLRS2 and years of
+#|   education. RRS2 is the reproductive success of respondent's offspring relative to the success of other offspring. Controls include 10 principal components of genetic array
+#|   data. Dashed lines show linear regressions.
+#| fig-subcap:
+#|   - ""
+#|   - ""
+#| layout-nrow: 2
+#| column: page
+
+for (eth in eths) {
+  ggp <-  
+    tidy_pgs_gkids[[eth]] |> 
+    inner_join(tidy_pgs_ed[[eth]], 
+               by = join_by(term), suffix = c(".gkids", ".edyrs")) |> 
+    ggplot(aes(estimate.edyrs, estimate.gkids)) +
+      geom_hline(yintercept = 0, color = "grey80") +
+      geom_vline(xintercept = 0, color = "grey80") +
+      geom_point() +
+      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
+                  linetype = "dashed", colour = "red", linewidth = 0.5) +
+      coord_cartesian(xlim = c(-.3, .75)) +
+      theme_light() + 
+      theme(panel.grid = element_blank()) +
+      labs(
+          title = eth,
+          x = "Coefficient on years of education",
+          y = "Coefficient on RRS2"
+      )
+  print(ggp)
+}
 
@@
 #| label: fig-check-correction
+#| eval: false
 #| fig-cap: |

### 2023-11-29 hrs-selection.qmd
Small commits before merging Tobias stuff

--- 2023-10-02 hrs-selection.qmd
@@
          all_of(pcs), weight, raevbrn, rlrs,
-         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
+         raedyrs, r10sibs, edu, income_resid, income_med, married, 
+         agefb, agefbn,
          rabyear, born, raehsamp, raestrat, parent_weight, ragender,

### 2023-11-29 explore-income.R

@@ -51,6 +51,7 @@ rw_l <- rw$variables |>
 
 pgs_sig <- tidy_inc$pgs[tidy_inc$sig]
 
+library(ggdark)
 rw_l |>
   filter(pgs %in% pgs_sig) |>
   mutate(
@@ -61,10 +62,12 @@ rw_l |>
   geom_point(alpha = 0.1, size = 0.1) +
   geom_smooth(aes(color = pgs), method = "loess") +
   coord_cartesian( ylim = c(-1, 1), xlim = c(-2, 2)) +
+  dark_theme_minimal() +
   theme(legend.position = "none") +
   labs(
-    title = "Scatterplots of income against significant PGS with loess smoother",
+    title = "Scatterplots of income against significant PGS (fdr < 0.05) with loess smoother",
     subtitle = "Plots show the centre of the distribution but smoother uses all data",
+    caption = "Source: Health and Retirement Study (US)",
     x = "Income (residualized and standardized)",
     y = "PGS"
   )

### 2023-11-29 hrs-selection.qmd
Improved coding style

--- 2023-11-22 hrs-selection.qmd
@@
 )
-
 nice_names <- c(
@@
   GCOG2 = "Gen. cognition",
-  ACTIVITY="Physical Activity", #pgirep here
-  ADHD="ADHD",
-  ADVENTURE="Adventurous",
-  AFB="Age First Birth",
-  ALLERGYCAT="Cat Allergy",
-  ALLERGYDUST="Dust Allergy",
-  ALLERGYPOLLEN="Pollen Allergy",
-  ASTECZRHI="Asthma/Eczema/Rhinitis",
-  ASTHMA="Asthma",
-  AUDIT="Alcohol Misuse",
-  BMI="Body Mass Index",
-  CANNABISpgirep="Cannabis Use",
-  COGEMP="Cognitive Empathy",
-  COPD="Chronic Obstructive Pulmonary Disease",
-  CPD="Cigarettes per Day",
-  CP="Cognitive Performance",
-  DELAYDISC="Delay Discounting",
-  DEP="Depressive Symptoms",
-  DPWpgirep="Drinks per Week",
-  EA="Educational Attainment",
-  EVERSMOKE="Ever Smoker",
-  EXTRA="Exraversion",
-  FAMSAT="Satisfaction with Family",
-  FINSAT="Satisfaction with Finances",
-  FRIENDSAT="Satisfaction with Friends",
-  HAYFEVER="Hayfever",
-  HEIGHTpgirep="Height",
-  HIGHMATH="Highest Math",
-  LEFTOUT="Left Out of Social Activity",
-  LONELY="Loneliness",
-  MENARCHE="Age First Menses",
-  MIGRAINE="Migraine",
-  MORNING="Morning Person",
-  NARCIS="Narcissism",
-  NEARSIGHTED="Nearsightedness",
-  NEBmen="Number of Children (men)",
-  NEBwomen="Number of Children (women)",
-  NEURO="Neuroticism",
-  OPEN="Openness",
-  READING="Age Started Reading",
-  RELIGATT="Religious Attendance",
-  RISK="Risk Tolerance",
-  SELFHEALTH="Self-Rated Health",
-  SELFMATH="Self-Rated Math Ability",
-  SWB="Subjective Well-Being",
-  VOICEDEEP="Age Voice Deepened",
-  WORKSAT="Satisfaction with Work"
+  ACTIVITY = "Physical Activity", # pgirep here
+  ADHD = "ADHD",
+  ADVENTURE = "Adventurous",
+  AFB = "Age First Birth",
+  ALLERGYCAT = "Cat Allergy",
+  ALLERGYDUST = "Dust Allergy",
+  ALLERGYPOLLEN = "Pollen Allergy",
+  ASTECZRHI = "Asthma/Eczema/Rhinitis",
+  ASTHMA = "Asthma",
+  AUDIT = "Alcohol Misuse",
+  BMI = "Body Mass Index",
+  CANNABISpgirep = "Cannabis Use",
+  COGEMP = "Cognitive Empathy",
+  COPD = "Chronic Obstructive Pulmonary Disease",
+  CPD = "Cigarettes per Day",
+  CP = "Cognitive Performance",
+  DELAYDISC = "Delay Discounting",
+  DEP = "Depressive Symptoms",
+  DPWpgirep = "Drinks per Week",
+  EA = "Educational Attainment",
+  EVERSMOKE = "Ever Smoker",
+  EXTRA = "Exraversion",
+  FAMSAT = "Satisfaction with Family",
+  FINSAT = "Satisfaction with Finances",
+  FRIENDSAT = "Satisfaction with Friends",
+  HAYFEVER = "Hayfever",
+  HEIGHTpgirep = "Height",
+  HIGHMATH = "Highest Math",
+  LEFTOUT = "Left Out of Social Activity",
+  LONELY = "Loneliness",
+  MENARCHE = "Age First Menses",
+  MIGRAINE = "Migraine",
+  MORNING = "Morning Person",
+  NARCIS = "Narcissism",
+  NEARSIGHTED = "Nearsightedness",
+  NEBmen = "Number of Children (men)",
+  NEBwomen = "Number of Children (women)",
+  NEURO = "Neuroticism",
+  OPEN = "Openness",
+  READING = "Age Started Reading",
+  RELIGATT = "Religious Attendance",
+  RISK = "Risk Tolerance",
+  SELFHEALTH = "Self-Rated Health",
+  SELFMATH = "Self-Rated Math Ability",
+  SWB = "Subjective Well-Being",
+  VOICEDEEP = "Age Voice Deepened",
+  WORKSAT = "Satisfaction with Work"
 )
@@
               # r = 0.6 let's keep them
-              # we suppose it's equivalent to having CP and EA
-              "BUNTE", #trans ethnic BUNTE is kept 
+              # I suppose it's equivalent to having CP and EA
+              "BUNTE",  
               "CKDTE", 
@@
  	            "EGFR",
-	          "HBA1CEA" #African ancestry vs European ancestry. No TE
-              )
-
-dupe_PGS_black <- c(
-              "PTSDAA", # African    
-              "PTSDC", # Combined Euro + African.
-              #"ANXFS"  ,"ANXCC", continuous vs case-control anxiety 
-              # r = 0.6 let's keep them
-              # we suppose it's equivalent to having CP and EA
-              "BUNTE", #trans ethnic BUNTE is kept 
+	            "HBA1CEA" 
+              )
+
+dupe_PGS_black <- c("PTSDAA",   
+              "PTSDC", 
+              "BUNTE", 
               "CKDTE", 
@@
               "ALZ_01AD2NA",
-              "GWAD2WA", # Three alzheimers deleted. Ones using p value 1 threshold
-              # and alzheimers that remove APOE variant region
+              "GWAD2WA", 
  	            "EGFR",
-	          "HBA1CEA" #African ancestry vs European ancestry. No TE
-              )
-
-pgen4e_new_names <- select(pgen4e_new_names,-hhid,-pn,-any_of(dupe_PGS),-any_of(old_pgs))
+	            "HBA1CEA" 
+              )
+
+pgen4e_new_names <- select(pgen4e_new_names, -hhid, -pn, -any_of(dupe_PGS), -any_of(old_pgs))
 pgs_white <- names(pgen4e_new_names)
-pgs_white[which(pgs_white=="HEIGHT")] <- "HEIGHTpgirep"
+pgs_white[which(pgs_white == "HEIGHT")] <- "HEIGHTpgirep"
 
@@
 rand$div_h10gkid <- rand$h10gkid/rand$raevbrn
-
 # TE 283 people report having grandkids, but having had no children. Weird!
-gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn==0,1,0)))
+gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))
 
@@
   )
-# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. we don't think this is necessary
+# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary
 
@@
 
-# Child dataset testing
-# TE sample size not too small... But mum and dad often report different values of grandkids
-# and big questions of sampling bias. Given the weakness of the data I don't think the approach can be justified
-
-#rand_male <- subset(rand,ragender == 1)
-#rand_female <- subset(rand,ragender == 2)
-
-#rand_mid <- merge(rand_male,rand_female, by = "hhid",suffixes=c(".m",".f"),all.x=FALSE,all.y=FALSE)
-#rand_mid <- subset(rand_mid,ethnicity.m=="white"))
-# 3588
-
-#rand_mid$EA <- rand_mid$EA.m + rand_mid$EA.f
-#install.packages("fixest")
-#library(fixest)
-
-#cor(rand_mid$EA3.f,rand_mid$EA3.m)
-#reg <- lm(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
-#reg <- feols(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
-#reg <- feols(rlrs.m ~ EA3.m, data=rand_mid)
-#rand_mid$gkid <- rand_mid$h10gkid.m/rand_mid$raevbrn.m
-#rand_mid$gkidf <- rand_mid$h10gkid.f/rand_mid$raevbrn.f
-
-#reg <- feols(gkid ~ EA, data=rand_mid, ~ child_weight.m)
-
-#reg <- feols(gkidf ~ EA, data=rand_mid, ~ child_weight.f)
-
-#reg <- feols(h10gkid.m ~ EA, data=rand_mid, ~ child_weight.m)
-#reg <- feols(h10gkid.f ~ EA, data=rand_mid, ~ child_weight.f)
-
 ```
@@
          all_of(pcs), weight, raevbrn, rlrs,
-         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
-         rabyear, born, raehsamp, raestrat, parent_weight, ragender,
-         adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
-         anxiety, r10cogtot, rlrs_div_h10gkid, child_weight)
-# added my grandkid rlrs
+         raedyrs, r10sibs, edu, income_resid, income_med, married, 
+         agefb, agefbn, rabyear, born, raehsamp, raestrat, 
+         parent_weight, ragender, adhd_score, neurot, r10diabe, 
+         r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
+         rlrs_div_h10gkid, child_weight)
 
@@
 
-# TE we don't think it makes sense to include black estimates in the calculation. Since we know PGS in blacks are extremely weak
-# and the tests overlap near 1:1 with the polygenic scores in whites
-# Could just report the results for whites? 
+# TE bonferroni correction might make more sense focusing on white tests
+# Either way seems to produce the same result
+
 n_tests_white <- nrow(tidy_pgs$white)
@@
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
-Survey bootstraps [\@canty1999resampling] are used so as to make inferences from the sample of respondents.
+Survey bootstraps \[\@canty1999resampling\] are used so as to make inferences from the sample of respondents.
 
@@
 cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
-#if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
+# if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
 # TE correlation without outliers is not NEG
@@
 
-correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx",na = "NA", range="B1:U53")
+correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
+                                            range="B1:U53")
 
 # Where h2 and R2 are missing, we take values from a similar sample - the WLS
-correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2) , correct_scores$WLS_h2,
+correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2), correct_scores$WLS_h2,
                                  correct_scores$HRS_h2)
 
-correct_scores$R2 <- ifelse(is.na(correct_scores$HRS_R2) , correct_scores$WLS_R2,
+correct_scores$R2      <- ifelse(is.na(correct_scores$HRS_R2), correct_scores$WLS_R2,
                                  correct_scores$HRS_R2)
@@
 
-correct_scores[which(correct_scores[,"pgs"]=="CANNABIS"),"pgs"] <- "CANNABISpgirep"
-correct_scores[which(correct_scores[,"pgs"]=="DPW"),"pgs"] <- "DPWpgirep"
-correct_scores[which(correct_scores[,"pgs"]=="HEIGHT"),"pgs"] <- "HEIGHTpgirep"
+correct_scores[which(correct_scores[, "pgs"] == "CANNABIS"), "pgs"] <- "CANNABISpgirep"
+correct_scores[which(correct_scores[, "pgs"] == "DPW"), "pgs"]      <- "DPWpgirep"
+correct_scores[which(correct_scores[, "pgs"] == "HEIGHT"), "pgs"]   <- "HEIGHTpgirep"
 
@@
 # 0.15 was estiamted in the American MIDUS sample from the same paper. 
-Leading to slightly less absurd estimates.
-Perhaps more appropriate in the American context.
-Uncertainty regarding parameters means results should be interepreted with CAUTION!!!
-correct_scores[which(correct_scores$pgs=="AFB"),"twin_h2"] <- 0.15
-#adjustment to use Midus values for NEB is also used
-correct_scores[which(correct_scores$pgs=="NEBwomen"),"twin_h2"] <- 0.28
-
-# PGI correct is quite fast, but can be made faster if the dataset is made smaller
-#loop for chip results
+# Leading to slightly less absurd estimates. Perhaps more appropriate in the American context. 
+# Uncertainty regarding parameters means results should be interepreted with CAUTION!!!
+# adjustment to use Midus values for NEB is also used
+
+correct_scores[which(correct_scores$pgs == "AFB"), "twin_h2"]      <- 0.15
+correct_scores[which(correct_scores$pgs == "NEBwomen"), "twin_h2"] <- 0.28
+
 map_df(1:nrow(correct_scores), function(i) {
 
-system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var',correct_scores[i,"pgs"],' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2',correct_scores[i,"R2"],' --h2 ',correct_scores[i,"chip_h2"],' --weights "weight"',sep=" "))
+system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"chip_h2"], ' --weights "weight"', sep=" "))
 
 system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
-reg_snp <- read.table("reg_results",header=TRUE)
+reg_snp <- read.table("reg_results", header=TRUE)
 
 tibble(
-  trait = as.character(correct_scores[i,"pgs"]), #Ugly way of removing the column name...
+  trait = as.character(correct_scores[i,"pgs"]), 
   beta  = as.numeric(reg_snp[1,4]),
@@
 
-system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var',correct_scores[i,"pgs"],' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2',correct_scores[i,"R2"],' --h2 ',correct_scores[i,"twin_h2"],' --weights "weight"',sep=" "))
+system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"twin_h2"], ' --weights "weight"', sep=" "))
 
@@
 
-# Deleting data created for PGI correct
 system("rm pgi_correct_data.txt")
 
-corrected_effects <- merge(corrected_effects, correct_scores, by.x="trait", by.y="Abr",prefix=c("",".y"))
+corrected_effects <- merge(corrected_effects, correct_scores, 
+                           by.x="trait", by.y="Abr",prefix=c("",".y"))
 
@@
                                         "Anthropometric" ~ "Health",
-         "Cognition and education" ~ "Cognition")
+                                        "Cognition and education" ~ "Cognition"
+                                       )
 
@@
   corrected_effects$TwinH2_Source <- corrected_effects$Source
-  corrected_effects <- corrected_effects[,c("trait","beta","SE","Heritability","chip_h2",
-                                            "twin_h2","R2","TwinH2_Source","PoldermanICF_ICD10_subchapter","Category")]
+
+  corrected_effects <- corrected_effects[, c("trait", "beta", "SE", "Heritability",
+                                            "chip_h2", "twin_h2", "R2", 
+                                            "TwinH2_Source", "PoldermanICF_ICD10_subchapter",
+                                            "Category")]
 

### 2023-11-30 hrs-selection.qmd
Update renv

--- 2023-11-29 hrs-selection.qmd
@@
 |               | Low group             | High group            | Difference            |
-|------------------|------------------|------------------|------------------|
+|---------------|-----------------------|-----------------------|-----------------------|
 | Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |

### 2023-11-30 hrs-selection.qmd
Further data cleaning

--- 2023-11-29 hrs-selection.qmd
@@
 # TE this is average number of kids that your kids have
-rand$div_h10gkid <- rand$h10gkid/rand$raevbrn
+rand$div_h10gkid <- rand$h10gkid / rand$raevbrn
+
 # TE 283 people report having grandkids, but having had no children. Weird!
+# This ugly code subsets the data by ethnicity then counts all observations
+# where grandkids is greater than 0 but number of kids is 0
 gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))
 
-gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn==0,1,0)))
-
-# If grandkids but no kids div_gkid is infinite. Replacing with NA
+gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn == 0, 1, 0)))
+
+# TE if grandkids but no kids div_gkid is infinite. Replacing with NA
 rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA
@@
   )
+
 # TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary
@@
 rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
+
 # TE we can delete more
@@
 
+####################################
+# Formatting parameter values to use
+####################################
+
+# correct_scores holds h2 and R2 values, as well as useful information about the PGS
+# such as group categories taken from the PGI repository and simplified names
 correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
@@
 
+####################################
+# Calculating error corrected estimates
+####################################
+
+# Loop for for calculating results with SNP heritabilities
 map_df(1:nrow(correct_scores), function(i) {
@@
 
+# This removes "junk" rows in the way of our regression betas
 system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
@@
 
-# Loop for twin results
+# Loop for for calculating results with twin heritabilities
 map_df(1:nrow(correct_scores), function(i) {
@@
 
-# TE TODO
-#opts_chunk$set(fig.width = 12, fig.height = 8)
-#570, 600
-# or 5.98 and 6.79 inches
-
 # This removes ADHD due to the low power of the score. Note ADHD predict RLRS after bonferroni! 
@@
 # is much lower than what the heritability of NEB should be.
-# The low R2 of NEB causes it to be removed, but the genetics for NEB could just be a little different
-# in HRS versus other samples!
-#NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]
+# The low R2 of NEB causes it to be removed, but the genetics for NEB 
+# could just be a little different in HRS versus other samples!
+# NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]
 
@@
 
+# Cleaning and saving corrected effects so people can have the exact figures
   corrected_effects$TwinH2_Source <- corrected_effects$Source

### 2023-12-01 hrs-selection.qmd
Clean up code, get pgi_correct working on DHJ's machine

--- 2023-11-30 hrs-selection.qmd
@@
 library(dplyr, warn.conflicts = FALSE)
+library(glue)
 library(stringr)
@@
 # TE edit PGI repository EUR only
-pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt", show_col_types = FALSE)
+pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt",
+                          show_col_types = FALSE)
 names(pgirep)[1:2] <- c("hhid","pn")
 
-# TE Maybe not the most efficient method, but this code removes MTAG PGS
-pgirep <- select(pgirep,"hhid","pn",matches("PC"),matches("single"))
-
-# TE Going to remove the NEB PGS here
-pgirep <- select(pgirep, -c(PGI_NEBwomen_single) )
-
-# TE this creates simple names for PGS in PGI Rep
-pgi_rep_names <- str_split_fixed(names(pgirep)[23:57],"_",n=3)[,2]
-names(pgirep)[23:57] <- pgi_rep_names
+pgirep <- pgirep |> 
+  select(hhid, pn, matches("single"), -PGI_NEBwomen_single) |> 
+  rename_with(
+    \(x) str_replace(x, "PGI_(.*)_single", "\\1")
+  )
 
@@
 # the approach taken with PGENscores
-pgirep[,3:ncol(pgirep)] <- scale(pgirep[,3:ncol(pgirep)])
+pgirep[, 3:ncol(pgirep)] <- scale(pgirep[, 3:ncol(pgirep)])
 
@@
 
-pgirep <- select(pgirep, -matches("^PC\\d"))
-#|>
-#|
-pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, by = join_by(hhid == hhid, pn == pn),suffix=c("","pgirep"))
-
-# TE Height is a duplicated name, but it's with the old PGS, so fixing this manually
-colnames(pgen4e_new_names)[which(names(pgen4e_new_names) == "HEIGHT")] <- "HEIGHTpgirep"
+pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, 
+                              by = join_by(hhid == hhid, pn == pn),
+                              suffix = c("", "pgirep"))
+
+# Height is a duplicated name, but it's with the old PGS, so fixing this manually
+pgen4e_new_names <- rename(pgen4e_new_names, 
+                           HEIGHTpgirep = HEIGHT)
 
@@
 
-pgen4e_new_names <- select(pgen4e_new_names, -hhid, -pn, -any_of(dupe_PGS), -any_of(old_pgs))
+pgen4e_new_names <- select(pgen4e_new_names, 
+                           -hhid, -pn, -any_of(dupe_PGS), -any_of(old_pgs))
 pgs_white <- names(pgen4e_new_names)
-pgs_white[which(pgs_white == "HEIGHT")] <- "HEIGHTpgirep"
+pgs_white[pgs_white == "HEIGHT"] <- "HEIGHTpgirep"
 
@@
 
-# TE 283 people report having grandkids, but having had no children. Weird!
-# This ugly code subsets the data by ethnicity then counts all observations
+# 83 people report having grandkids, but having had no children. Weird!
+# This code subsets the data by ethnicity then counts all observations
 # where grandkids is greater than 0 but number of kids is 0
-gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))
-
-gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn == 0, 1, 0)))
-
-# TE if grandkids but no kids div_gkid is infinite. Replacing with NA
+gkids_unusable_white <- rand |> 
+  filter(ethnicity == "white", h10gkid > 0, raevbrn == 0) |> 
+  nrow()
+gkids_unusable_black <- rand |> 
+  filter(ethnicity == "black", h10gkid > 0, raevbrn == 0) |> 
+  nrow()
+
+# if grandkids but no kids div_gkid is infinite. Replacing with NA
 rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA
 
-# TE added some variables for grandkid analysis
-# TE I've also added sampling weights in calculation of mean values
-# note the child weight is just weight* number of kids. so mean_h10gkid 
-# and mean_div_h10gkid are actually the same
 rand <- rand |>
@@
     rlrs_h10gkid = h10gkid/mean_h10gkid,
-    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
+    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, 
+                            na.rm = TRUE), 
     rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
@@
 # TE we can delete more
-rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)
+rm(pgen, pgen4a, pgen4e, pgirep, trk_weights)
 
@@
 
-Multiple steps were taken to prune the polygenic scores.
 For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository.
@@
 We choose to use polygenic scores trained only on individuals of European ancestry.
-we discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
+We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
 we also discard PGS for number of children ever born (but keep scores for age at first birth).
@@
 |               | Low group             | High group            | Difference            |
-|------------------|------------------|------------------|------------------|
+|-----------------|-------------------|-------------------|-------------------|
 | Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
@@
 ```{r}
-#| label: TE-pgicorrect
+#| label: stat-pgicorrect
 #| cache: true
@@
 # Writing data for pgi_correct to use
-write.table(subset(rand$variables, ethnicity == "white"),"pgi_correct_data.txt",
-            quote=FALSE, sep="\t")
+readr::write_tsv(
+  rand$variables |> filter(ethnicity == "white"),
+  "pgi_correct_data.txt",
+  quote = "none"
+)
 
@@
 # such as group categories taken from the PGI repository and simplified names
-correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
-                                            range="B1:U53")
+correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx", 
+                                     na = "NA", range = "B1:U53")
 
 # Where h2 and R2 are missing, we take values from a similar sample - the WLS
-correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2), correct_scores$WLS_h2,
+correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2), 
+                                 correct_scores$WLS_h2,
                                  correct_scores$HRS_h2)
 
-correct_scores$R2      <- ifelse(is.na(correct_scores$HRS_R2), correct_scores$WLS_R2,
+correct_scores$R2 <- ifelse(is.na(correct_scores$HRS_R2),
+                            correct_scores$WLS_R2,
                             correct_scores$HRS_R2)
@@
 
-correct_scores[which(correct_scores[, "pgs"] == "CANNABIS"), "pgs"] <- "CANNABISpgirep"
-correct_scores[which(correct_scores[, "pgs"] == "DPW"), "pgs"]      <- "DPWpgirep"
-correct_scores[which(correct_scores[, "pgs"] == "HEIGHT"), "pgs"]   <- "HEIGHTpgirep"
+correct_scores$pgs <- case_match(correct_scores$pgs,
+  "CANNABIS" ~ "CANNABISpgirep",
+  "DPW"      ~ "DPWpgirep",
+  "HEIGHT"   ~ "HEIGHTpgirep",
+  .default = correct_scores$pgs
+)
 
 # some of the PGS have estimates rounded to zero and are thus useless
-correct_scores <- subset(correct_scores, chip_h2 > 0)
-correct_scores <- subset(correct_scores, R2 > 0)
+correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
 
@@
 
-correct_scores[which(correct_scores$pgs == "AFB"), "twin_h2"]      <- 0.15
-correct_scores[which(correct_scores$pgs == "NEBwomen"), "twin_h2"] <- 0.28
+correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
+correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
 
@@
 
-# Loop for for calculating results with SNP heritabilities
-map_df(1:nrow(correct_scores), function(i) {
-
-system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"chip_h2"], ' --weights "weight"', sep=" "))
+run_pgic_once <- function (pgs, r2, h2, h2_type) {
+  pgic_command <- glue(
+     'python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt', 
+     '--outcome "rlrs" --pgi-var {pgs}',
+     '--covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A"',
+     '"PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2 {r2} --h2 {h2}',
+     '--weights "weight"', 
+     .sep = " ")
+  system(pgic_command)
 
@@
   system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
-reg_snp <- read.table("reg_results", header=TRUE)
+  pgic_results <- readr::read_table("reg_results", col_names = TRUE,
+                                    col_types = "cdddd")
+  file.remove("reg_results")
 
   tibble(
-  trait = as.character(correct_scores[i,"pgs"]), 
-  beta  = as.numeric(reg_snp[1,4]),
-  SE    = as.numeric(reg_snp[1,5]),
-  Heritability = "Chip"
-)
-}
-) ->
-  corrected_effects_snp
-
-# Loop for for calculating results with twin heritabilities
-map_df(1:nrow(correct_scores), function(i) {
-
-system(paste('python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt --outcome "rlrs" --pgi-var', correct_scores[i,"pgs"], ' --covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A" "PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2', correct_scores[i,"R2"], ' --h2 ', correct_scores[i,"twin_h2"], ' --weights "weight"', sep=" "))
-
-system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
-reg_twin <- read.table("reg_results",header=TRUE)
-
-tibble(
-  trait = as.character(correct_scores[i,"pgs"]), 
-  beta  = as.numeric(reg_twin[1,4]),
-  SE    = as.numeric(reg_twin[1,5]),
-  Heritability = "Twin"
-)
-}
-) ->
-  corrected_effects_twin
+    trait = pgs,
+    beta  = pgic_results[[1, 4]],
+    SE    = pgic_results[[1, 5]],
+    Heritability = h2_type
+  )
+}
+
+run_pgic_data <- function (score_data, h2_column, h2_type) {
+  results <- map(1:nrow(score_data), 
+    \(i) run_pgic_once(pgs = score_data$pgs[i], 
+                       r2 = score_data$R2[i],
+                       h2 = score_data[i, h2_column, drop = TRUE],
+                       h2_type = h2_type)
+  )
+
+  list_rbind(results)
+}
+
+corrected_effects_snp <- run_pgic_data(correct_scores, h2_column = "chip_h2",
+                                       h2_type = "Chip")
+corrected_effects_twin <- run_pgic_data(correct_scores, h2_column = "twin_h2",
+                                        h2_type = "Twin")
 
 corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
-corrected_effects$conf.low  <- corrected_effects$beta - 1.96*corrected_effects$SE
-corrected_effects$conf.high <- corrected_effects$beta + 1.96*corrected_effects$SE
+corrected_effects$conf.low <- corrected_effects$beta - 1.96 *
+                              corrected_effects$SE
+corrected_effects$conf.high <- corrected_effects$beta + 1.96 *
+                               corrected_effects$SE
 
@@
 
-corrected_effects <- merge(corrected_effects, correct_scores, 
-                           by.x="trait", by.y="Abr",prefix=c("",".y"))
+corrected_effects <- left_join(corrected_effects, correct_scores,
+                               by = join_by(trait == Abr),
+                               suffix = c("", ".y"))
 
@@
 
-corrected_effects$Category <- case_match(corrected_effects$Category, 
+corrected_effects$Category <-
+  case_match(
+    corrected_effects$Category,
     "Health and health behaviours"     ~ "Health",
@@
   )
+
+# Cleaning and saving corrected effects so people can have the exact figures
+corrected_effects$TwinH2_Source <- corrected_effects$Source
+corrected_effects |> 
+  select(trait, beta, SE, Heritability, chip_h2, twin_h2, R2, 
+           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
+  readr::write_csv("Corrected_effects.csv")
 
@@
 
-# This removes ADHD due to the low power of the score. Note ADHD predict RLRS after bonferroni! 
-# It also removes NEB
-corrected_effects <- subset(corrected_effects, R2 > 0.005)
-
 # TE Note that under Fisher's fundamental theorem of natural selection, the selection
@@
 
-my_labeller <- function(variable, value) {
-  return(value)
-  }
-
 corrected_effects |>
+  # This removes ADHD due to the low power of the score. 
+  # Note ADHD predict RLRS after bonferroni! 
+  # It also removes NEB
+  filter(R2 > 0.005) |> 
   mutate(
@@
     trait = fct_reorder(trait, beta)
-  )|>
+  ) |>
   ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
@@
   theme_minimal() +
-  facet_grid(Category ~ ., scales = "free_y", space = "free_y", labeller = my_labeller)+
+  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
   theme(strip.text.y = element_text(angle = -90,face = "bold"))
-
-# Cleaning and saving corrected effects so people can have the exact figures
-  corrected_effects$TwinH2_Source <- corrected_effects$Source
-
-  corrected_effects <- corrected_effects[, c("trait", "beta", "SE", "Heritability",
-                                            "chip_h2", "twin_h2", "R2", 
-                                            "TwinH2_Source", "PoldermanICF_ICD10_subchapter",
-                                            "Category")]
-
- write.csv(corrected_effects,"Corrected_effects.csv")
 

### 2023-12-04 hrs-selection.qmd
Cleanups

@@
 
-@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from \@becker1976child. The theory has two components.
+@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.
 
@@
 
-Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey \[\@hrsrandfam2023; \@hrsrand2023\].
+Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-The HRS also provides precalculated polygenic scores \[\@hrspgs2020\] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.
+The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.
 
@@
 This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.
-Statistical tests are adjusted for clustering and stratification using the R "survey" package \[\@lumleysurvey2023\].
+Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].
 
@@
 These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
-Appendix \@fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
+Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
 Standard errors are large because of the relatively low sample sizes.
@@
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
-Survey bootstraps \[\@canty1999resampling\] are used so as to make inferences from the sample of respondents.
+Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.
 
@@
 |               | Low group             | High group            | Difference            |
-|-----------------|-------------------|-------------------|-------------------|
+|------------------|------------------|------------------|------------------|
 | Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
@@
 
-where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype \[\@becker2021resource\].
+where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype [@becker2021resource].
 Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters.
@@
 
-system("rm pgi_correct_data.txt")
+file.remove("pgi_correct_data.txt")
 

### 2024-01-08 hrs-selection.qmd
TE update Jan 8

@@
 ---
-title: "Natural selection in the Health and Retirement Study"
+title: "Natural Selection Across Three Generations of Americans"
 author: 
@@
   partially support the economic theory of fertility as an explanation for 
+  natural selection: among both black and white respondents,
   scores which correlate negatively (positively) with education are
@@
   and unmarried parents, but not among younger parents or those with less 
-  education. We also estimate effect sizes corrected for noise in the 
-  polygenic scores. 
-  natural selection: among both white and black respondents,
-date: "November 2023"
+  education. Using respondents' number of siblings and grandchildren we 
+  were able to examine selection across three generations, finding the 
+  results consistent over time. We also estimate effect sizes 
+  corrected for noise in the polygenic scores, finding selection
+  for some health traits to be similar in magnitude to cognitive traits. 
+date: "January 2024"
 format: 
@@
 #| label: setup
-
-# TE quick TODOs change reference to time being insignificant. Fix the footnote citations
+#| cache-lazy: false
+
+# TE One time I tried to render the document with cache on I got the error message on the first chunk: "Error in `lazyLoadDBinsertVariable()`: !
+long vectors not supported yet: connections.c:6028" Added cache-lazy: false which seems to fix things
+
 options(digits = 3)
@@
 
-# TE edit PGI repository EUR only
 pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt",
@@
 rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
+
 # only selected columns to avoid creating suffixes when we join with
 # rand.
-# Age of oldest kid; grandchildren; num siblings who died
-rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)
+# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died
+rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied)
 
@@
 
-# TE I'm replacing 0 weights with NA because it can create odd problems
+# Replacing 0 weights with NA because it can create odd problems
 rand$weight <- na_if(rand$weight, 0)
@@
 # TE this is average number of kids that your kids have
-rand$div_h10gkid <- rand$h10gkid / rand$raevbrn
-
-# 83 people report having grandkids, but having had no children. Weird!
+rand$div_h13gkid <- rand$h13gkid / rand$raevbrn
+
+# Some people report having grandkids, but having had no children. Weird!
 # This code subsets the data by ethnicity then counts all observations
@@
 gkids_unusable_white <- rand |> 
-  filter(ethnicity == "white", h10gkid > 0, raevbrn == 0) |> 
+  filter(ethnicity == "white", h13gkid > 0, raevbrn == 0) |> 
   nrow()
 gkids_unusable_black <- rand |> 
-  filter(ethnicity == "black", h10gkid > 0, raevbrn == 0) |> 
+  filter(ethnicity == "black", h13gkid > 0, raevbrn == 0) |> 
   nrow()
@@
 # if grandkids but no kids div_gkid is infinite. Replacing with NA
-rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA
+rand$div_h13gkid[is.infinite(rand$div_h13gkid)] <- NA
 
@@
     rlrs = raevbrn/mean_raevbrn,
-    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
-    rlrs_h10gkid = h10gkid/mean_h10gkid,
-    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, 
+    mean_h13gkid = mean(h13gkid*weight/mean_weight, na.rm = TRUE),
+    rlrs_h13gkid = h13gkid/mean_h13gkid,
+    mean_div_h13gkid = mean(div_h13gkid*child_weight/mean_child_weight, 
                             na.rm = TRUE), 
-    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
+    rlrs_div_h13gkid = div_h13gkid/mean_div_h13gkid
   ) |>
@@
 
-# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary
-
 rand$neurot <- rand |> 
@@
          r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
-         rlrs_div_h10gkid, child_weight)
+         rlrs_div_h13gkid, child_weight, h13ageokid)
 
@@
 		nest = TRUE,
-		data = rand |> drop_na(child_weight) 
+		data = rand |> drop_na(child_weight) |>
+		        filter(!is.na( rlrs_div_h13gkid), h13ageokid > 39)
 		)
@@
 
-@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.
+@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components:
 
@@
 
-<!--#  TE sources of pgs are not mentioned in the data section. I've added that since we are now using two sources -->
-
 Genotyping took place in 2006, 2008 and subsequent years.
@@
 For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository.
-For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry.
+For some traits, polygenic scores were created from both European ancestry and multiple ancestry GWAS.
 We choose to use polygenic scores trained only on individuals of European ancestry.
@@
 
-The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by mean number of children of people born in the same year.
+The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year.
 RLRS is calculated pooling ethnicities, i.e.
@@
 # cache seems to cause trouble even with cache-lazy: FALSE 
+options("survey.lonely.psu" = "average")
 
@@
     form_pgs_gkids <- map(pgs, 
-                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
+                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
   mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
@@
 
-# TE bonferroni correction might make more sense focusing on white tests
-# Either way seems to produce the same result
-
 n_tests_white <- nrow(tidy_pgs$white)
-n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests)
+n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests_white)
 
 n_tests_black <- nrow(tidy_pgs$black)
-n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests)
+n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests_black)
 
@@
 #| label: stat-bootstrap-eth-cor
-#| cache: true
+#| cache: false
 #| cache-lazy: false
@@
 
-<!--# TE edited because 3 scores are bonferroni significant in whites.
-As of writing the same same three scores were signficiant regardlesss of whether we pool the blacks and whites in the bonferroni calculation.-->
-
 We estimated coefficients of PGS on RLRS among black and white respondents separately.
@@
 Standard errors are large because of the relatively low sample sizes.
-`r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`.
-The scores are age at first birth, educational attainment and attention deficit disorder.
+In the white sample, `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
+The scores are age at first birth, educational attainment, ADHD, and self-rated health.
 But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.
@@
 #| label: stat-bootstraps
-#| cache: true
+#| cache: false
 #| cache-lazy: false
@@
   cors_gkids_edu[[eth]] <- 
-    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h10gkid")
+    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h13gkid")
 }
@@
 
-@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment.
+@fig-rlrs-edu plots each PGS's regression beta on RLRS against its regression beta on educational attainment.
 The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I.
@@
 # TE correlation without outliers is not NEG
+# But it is not negative anyway now
 ```
@@
 #| label: stat-bootstrap-gen-cor
-#| cache: true
+#| cache: false
 
@@
 #| label: stat-bootstrap-gen-cor-gkids
-#| cache: true
+#| cache: false
 
@@
 f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
-f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
+f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
 
@@
   m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
-  m_gkids <- map(f_gkids, \(x) lm(x, data, weights = w*raevbrn))
-
+  m_gkids <- map(f_gkids, \(x) lm(x, data |> filter(h13ageokid > 39), weights = w*raevbrn))
+  # data subset so respondent's kids have 1+ married kids
   coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
@@
 ```
-
-<!--# TE Added grandkids and fixed minor typos -->
 
@@
 There is a positive and significant correlation across generations, i.e.
-between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`,cbootstrapped 95% C.I.
+between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I.
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
@@
 In other words, we calculate the average number of children the respondent's children have.
-This measures the reproductive success in the second generation which we can call RS2.
-`r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having had no children.
+This indicates reproductive success in the second generation which we can call RLRS2.
+`r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having reported no children.
 These individuals were removed for this test.
 In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.
+
 Older grandparents will have more time for their number of grandkids to accumulate.
-To deal with this time trend, we divide RS2 by the mean RS2 of respondents born in the same year, creating relative RS2 (RRS2).
-
-We regress the respondent's (the grandparent's) PGS on RRS2.
+To better deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded.
+Appendix @fig-offspring-cohorts shows the distribution of the birth year of the oldest child, adjusted for sampling weights.
+A majority of the observations are after $1960$, the most recent year the respondents were born in.
+This reassures us that we are observing reproduction in later generations.
+
+We regress the respondent's (the grandparent's) PGS on RLRS2.
 The resulting coefficient is only a proxy for selection in the next generation, not a perfect measure.
@@
 
-Appendix @fig-gkids plots coefficients on RRS2 versus coefficients on years of education.
+Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education.
 Correlations are significant for whites but not blacks.
-Standard errors were moderate in whites, but large in blacks (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I.
+Standard errors were large in both ethnicites (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I.
 `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I.
 `r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`).
-There is a strong positive correlation between the parent and grandchild regressions, i.e.
-between PGS coefficients on RRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I.
-`r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`).
+There is a positive correlation between PGS coefficients on RLRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I.
+`r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.
 
@@
 #| label: stat-x-bootstrap-cors
-#| cache: true
+#| cache: false
 #| cache-lazy: false
@@
 #| label: stat-x-bootstrap-effect-size
-#| cache: true
+#| cache: false
 #| cache-lazy: false
@@
 n_reps_x_es <-  199
+pgs <- pgs_white
 
@@
 #| label: stat-bootstrap-agefb
-#| cache: true
+#| cache: false
 
 n_reps_agefb <-  199
+pgs <- pgs_white
 
@@
             ci_discordant[2] >= length(pgs)/2)
+# TE THEY ARE DISCORDANT
 
@@
 # selection effects controlling for age at first birth
-# 
+pgs <- pgs_white
+
 form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
@@
 #| label: stat-x-bootstrap-birthyear
-#| cache: true
+#| cache: false
 
 n_reps_birthyear <-  199
+pgs <- pgs_white
 
@@
 
+```{r}
+#| label: ADHD
+#| cache: false
+
+# Testing the R2 of the ADHD PGS
+rand$variables |>
+  with(cor.test(adhd_score, ADHD, use = "pairwise.complete")) -> adhd_cor
+
+```
+
 Lastly, we would like to know natural selection's effect sizes.
@@
 
-where \$h\^2\$ is the heritability of the PGS target phenotype and \$R\^2\_{PGS}\$ is the \$R\^2\$ of the measured PGS on the target phenotype [@becker2021resource].
-Moving from error correction in a univariate model to a multivariate model requires a slightly more complex correction, although it uses the same parameters.
-To do this we use the error-corrected estimator developed by \@becker2021resource.
+where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource].
+Moving from error correction in a univariate model to a multivariate model, controlling for principal components of genetic array data, requires a slightly more complex correction, although it uses the same parameters.
+To do this we use the error-corrected estimator developed by @becker2021resource.
 
@@
 To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2].
-We remove polygenic scores with $R^2_{PGS} > 0.005$ to focus on PGS with adequate power.
-This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction.
+We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power.
+This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^3].
 
@@
 For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].
+
+[^3]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms.
+These were PV001-PV018 in the 2016 core survey.
+They estimated an $R^2_{PGS}$ of $0.003$.
+When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`.
+Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I.
+`r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`.
+Given the great uncertainty over the true correlation we are unable to justify using the error-corrected estimator with the ADHD polygenic score.
 
@@
 Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
-Age of First Birth has a twin corrected heritability of 0.15, implying that it maybe under fast selection.
-The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true \$h\^2\$ , the true \$R\^2\_{PGS}\$, limitations of the within-sample phenotypes, noise from correlated environments, and for twin-heritability, different relationships with fertility among variants not measured on the chip.
+The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations.
+For twin-heritability, different relationships with fertility among variants not measured on the chip.
 Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
@@
 For these traits, the confidence intervals on twin heritabilities are enormous.
-Error-corrected effect sizes and the $R^2$ and $h^2$ parameters used can be found on the Github page for this paper.
-
-<!--# TE The lay public want to know how much IQ is changing. So I've added in a calculation of this. -->
+Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.
 
@@
 
-<!--# TE It's hard to keep in the sanity check since we are jumping from LDSC back to GCTA to calcualte snp heritability with pgenscores.
-The above also uses PGI repository scores only.
-I can add the sanity check back in.
-To be fair, I'm only aware of the method being tested with different discovery sample sizes using simulated data (link below) the sanity check is useul.
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10368647/-->
-
 ```{r}
 #| label: stat-pgicorrect
-#| cache: true
+#| cache: false
+
+# Testing the R2 of the ADHD PGS
+rand$variables |>
+  with(cor.test(adhd_score, ADHD,, use = "pairwise.complete")) -> adhd_cor
 
@@
 
+# This removes NEBwomen from the data
+correct_scores <- correct_scores[-which(correct_scores$pgs == "NEBwomen"),]
+
 correct_scores$pgs <- case_match(correct_scores$pgs,
@@
 correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
-correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
+#correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
 
@@
 
-file.remove("pgi_correct_data.txt")
+unlink("pgi_correct_data.txt")
 
@@
 #| label: stat-bootstrap-true-effects
-#| cache: true
+#| cache: false
 #| eval: false
@@
 The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason.
-But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.
-
-[^3]: See @balbo2013fertility for a broad review of fertility theories.
-
+But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^4] In this sense, results here are less supportive of the economic theory specifically.
+
+[^4]: See @balbo2013fertility for a broad review of fertility theories.
+
+We also examined natural selection in different groups.
+Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment.
+Patterns of selection appear consistent across the twentieth and early twenty-first century.
+The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time.
 The smaller black sample makes most tests inconclusive for this population: we can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.
-Lastly, estimated effect sizes of "true" polygenic scores are typically between 0-0.2 standard deviations.
-To know more, we must await more accurate scores.
+
+Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations.
+Our estimate of how much intelligence is changing owing to natural selection, is similar to estimates given in the prior literature.
+For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation.
+However, cognitive traits are only one part of the story of natural selection in humans.
+Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits.
+The selection differential for self-rated health is greater than those of education and cognitive performance.
+The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human].
+Future research should study the health and medical implications of natural selection, in addition to their social implications.
+To know more, we must await more accurate polygenic scores.
 
@@
 #| label: fig-cohorts
-#| fig-cap: Distribution of birth years for the sample
-
+#| fig-cap: Distribution of birth years adjusted for sampling weights.
+
+# TE old version saved here in case we want to revert
+
+#rand$variables |> 
+#    mutate(
+#      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
+#    ) |> 
+#    ggplot(aes(rabyear, fill = Ethnicity)) + 
+#      geom_bar(position = "identity") +
+#      labs(x = "Birth year", y = "Count") +
+#      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
+#      theme_minimal() +
+#      theme(legend.title = element_blank())
+
+# Various packages and functions exist for making a weighted barplot
+# But the y axes show the count for the population, not a representative subsample. So I've made the weighted plot mannually. This way the y axis has sensible values rather than millions
+
+rand$variables <- rand$variables |> 
+    mutate(
+      weights_sample = (n_white+n_black)*weight/sum(weight)
+    )
+
+year_counts <- tibble()
+
+for (eth in eths) {
   rand$variables |>
+    subset(ethnicity == eth) |>
+    count(rabyear, wt = weights_sample) |>
+    cbind(eth) |>
+    rbind(year_counts) -> year_counts
+}
+
+year_counts |>
   mutate(
-      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
+      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
     ) |> 
-    ggplot(aes(rabyear, fill = Ethnicity)) + 
-      geom_bar(position = "identity") +
+  ggplot(aes(x = rabyear, y = n, fill = Ethnicity)) + 
+      geom_bar(stat="identity") +
       labs(x = "Birth year", y = "Count") +
@@
 
+```
+
+```{r}
+#| label: fig-sibling-cohorts
+#| eval: false
+#| fig-cap: Distribution of birth years adjusted for parent weights.
+
+# Ideally we'd like to know the time period we're analysing when we look at selection in the parent's generation.
+Unfortuantely we don't have the parents' age of birth only the parent's age if they are still alive or the age at which they died.
+
+# The next best thing we can do is look at the respondent's birth year adjusted for the parent weight.
+But this isn't very interesting and almost identical to weight by the respondent weight so the graph is left unevaluated.
+
+r_parents$variables <- r_parents$variables |> 
+    mutate(
+      weights_sample = (n_white+n_black)*parent_weight/
+                        sum(parent_weight)
+    )
+
+year_counts <- tibble()
+
+for (eth in eths) {
+  r_parents$variables |>
+    subset(ethnicity == eth) |>
+    count(rabyear, wt = weights_sample) |>
+    cbind(eth) |>
+    rbind(year_counts) -> year_counts
+}
+
+year_counts |>
+   mutate(
+      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
+    ) |> 
+  ggplot(aes(x = rabyear, y = n, fill = Ethnicity )) + 
+      geom_bar(stat="identity") +
+      labs(x = "Birth year", y = "Count") +
+      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
+      theme_minimal() +
+      theme(legend.title = element_blank())
+
+```
+
+```{r}
+#| label: fig-offspring-cohorts
+#| fig-cap: Distribution of birth years for the oldest child of respondents, adjusted for sampling weights.
+
+n_oldest <- r_child$variables |> 
+     filter(!is.na(rlrs_div_h13gkid)) |>
+             nrow()
+
+child_graph_data <- r_child$variables |> 
+    filter(!is.na(rlrs_div_h13gkid)) |>
+    mutate(
+      yearageokid = 2016 - h13ageokid,
+      weights_sample = n_oldest*child_weight/
+                        sum(child_weight)
+      )
+
+year_counts <- tibble()
+
+for (eth in eths) {
+  child_graph_data |>
+    subset(ethnicity == eth) |>
+    count(yearageokid, wt = weights_sample) |>
+    cbind(eth) |>
+    rbind(year_counts) -> year_counts
+}
+
+year_counts |>
+  mutate(
+      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
+    )  |> 
+  ggplot(aes(x = yearageokid, y = n, fill = Ethnicity )) + 
+      geom_bar(stat="identity") +
+      labs(x = "Birth year", y = "Count") +
+      scale_fill_manual(values = c(Black = "orange2", White =
+      "steelblue4")) +
+      theme_minimal() +
+      theme(legend.title = element_blank())
+
+mean_yold_kid <- sum(year_counts$yearageokid*
+                       year_counts$n)/sum(year_counts$n)
 ```
@@
       labs(
-          title = eth,
+          title = titles[eth],
           x = "Coefficient on years of education",
@@
 #|   Scatterplot of PGS coefficients on RLRS2 and years of
-#|   education. RRS2 is the reproductive success of respondent's offspring relative to the success of other offspring. Controls include 10 principal components of genetic array
+#|   education. RLRS2 is the reproductive success of respondent's offspring relative to the success of other offspring. Controls include 10 principal components of genetic array
 #|   data. Dashed lines show linear regressions.
@@
       labs(
-          title = eth,
+          title = titles[eth],
           x = "Coefficient on years of education",
-          y = "Coefficient on RRS2"
+          y = "Coefficient on RLRS2"
       )

### 2024-02-16 response-to-reviewers-revision-1.qmd
Begin response to reviewers

---
title: "'Natural Selection Across Three Generations of Americans': response to reviewers"
format: docx
---

Dear Professor Medland,

Thank you for the opportunity to revise our manuscript. Thank you also to the
reviewers. We respond to their points below.

One major change is that I have asked Tobias Edwards to be a co-author. I
discovered that he was working on similar topics, so we merged our efforts.
I hope this is OK. One effect is just that we have more polygenic scores to
analyse. The original results are 
not much changed, which is encouraging. TODO and what else?
XXX Another is that we look in more detail at two extra generations,
using children and grandparents of the respondents. We've given the paper
a title which reflects that.

With best wishes,

David Hugh-Jones

Reviewer Comments:

# Reviewer 1

*Thanks for the opportunity to read the manuscript ‘Natural selection in the Health and Retirement Study’. This study is of great interest to the behavior genetics and population genetics community and beyond. It demonstrates ongoing natural selection in a descriptive manner and partly confirms ideas about the trade-off between childbearing and work. I have a few comments.*

*First, it would be good to state in the abstract the expectations about the tested interactions. Otherwise it reads as though interactions with low income etc have been tested arbitrarily.*

*Second, in the introduction, it is implied that people with larger human capital work more, but it remains unclear why. Only in the discussion section, it is elaborated on the idea of substitution effect, that larger human capital leads to larger income loss when work is reduced for alternative activities. Note that this phenomenon is also known as opportunity costs of larger human capital. This should be explained early and explicitly as in absolute terms, one might argue that larger human capital increases net income and therefore might lead to working less and having more children with the same income.*

*In a similar vein, it would be very welcome to reiterate reason why the expected effect is stronger amongst low income etc people.*

We now detail the theory more in the abstract, and we've described the substitution
effects and income effects straight away in the intro. We've tried to keep the
language as simple and intuitive as possible.

*I was wondering why the authors control for PCAs in the PGI analyses. As stated by the authors, this study is not a causal study but one of correlations. The consequence of a correlation (natural selection) should be independent of its causes.*

Note that controlling for principal components in a regression,
*y ~ PGIs + PCs*, gives the
same effect sizes as running an uncontrolled regression on PGIs which have
previously had the PCs residualized  out, *y ~ PGIs_resid*. We control rather than
residualizing because this gives the correct standard errors^[Because the controls are correctly accounted for when calculating degrees of freedom.], which might matter 
with our relatively low N.

Why residualize or control for principal components at all? I think for two reasons. First, we are ultimately interested in selection on the "true PGI". Stripping out PCs
is an admittedly imperfect way to control for confounding population stratification.
Second, PGIs are typically residualized in this way, so by examining the effect
of these we're following the "industry standard" and learning about scores in the
form they are actually used.

*In general, the authors state that they are less interested in significance but in effect sizes. First, this is very generous. Whilst I find this acceptable, it would be good to read one or two more sentences on why the authors think this is legitimate in this case. Second, I believe such a descriptive approach might be useful studying the patterns of natural selection, however, when it comes to the interactions, the author state that effects ‘Appear larger for unmarried’, for example. It remains unclear to me when an effect is considered larger or not, which becomes important as the author differentiates between interactions which are real and those which are not. What are the criteria here.*

First, one could argue that the literal null hypothesis is not really plausible here. 
If *any* PGI is under selection in the population (clearly true), then
some underlying variants must also be under selection. Since all PGIs are just 
weighted sums of underlying variants, it will then be a knife-edge
case for any other PGI to have exactly zero correlation with fertility. What
matters more is whether effect sizes are big, small or trivial.

More specifically, we're interested in the pattern of selection effects 
across different PGIs,
and how it correlates with the PGI's relationship to income. So we want to
include PGIs with small selection effects, which we predict to have a low 
absolute correlation to income. Given the sample size, these PGIs are very likely
to have insignificant individual selection effects, but we shouldn't
ignore them in the cross-PGI analyses.

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample. TODO: do we always?

*The authors state that the correlation between the PGI and fertility quantifies natural selection. However, to my knowledge it is the covariance following the Robert-Price Identity: Morrissey, M. B., L. E. B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's equation in observational studies of natural populations." Journal of evolutionary biology 23.11 (2010): 2277-2288.*

TODO: Tobias, can you check the relevant para? I guess that if both PGI and RLRS are
normalized then the covariance is the correlation; but actually isn't the correct
thing covariance with the not-normalized RLRS?

*The authors might want to discuss the expected selection in terms of relevance. They quantify it but they do not discuss these quantifications. This is very important shaping our expectation of ongoing natural selection. It might also be very interesting to know more about whether those trends follow the secular trends in phenotypes.*

We have now added a final paragraph to the Discussion section on this.
In particular we discuss PGIs for health phenotypes, as a corrective to
the existing literature which is perhaps a bit over-focused on cognition and
IQ.

TODO: Could we look at health trends in phenotypes? I'd guess
that health trends improve over the C20, but we could look at some specifics.
(Depression/BMI?) One big issue here is that maybe the PGIs have confounds...
but perhaps that's less serious for some of these phenotypes

*Finally, the statement ’in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (-0.319), and for 42 out of 58 PGS, coefficients have opposite signs. However, in bootstraps, these statistics are not significantly different from 0 and 29 respectively.’ was not entirely clear to me. What does it mean that 42 PGS had opposite sign effects? Does that mean for old versus younger parents’ coefficients were very different, even in the opposite direction? How does this relate to the first part of the sentence that coefficients were negative. Also, why has this been tested against 0 and 29?*

The earlier claim is not "coefficients are negative overall" but 
"coefficients among younger parents are negatively correlated with coefficients
among older parents". We are dealing with correlations of correlations: this
can get confusing.

Negatively correlated, and more strongly, opposite-signed, effects between
young and old parents support the theory that income effects dominate for
older parents while substitution effects dominate for younger parents. To
be a bit more explicit: if a PGS simply doesn't correlate with human capital,
we expect it to not be selected on among either group of parents. If a PGS
is correlated with human capital, that will lead to income and substitution effects
for those with more human capital. If the substitution effect dominates for
younger parents then the PGS will have a negative correlation with fertility;
if the income effect dominates for older parents, then the PGS will have a positive
correlation with fertility.

The test against 29 had the following logic: if all effects were really zero,
and estimates were pure noise with mean 0, then we'd expect opposite-signed and same-signed "effects" to be equally likely; so, 29 is half of the 58 scores. However,
this is a rather specific null. A more natural null is that selection effects
are non-zero, but the same among both groups of parents. Then 
young-parent effects would be equally likely to be higher, or lower, than old-parent
effects.

TODO: can we figure out a test for this?

Test: take difference between scores, times the sign of the low group score.
Then under the null we'd expect as many + as - differences.

Really we should be doing the right test always! If abs size isn't enough...!

# Reviewer 2

*The author sought to examine natural selection as indicated by polygenic scores in a contemporary representative sample of the U.S. and found that the results partially support the economic theory of fertility as a result of natural selection among both white and black individuals. The authors' hypotheses are based primarily on the economic theory of fertility, which emphasizes a trade-off between the number of children and income: i) higher income makes raising children costlier, but ii) higher income makes having more children more affordable. As far as I understand the results of the paper, the results support prediction 1, but are more ambiguous for prediction 2?*

*However, in my opinion the theoretical database, data and analysis are adequate, the problem I have is that the MS is a bit confusing. It should be made clearer why an analysis was done and what the results mean in terms of rejecting or accepting the theoretical assumptions. The possible link to biological selection should also be explained in more detail, as this has been done in the case of education, income and reproduction to some extent on the basis of comparable data (Kong et al. 2017, Beauchamp 2016, Fieder & Huber 2002) and also in many other studies.*

We have clarified the theory early on in the intro and abstract.
TODO: Write a paragraph on biol. selection. (Not sure what is meant here?)

In terms of accepting or rejecting our theory, our paragraph in the discussion
was:

> The results here support prediction 1 but are more ambiguous for prediction 2. PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents. But there is little evidence for larger coefficients among people with lower education, or younger parents. This may be due to the low sample size. But in the UK, the between-group differences were large; differences that big would surely have been visible here. The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason. But note that *any* theory with a negative relationship between education and fertility will give prediction 1. In this sense, results here are less supportive of the economic theory specifically.

We don't think these results give a simple thumbs up or down to the theory,
but we're open to suggestions to clarify what we say here.

*The division into a "black" and a "white" is necessary due to technical reasons, but further complicates the results/discussion. Presumably the results for the larger sample (I think in the case of the HRS this are the  whites) should be presented in the main text, and the results for the smaller population moved into a supplement?*

TODO: yes, I think that is probably fine.

*Also, there seem to be many associations of PGS with RLRS (Figure 4), such as ADHD, diabetes, etc., which I think are interesting findings, even if of low significance, worth at least a brief discussion.*

TODO: more than the below?

We now discuss some health-related traits in the conclusion:

> Whilst past research has focused on [cognitive] traits, we find selection 
differentials are often of similar magnitudes for health related traits. The 
selection differential for self-rated health is greater than those of education 
and cognitive performance. The most significant, positively selected trait was 
ADHD, which was also found in the UK [@hugh2022human]. Future research should 
study the health and medical implications of natural selection, in addition to 
their social implications.

*What is the data base for the mean number of children of people born in the same year - is pooling appropriate, as the means may vary considerably according to ethnicity?*

Means do vary (TODO how much by?). But note that we analyse the ethnicities
separately anyway, so the only effect of pooling is to scale all the selection effects
within each ethnicity.

As a matter of scientific correctness, we think it's mostly best to think of both
ethnic groups as one population. Interracial marriage was very low for much
of the twentieth century, but this was a fact about laws and social norms, not
about biology.

TODO: not sure about this. Possibly the guy is correct. The only time it matters
is when we care about absolute effect sizes. Then, using separate means is
like treating white people as a single population and asking how much means will
change in the white population.  Pooling means is like treating white people as
part of the whole US population. If we then make claims about how means for that
whole population will change, we are effectively assuming that effect sizes are 
the same across ethnicities, which seems unlikely.

### 2024-02-16 hrs-selection.qmd
Begin response to reviewers

@@
   partially support the economic theory of fertility as an explanation for 
-  natural selection: among both black and white respondents,
-  scores which correlate negatively (positively) with education are
-  selected for (against). Selection coefficients are larger among low-income
+  natural selection: scores which correlate negatively (positively) with
+  education are selected for (against). The theory predicts that
+  selection coefficients should be stronger among low-income, 
+  less educated, unmarried and younger parents, but these predictions are only
+  partly borne out: coefficients are larger among low-income
   and unmarried parents, but not among younger parents or those with less 
-  education. Using respondents' number of siblings and grandchildren we 
-  were able to examine selection across three generations, finding the 
-  results consistent over time. We also estimate effect sizes 
-  corrected for noise in the polygenic scores, finding selection
-  for some health traits to be similar in magnitude to cognitive traits. 
-date: "January 2024"
+  education. Using respondents' number of siblings and grandchildren, we 
+  examine selection across three generations, finding consistent results 
+  over time. We also estimate effect sizes corrected for noise in the
+  polygenic scores. Selection for some health traits is similar in magnitude
+  to that for cognitive traits. 
+date: "February 2024"
 format: 
@@
 There is a trade-off between time spent working and raising children.
-This leads people with more human capital and higher expected wages to have fewer children.
+This "substitution effect" leads people with more human capital and higher expected wages to have fewer children.
 Evidence for this is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e.
@@
 The trade-off is sharper for low-income people, people with low human capital, and single parents.
+Because these groups value income more at the margin, the substitution effect is stronger for them.
+In other groups, the substitution effect is balanced by the "income effect", that children become more affordable when you get richer.
 As a result, natural selection is stronger among these groups.
@@
 There are no large differences in absolute effect size between younger and older parents.
-As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs.
+Just as in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs.
 However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively.
@@
 Lastly, we would like to know natural selection's effect sizes.
-The bivariate correlations of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
+The bivariate correlation of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
 Polygenic scores contain error, so estimated correlations are biased towards zero compared to the correlation of the true PGS.
@@
 The economic theory of fertility is driven by the trade-off between children and income.
-On the one hand, a higher hourly wage makes time raising children more costly (a "substitution effect").
+On the one hand, a higher hourly wage increases the opportunity cost of raising children (a "substitution effect").
 On the other hand, higher expected income makes children more affordable (an "income effect").
-Prediction 1 of the theory holds when substitution effects dominate income effects.
-Prediction 2 is driven by the specific form of individuals' preferences for income: when utility for income is sharply curved, the substitution effect is stronger for those who expect to earn less.
+Prediction 1 of the theory holds when substitution effects dominate income effects overall.
+Prediction 2 is driven by the specific form of individuals' preferences for income: when utility for income is sharply curved, i.e.
+marginal utility is high at low incomes but low at high incomes, the substitution effect is stronger for those who expect to earn less.
 

### 2024-02-16 response-to-reviewers-revision-1.qmd
specific todos and who todo

@@
 When we examine relationships across scores, we indeed test for
-significance, always by bootstrapping the sample. TODO: do we always?
+significance, always by bootstrapping the sample.
 
@@
 
-TODO: Tobias, can you check the relevant para? I guess that if both PGI and RLRS are
+TE TODO: Tobias, can you check the relevant para? I guess that if both PGI and RLRS are
 normalized then the covariance is the correlation; but actually isn't the correct
@@
 
-TODO: Could we look at health trends in phenotypes? I'd guess
+TE TODO: Could we look at health trends in phenotypes? I'd guess
 that health trends improve over the C20, but we could look at some specifics.
@@
 
-TODO: can we figure out a test for this?
+DHJ TODO: can we figure out a test for this?
 
-Test: take difference between scores, times the sign of the low group score.
-Then under the null we'd expect as many + as - differences.
-
-Really we should be doing the right test always! If abs size isn't enough...!
+* Test: take difference between scores, times the sign of the low group score.
+  Then under the null we'd expect as many + as - differences. Do this and compare
+  to 29, always, put it in table 1.
 
@@
 We have clarified the theory early on in the intro and abstract.
-TODO: Write a paragraph on biol. selection. (Not sure what is meant here?)
+TE TODO: Write a paragraph on biol. selection. (Not sure what is meant here?)
 
@@
 We don't think these results give a simple thumbs up or down to the theory,
-but we're open to suggestions to clarify what we say here.
+but we're open to suggestions that would help clarify what we say.
 
@@
 
-TODO: yes, I think that is probably fine.
+DHJ TODO: yes, I think that is probably fine.
 
@@
 
-TODO: not sure about this. Possibly the guy is correct. The only time it matters
-is when we care about absolute effect sizes. Then, using separate means is
-like treating white people as a single population and asking how much means will
-change in the white population.  Pooling means is like treating white people as
-part of the whole US population. If we then make claims about how means for that
-whole population will change, we are effectively assuming that effect sizes are 
-the same across ethnicities, which seems unlikely.
+DHJ/TE TODO: change it to be separate RLRS baseline for white and black; add 
+a couple of sentences discussing the implications. (Effectively examining a
+subpopulation as an imperfect proxy for the whole population.)

### 2024-02-16 response-to-reviewers-revision-1.qmd
More responses on tests and black/white subsamples

@@
 The test against 29 had the following logic: if all effects were really zero,
-and estimates were pure noise with mean 0, then we'd expect opposite-signed and same-signed "effects" to be equally likely; so, 29 is half of the 58 scores. However,
-this is a rather specific null. A more natural null is that selection effects
-are non-zero, but the same among both groups of parents. Then 
-young-parent effects would be equally likely to be higher, or lower, than old-parent
-effects.
+and estimates were pure noise with mean 0, then we'd expect opposite-signed and same-signed "effects" to be equally likely; so, 29 is half of the 58 scores.
+
+However, thinking about this issue a bit more made us change our tests more
+generally. The original situation was that we tested most pairs of coefficients
+across sample subgroups by comparing their absolute size. Coefficients for
+older and younger parents had similar absolute sizes, but we noticed they were
+opposite-signed and hence ran the tests above. This is not very satisfactory: it's 
+too like special pleading, and it also suggests the original tests weren't adequate.
+
+We now improve our tests in table 1 as follows. We take the null hypothesis
+to be "selection coefficients are the same in the different subgroups, for all
+PGIs". Under the null, the true difference between selection coefficients is zero; the 
+difference in our sample (and in bootstrapped subsamples) should then not
+be significantly different from zero. Our alternative hypothesis is that
+selection coefficients are smaller among the richer/more educated/older/married
+group. That is, if the coefficient among the poorer group is $x_p$, then
+
+$x_r \times sign(x_p) < x_p \times sign(x_p)$
+
+Note that, unlike tests of absolute values, the above always holds if 
+$x_r$ and $x_p$ are opposite signed.
+
+Under the null, we'd expect the above equation to hold for just 50%
+of the scores. If it holds for significantly more than 50% of the scores
+(using bootstrapped C.I.s) then we can reject the null in favour of the 
+alternative.
+
+We also try a simpler alternative approach. Using the few PGIs with the
+largest and most precisely estimated selection effects in the whole sample,
+we estimate selection effects in different subgroups and test for a significant
+difference between them, via a simple interaction term. We report these results
+too.
+
+TE TODO: the above^^^
 
 DHJ TODO: can we figure out a test for this?
-
-* Test: take difference between scores, times the sign of the low group score.
-  Then under the null we'd expect as many + as - differences. Do this and compare
-  to 29, always, put it in table 1.
 
@@
 
+We're OK with this, since the black sample is really too small to say much.
+
 DHJ TODO: yes, I think that is probably fine.
@@
 
-Means do vary (TODO how much by?). But note that we analyse the ethnicities
-separately anyway, so the only effect of pooling is to scale all the selection effects
-within each ethnicity.
+We thought about this, and agree that it is better to use mean number of 
+children of whites and of blacks separately for each ethnic group.
 
-As a matter of scientific correctness, we think it's mostly best to think of both
-ethnic groups as one population. Interracial marriage was very low for much
-of the twentieth century, but this was a fact about laws and social norms, not
-about biology.
+Neither option is perfect. If we calculate RLRS separately for whites and
+blacks, we are effectively treating them as separate breeding populations.
+Black and white people had low rates of intermarriage in the 20th century but
+that is a fact of law and culture, not of biology. Conversely, if we calculate
+RLRS by taking the mean number of children across the whole population, we
+end up effectively making claims about natural selection in the US population
+as a whole, by taking the white population as representative - which it very
+likely isn't.
+
+Overall we prefer the former option, so we have now done so and added a brief
+discussion. Effectively, the white US population is a "single data point" for
+theory testing. (Even if we could pool the whole US population, that is still
+not the whole human population; this point is not much discussed in the literature.)
 

### 2024-02-19 hrs-selection.qmd
Put black sample in appendix. Rework statistical tests and graphs.

@@
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.
+Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans.
+To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2.
+(TODO: or can we be stronger?)
 
@@
 we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
-The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.
+The resulting sample contains `r n_white` genotyped white participants.
+We focus on these because the sample size is large enough.
+The appendix reports the most basic analyses for the `r n_black` genotyped black participants.
 
@@
 we also discard PGS for number of children ever born (but keep scores for age at first birth).
-This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants.
-PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score.
+This leaves a total of `r length(pgs_white)` scores for the white participants.
+PGS are rescaled to zero mean and unit variance.
 In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.
+For our purposes, this is equivalent to residualizing on principal components [@wurm2014residualizing].
+Effectively, our analysis is of the residualized PGS that are widely used in practical applications.
+\[TODO actually we could just residualize; only the SE of the PCs is affected\]
 
@@
 
-We estimated coefficients of PGS on RLRS among black and white respondents separately.
+We estimate coefficients of PGS on RLRS.
 These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
-Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample.
+Appendix @fig-rlrs shows coefficients.
 Standard errors are large because of the relatively low sample sizes.
-In the white sample, `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
+`r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
 The scores are age at first birth, educational attainment, ADHD, and self-rated health.
@@
 #| label: stat-bootstraps
-#| cache: false
+#| cache: true
 #| cache-lazy: false
-#| cache-comments: false
+#| cache-comments: true
+
+# TODO set cache to false before final build
 
@@
 
-  if (eth == "white") {
-    pgs <- pgs_white
-  } else {
-    pgs <- pgs_black
-  }
+  pgs <- if (eth == "white") pgs_white else pgs_black
 
@@
 @fig-rlrs-edu plots each PGS's regression beta on RLRS against its regression beta on educational attainment.
-The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I.
-`r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I.
-`r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
+The relationship is negative (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I.
+`r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`).
 Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.
@@
 #|   data. Dashed lines are fitted from linear regressions.
-#| fig-subcap:
-#|   - ""
-#|   - ""
-#| layout-nrow: 2
-#| column: page
-
-for (eth in eths) {
-  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
+
+ggplot(tidy_pgs_joined$white, aes(estimate.edyrs, estimate.fert)) +
    geom_hline(yintercept = 0, color = "grey80") +
@@
    labs(
-            title = titles[eth],
      x = "Coefficient on years of education",
@@
    )
-  print(ggp)
-}
-
-```
-
-```{r}
-#| label: stat-siblings
-
-tmp <- 
-  tidy_pgs_sibs$black |> 
-  inner_join(tidy_pgs_ed$black, 
-             by = join_by(term), 
-             suffix = c(".sibs", ".edyrs")) |> 
-  filter(term != "GWAD2NA", term != "GWAD2WA")
-
-cor_no_outliers <- cor(tmp$estimate.sibs, tmp$estimate.edyrs)
-# if (! cor_no_outliers < 0) stop("Correlation without outliers IS NOT neg.")
-# TE correlation without outliers is not NEG
-# But it is not negative anyway now
+
 ```
@@
 #| label: stat-bootstrap-gen-cor
-#| cache: false
+#| cache: true
+#| cache-comments: true
+
+# TODO set cache to FALSE before submission
 
@@
 #| label: stat-bootstrap-gen-cor-gkids
-#| cache: false
+#| cache: true
+#| cache-comments: true
+
+# TODO set cache to FALSE before submission
 
@@
 Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
-Correlations are only significant and negative in the white respondents, although ethnicities have large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
-`r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I.
-`r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).
+Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`).
 There is a positive and significant correlation across generations, i.e.
-between PGS coefficients on number of siblings and number of children (among whites: `r gen_cor`, bootstrapped 95% C.I.
+between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I.
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
 
-To examine selection in the succeeding generation, we divide the respondents' number of grandchildren by their number of children.
-In other words, we calculate the average number of children the respondent's children have.
-This indicates reproductive success in the second generation which we can call RLRS2.
-`r gkids_unusable_white` white respondents and `r gkids_unusable_black` black respondents report having grandchildren despite reporting having reported no children.
-These individuals were removed for this test.
-In regression we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.
+To examine selection in the next generation, we divide the respondents' number of grandchildren by their number of children.
+In other words, we calculate the average number of children of the respondent's children.
+This indicates reproductive success in the second generation, which we call RLRS2.
+We removed `r gkids_unusable_white` respondents who report having grandchildren despite reporting having reported no children.
+In regressions we reweight respondents by the number of children they have, since more fecund grandparents account for a larger proportion of the next generation.
 
 Older grandparents will have more time for their number of grandkids to accumulate.
-To better deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded.
+To deal with this time trend, we subset the data to parents whose oldest child was $40$ years old or older in $2016$, when the number of grandchildren was recorded.
 Appendix @fig-offspring-cohorts shows the distribution of the birth year of the oldest child, adjusted for sampling weights.
@@
 The grandparent's PGS indicates, but does not determine the PGS of the parent.
-To know the expected parent's PGS, we would need both grandparent's PGS.
-Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, such an approach is untenable in our sample.
-We expect the effect of the grandparent's PGS to also be dependent upon the level of assortative mating for the trait.
+To know the expected parent's PGS, we would need both grandparents' PGS.
+Given that some respondents do not have a partner in the HRS and some have had children with multiple partners, that approach is untenable in our sample.
+We expect the effect of the grandparent's PGS to also depend upon the level of assortative mating for the trait.
 For traits with high assortative mating, the grandparent's PGS will more strongly predict the parent's PGS, leading to a greater regression slope.
@@
 Appendix @fig-gkids plots coefficients on RLRS2 versus coefficients on years of education.
-Correlations are significant for whites but not blacks.
-Standard errors were large in both ethnicites (whites: correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I.
-`r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`; blacks: correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I.
-`r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`).
-There is a positive correlation between PGS coefficients on RLRS2 and RLRS (among whites: `r gen_cor_gkids`, bootstrapped 95% C.I.
+Correlations are significant.
+Standard errors are large (correlation `r cors_gkids_edu$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`).
+There is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I.
 `r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.
+
+## Differences across social groups
 
@@
 Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split.
-From here on we only use the white subsample: there are too few black respondents to be informative.
-
-```{r}
-#| label: stat-x-bootstrap-cors
-#| cache: false
+We call the low-education, low-income, unmarried or younger-AFB group the "disadvantaged group".
+
+In @hugh2022human we could simply look for significant differences between coefficients on RLRS among the groups, for each polygenic score.
+That will not work here because of the smaller sample size.
+At the same time, scores are not independent and not a sample from a population.
+To get round this, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
+Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education.
+So, we run regressions of the form
+
+$$
+RLRS_i=\alpha+\beta PGS_i + \gamma DIS_i + \delta (PGS_i \times DIS_i) + \varepsilon_i
+$$
+
+where $DIS_i$ is a dummy for $i$'s membership in the disadvantaged group and where $PGS_i$ has been sign-flipped to correlate negatively with education.
+Under the null, the estimated $\delta$s will be distributed around zero; in expectation half will be positive and half negative.
+Under the alternative, more than half the $\delta$s will be positive.
+We bootstrap respondents and count the number of positive $\delta$s in each resample.
+If 95% confidence intervals are above `r length(pgs_white)`/2 = `r length(pgs_white)/2`, we reject the null in favour of the alternative.
+\[TODO: what if some are negative and some are positive?
+I think this is fine because it makes our test conservative?
+Maybe??\]
+
+```{r}
+#| label: stat-x-bootstrap-effect-size
+#| cache: true
 #| cache-lazy: false
-#| eval: false
-
-old_opts <- options(survey.multicore = TRUE)
-
-# all confindence intervals were too wide to be informative
-n_reps_x <-  199
-
-calc_diff_cor_ed <- function (w, data, term) {
-  groups <- switch(term,
-                edu = c("0-12", "13-17"),
-                income_med = c("Low", "High"),
-                married = c("Other", "Married"),
-                agefb = c("Low", "High"))
-  groups <- paste0(":", term, groups) # colon for the cross term
-  low_group <- groups[1]
-  high_group <- groups[2]
-
-  forms <- paste0(pgs, ":", term) |>
-    map(\(x) reformulate(c(x, term, pcs), response = "rlrs"))
-  mods <- map(forms, \(x) lm(x, data, weights = w))
-  cfs <- map(mods, coef)
-  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
-  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])
-
-  forms_ed <- paste0(pgs, ":", term)  |> 
-    map(\(x) reformulate(c(x, term, pcs), response = "raedyrs"))
-  mods_ed <- forms_ed |> map(\(x) lm(x, data, weights = w))
-  cfs_ed <- map(mods_ed, coef)
-  cfs_ed_low <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), low_group)])
-  cfs_ed_high <- map_dbl(cfs_ed, \(x) x[str_detect(names(x), high_group)])
-
-  cor_ed_low <- cor(cfs_low, cfs_ed_low)
-  cor_ed_high <- cor(cfs_high, cfs_ed_high)
-
-  c(low = cor_ed_low, high = cor_ed_high, diff = cor_ed_low - cor_ed_high)
-}
-
-diff_cor_ed <- ci <- est <- tbl <- list()
-
-terms <- c("edu", "income_med", "married", "agefb")
-for (eth in eths) {
-  rand_boot <- rand |> 
-    subset(ethnicity == eth) |> 
-    as.svrepdesign(type = "bootstrap", replicates = n_reps_x)
-
-  for (term in terms) {
-    diff_cor_ed[[eth]][[term]] <- 
-      withReplicates(rand_boot, theta = calc_diff_cor_ed, term = term)  
-    ci[[eth]][[term]] <- confint(diff_cor_ed[[eth]][[term]])
-    est[[eth]][[term]] <- as.numeric(diff_cor_ed[[eth]][[term]])
-  }
-}
-
-options(old_opts)
-
-```
-
-```{r}
-#| label: stat-x-bootstrap-effect-size
-#| cache: false
-#| cache-lazy: false
-#| cache-comments: false
+#| cache-comments: true
+
+# TODO set cache to false before submission
 
 n_reps_x_es <-  199
-pgs <- pgs_white
-
-calc_diff_effect_size <- function (w, data, term) {
+
+calc_delta_coefs <- function (w, data, term) {
   groups <- switch(term,
@@
 
-  forms <- paste0(pgs, ":", term) |>
-    map(\(x) reformulate(c(x, term, pcs), response = "rlrs"))
+  # We estimate separate betas then take the difference,
+  # rather than explicitly using a cross term.
+  # This is simpler.
+  cross_terms <- paste0(pgs_white, ":", term)
+
+  # "rlrs ~ pgs:term + term + pc1 + pc2 +..."
+  forms <-  map(cross_terms, 
+                \(x) reformulate(c(x, term, pcs), response = "rlrs"))
   mods <- map(forms, \(x) lm(x, data, weights = w))
@@
 
-  abs_eff_low <- mean(abs(cfs_low))
-  abs_eff_high <- mean(abs(cfs_high))
-  diff_abs_eff <- abs_eff_low - abs_eff_high
-  c(low = abs_eff_low, high = abs_eff_high, diff = diff_abs_eff)
-}
-
-effect_sizes <- ci <- est <- tbl <- list()
-
-terms <- c("edu", "income_med", "married", "agefb", "born")
+  # \delta in the estimating equation above.
+  delta_coefs <- cfs_low - cfs_high
+
+  setNames(delta_coefs, pgs_white)
+}
+
+calc_num_positive_deltas <- function (w, data, term) {
+  delta_coefs <- calc_delta_coefs(w, data, term)
+  # the name is required to workaround a bug in survey
+  c(n_pos = sum(delta_coefs > 0))
+}
 
@@
 
+# sign-flip the PGS to correlate negatively with education
+for (pgs_name in pgs_white) {
+  pgs <- rand_boot$variables[[pgs_name]]
+  sign_cor_pgs_edu <- sign(cor(pgs, rand_boot$variables$raedyrs, 
+                               use = "complete"))
+  rand_boot$variables[[pgs_name]] <- pgs * -1 * sign_cor_pgs_edu
+}
+
+# to get the non-bootstrapped deltas do, e.g.:
+# calc_delta_coefs(weights(rand_boot, type = "sampling"), 
+#                  rand_boot$variables, "edu")
+
+n_pos_deltas <- ci <- est <- tbl <- list()
+
+terms <- c("edu", "income_med", "married", "agefb", "born")
+
 for (term in terms) {
-  effect_sizes[[term]] <- 
-    withReplicates(rand_boot, theta = calc_diff_effect_size, term = term)  
-  ci[[term]] <- confint(effect_sizes[[term]])
-  est[[term]] <- as.numeric(effect_sizes[[term]])
-  tbl[[term]] <- sprintf("%.3f (%.3f to %.3f)", est[[term]], 
+  n_pos_deltas[[term]] <- 
+    withReplicates(rand_boot, theta = calc_num_positive_deltas, term = term)
+
+  ci[[term]] <- confint(n_pos_deltas[[term]])
+  est[[term]] <- round(n_pos_deltas[[term]])
+  tbl[[term]] <- sprintf("%d (%.1f to %.1f)", est[[term]], 
                          ci[[term]][, 1], ci[[term]][, 2])
@@
 
-```
-
-```{r}
-#| label: stat-bootstrap-agefb
-#| cache: false
-
-n_reps_agefb <-  199
-pgs <- pgs_white
-
-calc_cor_agefb <- function (w, data) {
-  forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
-  mods_low <- map(forms, \(x) lm(x, data, subset = agefb == "Low",
-                                 weights = w))
-  mods_high <- map(forms, \(x) lm(x, data, subset = agefb == "High",
-                                  weights = w))
-  coefs_low <- map2_dbl(mods_low, pgs, \(x, y) coef(x)[[y]])
-  coefs_high <- map2_dbl(mods_high, pgs, \(x, y) coef(x)[[y]])
-
-  c(
-    cor = cor(coefs_low, coefs_high), 
-    discordant = sum(sign(coefs_low) != sign(coefs_high)) # discordant pairs
-  )
-}
-
-rand_boot <- rand |> 
-  subset(ethnicity == "white") |> 
-  as.svrepdesign(type = "bootstrap", replicates = n_reps_agefb)
-
-cor_agefb <- withReplicates(rand_boot, theta = calc_cor_agefb)
-
-stopifnot(insignificant(confint(cor_agefb)[1,]))
-ci_discordant <- confint(cor_agefb)[2,]
-# significantly different from 1/2?
-stopifnot(ci_discordant[1] <= length(pgs)/2 && 
-            ci_discordant[2] >= length(pgs)/2)
-# TE THEY ARE DISCORDANT
-
-```
-
-```{r}
-#| label: stat-post-x-bootstrap
-
-stopifnot(insignificant(ci$edu[3,]))
-stopifnot(significant(ci$income_med[3,]))
-stopifnot(significant(ci$married[3,]))
-stopifnot(insignificant(ci$agefb[3,]))
-
-tidy_agefb_wide <- tidy_pgs_x_agefb$white |> 
-  pivot_wider(id_cols = pgs, names_from = agefb, values_from = estimate)
-
-cor_agefb <- cor(tidy_agefb_wide$Low, tidy_agefb_wide$High)
-n_discordant <- sum(sign(tidy_agefb_wide$Low) != sign(tidy_agefb_wide$High))
-```
-
-Appendix @fig-x-edu shows essentially no difference in the distribution of absolute coefficient sizes between respondents with more and less education.
-Appendix @fig-x-inc, on the other hand, shows that coefficients appear larger among lower-income respondents.
-Appendix @fig-x-mar shows absolute coefficients for respondents who were married in 2010 against all other statuses.
-Coefficients appear larger for unmarried respondents.
-
-Appendix @fig-x-agefb splits respondents by age at first birth (median within each gender).
-Note that the N is lower here due to missing values.
-There are no large differences in absolute effect size between younger and older parents.
-Just as in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs.
-However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively.
-Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with the uncontrolled coefficients; here, that isn't true, with coefficients remaining positively correlated.
+# calculate non-bootstrapped deltas
+sampling_weights <- weights(rand_boot, type = "sampling")
+
+deltas <- map(terms, \(term) {
+                calc_delta_coefs(sampling_weights,
+                                 rand_boot$variables, 
+                                 term)
+              })
+
+deltas <- deltas |> 
+  setNames(terms) |> 
+  as.data.frame() |> 
+  as_tibble(rownames = "PGS") |> 
+  pivot_longer(-PGS, names_to = "group", values_to = "delta")
+
+rm(rand_boot) # avoid messing with subsequent code
+
+```
+
+```{r}
+#| label: fig-deltas
+#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
+#|   polygenic scores 
+#|   on RLRS among different social groups. A positive $\delta$ means that
+#|   the coefficient was larger for the disadvantaged group, where the score
+#|   was sign-flipped to correlate negatively with education. Horizontal lines
+#|   show medians.
+deltas |> 
+  filter(
+    group != "born"
+  ) |> 
+  mutate(
+    med_d = median(delta),
+    Group = case_match(group,
+              "agefb"      ~ "AFB",
+              "edu"        ~ "Education",
+              "income_med" ~ "Income",
+              "married"    ~ "Marriage"
+            ),
+    outliers = ifelse(rank(delta) %in% c(1:3, 66:68), PGS, ""),
+    .by = group
+  ) |> 
+  ggplot(aes(Group, delta, colour = Group)) + 
+    geom_hline(yintercept = 0, color='grey70') + 
+    geom_point(alpha = 0.7, position = position_jitter(width = 0.1)) +
+    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
+    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
+    theme_minimal() + 
+    theme(
+      panel.grid = element_blank(),
+      legend.position = "none"
+    )
+```
+
+Figure @fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping.
+The majority of $\delta$s are positive for age at first birth, income and marriage but not for education.
+In other words, for younger parents, poorer participants, and unmarried participants, most PGS have larger effects on RLRS, but this is not true for less educated participants.
 
@@
 
-@tbl-x-bootstraps shows means and 95% confidence intervals for mean absolute coefficient sizes in each pair of groups, and the difference between them, using bootstraps.
-Differences for income and marriage are significant.
-
-|               | Low group             | High group            | Difference            |
-|------------------|------------------|------------------|------------------|
-| Education     | `r tbl$edu[1]`        | `r tbl$edu[2]`        | `r tbl$edu[3]`        |
-| Income        | `r tbl$income_med[1]` | `r tbl$income_med[2]` | `r tbl$income_med[3]` |
-| Marriage      | `r tbl$married[1]`    | `r tbl$married[2]`    | `r tbl$married[3]`    |
-| Age 1st birth | `r tbl$agefb[1]`      | `r tbl$agefb[2]`      | `r tbl$agefb[3]`      |
-| Birth year    | `r tbl$born[1]`       | `r tbl$born[2]`       | `r tbl$born[3]`       |
-
-: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents.
+@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income.
+
+|                    | Estimate (95% C.I.) |
+|--------------------|---------------------|
+| Education          | `r tbl$edu`         |
+| Income             | `r tbl$income_med`  |
+| Marriage           | `r tbl$married`     |
+| Age at first birth | `r tbl$agefb`       |
+| Birth year         | `r tbl$born`        |
+
+: Estimates and bootstrap 95% confidence intervals for numbers of PGS out of `r length(pgs_white)` where the disadvantaged group had a larger selection coefficient than the advantaged group.
 Groups are: 0-12 years education vs.
@@
 `r n_reps_x_es` bootstraps.
-{#tbl-x-bootstraps tbl-colwidths="\[19,28,28,33\]"}
-
-The economic theory of fertility also implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups.
-we tested this with bootstraps, but confidence intervals were always too wide to be informative.
-
-[^1]: See @hugh2022human equation (6) and following.
 
@@
 @goldscheider1969minority, @johnson1979minority, @yang2003big).
-And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu).
+And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
 But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
@@
 We also perform error correction with twin heritabilites.
-To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^2].
+To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1].
 We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power.
-This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^3].
-
-[^2]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin.
+This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].
+
+[^1]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin.
 For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].
 
-[^3]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms.
+[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms.
 These were PV001-PV018 in the 2016 core survey.
@@
 The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason.
-But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^4] In this sense, results here are less supportive of the economic theory specifically.
-
-[^4]: See @balbo2013fertility for a broad review of fertility theories.
+But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.
+
+[^3]: See @balbo2013fertility for a broad review of fertility theories.
 
@@
 Code to reproduce this paper is available at <https://github.com/hughjonesd/hrs-selection>.
+
+## Black respondents
+
+There are `r n_black` genotyped black survey participants.
+Among this sample, the relationship between polygenic scores' coefficients on RLRS and on education is negative but insignificant and imprecisely estimated (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I.
+`r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
+Correlations between number of siblings and education are insignificant (correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).
+Similarly, coefficients on RLRS2 (children's average number of children) were insignificantly correlated with coefficients on years of education, with large standard errors (correlation `r cors_gkids_edu$black[1]`, bootstrapped 95% C.I.
+`r ci_boot_gkids$black[1]` to `r ci_boot_gkids$black[2]`; removed `r gkids_unusable_black` respondents reporting grandchildren but no children).
 
@@
 #| fig-cap: |
-#|  Coefficients of PGS on RLRS among white respondents, controlling for 10
-#|  principal components of
+#|  Coefficients of PGS on RLRS, controlling for 10 principal components of
 #|  genomic array data. Lines are 95% confidence intervals.
@@
 #|   data. Dashed lines show linear regressions.
-#| fig-subcap:
-#|   - ""
-#|   - ""
-#| layout-nrow: 2
-#| column: page
-
-for (eth in eths) {
-  ggp <-  
-    tidy_pgs_sibs[[eth]] |> 
-    inner_join(tidy_pgs_ed[[eth]], 
+
+tidy_pgs_sibs$white |> 
+  inner_join(tidy_pgs_ed$white, 
              by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
@@
     labs(
-          title = titles[eth],
         x = "Coefficient on years of education",
@@
     )
-  print(ggp)
-}
+
 ```
@@
 #|   Scatterplot of PGS coefficients on RLRS2 and years of
-#|   education. RLRS2 is the reproductive success of respondent's offspring relative to the success of other offspring. Controls include 10 principal components of genetic array
-#|   data. Dashed lines show linear regressions.
-#| fig-subcap:
-#|   - ""
-#|   - ""
-#| layout-nrow: 2
-#| column: page
-
-for (eth in eths) {
-  ggp <-  
-    tidy_pgs_gkids[[eth]] |> 
-    inner_join(tidy_pgs_ed[[eth]], 
+#|   education. RLRS2 is the reproductive success of respondent's offspring
+#|   relative to the success of other offspring. Controls include 10 principal
+#|   components of genetic array data. Dashed lines show linear regressions.
+
+tidy_pgs_gkids$white |> 
+  inner_join(tidy_pgs_ed$white, 
              by = join_by(term), suffix = c(".gkids", ".edyrs")) |> 
@@
     labs(
-          title = titles[eth],
       x = "Coefficient on years of education",
@@
     )
-  print(ggp)
-}
-
-```
-
-```{r}
-#| label: prog-x-plots
-
-geom_x_boxplot <- geom_boxplot(color = "grey50", outlier.shape = NA)
-geom_x_point <- geom_point(aes(shape = pgs=="EA3"), 
-                           position = position_jitter(width = 0.03), 
-                           alpha = 0.5)
-
-themes_x <- theme_minimal() +
-      theme(
-          legend.position = "none",
-          panel.grid.major.x = element_blank()
-      ) 
-```
-
-```{r}
-#| label: fig-x-edu
-#| fig-cap: |
-#|   Boxplots of polygenic score coefficients on RLRS 
-#|   among white respondents, controlling for 10 principal 
-#|   components of genetic array data, estimated within the low/high
-#|   education group. Boxes show quartiles.
-
-tidy_pgs_x_ed$white |> 
-  mutate(
-    Education = str_remove(edu, "edu")
-  ) |> 
-  ggplot(aes(Education, abs(estimate), color = Education)) +
-    geom_x_boxplot + geom_x_point +
-    labs(
-      x = "Years of education",
-      y = "Absolute coefficient",
-      title = paste("PGS coefficients on RLRS by education", 
-                    calc_n(mod_pgs_x_ed))
-    ) +
-    themes_x
-```
-
-```{r}
-#| label: fig-x-inc
-#| fig-cap: |
-#|   Boxplots of polygenic score coefficients on RLRS 
-#|   among white respondents, controlling for 10 principal 
-#|   components of genetic array data, estimated within the below/above
-#|   median income group. Boxes show quartiles.
-
-tidy_pgs_x_inc$white |> 
-  mutate(
-    Income = fct_relevel(income_med, c("Low", "High"))
-  ) |> 
-  ggplot(aes(Income, abs(estimate), color = Income)) +
-    geom_x_boxplot + geom_x_point +
-    labs(
-      x = "Income",
-      y = "Absolute coefficient",
-      title = paste("PGS coefficients on RLRS by income", 
-                    calc_n(mod_pgs_x_inc))
-    ) +
-    themes_x
-```
-
-```{r}
-#| label: fig-x-mar
-#| fig-cap: |
-#|   Boxplots of polygenic score coefficients on RLRS 
-#|   among white respondents, controlling for 10 principal 
-#|   components of genetic array data, estimated within married 
-#|   and other respondents. Boxes show quartiles.
-
-tidy_pgs_x_mar$white |> 
-    mutate(
-      married = fct_relevel(married, "Other", "Married")
-    ) |> 
-    ggplot(aes(married, abs(estimate), color = married)) +
-      geom_x_boxplot + geom_x_point +
-      labs(
-          x = "Married",
-          y = "Absolute coefficient",
-          title = paste("PGS coefficients on RLRS by marital status", 
-                        calc_n(mod_pgs_x_mar))
-      ) +
-      themes_x
-
-```
-
-```{r}
-#| label: fig-x-agefb
-#| fig-cap: |
-#|   Boxplots of polygenic score coefficients on RLRS 
-#|   among white respondents, controlling for 10 principal 
-#|   components of genetic array data, estimated within below/above
-#|   median age at first birth respondents. Boxes show quartiles.
-
-tidy_pgs_x_agefb$white |> 
-    mutate(
-      agefb = fct_relevel(agefb, "Low", "High")
-    ) |> 
-    ggplot(aes(agefb, abs(estimate), color = agefb)) +
-      geom_x_boxplot + geom_x_point + 
-      labs(
-          x = "Age at first birth",
-          y = "Absolute coefficient",
-          title = paste("PGS coefficients on RLRS by age at first birth", 
-                        calc_n(mod_pgs_x_agefb))
-      ) +
-      themes_x
 

### 2024-02-19 response-to-reviewers-revision-1.qmd
Discuss separate vs pooled RLRS. Delete some out-of-date bits.

@@
 format: docx
+editor: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Thank you for the opportunity to revise our manuscript. Thank you also to the
-reviewers. We respond to their points below.
+Thank you for the opportunity to revise our manuscript. Thank you also
+to the reviewers. We respond to their points below.
 
-One major change is that I have asked Tobias Edwards to be a co-author. I
-discovered that he was working on similar topics, so we merged our efforts.
-I hope this is OK. One effect is just that we have more polygenic scores to
-analyse. The original results are 
-not much changed, which is encouraging. TODO and what else?
-XXX Another is that we look in more detail at two extra generations,
-using children and grandparents of the respondents. We've given the paper
-a title which reflects that.
+One major change is that I have asked Tobias Edwards to be a co-author.
+I discovered that he was working on similar topics, so we merged our
+efforts. One effect is just that we have more
+polygenic scores to analyse. The original results are not much changed,
+which is encouraging. Another is that we look in
+more detail at two extra generations, using children and grandparents of
+the respondents. We've given the paper a title which reflects that. Lastly,
+we have more careful and accurate estimates of effect sizes of "true" and/or
+"chip-wide" scores. I hope this is OK.
 
@@
 
-*Thanks for the opportunity to read the manuscript ‘Natural selection in the Health and Retirement Study’.
-This study is of great interest to the behavior genetics and population genetics community and beyond.
-It demonstrates ongoing natural selection in a descriptive manner and partly confirms ideas about the trade-off between childbearing and work.
+*Thanks for the opportunity to read the manuscript ‘Natural selection in
+the Health and Retirement Study’. This study is of great interest to the
+behavior genetics and population genetics community and beyond. It
+demonstrates ongoing natural selection in a descriptive manner and
+partly confirms ideas about the trade-off between childbearing and work.
 I have a few comments.*
 
-*First, it would be good to state in the abstract the expectations about the tested interactions. Otherwise it reads as though interactions with low income etc have been tested arbitrarily.*
+*First, it would be good to state in the abstract the expectations about
+the tested interactions. Otherwise it reads as though interactions with
+low income etc have been tested arbitrarily.*
 
-*Second, in the introduction, it is implied that people with larger human capital work more, but it remains unclear why.
-Only in the discussion section, it is elaborated on the idea of substitution effect, that larger human capital leads to larger income loss when work is reduced for alternative activities.
-Note that this phenomenon is also known as opportunity costs of larger human capital.
-This should be explained early and explicitly as in absolute terms, one might argue that larger human capital increases net income and therefore might lead to working less and having more children with the same income.*
+*Second, in the introduction, it is implied that people with larger
+human capital work more, but it remains unclear why. Only in the
+discussion section, it is elaborated on the idea of substitution effect,
+that larger human capital leads to larger income loss when work is
+reduced for alternative activities. Note that this phenomenon is also
+known as opportunity costs of larger human capital. This should be
+explained early and explicitly as in absolute terms, one might argue
+that larger human capital increases net income and therefore might lead
+to working less and having more children with the same income.*
 
-*In a similar vein, it would be very welcome to reiterate reason why the expected effect is stronger amongst low income etc people.*
+*In a similar vein, it would be very welcome to reiterate reason why the
+expected effect is stronger amongst low income etc people.*
 
-We now detail the theory more in the abstract, and we've described the substitution
-effects and income effects straight away in the intro. We've tried to keep the
-language as simple and intuitive as possible.
+We now detail the theory more in the abstract, and we've described the
+substitution effects and income effects straight away in the intro.
+We've tried to keep the language as simple and intuitive as possible.
 
 *I was wondering why the authors control for PCAs in the PGI analyses.
-As stated by the authors, this study is not a causal study but one of correlations.
-The consequence of a correlation (natural selection) should be independent of its causes.*
+As stated by the authors, this study is not a causal study but one of
+correlations. The consequence of a correlation (natural selection)
+should be independent of its causes.*
 
-Note that controlling for principal components in a regression,
-*y ~ PGIs + PCs*, gives the
-same effect sizes as running an uncontrolled regression on PGIs which have
-previously had the PCs residualized  out, *y ~ PGIs_resid*. We control rather than
-residualizing because this gives the correct standard errors^[Because the controls are correctly accounted for when calculating degrees of freedom.], which might matter 
-with our relatively low N.
+Note that controlling for principal components in a regression, *y \~
+PGIs + PCs*, gives the same effect sizes as running an uncontrolled
+regression on PGIs which have previously had the PCs residualized out,
+*y \~ PGIs_resid*. We control rather than residualizing because this
+gives the correct standard errors[^1], which might matter with our
+relatively low N.
 
-Why residualize or control for principal components at all? I think for two reasons. First, we are ultimately interested in selection on the "true PGI". Stripping out PCs
-is an admittedly imperfect way to control for confounding population stratification.
-Second, PGIs are typically residualized in this way, so by examining the effect
-of these we're following the "industry standard" and learning about scores in the
-form they are actually used.
+[^1]: Because the controls are correctly accounted for when calculating
+    degrees of freedom.
 
-*In general, the authors state that they are less interested in significance but in effect sizes.
-First, this is very generous.
-Whilst I find this acceptable, it would be good to read one or two more sentences on why the authors think this is legitimate in this case.
-Second, I believe such a descriptive approach might be useful studying the patterns of natural selection, however, when it comes to the interactions, the author state that effects ‘Appear larger for unmarried’, for example.
-It remains unclear to me when an effect is considered larger or not, which becomes important as the author differentiates between interactions which are real and those which are not.
-What are the criteria here.*
+Why residualize or control for principal components at all? I think for
+two reasons. First, we are ultimately interested in selection on the
+"true PGI". Stripping out PCs is an admittedly imperfect way to control
+for confounding population stratification. Second, PGIs are typically
+residualized in this way, so by examining the effect of these we're
+following the "industry standard" and learning about scores in the form
+they are actually used.
 
-First, one could argue that the literal null hypothesis is not really plausible here. 
-If *any* PGI is under selection in the population (clearly true), then
-some underlying variants must also be under selection. Since all PGIs are just 
-weighted sums of underlying variants, it will then be a knife-edge
-case for any other PGI to have exactly zero correlation with fertility. What
-matters more is whether effect sizes are big, small or trivial.
+*In general, the authors state that they are less interested in
+significance but in effect sizes. First, this is very generous. Whilst I
+find this acceptable, it would be good to read one or two more sentences
+on why the authors think this is legitimate in this case. Second, I
+believe such a descriptive approach might be useful studying the
+patterns of natural selection, however, when it comes to the
+interactions, the author state that effects ‘Appear larger for
+unmarried’, for example. It remains unclear to me when an effect is
+considered larger or not, which becomes important as the author
+differentiates between interactions which are real and those which are
+not. What are the criteria here.*
+
+First, one could argue that the literal null hypothesis is not really
+plausible here. If *any* PGI is under selection in the population
+(clearly true), then some underlying variants must also be under
+selection. Since all PGIs are just weighted sums of underlying variants,
+it will then be a knife-edge case for any other PGI to have exactly zero
+correlation with fertility. What matters more is whether effect sizes
+are big, small or trivial.
 
 More specifically, we're interested in the pattern of selection effects
-across different PGIs,
-and how it correlates with the PGI's relationship to income. So we want to
-include PGIs with small selection effects, which we predict to have a low 
-absolute correlation to income. Given the sample size, these PGIs are very likely
-to have insignificant individual selection effects, but we shouldn't
-ignore them in the cross-PGI analyses.
+across different PGIs, and how it correlates with the PGI's relationship
+to income. So we want to include PGIs with small selection effects,
+which we predict to have a low absolute correlation to income. Given the
+sample size, these PGIs are very likely to have insignificant individual
+selection effects, but we shouldn't ignore them in the cross-PGI
+analyses.
 
@@
 
-*The authors state that the correlation between the PGI and fertility quantifies natural selection.
-However, to my knowledge it is the covariance following the Robert-Price Identity: Morrissey, M.
-B., L.
-E.
-B.
-Kruuk, and Alastair J.
-Wilson.
-"The danger of applying the breeder's equation in observational studies of natural populations." Journal of evolutionary biology 23.11 (2010): 2277-2288.*
+*The authors state that the correlation between the PGI and fertility
+quantifies natural selection. However, to my knowledge it is the
+covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
+B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
+equation in observational studies of natural populations." Journal of
+evolutionary biology 23.11 (2010): 2277-2288.*
 
-TE TODO: Tobias, can you check the relevant para? I guess that if both PGI and RLRS are
-normalized then the covariance is the correlation; but actually isn't the correct
-thing covariance with the not-normalized RLRS?
+TE TODO: Tobias, can you check the relevant para? I guess that if both
+PGI and RLRS are normalized then the covariance is the correlation; but
+actually isn't the correct thing covariance with the not-normalized
+RLRS?
 
-*The authors might want to discuss the expected selection in terms of relevance.
-They quantify it but they do not discuss these quantifications.
-This is very important shaping our expectation of ongoing natural selection.
-It might also be very interesting to know more about whether those trends follow the secular trends in phenotypes.*
+*The authors might want to discuss the expected selection in terms of
+relevance. They quantify it but they do not discuss these
+quantifications. This is very important shaping our expectation of
+ongoing natural selection. It might also be very interesting to know
+more about whether those trends follow the secular trends in
+phenotypes.*
 
@@
 In particular we discuss PGIs for health phenotypes, as a corrective to
-the existing literature which is perhaps a bit over-focused on cognition and
-IQ.
+the existing literature which is perhaps a bit over-focused on cognition
+and IQ.
 
-TE TODO: Could we look at health trends in phenotypes? I'd guess
-that health trends improve over the C20, but we could look at some specifics.
-(Depression/BMI?) One big issue here is that maybe the PGIs have confounds...
-but perhaps that's less serious for some of these phenotypes
+TE TODO: Could we look at health trends in phenotypes? I'd guess that
+health trends improve over the C20, but we could look at some specifics.
+(Depression/BMI?) One big issue here is that maybe the PGIs have
+confounds... but perhaps that's less serious for some of these
+phenotypes
 
-*Finally, the statement ’in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (-0.319), and for 42 out of 58 PGS, coefficients have opposite signs.
-However, in bootstraps, these statistics are not significantly different from 0 and 29 respectively.’ was not entirely clear to me.
-What does it mean that 42 PGS had opposite sign effects?
-Does that mean for old versus younger parents’ coefficients were very different, even in the opposite direction?
-How does this relate to the first part of the sentence that coefficients were negative.
-Also, why has this been tested against 0 and 29?*
+*Finally, the statement ’in the UK Biobank sample, correlations of PGS
+coefficients between younger and older parents are negative (-0.319),
+and for 42 out of 58 PGS, coefficients have opposite signs. However, in
+bootstraps, these statistics are not significantly different from 0 and
+29 respectively.’ was not entirely clear to me. What does it mean that
+42 PGS had opposite sign effects? Does that mean for old versus younger
+parents’ coefficients were very different, even in the opposite
+direction? How does this relate to the first part of the sentence that
+coefficients were negative. Also, why has this been tested against 0 and
+29?*
 
 The earlier claim is not "coefficients are negative overall" but
-"coefficients among younger parents are negatively correlated with coefficients
-among older parents". We are dealing with correlations of correlations: this
-can get confusing.
+"coefficients among younger parents are negatively correlated with
+coefficients among older parents". We are dealing with correlations of
+correlations: this can get confusing.
 
-Negatively correlated, and more strongly, opposite-signed, effects between
-young and old parents support the theory that income effects dominate for
-older parents while substitution effects dominate for younger parents. To
-be a bit more explicit: if a PGS simply doesn't correlate with human capital,
-we expect it to not be selected on among either group of parents. If a PGS
-is correlated with human capital, that will lead to income and substitution effects
-for those with more human capital. If the substitution effect dominates for
-younger parents then the PGS will have a negative correlation with fertility;
-if the income effect dominates for older parents, then the PGS will have a positive
+Negatively correlated, and more strongly, opposite-signed, effects
+between young and old parents support the theory that income effects
+dominate for older parents while substitution effects dominate for
+younger parents. To be a bit more explicit: if a PGS simply doesn't
+correlate with human capital, we expect it to not be selected on among
+either group of parents. If a PGS is correlated with human capital, that
+will lead to income and substitution effects for those with more human
+capital. If the substitution effect dominates for younger parents then
+the PGS will have a negative correlation with fertility; if the income
+effect dominates for older parents, then the PGS will have a positive
 correlation with fertility.
 
-The test against 29 had the following logic: if all effects were really zero,
-and estimates were pure noise with mean 0, then we'd expect opposite-signed and same-signed "effects" to be equally likely; so, 29 is half of the 58 scores.
+The test against 29 had the following logic: if all effects were really
+zero, and estimates were pure noise with mean 0, then we'd expect
+opposite-signed and same-signed "effects" to be equally likely; so, 29
+is half of the 58 scores.
 
-However, thinking about this issue a bit more made us change our tests more
-generally. The original situation was that we tested most pairs of coefficients
-across sample subgroups by comparing their absolute size. Coefficients for
-older and younger parents had similar absolute sizes, but we noticed they were
-opposite-signed and hence ran the tests above. This is not very satisfactory: it's 
-too like special pleading, and it also suggests the original tests weren't adequate.
+However, thinking about this issue a bit more made us change our tests
+more generally. The original situation was that we tested most pairs of
+coefficients across sample subgroups by comparing their absolute size.
+Coefficients for older and younger parents had similar absolute sizes,
+but we noticed they were opposite-signed and hence ran the tests above.
+This is not very satisfactory: it's too like special pleading, and it
+also suggests the original tests weren't adequate.
 
-We now improve our tests in table 1 as follows. We take the null hypothesis
-to be "selection coefficients are the same in the different subgroups, for all
-PGIs". Under the null, the true difference between selection coefficients is zero; the 
-difference in our sample (and in bootstrapped subsamples) should then not
-be significantly different from zero. Our alternative hypothesis is that
-selection coefficients are smaller among the richer/more educated/older/married
-group. That is, if the coefficient among the poorer group is $x_p$, then
+We now improve our tests in table 1 as follows. We take the null
+hypothesis to be "selection coefficients are the same in the different
+subgroups, for all PGIs". Under the null, the true difference between
+selection coefficients is zero; the difference in our sample (and in
+bootstrapped subsamples) should then not be significantly different from
+zero. Our alternative hypothesis is that, for at least some scores,
+selection coefficients against human capital are smaller among the
+richer/more educated/older/married group. That is, if the selection
+coefficient among the poorer and richer groups are $x_p,x_r$, and
+$sign(corr)$ is the sign of the PGI's correlation with human capital,
+then
 
-$x_r \times sign(x_p) < x_p \times sign(x_p)$
+$x_r \times -sign(corr) < x_p \times -sign(corr)$
 
-Note that, unlike tests of absolute values, the above always holds if 
-$x_r$ and $x_p$ are opposite signed.
-
-Under the null, we'd expect the above equation to hold for just 50%
-of the scores. If it holds for significantly more than 50% of the scores
+Under the null, we'd expect the above equation to hold for just 50% of
+the scores. If it holds for significantly more than 50% of the scores
 (using bootstrapped C.I.s) then we can reject the null in favour of the
@@
 We also try a simpler alternative approach. Using the few PGIs with the
-largest and most precisely estimated selection effects in the whole sample,
-we estimate selection effects in different subgroups and test for a significant
-difference between them, via a simple interaction term. We report these results
-too.
+largest and most precisely estimated selection effects in the whole
+sample, we estimate selection effects in different subgroups and test
+for a significant difference between them, via a simple interaction
+term. We report these results too.
 
-TE TODO: the above^^^
-
-DHJ TODO: can we figure out a test for this?
+TE TODO: the above
 
@@
 
-*The author sought to examine natural selection as indicated by polygenic scores in a contemporary representative sample of the U.S.
-and found that the results partially support the economic theory of fertility as a result of natural selection among both white and black individuals.
-The authors' hypotheses are based primarily on the economic theory of fertility, which emphasizes a trade-off between the number of children and income: i) higher income makes raising children costlier, but ii) higher income makes having more children more affordable.
-As far as I understand the results of the paper, the results support prediction 1, but are more ambiguous for prediction 2?*
+*The author sought to examine natural selection as indicated by
+polygenic scores in a contemporary representative sample of the U.S. and
+found that the results partially support the economic theory of
+fertility as a result of natural selection among both white and black
+individuals. The authors' hypotheses are based primarily on the economic
+theory of fertility, which emphasizes a trade-off between the number of
+children and income: i) higher income makes raising children costlier,
+but ii) higher income makes having more children more affordable. As far
+as I understand the results of the paper, the results support prediction
+1, but are more ambiguous for prediction 2?*
 
-*However, in my opinion the theoretical database, data and analysis are adequate, the problem I have is that the MS is a bit confusing.
-It should be made clearer why an analysis was done and what the results mean in terms of rejecting or accepting the theoretical assumptions.
-The possible link to biological selection should also be explained in more detail, as this has been done in the case of education, income and reproduction to some extent on the basis of comparable data (Kong et al.
-2017, Beauchamp 2016, Fieder & Huber 2002) and also in many other studies.*
+*However, in my opinion the theoretical database, data and analysis are
+adequate, the problem I have is that the MS is a bit confusing. It
+should be made clearer why an analysis was done and what the results
+mean in terms of rejecting or accepting the theoretical assumptions. The
+possible link to biological selection should also be explained in more
+detail, as this has been done in the case of education, income and
+reproduction to some extent on the basis of comparable data (Kong et al.
+2017, Beauchamp 2016, Fieder & Huber 2002) and also in many other
+studies.*
 
-We have clarified the theory early on in the intro and abstract.
-TE TODO: Write a paragraph on biol. selection. (Not sure what is meant here?)
+We have clarified the theory early on in the intro and abstract. TE
+TODO: Write a paragraph on biol. selection. (Not sure what is meant
+here?)
 
-In terms of accepting or rejecting our theory, our paragraph in the discussion
-was:
+In terms of accepting or rejecting our theory, our paragraph in the
+discussion was:
 
-> The results here support prediction 1 but are more ambiguous for prediction 2.
-PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents.
-But there is little evidence for larger coefficients among people with lower education, or younger parents.
-This may be due to the low sample size.
-But in the UK, the between-group differences were large; differences that big would surely have been visible here.
-The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason.
-But note that *any* theory with a negative relationship between education and fertility will give prediction 1.
-In this sense, results here are less supportive of the economic theory specifically.
+> The results here support prediction 1 but are more ambiguous for
+> prediction 2. PGS coefficients on RLRS also appear larger for
+> low-income groups and unmarried respondents. But there is little
+> evidence for larger coefficients among people with lower education, or
+> younger parents. This may be due to the low sample size. But in the
+> UK, the between-group differences were large; differences that big
+> would surely have been visible here. The theory can accommodate this
+> non-result, if preferences for income are less curved in the US for
+> whatever reason. But note that *any* theory with a negative
+> relationship between education and fertility will give prediction 1.
+> In this sense, results here are less supportive of the economic theory
+> specifically.
 
-We don't think these results give a simple thumbs up or down to the theory,
-but we're open to suggestions that would help clarify what we say.
+We don't think these results give a simple thumbs up or down to the
+theory, but we're open to suggestions that would help clarify what we
+say.
 
-*The division into a "black" and a "white" is necessary due to technical reasons, but further complicates the results/discussion.
-Presumably the results for the larger sample (I think in the case of the HRS this are the  whites) should be presented in the main text, and the results for the smaller population moved into a supplement?*
+*The division into a "black" and a "white" is necessary due to technical
+reasons, but further complicates the results/discussion. Presumably the
+results for the larger sample (I think in the case of the HRS this are
+the whites) should be presented in the main text, and the results for
+the smaller population moved into a supplement?*
 
-We're OK with this, since the black sample is really too small to say much.
+Done. Indeed, the black sample is really too small to say much.
 
-DHJ TODO: yes, I think that is probably fine.
+*Also, there seem to be many associations of PGS with RLRS (Figure 4),
+such as ADHD, diabetes, etc., which I think are interesting findings,
+even if of low significance, worth at least a brief discussion.*
 
-*Also, there seem to be many associations of PGS with RLRS (Figure 4), such as ADHD, diabetes, etc., which I think are interesting findings, even if of low significance, worth at least a brief discussion.*
+We agree. We now discuss some health-related traits in the conclusion:
 
-TODO: more than the below?
+> Whilst past research has focused on \[cognitive\] traits, we find
+> selection differentials are often of similar magnitudes for health
+> related traits. The selection differential for self-rated health is
+> greater than those of education and cognitive performance. The most
+> significant, positively selected trait was ADHD, which was also found
+> in the UK [@hugh2022human]. Future research should study the health
+> and medical implications of natural selection, in addition to their
+> social implications.
 
-We now discuss some health-related traits in the conclusion:
+*What is the data base for the mean number of children of people born in
+the same year - is pooling appropriate, as the means may vary
+considerably according to ethnicity?*
 
-> Whilst past research has focused on [cognitive] traits, we find selection 
-differentials are often of similar magnitudes for health related traits. The 
-selection differential for self-rated health is greater than those of education 
-and cognitive performance. The most significant, positively selected trait was 
-ADHD, which was also found in the UK [@hugh2022human]. Future research should 
-study the health and medical implications of natural selection, in addition to 
-their social implications.
+We thought about this, and now agree that it is better to use mean number of
+children of whites and of blacks separately for each ethnic group:
 
-*What is the data base for the mean number of children of people born in the same year - is pooling appropriate, as the means may vary considerably according to ethnicity?*
+> "RLRS is calculated separately by ethnicity.
+This is not ideal, because it treats the ethnicities as separate breeding populations.
+But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
+We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights."
 
-We thought about this, and agree that it is better to use mean number of 
-children of whites and of blacks separately for each ethnic group.
+Effectively, the white US population is a "single data
+point" for theory testing. Note that since we don't compare the ethnicities,
+the only difference this makes is in estimating the absolute size of effects.
 
-Neither option is perfect. If we calculate RLRS separately for whites and
-blacks, we are effectively treating them as separate breeding populations.
-Black and white people had low rates of intermarriage in the 20th century but
-that is a fact of law and culture, not of biology. Conversely, if we calculate
-RLRS by taking the mean number of children across the whole population, we
-end up effectively making claims about natural selection in the US population
-as a whole, by taking the white population as representative - which it very
-likely isn't.
-
-Overall we prefer the former option, so we have now done so and added a brief
-discussion. Effectively, the white US population is a "single data point" for
-theory testing. (Even if we could pool the whole US population, that is still
-not the whole human population; this point is not much discussed in the literature.)
-
-DHJ/TE TODO: change it to be separate RLRS baseline for white and black; add 
-a couple of sentences discussing the implications. (Effectively examining a
-subpopulation as an imperfect proxy for the whole population.)
+TE TODO: change it to be separate RLRS baseline for white and black.

### 2024-02-19 hrs-selection.qmd
Discuss separate vs pooled RLRS. Delete some out-of-date bits.

@@
 The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year.
-RLRS is calculated pooling ethnicities, i.e.
-treating them as members of the same biological population.
-The mean number of children of people born in the same year was calculated using sampling weights.
+RLRS is calculated separately by ethnicity.
+This is not ideal, because it treats the ethnicities as separate breeding populations.
+But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
+We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights.
 
@@
 This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.
-Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].
+Statistical tests are adjusted for clustering and stratification using the R *survey* package [@lumleysurvey2023].
 
@@
 #|   show medians.
+
 deltas |> 
@@
             ),
-    outliers = ifelse(rank(delta) %in% c(1:3, 66:68), PGS, ""),
+    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
     .by = group
@@
     geom_hline(yintercept = 0, color='grey70') + 
-    geom_point(alpha = 0.7, position = position_jitter(width = 0.1)) +
+    geom_point(alpha = 0.7, 
+               position = position_jitter(width = 0.1, seed = 12345)) +
     geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
     geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
+    geom_text(aes(label = outliers), size = 2, colour = "black", 
+              position = position_jitter(width = 0.1, seed = 12345), 
+              hjust = -0.2) +
     theme_minimal() + 
@@
 `r n_reps_x_es` bootstraps.
+{#tbl-x-bootstraps}
 
@@
   geom_pointrange(aes(ymax = EA3 + 1.96 * se, ymin = EA3 - 1.96 * se))
-
-```
-
-```{r}
-#| label: stat-x-bootstrap-birthyear
-#| cache: false
-
-n_reps_birthyear <-  199
-pgs <- pgs_white
-
-forms <- pgs |> map(\(x) reformulate(c(x, pcs), response = "raevbrn"))
-
-calc_birthyear <- function (w, data) {
-  data$w <- w
-  mods_young <- map(forms, \(x) lm(x, data, subset = rabyear > 1945,
-                                 weights = w))
-  mods_old <- map(forms, \(x) lm(x, data, subset = rabyear <= 1945,
-                                  weights = w))
-  coefs_young <- map2_dbl(mods_young, pgs, \(x, y) coef(x)[[y]])
-  coefs_old <- map2_dbl(mods_old, pgs, \(x, y) coef(x)[[y]])
-
-  c(cor = cor(coefs_young, coefs_old))
-}
-
-rand_boot <- rand |> 
-  subset(ethnicity == "white") |> 
-  as.svrepdesign(type = "bootstrap", replicates = n_reps_birthyear)
-
-cor_agefb <- withReplicates(rand_boot, theta = calc_birthyear)
 
@@
 But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
-Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality.
 Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
-The last line of @tbl-x-bootstraps shows that effect sizes are larger for respondents born after 1942, but the difference is imprecisely estimated and not significant.
 
@@
 #| label: ADHD
-#| cache: false
 
 # Testing the R2 of the ADHD PGS
-rand$variables |>
-  with(cor.test(adhd_score, ADHD, use = "pairwise.complete")) -> adhd_cor
+
+adhd_cor <- cor.test(rand$variables$adhd_score, rand$variables$ADHD, 
+                     use = "pairwise.complete")
 

### 2024-02-20 response-to-reviewers-revision-1.qmd
rlrs by race, errors fixed, explained covariance problem, explored top PGI

@@
     wrap: 72
+bibliography: bibliography-hrs.bib
 ---
@@
 I discovered that he was working on similar topics, so we merged our
-efforts. One effect is just that we have more
-polygenic scores to analyse. The original results are not much changed,
-which is encouraging. Another is that we look in
-more detail at two extra generations, using children and grandparents of
-the respondents. We've given the paper a title which reflects that. Lastly,
-we have more careful and accurate estimates of effect sizes of "true" and/or
+efforts. One effect is just that we have more polygenic scores to
+analyse. The original results are not much changed, which is
+encouraging. Another is that we look in more detail at two extra
+generations, using children and grandparents of the respondents. We've
+given the paper a title which reflects that. Lastly, we have more
+careful and accurate estimates of effect sizes of "true" and/or
 "chip-wide" scores. I hope this is OK.
@@
 
-*Thanks for the opportunity to read the manuscript ‘Natural selection in
-the Health and Retirement Study’. This study is of great interest to the
+*Thanks for the opportunity to read the manuscript 'Natural selection in
+the Health and Retirement Study'. This study is of great interest to the
 behavior genetics and population genetics community and beyond. It
@@
 patterns of natural selection, however, when it comes to the
-interactions, the author state that effects ‘Appear larger for
-unmarried’, for example. It remains unclear to me when an effect is
+interactions, the author state that effects 'Appear larger for
+unmarried', for example. It remains unclear to me when an effect is
 considered larger or not, which becomes important as the author
@@
 
+We quantify natural selection with the regression coefficient of the PGI
+on relative lifetime reproductive success. At times in the original
+manuscript this was wrongly referred to as a correlation but has now
+been updated. The polygenic scores are standardized to have a standard
+deviation of 1. Under a univariate model, the regression coefficient is
+the covariance divided by the variance in the independent variable.
+Given the variance is 1, the regression coefficient is equal to the
+covariance and thus the two are interchangeable.
+
+The story is slightly complicated by the fact that we control for
+principal components. Thus our estimates are the partial covariance
+divided by the partial variance, the partial variance being slightly
+less than 1. Slightly biasing the effect size upwards in magnitude.
+Nevertheless, this is the standard approach used by others
+[@kong2017selection].
+
+It would be possible to residualise the polygenic score on the principal
+components, before restandardizing it have a variance of 1. That we
+might think we have something closer to the true polygenic score
+unafflicted by population stratification, with the right variance. In
+practice we've found this to only change estimates for the effect of the
+EA PGI by \~1%. The area of our analysis where the exact estimate is key
+is the error-corrected estimates. Since the estimates need to be
+adjusted for the fact that the error-corrected polygenic scores will
+covary differently with the principal components, making such a similar
+adjustment would be rather difficult.
+
 TE TODO: Tobias, can you check the relevant para? I guess that if both
@@
 
-*Finally, the statement ’in the UK Biobank sample, correlations of PGS
+*Finally, the statement 'in the UK Biobank sample, correlations of PGS
 coefficients between younger and older parents are negative (-0.319),
@@
 bootstraps, these statistics are not significantly different from 0 and
-29 respectively.’ was not entirely clear to me. What does it mean that
+29 respectively.' was not entirely clear to me. What does it mean that
 42 PGS had opposite sign effects? Does that mean for old versus younger
-parents’ coefficients were very different, even in the opposite
+parents' coefficients were very different, even in the opposite
 direction? How does this relate to the first part of the sentence that
@@
 
-We thought about this, and now agree that it is better to use mean number of
-children of whites and of blacks separately for each ethnic group:
+We thought about this, and now agree that it is better to use mean
+number of children of whites and of blacks separately for each ethnic
+group:
 
-> "RLRS is calculated separately by ethnicity.
-This is not ideal, because it treats the ethnicities as separate breeding populations.
-But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
-We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights."
+> "RLRS is calculated separately by ethnicity. This is not ideal,
+> because it treats the ethnicities as separate breeding populations.
+> But the alternative of calculating pooled RLRS would effectively be
+> estimating natural selection in the whole US population by treating
+> whites as representative. We therefore focus on the white US
+> population, with the caveat that this"one data point" may not be
+> representative of the US population (still less the species!) The mean
+> number of children of people born in each year was calculated using
+> sampling weights."
 
-Effectively, the white US population is a "single data
-point" for theory testing. Note that since we don't compare the ethnicities,
-the only difference this makes is in estimating the absolute size of effects.
+Effectively, the white US population is a "single data point" for theory
+testing. Note that since we don't compare the ethnicities, the only
+difference this makes is in estimating the absolute size of effects.
 

### 2024-02-20 hrs-selection.qmd
rlrs by race, errors fixed, explained covariance problem, explored top PGI

@@
   partially support the economic theory of fertility as an explanation for 
-  natural selection: scores which correlate negatively (positively) with
   education are selected for (against). The theory predicts that
@@
   less educated, unmarried and younger parents, but these predictions are only
-  partly borne out: coefficients are larger among low-income
   and unmarried parents, but not among younger parents or those with less 
@@
   to that for cognitive traits. 
+  natural selection: scores which correlate negatively (positively) with
+  partly borne out: coefficients are larger among low-income
 date: "February 2024"
@@
 rand <- rand |>
-  mutate(.by = rabyear, # calculate this pooling ethnicity
+  mutate(.by = c(ethnicity, rabyear), # by year and ethnicity
     mean_weight = mean(weight, na.rm = TRUE),
@@
 Standard errors are large because of the relatively low sample sizes.
-`r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
+`r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
 The scores are age at first birth, educational attainment, ADHD, and self-rated health.
@@
 
+```{r}
+#| label: Top PGS
+#| eval: false
+
+# Is it ok to residualise PGS on PCS vs control for them?
+# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
+# if the regression beta is to be the covariance for the Price Equation
+# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs.
+The difference is marginal (~1% lower in the below example).
+Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.
+
+# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs.
+Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs?
+I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs.
+I'd have to think very carefully about how to do this.
+
+# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
+# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.
+
+reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
+reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
+# beta = -0.027651
+reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
+
+reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
+
+reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
+                               reg_pc$model)
+# beta = -0.027651
+reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
+                       reg_pc$model)
+# beta = -0.027358
+
+reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
+                       reg_pc$model)
+# beta = -0.027358
+
+# Below I get the estimate of the interactions and their p values. 
+# EA is only significant for income at p = 0.0106
+# There are always traits significant after bonferroni. But usually it's autism
+# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.
+
+for (eth in "white") {
+  r_eth <- rand |> subset(ethnicity == eth)
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
+  # the colon interaction estimates the effect separately within each group
+  form_pgs_x_ed <- paste0(pgs, "*edu") |>
+    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
+  mod_pgs_x_ed[[eth]]  <-
+    map(form_pgs_x_ed, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_ed[[eth]] <-
+    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_ed[[eth]]) <- pgs
+  tidy_pgs_x_ed[[eth]] <-
+    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(edu = str_remove(term, ".*:"))
+
+  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
+  mod_pgs_x_inc[[eth]]  <-
+    map(form_pgs_x_inc, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_inc[[eth]] <-
+    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_inc[[eth]]) <- pgs
+  tidy_pgs_x_inc[[eth]] <-
+    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(income_med = str_remove(term, ".*:income_med"))
+
+  form_pgs_x_mar <- paste0(pgs, "*married") |>
+    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
+  mod_pgs_x_mar[[eth]]  <-
+    map(form_pgs_x_mar, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_mar[[eth]] <-
+    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_mar[[eth]]) <- pgs
+  tidy_pgs_x_mar[[eth]] <-
+    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(
+      married = str_remove(term, ".*:married")
+    )
+
+  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
+  mod_pgs_x_agefb[[eth]]  <-
+    map(form_pgs_x_agefb, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_agefb[[eth]] <-
+    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
+    setNames(pgs) |> 
+    list_rbind(names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(
+      agefb = str_remove(term, ".*:agefb")
+    ) 
+}
+
+```
+
 Why does the US data show fewer differences by socio-economic status (SES) than the UK?

### 2024-02-20 hrs-selection.qmd
Residualised PGS on PCs then restandardized using sampling weights

@@
 
+##############################################################
+# Residualising then standardising PGS, with sampling weights
+##############################################################
+
+dat_eth_list <- list()
+
+for (eth in eths) {
+
+  r_eth <- rand |> subset(ethnicity == eth)
+
+   if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
+for (pg in pgs) {
+
+  reg <- lm(get(pg) ~ PC1_5A + PC1_5B + PC1_5C + PC1_5D + PC1_5E +
+                    PC6_10A + PC6_10B + PC6_10C + PC6_10D +PC6_10E, weights = weight, 
+                    data = r_eth)
+
+  r_eth <- r_eth |> mutate(
+                    !!pg := get(pg) - predict(reg, r_eth),
+                    mean_weight = mean(weight, na.rm=TRUE),
+                    tot_weight = sum(weight, na.rm=TRUE),
+                    mean_res_pgs = sum(get(pg)*weight, na.rm=TRUE)/tot_weight,
+                    w.dev = weight*(get(pg) - mean_res_pgs)^2,
+                    w.var.pgs = sum(w.dev, na.rm =TRUE)/tot_weight,
+                    deman_pgs = get(pg) - mean_res_pgs,
+                    !!pg := deman_pgs/sqrt(w.var.pgs)
+  )
+}
+
+  dat_eth_list[[eth]] <- r_eth
+
+}
+
+rand <- dat_eth_list |> list_rbind()
+pgs <- unique(c(pgs_white,pgs_black))
+
+# SD of EA before and after this process goes from 1 to 1.01, implying that the HRS is slightly more diverse in EA than the population and that Z scoring using sampling weights doesn't really matter
+
 ```
@@
 
-  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
+  form_pgs <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
   mod_pgs[[eth]] <- map(form_pgs, 
@@
 
-  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
+  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x),
                                               response = "raedyrs"))
@@
   form_pgs_sibs <- map(pgs, 
-                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
+                       \(x) reformulate(c(x), response = "r10sibs"))
   mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
@@
     form_pgs_gkids <- map(pgs, 
-                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
+                       \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))
   mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
@@
   form_pgs_x_ed <- paste0(pgs, ":edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
   mod_pgs_x_ed[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
-    filter(!term %in% c("(Intercept)", "edu13-17", pcs)) |>
+    filter(!term %in% c("(Intercept)", "edu13-17")) |>
     mutate(edu = str_remove(term, ".*:"))
@@
   form_pgs_x_inc <- paste0(pgs, ":income_med") |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
   mod_pgs_x_inc[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
-    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
+    filter(!term %in% c("(Intercept)", "income_medHigh")) |>
     mutate(income_med = str_remove(term, ".*:income_med"))
@@
   form_pgs_x_mar <- paste0(pgs, ":married") |>
-    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
   mod_pgs_x_mar[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
-    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
+    filter(! term %in% c("(Intercept)", "marriedMarried")) |>
     mutate(
@@
   form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
-    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
   mod_pgs_x_agefb[[eth]]  <-
@@
     list_rbind(names_to = "pgs") |>
-    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
+    filter(! term %in% c("(Intercept)", "agefbHigh")) |>
     mutate(
@@
   coef_pgs_white <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
+        map(\(x) reformulate(c(x), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
   coef_pgs_black <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
+        map(\(x) reformulate(c(x), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
   coef_pgs_fert <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = var)) |> 
+        map(\(x) reformulate(c(x), response = var)) |> 
         map(\(x) lm(x, data, weights = weights)) |> 
@@
   coef_pgs_edu <- pgs |> 
-        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
+        map(\(x) reformulate(c(x), response = "raedyrs")) |> 
         map(\(x) lm(x, data, weights = weights)) |> 
@@
 
-f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
-f_sibs <- map(pgs, \(x) reformulate(c(x, pcs), response = "r10sibs"))
+f_resp <- map(pgs, \(x) reformulate(c(x), response = "raevbrn"))
+f_sibs <- map(pgs, \(x) reformulate(c(x), response = "r10sibs"))
 
@@
 
-f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
-f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
+f_resp <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
+f_gkids <- map(pgs, \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))
 
@@
   forms <-  map(cross_terms, 
-                \(x) reformulate(c(x, term, pcs), response = "rlrs"))
+                \(x) reformulate(c(x, term), response = "rlrs"))
   mods <- map(forms, \(x) lm(x, data, weights = w))
@@
 
-form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
+form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn"), 
                                                     response = "rlrs"))
@@
 
-reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
-reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+reg_simple <- lm(rlrs ~ EA ,rand$variables)
+reg_pc <- lm(rlrs ~ EA + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
              PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
@@
   form_pgs_x_ed <- paste0(pgs, "*edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
   mod_pgs_x_ed[[eth]]  <-
@@
   form_pgs_x_inc <- paste0(pgs, "*income_med") |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
   mod_pgs_x_inc[[eth]]  <-
@@
   form_pgs_x_mar <- paste0(pgs, "*married") |>
-    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
   mod_pgs_x_mar[[eth]]  <-
@@
   form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
-    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
+    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
   mod_pgs_x_agefb[[eth]]  <-
@@
 correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
+correct_scores <- correct_scores[,c("pgs",
+                                    "chip_h2",
+                                    "twin_h2",
+                                    "R2",
+                                    "Source", 
+                                    "PoldermanICF_ICD10_subchapter",
+                                    "Category")]
 
@@
 #correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
+pgs_correct <- correct_scores$pgs
 
 ####################################
-# Calculating error corrected estimates
+# Running the regressions
 ####################################
 
-run_pgic_once <- function (pgs, r2, h2, h2_type) {
-  pgic_command <- glue(
-     'python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt', 
-     '--outcome "rlrs" --pgi-var {pgs}',
-     '--covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A"',
-     '"PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2 {r2} --h2 {h2}',
-     '--weights "weight"', 
-     .sep = " ")
-  system(pgic_command)
-
-  # This removes "junk" rows in the way of our regression betas
-  system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
-  pgic_results <- readr::read_table("reg_results", col_names = TRUE,
-                                    col_types = "cdddd")
-  file.remove("reg_results")
-
-  tibble(
-    trait = pgs,
-    beta  = pgic_results[[1, 4]],
-    SE    = pgic_results[[1, 5]],
-    Heritability = h2_type
-  )
-}
-
-run_pgic_data <- function (score_data, h2_column, h2_type) {
-  results <- map(1:nrow(score_data), 
-    \(i) run_pgic_once(pgs = score_data$pgs[i], 
-                       r2 = score_data$R2[i],
-                       h2 = score_data[i, h2_column, drop = TRUE],
-                       h2_type = h2_type)
-  )
-
-  list_rbind(results)
-}
-
-corrected_effects_snp <- run_pgic_data(correct_scores, h2_column = "chip_h2",
-                                       h2_type = "Chip")
-corrected_effects_twin <- run_pgic_data(correct_scores, h2_column = "twin_h2",
-                                        h2_type = "Twin")
+form_pgs_correct <- map(pgs_correct, \(x) reformulate(c(x), response = "rlrs"))
+
+corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
+    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
+    list_rbind() |>
+    filter(term %in% pgs) |>
+    rename(pgs = term) |>
+    left_join(correct_scores)
+
+corrected_effects_snp <- corrected_effects |>
+    mutate(estimate = estimate*sqrt(chip_h2/R2),
+           std.error = std.error*sqrt(chip_h2/R2),
+           conf.low = conf.low*sqrt(chip_h2/R2),
+           conf.high = conf.high*sqrt(chip_h2/R2),
+           heritability = "Chip")
+
+corrected_effects_twin <- corrected_effects |>
+    mutate(estimate = estimate*sqrt(twin_h2/R2),
+           std.error = std.error*sqrt(twin_h2/R2),
+           conf.low = conf.low*sqrt(twin_h2/R2),
+           conf.high = conf.high*sqrt(twin_h2/R2),
+           heritability = "Twin")
 
 corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
-corrected_effects$conf.low <- corrected_effects$beta - 1.96 *
-                              corrected_effects$SE
-corrected_effects$conf.high <- corrected_effects$beta + 1.96 *
-                               corrected_effects$SE
-
-unlink("pgi_correct_data.txt")
-
-corrected_effects <- left_join(corrected_effects, correct_scores,
-                               by = join_by(trait == Abr),
-                               suffix = c("", ".y"))
-
-# The odd looking results are for traits with large missing heritability e.g.
-Asthma.
-My guess is that the rare variants act somewhat differently as DHJ as argued, meaning traits with large missing heritability show nonsense twin heritability estimates.
 
@@
 
-# Cleaning and saving corrected effects so people can have the exact figures
-corrected_effects$TwinH2_Source <- corrected_effects$Source
+corrected_effects <- corrected_effects |> 
+  rename(trait = pgs,
+         TwinH2_Source = Source,
+         Heritability = heritability)
+
 corrected_effects |>
-  select(trait, beta, SE, Heritability, chip_h2, twin_h2, R2, 
+  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
            TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
@@
 
-# TE Note that under Fisher's fundamental theorem of natural selection, the selection
-# for fitness should be equal to it's heritability... But in HRS the
-# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
-# is much lower than what the heritability of NEB should be.
-# The low R2 of NEB causes it to be removed, but the genetics for NEB 
-# could just be a little different in HRS versus other samples!
-# NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]
-
 corrected_effects |>
@@
     trait = nice_names[trait],
-    trait = fct_reorder(trait, beta)
+    trait = fct_reorder(trait, estimate)
   ) |>
-  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
+  ggplot(aes(y = trait, x = estimate, xmin = conf.low, xmax = conf.high,
            color = Heritability)) +
@@
 
-  fs_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
-  fs_phen <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))
+  fs_fert <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
+  fs_phen <- map(pgs, \(x) reformulate(c(x), response = phenotype))
 

### 2024-02-21 response-to-reviewers-revision-1.qmd
Text edits for reviewers and to adjust for new results

@@
 regression on PGIs which have previously had the PCs residualized out,
-*y \~ PGIs_resid*. We control rather than residualizing because this
-gives the correct standard errors[^1], which might matter with our
-relatively low N.
+<!--# TE not an expert here, but I think it should say y_resid here.
+I've residualised PGS but not rlrs, because it would complicate the interpetation of our beta under the Price equation.
+I think we should just remove this paragraph.
+We can explain our approach by saying the residualisation get's us closer to an unbiased PGS and cite others who have done the same thing.-->*y
+\~ PGIs_resid*. We control rather than residualizing because this gives
+the correct standard errors[^1], which might matter with our relatively
+low N.
 
@@
 they are actually used.
+
+To simplify interpretation, we've decided to residualise and then
+standardize the PGIs on the principal components, instead of control for
+them.
 
@@
 on relative lifetime reproductive success. At times in the original
-manuscript this was wrongly referred to as a correlation but has now
-been updated. The polygenic scores are standardized to have a standard
-deviation of 1. Under a univariate model, the regression coefficient is
-the covariance divided by the variance in the independent variable.
-Given the variance is 1, the regression coefficient is equal to the
-covariance and thus the two are interchangeable.
+manuscript this was wrongly referred to as a correlation but thus has
+now been updated. The polygenic scores are standardized to have a
+standard deviation of 1. Under a univariate model, the regression
+coefficient is the covariance divided by the variance in the independent
+variable. The variance is standardized to 1, so the regression
+coefficient is equal to the covariance and thus the two are
+interchangeable.
 
-The story is slightly complicated by the fact that we control for
-principal components. Thus our estimates are the partial covariance
-divided by the partial variance, the partial variance being slightly
-less than 1. Slightly biasing the effect size upwards in magnitude.
-Nevertheless, this is the standard approach used by others
-[@kong2017selection].
-
-It would be possible to residualise the polygenic score on the principal
-components, before restandardizing it have a variance of 1. That we
-might think we have something closer to the true polygenic score
-unafflicted by population stratification, with the right variance. In
-practice we've found this to only change estimates for the effect of the
-EA PGI by \~1%. The area of our analysis where the exact estimate is key
-is the error-corrected estimates. Since the estimates need to be
-adjusted for the fact that the error-corrected polygenic scores will
-covary differently with the principal components, making such a similar
-adjustment would be rather difficult.
+In the previous manuscript we used a multivariate model, controlling for
+genetic principal components. We now no longer control for the PCs. Now
+we residualise the polygenic scores on the PCs before standardizing the
+polygenic scores.
 
@@
 
-We have clarified the theory early on in the intro and abstract. TE
-TODO: Write a paragraph on biol. selection. (Not sure what is meant
-here?)
+We have added a sentence explicitly addressing our motivation in the
+introduction. We have also clarified the biological theory in the Data
+section. We provide an explanation of the intuition for our approach and
+how it is established in biology, through the Robertson-Price identity,
+to estimate the selection differential. TE TODO: Write a paragraph on
+biol. selection. (Not sure what is meant here?)
 

### 2024-02-21 hrs-selection.qmd
Text edits for reviewers and to adjust for new results

@@
 The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
+The motivation for our study is to establish the direction and magnitude of natural selection upon a range of traits in Americans and to test whether the economic theory can explain the selection.
 Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans.
@@
 
+The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection.
+However the approach is also based in quantitative genetic theory and is able to yield an approximate estimate of genetic change from one generation to the next.
+
+The Robertson-Price Identity states that the change in a genetic trait between one generation and the next is equal to its covariance with relative fitness [@Price1970identity; @Robertson1966identity], assuming no mutations.
+In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness.
+As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score from one generation.
+We standardize our polygenic scores to have a mean of 0 and a standard deviation of 1.
+With this standardization the regression coefficient of RLRS on the PGS is equal to the covariance.
+Our approach to measuring natural selection is not novel, having been and detailed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].
+
 The HRS contains weights which match survey respondents to the US population.
@@
 This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012.
+Standardization of polygenic scores used estimates of the population mean and variance of the polygenic score, as estimated with sampling weights.
 Statistical tests are adjusted for clustering and stratification using the R *survey* package [@lumleysurvey2023].
@@
 `r n_sig_bonf_white` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests_white`.
-The scores are age at first birth, educational attainment, ADHD, and self-rated health.
+The scores are age at first birth, educational attainment, ADHD, self-rated health and having ever smoked.
 But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.
@@
 Lastly, we would like to know natural selection's effect sizes.
-The bivariate correlation of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
-Polygenic scores contain error, so estimated correlations are biased towards zero compared to the correlation of the true PGS.
+The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
+Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS.
 They can be scaled up by
@@
 where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource].
-Moving from error correction in a univariate model to a multivariate model, controlling for principal components of genetic array data, requires a slightly more complex correction, although it uses the same parameters.
-To do this we use the error-corrected estimator developed by @becker2021resource.
-
 SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository.
@@
 This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].
+Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.
 
@@
 When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`.
-Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I.
+Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not very significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I.
 `r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`.
-Given the great uncertainty over the true correlation we are unable to justify using the error-corrected estimator with the ADHD polygenic score.
+Given the great uncertainty over the true correlation we were unable to justify using the error-corrected estimator with the ADHD polygenic score.
 
@@
 The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations.
-For twin-heritability, different relationships with fertility among variants not measured on the chip.
+For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates.
 Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
 
-Results from a few traits show substantial uncertainty regarding the true effect size.
-For asthma, when chip heritability is used, the effect is positive and significant, but negative and significant when twin heritability is used!
-Such a large disparity is because @becker2021resource estimate the chip heritability at 0.015 and we use a twin heritability 0f 0.55.
-For personality traits, chip heritability is often low (e.g.
-$<0.05$) but twin heritabilities are often much higher ($≈0.4$) implying substantial missing heritability.
-For these traits, the confidence intervals on twin heritabilities are enormous.
-Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.
+<!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->
 
@@
 Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.
+
+<!--# Put in some discussion of the fact the most robust findings seem show opposite change in the phenotype during the 20th century.
+Negative Flynn Effect...
+Environmental variation across time as surely more important.-->
 
@@
     rename(pgs = term) |>
-    left_join(correct_scores)
+    left_join(correct_scores, by = join_by(pgs))
 

### 2024-02-22 response-to-reviewers-revision-1.qmd
Added EA example Interactions (Table unfinished). Finished up all other todos

@@
 coefficient is equal to the covariance and thus the two are
-interchangeable.
+interchangeable. A brief discussion of the issue is now given in the
+Data section.
 
@@
 we residualise the polygenic scores on the PCs before standardizing the
-polygenic scores.
+polygenic scores. This ensures our regression estimates are exactly the
+covariance between the polygenic scores and RLRS.
 
@@
 
-We have now added a final paragraph to the Discussion section on this.
-In particular we discuss PGIs for health phenotypes, as a corrective to
-the existing literature which is perhaps a bit over-focused on cognition
-and IQ.
+We have now elaborated on the error corrected results and added a final
+paragraph to the Discussion section on this. We estimate how selection
+against intelligence will affect the phenotype, finding the result is
+not especially large. In particular we discuss PGIs for health
+phenotypes, as a corrective to the existing literature which is perhaps
+a bit over-focused on cognition and IQ.
+
+We briefly mention that genetic change in traits is often opposite to
+the phenotypic change. This highlights that environmental change is
+likely to be of greater importance for phenotypic trends then genetic
+change.
 
@@
 how it is established in biology, through the Robertson-Price identity,
-to estimate the selection differential. TE TODO: Write a paragraph on
-biol. selection. (Not sure what is meant here?)
+to estimate the selection differential.
+
+TE TODO: Write a paragraph on biol. selection. (Not sure what is meant
+here?)
 

### 2024-02-22 hrs-selection.qmd
Added EA example Interactions (Table unfinished). Finished up all other todos

@@
 
+```{r}
+#| label: Top PGS
+#| eval: true
+
+for (eth in "white") {
+  r_eth <- rand |> subset(ethnicity == eth)
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
+  # the colon interaction estimates the effect separately within each group
+  form_pgs_x_ed <- paste0(pgs, "*edu") |>
+    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
+  mod_pgs_x_ed[[eth]]  <-
+    map(form_pgs_x_ed, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_ed[[eth]] <-
+    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_ed[[eth]]) <- pgs
+  tidy_pgs_x_ed[[eth]] <-
+    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
+    filter(pgs == "EA") |>
+    mutate(edu = str_remove(term, ".*:"))
+
+  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
+    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
+  mod_pgs_x_inc[[eth]]  <-
+    map(form_pgs_x_inc, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_inc[[eth]] <-
+    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_inc[[eth]]) <- pgs
+  tidy_pgs_x_inc[[eth]] <-
+    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
+    filter(pgs == "EA") |>
+    mutate(income_med = str_remove(term, ".*:income_med"))
+
+  form_pgs_x_mar <- paste0(pgs, "*married") |>
+    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
+  mod_pgs_x_mar[[eth]]  <-
+    map(form_pgs_x_mar, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_mar[[eth]] <-
+    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_mar[[eth]]) <- pgs
+  tidy_pgs_x_mar[[eth]] <-
+    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
+    filter(pgs == "EA") |>
+    mutate(
+      married = str_remove(term, ".*:married")
+    )
+
+  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
+    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
+  mod_pgs_x_agefb[[eth]]  <-
+    map(form_pgs_x_agefb, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_agefb[[eth]] <-
+    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
+    setNames(pgs) |> 
+    list_rbind(names_to = "pgs") |>
+    filter(pgs == "EA") |>
+    mutate(
+      agefb = str_remove(term, ".*:agefb")
+    ) 
+}
+
+eff_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"estimate"]
+se_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"std.error"]
+p_ea_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
+eff_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"estimate"]
+se_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"std.error"]
+p_ea_int_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"estimate"]
+se_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"std.error"]
+p_ea_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"estimate"]
+se_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"std.error"]
+p_ea_int_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"estimate"]
+se_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"std.error"]
+p_ea_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"estimate"]
+se_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"std.error"]
+p_ea_int_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"estimate"]
+se_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"std.error"]
+p_ea_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
+
+eff_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"estimate"]
+se_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"std.error"]
+p_ea_int_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.001, "***",
+            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.01, "**",
+            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.05, "*", "")))
+
+```
+
 ## Differences across social groups
@@
 
-In @hugh2022human we could simply look for significant differences between coefficients on RLRS among the groups, for each polygenic score.
-That will not work here because of the smaller sample size.
-At the same time, scores are not independent and not a sample from a population.
-To get round this, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
+The EA polygenic score captures a key element of human capital and was trained on a GWAS sample size of one million, making it a sensible PGS with which to test the economic theory.
+In TABLE XXX we report the main effect of the EA polygenic score and its interaction effect on RLRS, for each of the four putative moderators.
+Only for the case of income is the interaction effect significant at p \< 0.05.
+We might wonder whether the weak effects are because of insufficient power or because any real interaction is small in the sample.
+In @hugh2022human the interaction effects were large, often as big as the main effects.
+Yet in the HRS, the point estimates of the interactions appear generally smaller, suggesting the nulls may not just be a consequence of lower power.
+
+We should attain more statistical power if we utilise information from other polygenic scores at the same time.
+However, some caution is necessary.
+Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate.
+To perform, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
 Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education.
@@
 
-Figure @fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping.
+@fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping.
 The majority of $\delta$s are positive for age at first birth, income and marriage but not for education.
@@
 
-```{r}
-#| label: Top PGS
-#| eval: false
-
-# Is it ok to residualise PGS on PCS vs control for them?
-# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
-# if the regression beta is to be the covariance for the Price Equation
-# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs.
-The difference is marginal (~1% lower in the below example).
-Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.
-
-# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs.
-Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs?
-I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs.
-I'd have to think very carefully about how to do this.
-
-# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
-# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.
-
-reg_simple <- lm(rlrs ~ EA ,rand$variables)
-reg_pc <- lm(rlrs ~ EA + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
-# beta = -0.027651
-reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
-
-reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
-
-reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
-                               reg_pc$model)
-# beta = -0.027651
-reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
-                       reg_pc$model)
-# beta = -0.027358
-
-reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
-                       reg_pc$model)
-# beta = -0.027358
-
-# Below I get the estimate of the interactions and their p values. 
-# EA is only significant for income at p = 0.0106
-# There are always traits significant after bonferroni. But usually it's autism
-# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.
-
-for (eth in "white") {
-  r_eth <- rand |> subset(ethnicity == eth)
-
-  if (eth == "white") {
-    pgs <- pgs_white
-  } else {
-    pgs <- pgs_black
-  }
-
-  # the colon interaction estimates the effect separately within each group
-  form_pgs_x_ed <- paste0(pgs, "*edu") |>
-    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
-  mod_pgs_x_ed[[eth]]  <-
-    map(form_pgs_x_ed, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_ed[[eth]] <-
-    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_ed[[eth]]) <- pgs
-  tidy_pgs_x_ed[[eth]] <-
-    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(edu = str_remove(term, ".*:"))
-
-  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
-    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
-  mod_pgs_x_inc[[eth]]  <-
-    map(form_pgs_x_inc, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_inc[[eth]] <-
-    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_inc[[eth]]) <- pgs
-  tidy_pgs_x_inc[[eth]] <-
-    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(income_med = str_remove(term, ".*:income_med"))
-
-  form_pgs_x_mar <- paste0(pgs, "*married") |>
-    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
-  mod_pgs_x_mar[[eth]]  <-
-    map(form_pgs_x_mar, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_mar[[eth]] <-
-    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_mar[[eth]]) <- pgs
-  tidy_pgs_x_mar[[eth]] <-
-    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(
-      married = str_remove(term, ".*:married")
-    )
-
-  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
-    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
-  mod_pgs_x_agefb[[eth]]  <-
-    map(form_pgs_x_agefb, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_agefb[[eth]] <-
-    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
-    setNames(pgs) |> 
-    list_rbind(names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(
-      agefb = str_remove(term, ".*:agefb")
-    ) 
-}
-
-```
-
 Why does the US data show fewer differences by socio-economic status (SES) than the UK?
@@
 
+To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits.
+Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance.
+Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster.
+This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.
+
 <!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->
@@
 
-<!--# Put in some discussion of the fact the most robust findings seem show opposite change in the phenotype during the 20th century.
-Negative Flynn Effect...
-Environmental variation across time as surely more important.-->
+Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century.
+Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed.
+The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting.
+The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.
+
+<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
 

### 2024-02-23 hrs-selection.qmd
moved discussion of non-result to discussion section

--- 2024-02-20 hrs-selection.qmd
@@
   partially support the economic theory of fertility as an explanation for 
+  natural selection: scores which correlate negatively (positively) with
   education are selected for (against). The theory predicts that
@@
   less educated, unmarried and younger parents, but these predictions are only
-  and unmarried parents, but not among younger parents or those with less 
-  education. Using respondents' number of siblings and grandchildren, we 
+  partly borne out: coefficients are larger among low-income
+  parents, but not among other groups. Using respondents' number of 
+  siblings and grandchildren, we 
   examine selection across three generations, finding consistent results 
@@
   to that for cognitive traits. 
-  natural selection: scores which correlate negatively (positively) with
-  partly borne out: coefficients are larger among low-income
 date: "February 2024"
@@
 
-##############################################################
-# Residualising then standardising PGS, with sampling weights
-##############################################################
-
-dat_eth_list <- list()
-
-for (eth in eths) {
-
-  r_eth <- rand |> subset(ethnicity == eth)
-
-   if (eth == "white") {
-    pgs <- pgs_white
-  } else {
-    pgs <- pgs_black
-  }
-
-for (pg in pgs) {
-
-  reg <- lm(get(pg) ~ PC1_5A + PC1_5B + PC1_5C + PC1_5D + PC1_5E +
-                    PC6_10A + PC6_10B + PC6_10C + PC6_10D +PC6_10E, weights = weight, 
-                    data = r_eth)
-
-  r_eth <- r_eth |> mutate(
-                    !!pg := get(pg) - predict(reg, r_eth),
-                    mean_weight = mean(weight, na.rm=TRUE),
-                    tot_weight = sum(weight, na.rm=TRUE),
-                    mean_res_pgs = sum(get(pg)*weight, na.rm=TRUE)/tot_weight,
-                    w.dev = weight*(get(pg) - mean_res_pgs)^2,
-                    w.var.pgs = sum(w.dev, na.rm =TRUE)/tot_weight,
-                    deman_pgs = get(pg) - mean_res_pgs,
-                    !!pg := deman_pgs/sqrt(w.var.pgs)
-  )
-}
-
-  dat_eth_list[[eth]] <- r_eth
-
-}
-
-rand <- dat_eth_list |> list_rbind()
-pgs <- unique(c(pgs_white,pgs_black))
-
-# SD of EA before and after this process goes from 1 to 1.01, implying that the HRS is slightly more diverse in EA than the population and that Z scoring using sampling weights doesn't really matter
-
 ```
@@
 
-  form_pgs <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
+  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
   mod_pgs[[eth]] <- map(form_pgs, 
@@
 
-  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x),
+  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                               response = "raedyrs"))
@@
   form_pgs_sibs <- map(pgs, 
-                       \(x) reformulate(c(x), response = "r10sibs"))
+                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
   mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
@@
     form_pgs_gkids <- map(pgs, 
-                       \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))
+                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
   mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
@@
   form_pgs_x_ed <- paste0(pgs, ":edu") |>
-    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
   mod_pgs_x_ed[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
-    filter(!term %in% c("(Intercept)", "edu13-17")) |>
+    filter(!term %in% c("(Intercept)", "edu13-17", pcs)) |>
     mutate(edu = str_remove(term, ".*:"))
@@
   form_pgs_x_inc <- paste0(pgs, ":income_med") |>
-    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
   mod_pgs_x_inc[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
-    filter(!term %in% c("(Intercept)", "income_medHigh")) |>
+    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
     mutate(income_med = str_remove(term, ".*:income_med"))
@@
   form_pgs_x_mar <- paste0(pgs, ":married") |>
-    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
   mod_pgs_x_mar[[eth]]  <-
@@
     list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
-    filter(! term %in% c("(Intercept)", "marriedMarried")) |>
+    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
     mutate(
@@
   form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
-    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
   mod_pgs_x_agefb[[eth]]  <-
@@
     list_rbind(names_to = "pgs") |>
-    filter(! term %in% c("(Intercept)", "agefbHigh")) |>
+    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
     mutate(
@@
   coef_pgs_white <- pgs |> 
-        map(\(x) reformulate(c(x), response = "rlrs")) |> 
+        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
   coef_pgs_black <- pgs |> 
-        map(\(x) reformulate(c(x), response = "rlrs")) |> 
+        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
         map(\(x) lm(x, data, weights = weights, 
@@
   coef_pgs_fert <- pgs |> 
-        map(\(x) reformulate(c(x), response = var)) |> 
+        map(\(x) reformulate(c(x, pcs), response = var)) |> 
         map(\(x) lm(x, data, weights = weights)) |> 
@@
   coef_pgs_edu <- pgs |> 
-        map(\(x) reformulate(c(x), response = "raedyrs")) |> 
+        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
         map(\(x) lm(x, data, weights = weights)) |> 
@@
 
-f_resp <- map(pgs, \(x) reformulate(c(x), response = "raevbrn"))
-f_sibs <- map(pgs, \(x) reformulate(c(x), response = "r10sibs"))
+f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
+f_sibs <- map(pgs, \(x) reformulate(c(x, pcs), response = "r10sibs"))
 
@@
 
-f_resp <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
-f_gkids <- map(pgs, \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))
+f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
+f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
 
@@
   forms <-  map(cross_terms, 
-                \(x) reformulate(c(x, term), response = "rlrs"))
+                \(x) reformulate(c(x, term, pcs), response = "rlrs"))
   mods <- map(forms, \(x) lm(x, data, weights = w))
@@
 #|   on RLRS among different social groups. A positive $\delta$ means that
-#|   the coefficient was larger for the disadvantaged group, where the score
-#|   was sign-flipped to correlate negatively with education. Horizontal lines
-#|   show medians.
+#|   the selection coefficient was larger for the disadvantaged group, in 
+#|   the direction associated with less education. Horizontal lines
+#|   show median coefficients.
 
@@
               position = position_jitter(width = 0.1, seed = 12345), 
-              hjust = -0.2) +
+              hjust = -0.15) +
     theme_minimal() + 
@@
 
-form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn"), 
+form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
                                                     response = "rlrs"))
@@
 
-reg_simple <- lm(rlrs ~ EA ,rand$variables)
-reg_pc <- lm(rlrs ~ EA + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
+reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
              PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
@@
   form_pgs_x_ed <- paste0(pgs, "*edu") |>
-    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
   mod_pgs_x_ed[[eth]]  <-
@@
   form_pgs_x_inc <- paste0(pgs, "*income_med") |>
-    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
   mod_pgs_x_inc[[eth]]  <-
@@
   form_pgs_x_mar <- paste0(pgs, "*married") |>
-    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
   mod_pgs_x_mar[[eth]]  <-
@@
   form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
-    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
   mod_pgs_x_agefb[[eth]]  <-
@@
 
-Why does the US data show fewer differences by socio-economic status (SES) than the UK?
-One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority, @johnson1979minority, @yang2003big).
-And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
-But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
-Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
-
 ```{r}
@@
 correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
-correct_scores <- correct_scores[,c("pgs",
-                                    "chip_h2",
-                                    "twin_h2",
-                                    "R2",
-                                    "Source", 
-                                    "PoldermanICF_ICD10_subchapter",
-                                    "Category")]
 
@@
 #correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
-pgs_correct <- correct_scores$pgs
 
 ####################################
-# Running the regressions
+# Calculating error corrected estimates
 ####################################
 
-form_pgs_correct <- map(pgs_correct, \(x) reformulate(c(x), response = "rlrs"))
-
-corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
-    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
-    list_rbind() |>
-    filter(term %in% pgs) |>
-    rename(pgs = term) |>
-    left_join(correct_scores)
-
-corrected_effects_snp <- corrected_effects |>
-    mutate(estimate = estimate*sqrt(chip_h2/R2),
-           std.error = std.error*sqrt(chip_h2/R2),
-           conf.low = conf.low*sqrt(chip_h2/R2),
-           conf.high = conf.high*sqrt(chip_h2/R2),
-           heritability = "Chip")
-
-corrected_effects_twin <- corrected_effects |>
-    mutate(estimate = estimate*sqrt(twin_h2/R2),
-           std.error = std.error*sqrt(twin_h2/R2),
-           conf.low = conf.low*sqrt(twin_h2/R2),
-           conf.high = conf.high*sqrt(twin_h2/R2),
-           heritability = "Twin")
+run_pgic_once <- function (pgs, r2, h2, h2_type) {
+  pgic_command <- glue(
+     'python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt', 
+     '--outcome "rlrs" --pgi-var {pgs}',
+     '--covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A"',
+     '"PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2 {r2} --h2 {h2}',
+     '--weights "weight"', 
+     .sep = " ")
+  system(pgic_command)
+
+  # This removes "junk" rows in the way of our regression betas
+  system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
+  pgic_results <- readr::read_table("reg_results", col_names = TRUE,
+                                    col_types = "cdddd")
+  file.remove("reg_results")
+
+  tibble(
+    trait = pgs,
+    beta  = pgic_results[[1, 4]],
+    SE    = pgic_results[[1, 5]],
+    Heritability = h2_type
+  )
+}
+
+run_pgic_data <- function (score_data, h2_column, h2_type) {
+  results <- map(1:nrow(score_data), 
+    \(i) run_pgic_once(pgs = score_data$pgs[i], 
+                       r2 = score_data$R2[i],
+                       h2 = score_data[i, h2_column, drop = TRUE],
+                       h2_type = h2_type)
+  )
+
+  list_rbind(results)
+}
+
+corrected_effects_snp <- run_pgic_data(correct_scores, h2_column = "chip_h2",
+                                       h2_type = "Chip")
+corrected_effects_twin <- run_pgic_data(correct_scores, h2_column = "twin_h2",
+                                        h2_type = "Twin")
 
 corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
+corrected_effects$conf.low <- corrected_effects$beta - 1.96 *
+                              corrected_effects$SE
+corrected_effects$conf.high <- corrected_effects$beta + 1.96 *
+                               corrected_effects$SE
+
+unlink("pgi_correct_data.txt")
+
+corrected_effects <- left_join(corrected_effects, correct_scores,
+                               by = join_by(trait == Abr),
+                               suffix = c("", ".y"))
+
+# The odd looking results are for traits with large missing heritability e.g.
+Asthma.
+My guess is that the rare variants act somewhat differently as DHJ as argued, meaning traits with large missing heritability show nonsense twin heritability estimates.
 
@@
 
-corrected_effects <- corrected_effects |> 
-  rename(trait = pgs,
-         TwinH2_Source = Source,
-         Heritability = heritability)
-
+# Cleaning and saving corrected effects so people can have the exact figures
+corrected_effects$TwinH2_Source <- corrected_effects$Source
 corrected_effects |> 
-  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
+  select(trait, beta, SE, Heritability, chip_h2, twin_h2, R2, 
            TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
@@
 
+# TE Note that under Fisher's fundamental theorem of natural selection, the selection
+# for fitness should be equal to it's heritability... But in HRS the
+# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
+# is much lower than what the heritability of NEB should be.
+# The low R2 of NEB causes it to be removed, but the genetics for NEB 
+# could just be a little different in HRS versus other samples!
+# NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]
+
 corrected_effects |>
@@
     trait = nice_names[trait],
-    trait = fct_reorder(trait, estimate)
+    trait = fct_reorder(trait, beta)
   ) |>
-  ggplot(aes(y = trait, x = estimate, xmin = conf.low, xmax = conf.high,
+  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
            color = Heritability)) +
@@
 
-  fs_fert <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
-  fs_phen <- map(pgs, \(x) reformulate(c(x), response = phenotype))
+  fs_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
+  fs_phen <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))
 
@@
 
-```{r}
-#| label: fig-true-effects-old
-#| eval: false
-#| fig-cap: |
-#|   Estimated coefficients of true polygenic scores on RLRS calculated
-#|   using estimates of chip- and twin-heritability and the most recent
-#|   polygenic score.
-
-# TODO: maybe just use the polygenic score name not the phenotype?
-# TODO: why are some result bounds negative? e.g. ADHD? Presumably
-#. because can have negative effect on RLRS?
-# TODO: check the original phenotypes from which h2 ests were derived...
-#. - is CAD the same as coronary heart disease (r10hearte)?
-# TODO: why is ANXCC not predicting anxiety?
-corrected_effects |> 
-  mutate(
-    Phenotype = nice_names[source.pgs],
-    Heritability = target.pgs
-  ) |> 
-  filter(Heritability %in% c("Chip", "Twin"), ! source.pgs %in% old_pgs) |> 
-  ggplot(aes(y = Phenotype, x = mean, xmin = conf.low, xmax = conf.high,
-             color = Heritability)) +
-    geom_vline(xintercept = 0, color = "grey60") +
-    geom_pointrange(aes(linetype = Heritability == "Twin"),
-                    position = position_dodge(0.2)) +
-    scale_linetype(guide = "none") +
-    coord_cartesian(xlim = c(- 0.3, 0.3)) +
-    labs(x = "Corrected coefficient on RLRS", y = "") +
-    theme_minimal()
-
-```
-
 # Discussion
@@
 
-We also examined natural selection in different groups.
+Why does the US data show fewer differences by socio-economic status (SES) than the UK?
+One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
+The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
+@goldscheider1969minority, @johnson1979minority, @yang2003big).
+And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
+But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
+Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
+
 Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment.
@@
 The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time.
-The smaller black sample makes most tests inconclusive for this population: we can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.
 
 Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations.
-Our estimate of how much intelligence is changing owing to natural selection, is similar to estimates given in the prior literature.
+Our estimate of how much natural selection is changing intelligence is similar to previous work.
 For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation.
@@
 Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits.
-The selection differential for self-rated health is greater than those of education and cognitive performance.
+The selection differential for self-rated health is greater than those for education and cognitive performance.
 The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human].

### 2024-02-23 hrs-selection.qmd

--- 2024-02-22 hrs-selection.qmd
@@
 
-Why does the US data show fewer differences by socio-economic status (SES) than the UK?
-One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority, @johnson1979minority, @yang2003big).
-And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
-But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
-Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
+```{r}
+#| label: top-PGS
+#| eval: false
+
+# Is it ok to residualise PGS on PCS vs control for them?
+# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
+# if the regression beta is to be the covariance for the Price Equation
+# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs.
+The difference is marginal (~1% lower in the below example).
+Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.
+
+# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs.
+Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs?
+I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs.
+I'd have to think very carefully about how to do this.
+
+# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
+# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.
+
+reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
+reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
+# beta = -0.027651
+reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
+
+reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
+             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
+
+reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
+                               reg_pc$model)
+# beta = -0.027651
+reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
+                       reg_pc$model)
+# beta = -0.027358
+
+reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
+                       reg_pc$model)
+# beta = -0.027358
+
+# Below I get the estimate of the interactions and their p values. 
+# EA is only significant for income at p = 0.0106
+# There are always traits significant after bonferroni. But usually it's autism
+# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.
+
+for (eth in "white") {
+  r_eth <- rand |> subset(ethnicity == eth)
+
+  if (eth == "white") {
+    pgs <- pgs_white
+  } else {
+    pgs <- pgs_black
+  }
+
+  # the colon interaction estimates the effect separately within each group
+  form_pgs_x_ed <- paste0(pgs, "*edu") |>
+    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
+  mod_pgs_x_ed[[eth]]  <-
+    map(form_pgs_x_ed, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_ed[[eth]] <-
+    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_ed[[eth]]) <- pgs
+  tidy_pgs_x_ed[[eth]] <-
+    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(edu = str_remove(term, ".*:"))
+
+  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
+    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
+  mod_pgs_x_inc[[eth]]  <-
+    map(form_pgs_x_inc, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_inc[[eth]] <-
+    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_inc[[eth]]) <- pgs
+  tidy_pgs_x_inc[[eth]] <-
+    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(income_med = str_remove(term, ".*:income_med"))
+
+  form_pgs_x_mar <- paste0(pgs, "*married") |>
+    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
+  mod_pgs_x_mar[[eth]]  <-
+    map(form_pgs_x_mar, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_mar[[eth]] <-
+    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
+  names(tidy_pgs_x_mar[[eth]]) <- pgs
+  tidy_pgs_x_mar[[eth]] <-
+    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(
+      married = str_remove(term, ".*:married")
+    )
+
+  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
+    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
+  mod_pgs_x_agefb[[eth]]  <-
+    map(form_pgs_x_agefb, 
+               \(x) svyglm_quiet(x, r_eth))
+
+  tidy_pgs_x_agefb[[eth]] <-
+    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
+    setNames(pgs) |> 
+    list_rbind(names_to = "pgs") |>
+    filter(grepl(':', term)) |>
+    mutate(
+      agefb = str_remove(term, ".*:agefb")
+    ) 
+}
+
+```
 
@@
 $$
-\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}
-}$$
+\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}}
+$$
 

### 2024-02-23 hrs-selection.qmd
remove dupe top-pgs section

@@
 ```{r}
-#| label: top-PGS
-#| eval: false
-
-# Is it ok to residualise PGS on PCS vs control for them?
-# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
-# if the regression beta is to be the covariance for the Price Equation
-# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs.
-The difference is marginal (~1% lower in the below example).
-Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.
-
-# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs.
-Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs?
-I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs.
-I'd have to think very carefully about how to do this.
-
-# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
-# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.
-
-reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
-reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
-# beta = -0.027651
-reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
-
-reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
-             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)
-
-reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
-                               reg_pc$model)
-# beta = -0.027651
-reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
-                       reg_pc$model)
-# beta = -0.027358
-
-reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
-                       reg_pc$model)
-# beta = -0.027358
-
-# Below I get the estimate of the interactions and their p values. 
-# EA is only significant for income at p = 0.0106
-# There are always traits significant after bonferroni. But usually it's autism
-# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.
-
-for (eth in "white") {
-  r_eth <- rand |> subset(ethnicity == eth)
-
-  if (eth == "white") {
-    pgs <- pgs_white
-  } else {
-    pgs <- pgs_black
-  }
-
-  # the colon interaction estimates the effect separately within each group
-  form_pgs_x_ed <- paste0(pgs, "*edu") |>
-    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
-  mod_pgs_x_ed[[eth]]  <-
-    map(form_pgs_x_ed, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_ed[[eth]] <-
-    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_ed[[eth]]) <- pgs
-  tidy_pgs_x_ed[[eth]] <-
-    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(edu = str_remove(term, ".*:"))
-
-  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
-    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
-  mod_pgs_x_inc[[eth]]  <-
-    map(form_pgs_x_inc, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_inc[[eth]] <-
-    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_inc[[eth]]) <- pgs
-  tidy_pgs_x_inc[[eth]] <-
-    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(income_med = str_remove(term, ".*:income_med"))
-
-  form_pgs_x_mar <- paste0(pgs, "*married") |>
-    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
-  mod_pgs_x_mar[[eth]]  <-
-    map(form_pgs_x_mar, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_mar[[eth]] <-
-    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_mar[[eth]]) <- pgs
-  tidy_pgs_x_mar[[eth]] <-
-    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(
-      married = str_remove(term, ".*:married")
-    )
-
-  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
-    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
-  mod_pgs_x_agefb[[eth]]  <-
-    map(form_pgs_x_agefb, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_agefb[[eth]] <-
-    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
-    setNames(pgs) |> 
-    list_rbind(names_to = "pgs") |>
-    filter(grepl(':', term)) |>
-    mutate(
-      agefb = str_remove(term, ".*:agefb")
-    ) 
-}
-
-```
-
-```{r}
 #| label: ADHD

### 2024-02-23 hrs-selection.qmd
Move & rewrite tests for interaction effects with top coefficients. Test EA and ADHD. Add a figure showing the results.

@@
 
-```{r}
-#| label: Top PGS
-#| eval: true
-
-for (eth in "white") {
-  r_eth <- rand |> subset(ethnicity == eth)
-
-  if (eth == "white") {
-    pgs <- pgs_white
-  } else {
-    pgs <- pgs_black
-  }
-
-  # the colon interaction estimates the effect separately within each group
-  form_pgs_x_ed <- paste0(pgs, "*edu") |>
-    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
-  mod_pgs_x_ed[[eth]]  <-
-    map(form_pgs_x_ed, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_ed[[eth]] <-
-    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_ed[[eth]]) <- pgs
-  tidy_pgs_x_ed[[eth]] <-
-    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
-    filter(pgs == "EA") |>
-    mutate(edu = str_remove(term, ".*:"))
-
-  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
-    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
-  mod_pgs_x_inc[[eth]]  <-
-    map(form_pgs_x_inc, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_inc[[eth]] <-
-    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_inc[[eth]]) <- pgs
-  tidy_pgs_x_inc[[eth]] <-
-    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
-    filter(pgs == "EA") |>
-    mutate(income_med = str_remove(term, ".*:income_med"))
-
-  form_pgs_x_mar <- paste0(pgs, "*married") |>
-    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
-  mod_pgs_x_mar[[eth]]  <-
-    map(form_pgs_x_mar, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_mar[[eth]] <-
-    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
-  names(tidy_pgs_x_mar[[eth]]) <- pgs
-  tidy_pgs_x_mar[[eth]] <-
-    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
-    filter(pgs == "EA") |>
-    mutate(
-      married = str_remove(term, ".*:married")
-    )
-
-  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
-    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
-  mod_pgs_x_agefb[[eth]]  <-
-    map(form_pgs_x_agefb, 
-               \(x) svyglm_quiet(x, r_eth))
-
-  tidy_pgs_x_agefb[[eth]] <-
-    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
-    setNames(pgs) |> 
-    list_rbind(names_to = "pgs") |>
-    filter(pgs == "EA") |>
-    mutate(
-      agefb = str_remove(term, ".*:agefb")
-    ) 
-}
-
-eff_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"estimate"]
-se_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"std.error"]
-p_ea_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
-eff_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"estimate"]
-se_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"std.error"]
-p_ea_int_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"estimate"]
-se_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"std.error"]
-p_ea_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"estimate"]
-se_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"std.error"]
-p_ea_int_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"estimate"]
-se_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"std.error"]
-p_ea_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"estimate"]
-se_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"std.error"]
-p_ea_int_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"estimate"]
-se_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"std.error"]
-p_ea_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
-
-eff_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"estimate"]
-se_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"std.error"]
-p_ea_int_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.001, "***",
-            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.01, "**",
-            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.05, "*", "")))
-
-```
-
 ## Differences across social groups
@@
 
-@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income.
+@tbl-x-bootstraps shows estimates and bootstrapped 95% confidence intervals for the number of scores where the $\delta$ term is positive. Confidence intervals exclude 34 only for income and marriage.
 
@@
 {#tbl-x-bootstraps}
+
+To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD.
+Both of these showed large differences in @hugh2022human, and have significant main effects in this data.
+Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups.
+All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA.
+However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction.
+[XXX or we just say "nothing survives B.
+C.]
+
+```{r}
+#| label: stat-top-pgs-interactions
+
+run_interaction_model <- function (pgs, interaction) {
+  # This formulation gives you separate estimates of `pgs`
+  # for each value of  `interaction`, as well as the main effect
+  # for `interaction` itself. We can then test separately
+  # for the interaction term using e.g. coeftest()
+  # I prefer doing this because the coefficients are easy to interpret
+  cross_terms <- glue("{interaction} + {interaction}:{pgs}")
+  form_interaction <- reformulate(cross_terms, response = "rlrs")
+  rand_white <- rand |> subset(ethnicity == "white")
+  mod_interaction <- svyglm_quiet(form_interaction, rand_white)
+
+  mod_interaction
+}
+
+test_interaction_term <- function (mod) {
+  # All our models have two interaction terms. This finds them
+  # and tests if they are equal, using car::linearHypothesis.
+  x_terms <- matchCoefs(mod, ":") 
+  x_hypothesis <- glue("{x_terms[1]} = {x_terms[2]}")
+  car::linearHypothesis(mod, x_hypothesis)
+}
+
+interactions <- c(
+    Education = "edu", 
+    Income    = "income_med", 
+    Marriage  = "married",
+    AFB       = "agefb"
+  )
+
+top_pgs <- c("EA", "ADHD")
+
+regs_to_run <- expand_grid(pgs = top_pgs, interaction = interactions)
+
+# This creates a tibble of regression results
+tidy_regs <- regs_to_run |> 
+  # Using rowwise() like this preserves the variables pgs and interaction,
+  # but makes it easier to refer to variables in list-columns.
+  rowwise(pgs, interaction) |> 
+  mutate(
+    mod = list(run_interaction_model(pgs = pgs, interaction = interaction)),
+    tidy = list(broom::tidy(mod, conf.int = TRUE)),
+    # A test that the two interaction terms
+    # have the same coefficient:
+    test = list(test_interaction_term(mod)),
+    # P value of the test:
+    test.p.value = suppressWarnings(broom::tidy(test)) |> pull(p.value)
+  )
+
+# this gives all the coefficients for the models
+tidy_coefs <- tidy_regs |> 
+  unnest(tidy) |> 
+  ungroup()
+
+```
+
+```{r}
+#| label: fig-top-pgs-interactions
+#| fig-cap: |
+#|   Coefficients of RLRS across different groups for EA and ADHD 
+#|   polygenic scores. Lines show 95 per cent confidence intervals. Groups
+#|   are: 0-12 years education vs. 13-17 years; below vs. above median
+#|   income; all others vs. married; below vs. above median age at first
+#|   birth.
+
+pos_dod <- position_dodge(width = 0.2)
+
+tidy_coefs |> 
+  # Select only the interaction terms:
+  filter(grepl(":", term)) |> 
+  mutate(.by = c(pgs, interaction),
+    PGS = pgs,
+    # gets the pretty name of the group
+    Split = names(interactions)[match(interaction, interactions)],
+    Group = c("Unprivileged", "Privileged"),
+    Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
+                        "Difference n.s.")
+  ) |> 
+  ggplot(aes(estimate, Split, colour = Group)) + 
+    geom_linerange(aes(xmin = conf.low, xmax = conf.high), alpha = 0.5,
+                   position = pos_dod) +
+    geom_point(aes(shape = Group, colour = Group, fill = Difference),
+               position = pos_dod, 
+               shape = "circle filled", size = 2) +
+    facet_wrap(vars(PGS), ncol = 1) +
+    scale_fill_manual(values = c("n.s." = "white", 
+                                 "p < 0.05" = "orange"),
+                                 guide = guide_legend(title = NULL)) +
+    scale_color_manual(values = c("Privileged" = "steelblue4", 
+                                  "Unprivileged" = "red4")) +
+    theme_minimal() +
+    theme(
+      panel.grid.major.y = element_blank(),
+      panel.grid.minor.x = element_blank()
+    ) + 
+    labs(x = "Coefficient on RLRS", y = "")
+
+```
 

### 2024-02-23 hrs-selection.qmd
tweaks

@@
     Split = names(interactions)[match(interaction, interactions)],
-    Group = c("Unprivileged", "Privileged"),
+    Group = c("Disadvantaged", "Advantaged"),
     Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
@@
     facet_wrap(vars(PGS), ncol = 1) +
-    scale_fill_manual(values = c("n.s." = "white", 
-                                 "p < 0.05" = "orange"),
+    scale_fill_manual(values = c("Difference n.s." = "white", 
+                                 "Difference p < 0.05" = "orange"),
                                  guide = guide_legend(title = NULL)) +
-    scale_color_manual(values = c("Privileged" = "steelblue4", 
-                                  "Unprivileged" = "red4")) +
+    scale_color_manual(values = c("Advantaged" = "steelblue3", 
+                                  "Disadvantaged" = "red4")) +
     theme_minimal() +

### 2024-02-27 hrs-selection.qmd
Phenotypic change updated in text

@@
 However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction.
-[XXX or we just say "nothing survives B.
-C.]
+\[XXX or we just say "nothing survives B.
+C.\]
 
@@
 
+```{r}
+#| label: stat-pgicorrect
+#| cache: false
+
+####################################
+# Formatting parameter values to use
+####################################
+
+# correct_scores holds h2 and R2 values, as well as useful information about the PGS
+# such as group categories taken from the PGI repository and simplified names
+correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx", 
+                                     na = "NA", range = "B1:U53")
+
+# Where h2 and R2 are missing, we take values from a similar sample - the WLS
+correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2), 
+                                 correct_scores$WLS_h2,
+                                 correct_scores$HRS_h2)
+
+correct_scores$R2 <- ifelse(is.na(correct_scores$HRS_R2),
+                            correct_scores$WLS_R2,
+                            correct_scores$HRS_R2)
+
+correct_scores$twin_h2 <- correct_scores$TwinH2
+correct_scores$pgs     <- correct_scores$Abr
+
+# This removes NEBwomen from the data
+correct_scores <- correct_scores[-which(correct_scores$pgs == "NEBwomen"),]
+
+correct_scores$pgs <- case_match(correct_scores$pgs,
+  "CANNABIS" ~ "CANNABISpgirep",
+  "DPW"      ~ "DPWpgirep",
+  "HEIGHT"   ~ "HEIGHTpgirep",
+  .default = correct_scores$pgs
+)
+
+# some of the PGS have estimates rounded to zero and are thus useless
+correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
+correct_scores <- correct_scores[,c("pgs",
+                                    "chip_h2",
+                                    "twin_h2",
+                                    "R2",
+                                    "Source", 
+                                    "PoldermanICF_ICD10_subchapter",
+                                    "Category")]
+
+# TE I originally used 0.28 from the TED sample. But that leads to especially absurd estimates
+# 0.15 was estiamted in the American MIDUS sample from the same paper.
+# Leading to slightly less absurd estimates. Perhaps more appropriate in the American context.
+# Uncertainty regarding parameters means results should be interepreted with CAUTION!!!
+# adjustment to use Midus values for NEB is also used
+
+correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
+#correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
+pgs_correct <- correct_scores$pgs
+
+####################################
+# Running the regressions
+####################################
+
+form_pgs_correct <- map(pgs_correct, \(x) reformulate(c(x), response = "rlrs"))
+
+corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
+    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
+    list_rbind() |>
+    filter(term %in% pgs) |>
+    rename(pgs = term) |>
+    left_join(correct_scores, by = join_by(pgs))
+
+corrected_effects_snp <- corrected_effects |>
+    mutate(estimate = estimate*sqrt(chip_h2/R2),
+           std.error = std.error*sqrt(chip_h2/R2),
+           conf.low = conf.low*sqrt(chip_h2/R2),
+           conf.high = conf.high*sqrt(chip_h2/R2),
+           expected_phenotypic_change = estimate*sqrt(chip_h2),
+           heritability = "Chip")
+
+corrected_effects_twin <- corrected_effects |>
+    mutate(estimate = estimate*sqrt(twin_h2/R2),
+           std.error = std.error*sqrt(twin_h2/R2),
+           conf.low = conf.low*sqrt(twin_h2/R2),
+           conf.high = conf.high*sqrt(twin_h2/R2),
+           expected_phenotypic_change = estimate*sqrt(twin_h2),
+           heritability = "Twin")
+
+corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
+
+corrected_effects$Category <-
+  case_match(
+    corrected_effects$Category,
+    "Health and health behaviours"     ~ "Health",
+    "Personality and well-being"       ~ "Personality",
+    "Fertility and sexual development" ~ "Health",
+    "Anthropometric"                   ~ "Health",
+    "Cognition and education"          ~ "Cognition"
+  )
+
+corrected_effects <- corrected_effects |> 
+  rename(trait = pgs,
+         TwinH2_Source = Source,
+         Heritability = heritability)
+
+corrected_effects |>
+  select(trait, estimate, std.error, expected_phenotypic_change, Heritability, chip_h2, 
+         twin_h2, R2, TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
+  readr::write_csv("Corrected_effects.csv")
+
+####################################
+# Extracting Phenotypic change
+####################################
+
+IQgen_change <- corrected_effects_twin |> 
+                filter(pgs == "CP") |>
+                pull(estimate)
+
+IQphen_change_sd_units <- corrected_effects_twin |> 
+                filter(pgs == "CP") |>
+                pull(expected_phenotypic_change)
+IQphen_change <- IQphen_change_sd_units*15
+
+```
+
 Lastly, we would like to know natural selection's effect sizes.
@@
 
-<!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->
-
 To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability.
-For cognitive performance we have estimated a genetic change of $-0.066$ standard deviations per generation, assuming a twin heritability of $0.51$.
-This implies a phenotypic change of $-0.066\times0.51^\frac{1}{2}=-0.047$ standard deviations per generation, equivalent to $-0.71$ points in the units of IQ, where a standard deviation is equal to $15$ points.
+For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$.
+This implies a phenotypic change of $`r IQgen_change`\times0.51^\frac{1}{2}=`r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points.
 This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype.
@@
 <!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
-
-```{r}
-#| label: stat-pgicorrect
-#| cache: false
-
-# Testing the R2 of the ADHD PGS
-rand$variables |>
-  with(cor.test(adhd_score, ADHD,, use = "pairwise.complete")) -> adhd_cor
-
-# Writing data for pgi_correct to use
-readr::write_tsv(
-  rand$variables |> filter(ethnicity == "white"),
-  "pgi_correct_data.txt",
-  quote = "none"
-)
-
-####################################
-# Formatting parameter values to use
-####################################
-
-# correct_scores holds h2 and R2 values, as well as useful information about the PGS
-# such as group categories taken from the PGI repository and simplified names
-correct_scores <- readxl::read_excel("PGI_Heritabilities.xlsx", 
-                                     na = "NA", range = "B1:U53")
-
-# Where h2 and R2 are missing, we take values from a similar sample - the WLS
-correct_scores$chip_h2 <- ifelse(is.na(correct_scores$HRS_h2), 
-                                 correct_scores$WLS_h2,
-                                 correct_scores$HRS_h2)
-
-correct_scores$R2 <- ifelse(is.na(correct_scores$HRS_R2),
-                            correct_scores$WLS_R2,
-                            correct_scores$HRS_R2)
-
-correct_scores$twin_h2 <- correct_scores$TwinH2
-correct_scores$pgs     <- correct_scores$Abr
-
-# This removes NEBwomen from the data
-correct_scores <- correct_scores[-which(correct_scores$pgs == "NEBwomen"),]
-
-correct_scores$pgs <- case_match(correct_scores$pgs,
-  "CANNABIS" ~ "CANNABISpgirep",
-  "DPW"      ~ "DPWpgirep",
-  "HEIGHT"   ~ "HEIGHTpgirep",
-  .default = correct_scores$pgs
-)
-
-# some of the PGS have estimates rounded to zero and are thus useless
-correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
-correct_scores <- correct_scores[,c("pgs",
-                                    "chip_h2",
-                                    "twin_h2",
-                                    "R2",
-                                    "Source", 
-                                    "PoldermanICF_ICD10_subchapter",
-                                    "Category")]
-
-# TE I originally used 0.28 from the TED sample. But that leads to especially absurd estimates
-# 0.15 was estiamted in the American MIDUS sample from the same paper.
-# Leading to slightly less absurd estimates. Perhaps more appropriate in the American context.
-# Uncertainty regarding parameters means results should be interepreted with CAUTION!!!
-# adjustment to use Midus values for NEB is also used
-
-correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
-#correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
-pgs_correct <- correct_scores$pgs
-
-####################################
-# Running the regressions
-####################################
-
-form_pgs_correct <- map(pgs_correct, \(x) reformulate(c(x), response = "rlrs"))
-
-corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
-    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
-    list_rbind() |>
-    filter(term %in% pgs) |>
-    rename(pgs = term) |>
-    left_join(correct_scores, by = join_by(pgs))
-
-corrected_effects_snp <- corrected_effects |>
-    mutate(estimate = estimate*sqrt(chip_h2/R2),
-           std.error = std.error*sqrt(chip_h2/R2),
-           conf.low = conf.low*sqrt(chip_h2/R2),
-           conf.high = conf.high*sqrt(chip_h2/R2),
-           heritability = "Chip")
-
-corrected_effects_twin <- corrected_effects |>
-    mutate(estimate = estimate*sqrt(twin_h2/R2),
-           std.error = std.error*sqrt(twin_h2/R2),
-           conf.low = conf.low*sqrt(twin_h2/R2),
-           conf.high = conf.high*sqrt(twin_h2/R2),
-           heritability = "Twin")
-
-corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
-
-corrected_effects$Category <-
-  case_match(
-    corrected_effects$Category,
-    "Health and health behaviours"     ~ "Health",
-    "Personality and well-being"       ~ "Personality",
-    "Fertility and sexual development" ~ "Health",
-    "Anthropometric"                   ~ "Health",
-    "Cognition and education"          ~ "Cognition"
-  )
-
-corrected_effects <- corrected_effects |> 
-  rename(trait = pgs,
-         TwinH2_Source = Source,
-         Heritability = heritability)
-
-corrected_effects |>
-  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
-           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
-  readr::write_csv("Corrected_effects.csv")
-
-```
 
@@
 Our estimate of how much natural selection is changing intelligence is similar to previous work.
-For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation.
+For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation.
 However, cognitive traits are only one part of the story of natural selection in humans.

### 2024-02-28 hrs-selection.qmd
Parents education and improved discussion of generations

@@
 # rand.
-# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died
-rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied)
+# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died;
+# Mother's years of education; father's years of education
+rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied,
+                               rameduc, rafeduc) |>
+                        mutate(paedyrs = as.numeric(rameduc + rafeduc)/2) #parent ave education
 
@@
          r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
-         rlrs_div_h13gkid, child_weight, h13ageokid)
+         rlrs_div_h13gkid, child_weight, h13ageokid, paedyrs)
 
@@
 tidy_pgs_ed <- mod_pgs_edyrs <- list()
+tidy_pgs_ed_parent <- mod_pgs_edyrs_parent <- list()
 tidy_pgs_sibs <- mod_pgs_sibs <- list()
@@
 cors_sibs_edu <- list()
+cors_sibs_edu_parent <- list()
 cors_gkids_edu <- list()
@@
   tidy_pgs_ed[[eth]] <- mod_pgs_edyrs[[eth]] |>  
+    map(broom::tidy) |>
+    list_rbind() |> 
+    filter(term %in% pgs)
+
+  form_pgs_edyrs_parent <- map(pgs, \(x) reformulate(c(x),
+                                              response = "paedyrs"))
+  mod_pgs_edyrs_parent[[eth]] <- map(form_pgs_edyrs_parent, 
+                                     \(x) svyglm_quiet(x, r_parents_eth))
+  names(mod_pgs_edyrs_parent[[eth]]) <- pgs
+
+  tidy_pgs_ed_parent[[eth]] <- mod_pgs_edyrs_parent[[eth]] |>  
     map(broom::tidy) |>
@@
 
+  cors_sibs_edu_parent[[eth]] <- cor(tidy_pgs_sibs[[eth]]$estimate, 
+                     tidy_pgs_ed_parent[[eth]]$estimate)
+
     form_pgs_gkids <- map(pgs, 
@@
 # since it is close to 0:
-n_reps <-  599
+n_reps <-  30 #599
 
@@
         map(\(x) reformulate(c(x), response = "raedyrs")) |> 
+        map(\(x) lm(x, data, weights = weights)) |> 
+        map_dbl(\(x) coef(x)[[2]])
+
+  c(
+    cor = cor(coef_pgs_edu, coef_pgs_fert),
+    coef = coef(lm(coef_pgs_fert ~ coef_pgs_edu))["coef_pgs_edu"]
+  )
+}
+
+calc_cors_edu_parent <- function (weights, data, var) {
+  coef_pgs_fert <- pgs |> 
+        map(\(x) reformulate(c(x), response = var)) |> 
+        map(\(x) lm(x, data, weights = weights)) |> 
+        map_dbl(\(x) coef(x)[[2]])
+
+  coef_pgs_edu <- pgs |> 
+        map(\(x) reformulate(c(x), response = "paedyrs")) |> 
         map(\(x) lm(x, data, weights = weights)) |> 
@@
     withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
+  cors_sibs_edu_parent[[eth]] <- 
+    withReplicates(r_parents_boot, theta = calc_cors_edu_parent, var = "r10sibs")
   cors_gkids_edu[[eth]] <- 
@@
 ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
+ci_boot_sibs_parent <- map(cors_sibs_edu_parent, \(x) confint(x)[1,])
 ci_boot_gkids <- map(cors_gkids_edu, \(x) confint(x)[1,])
@@
 ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
+ci_boot_sibs_lm_parent <- map(cors_sibs_edu_parent, \(x) confint(x)[2,])
 ci_boot_gkids_lm <- map(cors_gkids_edu, \(x) confint(x)[2,])
@@
 if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
+if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$white insignificant")
+if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$black significant")
 if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
@@
 We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent.
-Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations.
+Parents of no siblings cannot be included, so coefficient sizes are not perfectly comparable across the generations.
 Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
@@
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
+
+The HRS has data on the years of education for the parents of respondents.
+We recalculate the coefficients on education using the parent's PGS, in case the genetics of education and human capital changed between generations.
+We then correlate these coefficients with the number of siblings, finding a significant result (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`).
+The correlation is only slightly smaller and the confidence intervals are only slightly wider.
+This is surprising since parent's education is likely recorded with error and the respondent's polygenic score is only a proxy for their parents.
 
@@
 Patterns of selection appear consistent across the twentieth and early twenty-first century.
-The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time.
+The result is consistent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929.
+The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.
 

### 2024-03-02 hrs-selection.qmd
Refined phenotypic change section

@@
   less educated, unmarried and younger parents, but these predictions are only
-  partly borne out: coefficients are larger among low-income
-  parents, but not among other groups. Using respondents' number of 
+  half borne out: coefficients are larger only among low-income
+  parents and unmarried parents. Using respondents' number of 
   siblings and grandchildren, we 
@@
   EVERSMOKE = "Ever Smoker",
-  EXTRA = "Exraversion",
+  EXTRA = "Extraversion",
   FAMSAT = "Satisfaction with Family",
@@
 This leaves a total of `r length(pgs_white)` scores for the white participants.
-PGS are rescaled to zero mean and unit variance.
-In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.
-For our purposes, this is equivalent to residualizing on principal components [@wurm2014residualizing].
-Effectively, our analysis is of the residualized PGS that are widely used in practical applications.
-\[TODO actually we could just residualize; only the SE of the PCs is affected\]
+PGS were residualised on the first ten within-ethnicity principal components of the DNA array data, to reduce bias from population stratification.
+PGS were then rescaled to zero mean and unit variance.
 
@@
 The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection.
-However the approach is also based in quantitative genetic theory and is able to yield an approximate estimate of genetic change from one generation to the next.
+However the approach is also based on quantitative genetic theory and is able to yield an estimate of genetic change from one generation to the next.
 
@@
 In humans, infant mortality is so low today that a person's reproductive success is a very close approximation of the fitness.
-As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score from one generation.
-We standardize our polygenic scores to have a mean of 0 and a standard deviation of 1.
-With this standardization the regression coefficient of RLRS on the PGS is equal to the covariance.
-Our approach to measuring natural selection is not novel, having been and detailed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].
+As such, the covariance between a polygenic score and RLRS yields the expected change in the mean polygenic score per generation.
+Because the polygenic scores are standardized to have a variance of 1, the regression coefficient of RLRS on the PGS is equal to the covariance.
+Our approach to measuring natural selection is not novel, having been employed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].
 
@@
 # since it is close to 0:
-n_reps <-  30 #599
+n_reps <-  599
 
@@
 if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
-if (significant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
+if (insignificant(ci_boot_fert$black)) warning("ci_boot_fert$black insignificant")
 if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
@@
 #|   coefficients on years of education. Each dot is one polygenic score.
-#|   Controls include 10 principal components of genetic array
-#|   data. Dashed lines are fitted from linear regressions.
+#|   Dashed lines are fitted from linear regressions.
 
@@
 The HRS has data on the years of education for the parents of respondents.
-We recalculate the coefficients on education using the parent's PGS, in case the genetics of education and human capital changed between generations.
-We then correlate these coefficients with the number of siblings, finding a significant result (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I.
+We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations.
+We then correlate these new coefficients with the number of siblings, finding a positive correlation again (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`).
-The correlation is only slightly smaller and the confidence intervals are only slightly wider.
-This is surprising since parent's education is likely recorded with error and the respondent's polygenic score is only a proxy for their parents.
+The estimate is slightly smaller as would be expected since respondents likely recall their parent's education with some error.
 
@@
 
-The EA polygenic score captures a key element of human capital and was trained on a GWAS sample size of one million, making it a sensible PGS with which to test the economic theory.
-In TABLE XXX we report the main effect of the EA polygenic score and its interaction effect on RLRS, for each of the four putative moderators.
-Only for the case of income is the interaction effect significant at p \< 0.05.
-We might wonder whether the weak effects are because of insufficient power or because any real interaction is small in the sample.
-In @hugh2022human the interaction effects were large, often as big as the main effects.
-Yet in the HRS, the point estimates of the interactions appear generally smaller, suggesting the nulls may not just be a consequence of lower power.
-
 We should attain more statistical power if we utilise information from other polygenic scores at the same time.
@@
 Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate.
-To perform, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
+We test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
 Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education.
@@
     geom_point(alpha = 0.7, 
-               position = position_jitter(width = 0.1, seed = 12345)) +
+               position = position_jitter(width = 0.125, height=0.001, seed = 12346)) +
     geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
@@
     geom_text(aes(label = outliers), size = 2, colour = "black", 
-              position = position_jitter(width = 0.1, seed = 12345), 
+              position = position_jitter(width = 0.125,  height=0.001, seed = 12346), 
               hjust = -0.15) +
@@
 The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
-Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS.
+Polygenic scores contain error, so estimated effects are biased towards zero compared to the effect of the true PGS.
 They can be scaled up by
@@
 Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century.
-Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed.
-The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting.
-The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.
-
-<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
+Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased in the late twentieth century, contrary to how we believe the genetics have changed.
+However, the prevalence of self-reported anxiety and depressive symptoms have increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism.
+The prevalence of asthma has also increased [@Moorman2007] in line with the polygenic score.
+The apparent paradox of phenotypes and nautral selection moving in opposite directions suggests that environmental change across history is a much greater force than that of natural selection.
 
@@
 
-Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment.
+Across three generations of Americans, we found estimates of selection to correlate over time and that selection is always more positive for traits that negatively predict educational attainment.
 Patterns of selection appear consistent across the twentieth and early twenty-first century.
-The result is consistent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929.
+The result is congruent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929.
 The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.
@@
 Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations.
-Our estimate of how much natural selection is changing intelligence is similar to previous work.
+Our estimate of how much natural selection is changing intelligence is slightly smaller than in previous work.
 For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation.
@@
 There are `r n_black` genotyped black survey participants.
-Among this sample, the relationship between polygenic scores' coefficients on RLRS and on education is negative but insignificant and imprecisely estimated (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I.
+Among this sample, the relationship between polygenic scores' coefficients on RLRS and on education is negative and significant (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I.
 `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`).
@@
 #| fig-cap: |
-#|  Coefficients of PGS on RLRS, controlling for 10 principal components of
-#|  genomic array data. Lines are 95% confidence intervals.
+#|  Coefficients of PGS on RLRS. Lines are 95% confidence intervals.
 #| fig-height: 6
@@
 #|   Scatterplot of PGS coefficients on number of live siblings and years of
-#|   education. Controls include 10 principal components of genetic array
-#|   data. Dashed lines show linear regressions.
+#|   education. Dashed lines show linear regressions.
 
@@
 #|   education. RLRS2 is the reproductive success of respondent's offspring
-#|   relative to the success of other offspring. Controls include 10 principal
-#|   components of genetic array data. Dashed lines show linear regressions.
+#|   relative to the success of other offspring. Dashed lines show linear regressions.
 

### 2024-03-03 hrs-selection.qmd
Small tweak before pulling

--- 2024-02-27 hrs-selection.qmd
@@
 However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction.
-\[XXX or we just say "nothing survives B.
-C.\]
+[XXX or we just say "nothing survives B.
+C.]
 
@@
   # I prefer doing this because the coefficients are easy to interpret
-  cross_terms <- glue("{interaction} + {interaction}:{pgs}")
-  form_interaction <- reformulate(cross_terms, response = "rlrs")
+  fml_x <- glue("rlrs ~ {interaction} + {interaction}:{pgs}")
+  fml_x <- as.formula(fml_x)
   rand_white <- rand |> subset(ethnicity == "white")
-  mod_interaction <- svyglm_quiet(form_interaction, rand_white)
-
-  mod_interaction
+  mod_x <- svyglm_quiet(fml_x, rand_white)
+
+  mod_x
 }
@@
     test = list(test_interaction_term(mod)),
-    # P value of the test:
+    # P value of the test. (broom throws a warning which I think we can 
+    # ignore.)
     test.p.value = suppressWarnings(broom::tidy(test)) |> pull(p.value)
@@
 
+Lastly, we would like to know natural selection's effect sizes.
+The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
+Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS.
+They can be scaled up by
+
+$$
+\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}}
+$$
+
+where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource].
+SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository.
+The authors use GCTA to estimate heritability.
+When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead.
+We also perform error correction with twin heritabilites.
+To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1].
+We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power.
+This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].
+Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.
+
+[^1]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin.
+For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].
+
+[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms.
+These were PV001-PV018 in the 2016 core survey.
+They estimated an $R^2_{PGS}$ of $0.003$.
+When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`.
+Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not very significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I.
+`r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`.
+Given the great uncertainty over the true correlation we were unable to justify using the error-corrected estimator with the ADHD polygenic score.
+
+@fig-true-effects reports the error corrected estimates of selection.
+As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
+Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
+The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations.
+For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates.
+Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
+
+To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits.
+Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance.
+Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster.
+This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.
+
+<!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->
+
+To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability.
+For cognitive performance we have estimated a genetic change of $-0.066$ standard deviations per generation, assuming a twin heritability of $0.51$.
+This implies a phenotypic change of $-0.066\times0.51^\frac{1}{2}=-0.047$ standard deviations per generation, equivalent to $-0.71$ points in the units of IQ, where a standard deviation is equal to $15$ points.
+This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype.
+Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.
+
+Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century.
+Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed.
+The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting.
+The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.
+
+<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
+
 ```{r}
@@
 #| cache: false
+
+# Testing the R2 of the ADHD PGS
+rand$variables |>
+  with(cor.test(adhd_score, ADHD,, use = "pairwise.complete")) -> adhd_cor
+
+# Writing data for pgi_correct to use
+readr::write_tsv(
+  rand$variables |> filter(ethnicity == "white"),
+  "pgi_correct_data.txt",
+  quote = "none"
+)
 
@@
            conf.high = conf.high*sqrt(chip_h2/R2),
-           expected_phenotypic_change = estimate*sqrt(chip_h2),
            heritability = "Chip")
@@
            conf.high = conf.high*sqrt(twin_h2/R2),
-           expected_phenotypic_change = estimate*sqrt(twin_h2),
            heritability = "Twin")
@@
 corrected_effects |>
-  select(trait, estimate, std.error, expected_phenotypic_change, Heritability, chip_h2, 
-         twin_h2, R2, TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
+  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
+           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
   readr::write_csv("Corrected_effects.csv")
 
-####################################
-# Extracting Phenotypic change
-####################################
-
-IQgen_change <- corrected_effects_twin |> 
-                filter(pgs == "CP") |>
-                pull(estimate)
-
-IQphen_change_sd_units <- corrected_effects_twin |> 
-                filter(pgs == "CP") |>
-                pull(expected_phenotypic_change)
-IQphen_change <- IQphen_change_sd_units*15
-
-```
-
-Lastly, we would like to know natural selection's effect sizes.
-The regression beta of RLRS on the PGS gives the change in one generation in the PGS due to natural selection, measured in standard deviations.
-Polygenic scores contain error, so estimated effects are biased towards zero compared to the correlation of the true PGS.
-They can be scaled up by
-
-$$
-\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}}
-$$
-
-where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the coefficient of determination of the measured PGS on the target phenotype [@becker2021resource].
-SNP- or chip-heritabilities and $R^2_{PGS}$ are calculated by @becker2021resource for scores in the Polygenic Index Repository.
-The authors use GCTA to estimate heritability.
-When the corresponding phenotype is not available in the HRS to estimate $R^2_{PGS}$, we use parameters estimated by the authors in the Wisconsin Longitudinal Study instead.
-We also perform error correction with twin heritabilites.
-To attain precise estimates we use heritabilities from a meta-analysis including over 14,000 twin pairs authored by @polderman2015twin[^1].
-We remove polygenic scores with $R^2_{PGS} < 0.005$ to focus on PGS with adequate power.
-This removed the ADHD PGS which significantly predicted RLRS, even after bonferroni correction[^2].
-Error-corrected effect sizes and the $R^2_{PGS}$ and $h^2$ parameters used can be found on the Github page for this paper.
-
-[^1]: For age of first birth, we could not find an appropriately close trait in @polderman2015twin.
-For this trait we used a twin heritability of 0.15 as estimated in the Midlife in the United States (MIDUS) sample [@briley2017fertility].
-
-[^2]: @becker2021resource estimate $R^2_{PGS}$ for ADHD using binary, yes or no, items of ADHD symptoms.
-These were PV001-PV018 in the 2016 core survey.
-They estimated an $R^2_{PGS}$ of $0.003$.
-When we attempt to replicate it we find an $r_{PGS}$ of `r adhd_cor$estimate`, implying an $R^2_{PGS}$ of `r adhd_cor$estimate^2`.
-Although this would pass our requirement of $R^2_{PGS} > 0.005$, the correlation was not very significant (p = `r adhd_cor$p.value`) and the confidence intervals were large 95% C.I.
-`r adhd_cor$conf.int[1]` to `r adhd_cor$conf.int[2]` owing to the fact that few respondents were given the ADHD items N = `r adhd_cor$parameter + 2`.
-Given the great uncertainty over the true correlation we were unable to justify using the error-corrected estimator with the ADHD polygenic score.
-
-@fig-true-effects reports the error corrected estimates of selection.
-As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
-Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
-The confidence bounds capture uncertainty from sampling variation, but not other sources, including uncertainty about the true $h^2$, the true $R^2_{PGS}$, limitations of the within-sample phenotypes, noise from correlated environments, and biases in the polygenic scores such as assortative mating, population stratification, and gene-environment correlations.
-For twin-heritability, different relationships with fertility among variants not measured on the chip may also bias the estimates.
-Given all this, the estimates mostly show the limits of our knowledge, and should be treated as best guesses only.
-
-To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits.
-Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance.
-Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster.
-This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.
-
-To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability.
-For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$.
-This implies a phenotypic change of $`r IQgen_change`\times0.51^\frac{1}{2}=`r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points.
-This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype.
-Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.
-
-Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century.
-Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed.
-The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting.
-The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.
-
-<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
+```
 
@@
 Our estimate of how much natural selection is changing intelligence is similar to previous work.
-For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation.
+For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation.
 However, cognitive traits are only one part of the story of natural selection in humans.

### 2024-03-03 response-to-reviewers-revision-1.qmd

@@
 
-Note that controlling for principal components in a regression, *y \~
-PGIs + PCs*, gives the same effect sizes as running an uncontrolled
-regression on PGIs which have previously had the PCs residualized out,
-<!--# TE not an expert here, but I think it should say y_resid here.
-I've residualised PGS but not rlrs, because it would complicate the interpetation of our beta under the Price equation.
-I think we should just remove this paragraph.
-We can explain our approach by saying the residualisation get's us closer to an unbiased PGS and cite others who have done the same thing.-->*y
-\~ PGIs_resid*. We control rather than residualizing because this gives
-the correct standard errors[^1], which might matter with our relatively
-low N.
-
-[^1]: Because the controls are correctly accounted for when calculating
-    degrees of freedom.
-
 Why residualize or control for principal components at all? I think for
@@
 
-To simplify interpretation, we've decided to residualise and then
-standardize the PGIs on the principal components, instead of control for
+To simplify interpretation, we've decided to residualise the PGIs on the
+principal components and then standardize them, instead of control for
 them.
@@
 covariance between the polygenic scores and RLRS.
-
-TE TODO: Tobias, can you check the relevant para? I guess that if both
-PGI and RLRS are normalized then the covariance is the correlation; but
-actually isn't the correct thing covariance with the not-normalized
-RLRS?
 

### 2024-03-04 response-to-reviewers-revision-1.qmd
Tweaks before resubmission: - edit response letter - foreground parents' education correlation

@@
 careful and accurate estimates of effect sizes of "true" and/or
-"chip-wide" scores. I hope this is OK.
+"chip-wide" scores.
+
+I hope this is OK. I don't want to bait and switch. But doing this seems fair to Tobias, and I think it has also improved the paper.
 
@@
 
-Why residualize or control for principal components at all? I think for
-two reasons. First, we are ultimately interested in selection on the
+I think for two reasons. First, we are ultimately interested in selection on the
 "true PGI". Stripping out PCs is an admittedly imperfect way to control
@@
 the phenotypic change. This highlights that environmental change is
-likely to be of greater importance for phenotypic trends then genetic
+likely to be of greater importance for phenotypic trends than genetic
 change.
-
-TE TODO: Could we look at health trends in phenotypes? I'd guess that
-health trends improve over the C20, but we could look at some specifics.
-(Depression/BMI?) One big issue here is that maybe the PGIs have
-confounds... but perhaps that's less serious for some of these
-phenotypes
 
@@
 between young and old parents support the theory that income effects
-dominate for older parents while substitution effects dominate for
+dominate for older parents, while substitution effects dominate for
 younger parents. To be a bit more explicit: if a PGS simply doesn't
@@
 
-Under the null, we'd expect the above equation to hold for just 50% of
-the scores. If it holds for significantly more than 50% of the scores
-(using bootstrapped C.I.s) then we can reject the null in favour of the
-alternative.
+Under the null, we'd expect the above equation to hold in the sample 
+for just 50% of the scores. If it holds for significantly more than 50% of 
+the scores (using bootstrapped C.I.s) then we can reject the null in favour 
+of the alternative.
 
-We also try a simpler alternative approach. Using the few PGIs with the
+We also try a simpler approach. Using the PGIs with the
 largest and most precisely estimated selection effects in the whole
 sample, we estimate selection effects in different subgroups and test
-for a significant difference between them, via a simple interaction
+for a significant difference between them via a simple interaction
 term. We report these results too.
-
-TE TODO: the above
 
@@
 to estimate the selection differential.
-
-TE TODO: Write a paragraph on biol. selection. (Not sure what is meant
-here?)
 
@@
 > in the UK [@hugh2022human]. Future research should study the health
-> and medical implications of natural selection, in addition to their
+> and medical implications of natural selection, in addition to its
 > social implications.
@@
 
-> "RLRS is calculated separately by ethnicity. This is not ideal,
-> because it treats the ethnicities as separate breeding populations.
-> But the alternative of calculating pooled RLRS would effectively be
-> estimating natural selection in the whole US population by treating
-> whites as representative. We therefore focus on the white US
-> population, with the caveat that this"one data point" may not be
-> representative of the US population (still less the species!) The mean
-> number of children of people born in each year was calculated using
-> sampling weights."
-
-Effectively, the white US population is a "single data point" for theory
-testing. Note that since we don't compare the ethnicities, the only
-difference this makes is in estimating the absolute size of effects.
-
-TE TODO: change it to be separate RLRS baseline for white and black.
+> 'RLRS is calculated separately by ethnicity. This is not ideal, because it 
+> treats the ethnicities as separate breeding populations. But the alternative of
+> calculating pooled RLRS would effectively be estimating natural selection in 
+> the whole US population by treating whites as representative. We therefore 
+> focus on the white US population, with the caveat that results from this 
+> "one data point" may not replicate in other ethnicities or countries.'

### 2024-03-04 hrs-selection.qmd
Tweaks before resubmission: - edit response letter - foreground parents' education correlation

@@
   We investigate natural selection on polygenic scores
-  in the contemporary US, using the Health and Retirement Study. Results
+  in the contemporary US, using the Health and Retirement Study. Across
+  three generations, scores which correlate negatively (positively) with
+  education are selected for (against). However, results only
   partially support the economic theory of fertility as an explanation for 
-  natural selection: scores which correlate negatively (positively) with
-  education are selected for (against). The theory predicts that
+  natural selection. The theory predicts that
   selection coefficients should be stronger among low-income, 
@@
   half borne out: coefficients are larger only among low-income
-  parents and unmarried parents. Using respondents' number of 
-  siblings and grandchildren, we 
-  examine selection across three generations, finding consistent results 
-  over time. We also estimate effect sizes corrected for noise in the
-  polygenic scores. Selection for some health traits is similar in magnitude
-  to that for cognitive traits. 
+  parents and unmarried parents. We also estimate effect sizes corrected 
+  for noise in the polygenic scores. Selection for some health traits is 
+  similar in magnitude to that for cognitive traits. 
 date: "February 2024"
@@
 		nest = TRUE,
-		data = rand |> drop_na(weight) 
-		)
+		data = rand |> drop_na(weight))
 
@@
 Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023].
-The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper.
-The motivation for our study is to establish the direction and magnitude of natural selection upon a range of traits in Americans and to test whether the economic theory can explain the selection.
+The motivation is to establish the direction and magnitude of natural selection on a range of traits in Americans, and to test whether the economic theory can explain the selection.
 Using information on respondents' siblings and grandchildren, we can also extend the analysis to three generations of Americans.
-To preview our results, we confirm point 1 above, but only see partial and ambiguous support for point 2.
-(TODO: or can we be stronger?)
+This is interesting first because selection effects may accumulate over time, and second because an possible alternative explanation for existing findings is the development of the welfare state, which happened in the US during the "Great Society" programs of the 1960s (i.e.
+after the respondents' parents' generation).
+To preview our results, we confirm point 1 above across all three generations.
+But we only see partial and ambiguous support for point 2.
 
@@
 The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants.
-we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
+We include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.
 The resulting sample contains `r n_white` genotyped white participants.
 We focus on these because the sample size is large enough.
-The appendix reports the most basic analyses for the `r n_black` genotyped black participants.
+The appendix reports some basic analyses for the `r n_black` genotyped black participants.
 
@@
 We discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype.
-we also discard PGS for number of children ever born (but keep scores for age at first birth).
+We also discard PGS for number of children ever born (but keep scores for age at first birth).
 This leaves a total of `r length(pgs_white)` scores for the white participants.
@@
 But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
-We therefore focus on the white US population, with the caveat that this "one data point" may not be representative of the US population (still less the species!) The mean number of children of people born in each year was calculated using sampling weights.
-
-The intuition behind our modelling is very simple; if a polygenic score predicts more reproductive success then people higher in the trait will reproduce more than others causing that trait to be favored by natural selection.
+We therefore focus on the white US population, with the caveat that results from this "one data point" may not replicate in other ethnicities or countries.
+The mean number of children of people born in each year was calculated using sampling weights.
+
+The intuition behind our analysis is simple: if a polygenic score predicts more reproductive success, then people higher in the PGS will reproduce more than others, causing scores to increase in the population.
 However the approach is also based on quantitative genetic theory and is able to yield an estimate of genetic change from one generation to the next.
@@
 Because the polygenic scores are standardized to have a variance of 1, the regression coefficient of RLRS on the PGS is equal to the covariance.
-Our approach to measuring natural selection is not novel, having been employed in the literature before [@Beauchamp2016selection; @hugh2022human; @kong2017selection].
+This approach is standard in the literature [@Beauchamp2016selection; @hugh2022human; @kong2017selection].
 
@@
 #| label: stat-regressions
-
 # cache seems to cause trouble even with cache-lazy: FALSE
+
 options("survey.lonely.psu" = "average")
@@
 We estimate coefficients of PGS on RLRS.
-These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
+These do not identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility.
 Appendix @fig-rlrs shows coefficients.
@@
 #|   coefficients on years of education. Each dot is one polygenic score.
-#|   Dashed lines are fitted from linear regressions.
+#|   The dashed line shows a linear regression.
 
@@
 Parents of no siblings cannot be included, so coefficient sizes are not perfectly comparable across the generations.
-Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education.
-Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
+The HRS has data on the years of education for the parents of respondents.
+We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations.
+Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on parents' years of education.
+The correlation is negative and significant (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I.
+`r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`).
+The same holds if we use coefficients on own years of education (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I.
 `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`).
@@
 `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).
-
-The HRS has data on the years of education for the parents of respondents.
-We recalculate the coefficients on education using values for the parents, in case the genetics of education and human capital changed between generations.
-We then correlate these new coefficients with the number of siblings, finding a positive correlation again (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I.
-`r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`).
-The estimate is slightly smaller as would be expected since respondents likely recall their parent's education with some error.
 
@@
 `r ci_boot_gkids$white[1]` to `r ci_boot_gkids$white[2]`).
-There is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I.
-`r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating some stability in selection over time.
+Again, there is a positive correlation between PGS coefficients on RLRS2 and RLRS (`r gen_cor_gkids`, bootstrapped 95% C.I.
+`r ci_gen_cor_gkids[1,1]` to `r ci_gen_cor_gkids[1,2]`), indicating stability in selection over time.
 
@@
 
-We should attain more statistical power if we utilise information from other polygenic scores at the same time.
-However, some caution is necessary.
-Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate.
 We test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group.
@@
 If 95% confidence intervals are above `r length(pgs_white)`/2 = `r length(pgs_white)/2`, we reject the null in favour of the alternative.
-\[TODO: what if some are negative and some are positive?
-I think this is fine because it makes our test conservative?
-Maybe??\]
 
@@
 Both of these showed large differences in @hugh2022human, and have significant main effects in this data.
-Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups.
-All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA.
-However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction.
-\[XXX or we just say "nothing survives B.
-C.\]
+@fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups.
+7 out of 8 differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA.
+However, no differences remain significant after Bonferroni correction.
 
@@
     Split = names(interactions)[match(interaction, interactions)],
-    Group = c("Disadvantaged", "Advantaged"),
+    Group = c("Disadvantaged group", "Advantaged group"),
     Difference = ifelse(test.p.value < 0.05, "Difference p < 0.05",
@@
     facet_wrap(vars(PGS), ncol = 1) +
-    scale_fill_manual(values = c("Difference n.s." = "white", 
+    scale_fill_manual(
+      values = c("Difference n.s."     = "white", 
                  "Difference p < 0.05" = "orange"),
       guide = guide_legend(title = NULL)) +
-    scale_color_manual(values = c("Advantaged" = "steelblue3", 
-                                  "Disadvantaged" = "red4")) +
+    scale_color_manual(
+      values = c("Advantaged group"    = "steelblue3", 
+                 "Disadvantaged group" = "red4"),
+      guide = guide_legend(title = NULL)) +
     theme_minimal() +
@@
 
+## Corrected effect sizes
+
 Lastly, we would like to know natural selection's effect sizes.
@@
 @fig-true-effects reports the error corrected estimates of selection.
-As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean.
+As a rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "substantive": about 54% of the new generation will be below the parents' mean.
 Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero.
@@
 To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits.
-Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance.
-Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster.
-This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.
+The point estimates for many traits, including depressive symptoms, asthma, age of first birth and self-rated health, exceed point estimates for educational attainment and cognitive performance.
+This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also relevant to population health.
 
@@
 For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$.
-This implies a phenotypic change of $`r IQgen_change`\times0.51^\frac{1}{2}=`r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points.
+This implies a phenotypic change of $`r IQgen_change` \times 0.51^\frac{1}{2} = `r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points.
 This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype.
@@
 Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased in the late twentieth century, contrary to how we believe the genetics have changed.
-However, the prevalence of self-reported anxiety and depressive symptoms have increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism.
+However, the prevalence of self-reported anxiety and depressive symptoms has increased [@Witters2023; @Twenge2000], paralleling the polygenic scores for depression and neuroticism.
 The prevalence of asthma has also increased [@Moorman2007] in line with the polygenic score.
-The apparent paradox of phenotypes and nautral selection moving in opposite directions suggests that environmental change across history is a much greater force than that of natural selection.
+The apparent paradox of phenotypes and natural selection moving in opposite directions suggests that environmental change is a much stronger force than that of natural selection across recent history.
 
@@
 #| eval: false
+
+# Now we just do this using the python tool
 
@@
 
+Across three generations of Americans, estimates of selection correlate over time, and selection is always more positive for traits that negatively predict educational attainment.
+Patterns of selection appear consistent across the twentieth and early twenty-first century.
+This result fits that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a sample of Swedish twins.
+The underlying causes of selection seem to have stayed the same over time, so that the effects of selection have accumulated.
+In particular, selection against traits associated with education clearly predates the development of the US welfare state.
+
 The economic theory of fertility is driven by the trade-off between children and income.
@@
 Prediction 1 of the theory holds when substitution effects dominate income effects overall.
-Prediction 2 is driven by the specific form of individuals' preferences for income: when utility for income is sharply curved, i.e.
-marginal utility is high at low incomes but low at high incomes, the substitution effect is stronger for those who expect to earn less.
+Prediction 2 is driven by the specific form of individuals' preferences for income: when marginal utility is higher at lower income levels, the substitution effect is stronger for those who expect to earn less.
 
 The results here support prediction 1 but are more ambiguous for prediction 2.
-PGS coefficients on RLRS also appear larger for low-income groups and unmarried respondents.
+PGS coefficients on RLRS appear larger for low-income groups and unmarried respondents.
 But there is little evidence for larger coefficients among people with lower education, or younger parents.
@@
 But in the UK, the between-group differences were large [@hugh2022human]; differences that big would surely have been visible here.
-The theory can accommodate this non-result, if preferences for income are less curved in the US for whatever reason.
+The theory can accommodate this non-result, if preferences for income are less curved in the US for any reason.
 But note that *any* theory with a negative relationship between education and fertility will give prediction 1.[^3] In this sense, results here are less supportive of the economic theory specifically.
@@
 Why does the US data show fewer differences by socio-economic status (SES) than the UK?
-One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
+One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK.
+The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents [blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
 `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority, @johnson1979minority, @yang2003big).
+@goldscheider1969minority; @johnson1979minority; @yang2003big; @maralani2013demography].
 And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
 But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative.
-Another possibility is that the US cohort were exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
-
-Across three generations of Americans, we found estimates of selection to correlate over time and that selection is always more positive for traits that negatively predict educational attainment.
-Patterns of selection appear consistent across the twentieth and early twenty-first century.
-The result is congruent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929.
-The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.
+Another possibility is that the US cohort was exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.
 
@@
 The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human].
-Future research should study the health and medical implications of natural selection, in addition to their social implications.
+Future research should study the health and medical implications of natural selection, in addition to its social implications.
 To know more, we must await more accurate polygenic scores.
@@
 #|   Scatterplot of PGS coefficients on number of live siblings and years of
-#|   education. Dashed lines show linear regressions.
+#|   education for parents. The dashed line shows a linear regression.
 
 tidy_pgs_sibs$white |> 
-  inner_join(tidy_pgs_ed$white, 
+  inner_join(tidy_pgs_ed_parent$white, 
              by = join_by(term), suffix = c(".sibs", ".edyrs")) |> 
@@
 #|   education. RLRS2 is the reproductive success of respondent's offspring
-#|   relative to the success of other offspring. Dashed lines show linear regressions.
+#|   relative to the success of other offspring. The dashed line shows a 
+#|   linear regression.
 
@@
 
-```{r}
-#| label: fig-check-correction
-#| eval: false
-#| fig-cap: |
-#|   Estimated coefficients of series of PGS on fertility 
-#|   using errors-in-variables correction. "Own" is the uncorrected estimate
-#|   using the PGS itself. Other estimates are from earlier PGS targeting
-#|   the same phenotype, corrected via the method in @becker2021resource.
-#| fig-height: 6
-
-corrected_effects |> 
-  filter(length(unique(source.pgs)) > 1, .by = phenotype) |> 
-  filter(! target.pgs %in% c("Chip", "Twin")) |> 
-  filter(
-    ! target.pgs %in% old_pgs
-  ) |> 
-  mutate(
-    Phenotype = case_match(phenotype,
-                           "adhd_score" ~ "ADHD",
-                           "r10cesd" ~ "Depression",
-                           "r10cogtot" ~ "Cognition",
-                           "raedyrs" ~ "Educ. attainment"
-                           ),
-    Target = target.pgs,
-    Source = ifelse(source.pgs == target.pgs, "Own", source.pgs),
-  ) |> 
-  ggplot(aes(y = Target, x = mean, xmin = conf.low, xmax = conf.high,
-             color = Source)) +
-    geom_vline(xintercept = 0, color = "grey60") +
-    geom_pointrange(position = position_dodge(0.7)) +
-    geom_text(aes(x = mean, label = Source), hjust = "center", 
-              vjust = 1.5,
-              position = position_dodge(0.7), size = 2.5) + 
-    scale_color_manual(values = c("Own" = "red"), guide = "none") +
-    facet_grid(rows = vars(Phenotype), scales = "free", space = "free") +
-    labs(x = "Corrected coefficient on RLRS", y = "Target PGS") +
-    coord_cartesian(xlim = c(-0.05, 0.05)) + 
-    theme_minimal()
-
-```
-
 \FloatBarrier

### 2024-03-04 hrs-selection.qmd
Rerun uncached

@@
 #| label: stat-bootstrap-gen-cor
-#| cache: true
+#| cache: false
 #| cache-comments: true
@@
 #| label: stat-bootstrap-gen-cor-gkids
-#| cache: true
+#| cache: false
 #| cache-comments: true
@@
 One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents [blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
 `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
 `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority; @johnson1979minority; @yang2003big; @maralani2013demography].
+@goldscheider1969minority, @johnson1979minority, @yang2003big, @maralani2013demography).
 And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.

### 2024-03-04 response-to-reviewers-revision-1.qmd
Changed letter to editor

@@
 
-I hope this is OK. I don't want to bait and switch. But doing this seems fair to Tobias, and I think it has also improved the paper.
+I hope this is OK. I think it has appreciably improved the paper.
 
@@
 
-I think for two reasons. First, we are ultimately interested in selection on the
-"true PGI". Stripping out PCs is an admittedly imperfect way to control
-for confounding population stratification. Second, PGIs are typically
-residualized in this way, so by examining the effect of these we're
-following the "industry standard" and learning about scores in the form
-they are actually used.
+I think for two reasons. First, we are ultimately interested in
+selection on the "true PGI". Stripping out PCs is an admittedly
+imperfect way to control for confounding population stratification.
+Second, PGIs are typically residualized in this way, so by examining the
+effect of these we're following the "industry standard" and learning
+about scores in the form they are actually used.
 
@@
 polygenic scores. This ensures our regression estimates are exactly the
-covariance between the polygenic scores and RLRS.
+covariance between our polygenic scores and RLRS.
 
@@
 
-Under the null, we'd expect the above equation to hold in the sample 
-for just 50% of the scores. If it holds for significantly more than 50% of 
-the scores (using bootstrapped C.I.s) then we can reject the null in favour 
-of the alternative.
+Under the null, we'd expect the above equation to hold in the sample for
+just 50% of the scores. If it holds for significantly more than 50% of
+the scores (using bootstrapped C.I.s) then we can reject the null in
+favour of the alternative.
 
-We also try a simpler approach. Using the PGIs with the
-largest and most precisely estimated selection effects in the whole
-sample, we estimate selection effects in different subgroups and test
-for a significant difference between them via a simple interaction
-term. We report these results too.
+We also try a simpler approach. Using the PGIs with the largest and most
+precisely estimated selection effects in the whole sample, we estimate
+selection effects in different subgroups and test for a significant
+difference between them via a simple interaction term. We report these
+results too.
 
@@
 
-> 'RLRS is calculated separately by ethnicity. This is not ideal, because it 
-> treats the ethnicities as separate breeding populations. But the alternative of
-> calculating pooled RLRS would effectively be estimating natural selection in 
-> the whole US population by treating whites as representative. We therefore 
-> focus on the white US population, with the caveat that results from this 
-> "one data point" may not replicate in other ethnicities or countries.'
+> 'RLRS is calculated separately by ethnicity. This is not ideal,
+> because it treats the ethnicities as separate breeding populations.
+> But the alternative of calculating pooled RLRS would effectively be
+> estimating natural selection in the whole US population by treating
+> whites as representative. We therefore focus on the white US
+> population, with the caveat that results from this "one data point"
+> may not replicate in other ethnicities or countries.'

### 2024-03-21 response-to-reviewers-revision-1.qmd
Final tweaks for resubmission

@@
 title: "'Natural Selection Across Three Generations of Americans': response to reviewers"
-format: docx
+format: pdf
 editor: 
@@
 on relative lifetime reproductive success. At times in the original
-manuscript this was wrongly referred to as a correlation but thus has
+manuscript this was wrongly referred to as a correlation but this has
 now been updated. The polygenic scores are standardized to have a

### 2024-06-04 hrs-selection.qmd
Small wording tweaks; add some dependencies to renv

--- 2024-03-04 hrs-selection.qmd
@@
 
-titles <- c(white = "White", black = "Black")
+titles <- c(white = "White participants", black = "Black participants")
 
@@
 The appendix reports some basic analyses for the `r n_black` genotyped black participants.
+(Throughout, "white" and "black" refer to participants who (a) self-identified as non-Hispanic and as "White/Caucasian" and "Black/African-American" respectively, and who (b) had principal components of SNP data close to the respective population mean.
+See @hrspgs2020.)
 
@@
 RLRS is calculated separately by ethnicity.
-This is not ideal, because it treats the ethnicities as separate breeding populations.
-But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
+This is not ideal, because it treats the ethnicities as "separate breeding populations" (i.e.
+subpopulations who rarely interbreed; black and white people intermarried rarely in the HRS generation, but that is due to laws and norms against intermarriage which have now mostly disappeared.) But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative.
 We therefore focus on the white US population, with the caveat that results from this "one data point" may not replicate in other ethnicities or countries.
@@
 #| label: stat-bootstrap-eth-cor
-#| cache: false
+#| cache: true
 #| cache-lazy: false
@@
 #| label: stat-pgicorrect
-#| cache: false
+#| cache: true
 
@@
 #| label: stat-bootstrap-true-effects
-#| cache: false
+#| cache: true
 #| eval: false
@@
 One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK.
-The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents [blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
-`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
+The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (black respondents: `r tidy_fert_edu_phen$black$estimate`, 95% C.I.
+`r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, white respondents: `r tidy_fert_edu_phen$white$estimate`, 95% C.I.
 `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf.
-@goldscheider1969minority; @johnson1979minority; @yang2003big; @maralani2013demography].
+@goldscheider1969minority, @johnson1979minority, @yang2003big, @maralani2013demography).
 And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated.
@@
 #    mutate(
-#      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
+#      Ethnicity = fct_relevel(titles[ethnicity], "White participants",
+#                              "Black participants")
 #    ) |> 
@@
   mutate(
-      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
+      Ethnicity = fct_relevel(titles[eth], "White participants",
+                              "Black participants"),
     ) |> 
@@
       labs(x = "Birth year", y = "Count") +
-      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
+      scale_fill_manual(values = c("Black participants" = "orange2",
+                                   "White participants" = "steelblue4")) +
       theme_minimal() +
@@
    mutate(
-      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
+      Ethnicity = fct_relevel(titles[eth], "White participants", 
+                              "Black participants"),
     ) |> 
@@
   mutate(
-      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
+      Ethnicity = fct_relevel(titles[eth], "White participants",
+                              "Black participants"),
     )  |> 
@@
       labs(x = "Birth year", y = "Count") +
-      scale_fill_manual(values = c(Black = "orange2", White =
-      "steelblue4")) +
+      scale_fill_manual(values = c("Black participants" = "orange2",
+                                   "White participants" = "steelblue4")) +
       theme_minimal() +

### 2024-06-04 renv-extra-dependencies.R

@@ -1,0 +1,5 @@
+# This file exists for renv to read, so that quarto and quarto cache dependencies
+# are recorded
+library(quarto)
+library(furrr)
+library(future)

### 2024-06-04 hrs-selection.qmd
Turn off cache

@@
 #| label: stat-pgicorrect
-#| cache: true
+#| cache: false
 
